### 2022-06-08 Code/PaperMelanoma.Rmd
Initial

---
title: "PaperMelanoma"
author: "Carolien C.H.M. Maas"
date: "3-6-2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Development and validation of a clinial prediction model to predict recurrence and disease-specific mortality in patients with melanoma after sentinel lymph node biopsy

This file includes the code and results for the above mentioned paper.

# Data preparation
### Load packages, functions, and data 
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# load libraries
library(foreign)   # Reading external (spps) data
library(mice)      # Multipe imputation
library(survminer) # Kaplan-Meier plot
library(prodlim)   # median follow-up time
library(glmnet)    # LASSO
library(rms)       # rcs
library(Hmisc)     # rcorr.cens and fit.mult.impute
library(mitools)   # MIcombine
library(arsenal)   # descriptives table
library(png)       # reading PNG images
set.seed(100)
# import function to create table of missing values
source("Z:/Project Melanoom/PaperMelanoma/Code/table.missing.values.R") 
# import function to plot calibration
source("Z:/Project Melanoom/PaperMelanoma/Code/val_surv_mi_v3.R")
# helper functions for calibration plots
km <- function(sel, S.km, horizon){
  S.km.sel <- S.km[sel,]
  sf <- survfit(S.km.sel~1)
  year1 <- max(sf$time[sf$time<=horizon])
  1-sf$surv[sf$time==year1]
}
km.lower <- function(sel, S.km, horizon){
  S.km.sel <- S.km[sel,]
  sf <- survfit(S.km.sel~1)
  year1 <- max(sf$time[sf$time<=horizon])
  1-sf$lower[sf$time==year1]
}
km.upper <- function(sel, S.km, horizon){
  S.km.sel <- S.km[sel,]
  sf <- survfit(S.km.sel~1)
  year1 <- max(sf$time[sf$time<=horizon])
  1-sf$upper[sf$time==year1]
}
fun.event <- function(lp, h0)
{
  h <- h0*exp(lp)
  p <- 1-exp(-h)
  return(p)
}

# load data file
original.data <- foreign::read.spss(file="Z:/Project Melanoom/PaperMelanoma/Data/Dataset Predictiemodel Stassen.sav",
                                    to.data.frame=TRUE, origin="01-01-1970")

# Select necessary variables
selected.vars <- c("EXCLUSIE", "DatabaseNR", "Center", "Sex", "Age", "DOB",
                   "Ulceration", "Loc_CAT",            # OLD data set: Loc_CAT2
                   "Histology", "Breslow", "Tot_nr_SNs",
                   "nr_SNs_Field_1", "nr_SNs_Field_2", "nr_SNs_Field_3",
                   "Mitosis",                          # needed for AJCC.7.i
                   "SNdate", "Recurrence", "Date_1st_Rec", "Last_FU",
                   "dead", "Status",
                   "SNstatus",                         # distinguish between positive and negative SN patients
                   "Tot_SNs_pos", "Rdamcrit", "Dewar") # additional for positive SN patients
```

### Exclude patients
Exclude when Exclusion = 2: based on duplicates, urogenital melanoma, etc.
Leave the patients included that have EXCLUSIE = 1 and Dbstatus = 8,
even though diagnosis and operation date are very close.
```{r, eval=TRUE}
sel.data <- original.data[original.data$EXCLUSIE!="Exclusion", selected.vars]
cat("Number of patients excluded:", nrow(original.data)-nrow(sel.data), "\n")
```
Omit patient A0523 - no last follow-up date
```{r, eval=TRUE}
sel.data <- sel.data[sel.data$DatabaseNR!="A0523               ",]
```

### Recode
Set Rdamcrit of negative SN patients equal to 0
```{r, eval=TRUE}
sel.data$Rdamcrit[sel.data$SNstatus=="Negative"] <- 1 # since log(1) = 0
```

Merge single cells with subcap and set Dewar for negative patients to subcap
```{r, eval=TRUE}
sel.data$Dewar[sel.data$Dewar=="single cells"] <- "subcap"
sel.data$Dewar[sel.data$SNstatus=="Negative"] <- "subcap"
```

Merge location other with trunk
```{r, eval=TRUE}
sel.data$Loc_CAT[sel.data$Loc_CAT=="other"] <- "trunk"
```

Merge LMM other with other
```{r, eval=TRUE}
sel.data$Histology[sel.data$Histology=="LMM"] <- "other"
```

Melanoma specific mortality (MSM)
```{r, eval=TRUE}
sel.data$MSM<-1*(sel.data$Status=="DOD")
```

Define multiple fields
```{r, eval=TRUE}
sel.data$multiple.fields <- ifelse(!is.na(sel.data$nr_SNs_Field_2)&sel.data$nr_SNs_Field_2>0,1,0)
```

Create number of negative nodes removed
```{r, eval=TRUE}
sel.data$Tot_SNs_neg <- sel.data$Tot_nr_SNs - sel.data$Tot_SNs_pos
```

Use age at SN instead of age at diagnosis
```{r, eval=TRUE}
sel.data$Age.SN <- floor((sel.data$SNdate-sel.data$DOB)/(24*60*60)/365.25)
```

Define time since diagnosis in years (follow-up time)
```{r, eval=TRUE}
sel.data$FU.time <- (sel.data$Last_FU-sel.data$SNdate)/(24*60*60)/365.25
```

Define recurrence time in years
```{r, eval=TRUE}
sel.data$Rec.time<- (sel.data$Date_1st_Rec-sel.data$SNdate)/(24*60*60)/365.25
```

Omit patients with Rec date before SN date
```{r, eval=TRUE}
old.n <- nrow(sel.data)
sel.data <- sel.data[!(sel.data$Rec.time<0&!is.na(sel.data$Rec.time)),]
cat("Number of patients excluded:", old.n-nrow(sel.data), "\n")
```

Set recurrence time equal to follow-up time for those without recurrence
```{r, eval=TRUE}
sel.data$Rec.time[sel.data$Recurrence=="no recurrence"|is.na(sel.data$SNdate)]<-sel.data$FU.time[sel.data$Recurrence=="no recurrence"|is.na(sel.data$SNdate)]
```

Set Mitosis to Unknown if missing
```{r, eval=TRUE}
sel.data$Mitosis[is.na(sel.data$Mitosis)] <- "Unknown" # mitosis unknown for 80% of patients
```

Replace "Unknown" by NA
```{r, eval=TRUE}
sel.data$Ulceration[sel.data$Ulceration=="unknown"]<-NA
sel.data$Loc_CAT[sel.data$Loc_CAT=="unknown"]<-NA
sel.data$Histology[sel.data$Histology=="unknown"]<-NA
sel.data$Dewar[sel.data$Dewar=="unknown"] <- NA
sel.data$Mitosis[sel.data$Mitosis=="Unknown"] <- NA
```

Delete unused levels
```{r, eval=TRUE}
sel.data <- droplevels(sel.data)
```

# Multiple imputation
```{r, eval=TRUE}
variables.to.be.imputed <- c("Center", "SNstatus", "Sex", "Age.SN", "Ulceration",
                             "Loc_CAT", "Histology", "Breslow", "multiple.fields",
                             "Tot_SNs_neg", "Tot_SNs_pos", "Rdamcrit", "Dewar",
                             "Mitosis",
                             "Recurrence", "dead", "MSM")
data.to.be.imputed <- sel.data[,variables.to.be.imputed]
```
Save distributions of all variables, necessary for Harrell summaries
```{r, eval=TRUE}
dd<-datadist(data.to.be.imputed)
options(datadist='dd')
options(digits=8)
```

Relation between outcome and covariates
```{r, eval=TRUE}
data.to.be.imputed$logRec.time <- log(sel.data$Rec.time+1/365.25)
data.to.be.imputed$logFU.time <- log(sel.data$FU.time+1/365.25)
missings.table(data.to.be.imputed)
save(data.to.be.imputed,file='Z:/Project Melanoom/PaperMelanoma/Data/descr.data.Rdata', compress=TRUE)
```

Number of imputations
```{r, eval=TRUE}
m <- 5
```

Perform multiple imputation
```{r, eval=FALSE}
imputed.data <- mice::mice(data.to.be.imputed, m=m, maxit=5, print=FALSE, seed=500)
print(imputed.data$loggedEvents)
save(imputed.data,file='Z:/Project Melanoom/PaperMelanoma/Data/imputed.data.RData', compress=TRUE)
```

Load imputed data set
```{r, eval=TRUE}
load('Z:/Project Melanoom/PaperMelanoma/Data/imputed.data.RData')
complete.imputed.data <- mice::complete(imputed.data, m)
summary(complete.imputed.data)
missings.table(complete.imputed.data)
```

# Set as survival data
```{r, eval=TRUE}
S.Rec <- survival::Surv(exp(complete.imputed.data$logRec.time), sel.data$Recurrence=="recurred")
S.MSM <- survival::Surv(exp(complete.imputed.data$logFU.time), sel.data$MSM)
```

### Kaplan-Meier plots
```{r, eval=TRUE}
center.Rec.KM <- survival::survfit(S.Rec ~ Center, data=sel.data)
survminer::ggsurvplot(center.Rec.KM, conf.int=TRUE, risk.table=TRUE)
center.MSM.KM <- survival::survfit(S.MSM ~ Center, data=sel.data)
survminer::ggsurvplot(center.MSM.KM, conf.int=TRUE, risk.table=TRUE)
```

# Median follow-up time
```{r, eval=TRUE}
median.FU.time.total <- stats::quantile(prodlim::prodlim(Hist(exp(logFU.time), MSM)~1, data=complete.imputed.data, reverse=TRUE))
tab.FU.time <- c("total", paste(sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "quantile"]),
                 paste0("[", sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "lower"]), "; ",
                        sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "upper"]), "]")))
median.FU.time <- stats::quantile(prodlim::prodlim(Hist(exp(logFU.time), MSM)~Center, data=complete.imputed.data, reverse=TRUE))
for (center in sort(as.character(unique(sel.data$Center)))){
  med.FU.table <- eval(parse(text=paste0("median.FU.time$`Center=", center, "`")))
  med.FU <- med.FU.table[3,"quantile"]
  lower.FU <- med.FU.table[3,"lower"]
  upper.FU <- med.FU.table[3,"upper"]
  tab.FU.time <- rbind(tab.FU.time, c(center, paste(sprintf("%.2f", med.FU), paste0("[", sprintf("%.2f", lower.FU), "; ", sprintf("%.2f", upper.FU), "]"))))
}
write.table(as.data.frame(tab.FU.time),file='Z:/Project Melanoom/PaperMelanoma/Results/med.FU.table.txt',
            row.names=FALSE)
tab.FU.time
```

# Recurrence at 5 years
```{r, eval=TRUE}
Rec.table <- c("Total", sum(sel.data$Recurrence=="recurred"), sprintf("%.1f", sum(sel.data$Recurrence=="recurred")/nrow(sel.data)*100))
Rec.table <- rbind(Rec.table, c("Positive", sum(sel.data$Recurrence=="recurred"&sel.data$SNstatus=="Positive"),
                                sprintf("%.1f", sum(sel.data$Recurrence=="recurred"&sel.data$SNstatus=="Positive")/sum(sel.data$SNstatus=="Positive")*100)))
Rec.table <- rbind(Rec.table, c("Negative", sum(sel.data$Recurrence=="recurred"&sel.data$SNstatus=="Negative"),
                                sprintf("%.1f", sum(sel.data$Recurrence=="recurred"&sel.data$SNstatus=="Negative")/sum(sel.data$SNstatus=="Negative")*100)))
utils::write.table(as.data.frame(Rec.table),file='Z:/Project Melanoom/PaperMelanoma/Results/Rec.table.txt',
                   row.names=FALSE)
Rec.table
```

# Melanoma Specific Mortality at 5 years
```{r, eval=TRUE}
MSM.table <- c("Total", sum(sel.data$MSM==1), sprintf("%.1f", sum(sel.data$MSM==1)/nrow(sel.data)*100))
MSM.table <- rbind(MSM.table, c("Positive", sum(sel.data$MSM==1&sel.data$SNstatus=="Positive"),
                                sprintf("%.1f", sum(sel.data$MSM==1&sel.data$SNstatus=="Positive")/sum(sel.data$SNstatus=="Positive")*100)))
MSM.table <- rbind(MSM.table, c("Negative", sum(sel.data$MSM==1&sel.data$SNstatus=="Negative"),
                                sprintf("%.1f", sum(sel.data$MSM==1&sel.data$SNstatus=="Negative")/sum(sel.data$SNstatus=="Negative")*100)))
utils::write.table(as.data.frame(MSM.table),file='Z:/Project Melanoom/PaperMelanoma/Results/MSM.table.txt',
                   row.names=FALSE)
MSM.table
```

### Restrict survival at 5 years
```{r, eval=TRUE}
S.Rec[S.Rec[,1]>5, 1] <- 5 # set recurrence time to 5
S.Rec[S.Rec[,1]>5, 2] <- 0 # set censoring
S.MSM[S.MSM[,1]>5, 1] <- 5 # set follow-up time to 5
S.MSM[S.MSM[,1]>5, 2] <- 0 # set censoring
```

# Obtain C-index of AJCC7 and AJCC8
```{r, eval=TRUE}
TNM.7 <- c()
AJCC.7 <- c()
cindex.AJCC.7 <- rep(0, m)
cindex.AJCC.7.se <- rep(0, m)
cindex.AJCC.7.pos <- rep(0, m)
cindex.AJCC.7.se.pos <- rep(0, m)
cindex.AJCC.7.neg <- rep(0, m)
cindex.AJCC.7.se.neg <- rep(0, m)
for (i in 1:m){
  imputation.i <- mice::complete(imputed.data,i)
  TNM.7.i <- rep("", nrow(imputation.i))
  AJCC.7.i <- rep(0, nrow(imputation.i))
  for (patient in 1:nrow(imputation.i)){
    breslow.patient <- imputation.i[patient, "Breslow"]
    ulceration.patient <- as.character(imputation.i[patient, "Ulceration"])
    mitosis.patient <- as.character(imputation.i[patient, "Mitosis"])
    nodes.patient <- imputation.i[patient, "Tot_SNs_pos"]

    snstatus.patient <- imputation.i[patient, "SNstatus"]

    # T stage
    if (breslow.patient <= 1){
      if (ulceration.patient=="No" & mitosis.patient=="No"){
        TNM.7.i[patient] <- paste0(TNM.7.i[patient], "T1a")
      } else if (ulceration.patient=="Yes" | mitosis.patient=="Yes"){
        TNM.7.i[patient] <- paste0(TNM.7.i[patient], "T1b")
      }
    } else if (breslow.patient>1 & breslow.patient <= 2){
      if (ulceration.patient=="No"){
        TNM.7.i[patient] <- paste0(TNM.7.i[patient], "T2a")
      } else if (ulceration.patient=="Yes"){
        TNM.7.i[patient] <- paste0(TNM.7.i[patient], "T2b")
      }
    } else if (breslow.patient>2 & breslow.patient <= 4){
      if (ulceration.patient=="No"){
        TNM.7.i[patient] <- paste0(TNM.7.i[patient], "T3a")
      } else if (ulceration.patient=="Yes"){
        TNM.7.i[patient] <- paste0(TNM.7.i[patient], "T3b")
      }
    } else if (breslow.patient>4){
      if (ulceration.patient=="No"){
        TNM.7.i[patient] <- paste0(TNM.7.i[patient], "T4a")
      } else if (ulceration.patient=="Yes"){
        TNM.7.i[patient] <- paste0(TNM.7.i[patient], "T4b")
      }
    }

    # N stage
    if (nodes.patient == 0){
      TNM.7.i[patient] <- paste0(TNM.7.i[patient], "N0")
    } else if (nodes.patient==1){
      TNM.7.i[patient] <- paste0(TNM.7.i[patient], "N1a")
    } else if (nodes.patient==2 | nodes.patient==3){
      TNM.7.i[patient] <- paste0(TNM.7.i[patient], "N2a")
    } else if (nodes.patient>3){
      TNM.7.i[patient] <- paste0(TNM.7.i[patient], "N3a")
    }

    # M stage for all patients in our data set
    TNM.7.i[patient] <- paste0(TNM.7.i[patient], "M0")

    # determine AJCC.7.i
    if (TNM.7.i[patient]=="T1aN0M0"){
      AJCC.7.i[patient] <- 1.1
    } else if (TNM.7.i[patient]=="T1bN0M0" | TNM.7.i[patient]=="T2aN0M0"){
      AJCC.7.i[patient] <- 1.2
    } else if (TNM.7.i[patient]=="T2bN0M0" | TNM.7.i[patient]=="T3aN0M0"){
      AJCC.7.i[patient] <- 2.1
    } else if (TNM.7.i[patient]=="T3bN0M0" | TNM.7.i[patient]=="T4aN0M0"){
      AJCC.7.i[patient] <- 2.2
    } else if (TNM.7.i[patient]=="T4bN0M0"){
      AJCC.7.i[patient] <- 2.3
    } else if (TNM.7.i[patient]=="T1aN1aM0" | TNM.7.i[patient]=="T2aN1aM0" | # N1a
               TNM.7.i[patient]=="T3aN1aM0" | TNM.7.i[patient]=="T4aN1aM0" |
               TNM.7.i[patient]=="T1aN2aM0" | TNM.7.i[patient]=="T2aN2aM0" | # N2a
               TNM.7.i[patient]=="T3aN2aM0" | TNM.7.i[patient]=="T4aN2aM0"){
      AJCC.7.i[patient] <- 3.1
    } else if (TNM.7.i[patient]=="T1bN1aM0" | TNM.7.i[patient]=="T2bN1aM0" | # N1a
               TNM.7.i[patient]=="T3bN1aM0" | TNM.7.i[patient]=="T4bN1aM0" |
               TNM.7.i[patient]=="T1bN2aM0" | TNM.7.i[patient]=="T2bN2aM0" | # N2a
               TNM.7.i[patient]=="T3bN2aM0" | TNM.7.i[patient]=="T4bN2aM0" |
               TNM.7.i[patient]=="T1aN1bM0" | TNM.7.i[patient]=="T2aN1bM0" | # N1b
               TNM.7.i[patient]=="T3aN1bM0" | TNM.7.i[patient]=="T4aN1bM0" |
               TNM.7.i[patient]=="T1aN2bM0" | TNM.7.i[patient]=="T2aN2bM0" | # N2b
               TNM.7.i[patient]=="T3aN2bM0" | TNM.7.i[patient]=="T4aN2bM0" |
               TNM.7.i[patient]=="T1aN2cM0" | TNM.7.i[patient]=="T2aN2cM0" | # N2c
               TNM.7.i[patient]=="T3aN2cM0" | TNM.7.i[patient]=="T4aN2cM0"){
      AJCC.7.i[patient] <- 3.2
    } else {
      AJCC.7.i[patient] <- 3.3
    }
  }

  # save TNM.7 and AJCC.7 for each imputated data set
  TNM.7 <- cbind(TNM.7, TNM.7.i)
  AJCC.7 <- cbind(AJCC.7, AJCC.7.i)

  # compute C-index
  rc <- Hmisc::rcorr.cens(-AJCC.7.i, S.Rec) # add a minus to revert class order: from worst to best
  cindex.AJCC.7[i]<-rc["C Index"]
  cindex.AJCC.7.se[i]<-rc["S.D."]/2

  # compute C-index for only positive sentinel node patients
  rc.pos <- Hmisc::rcorr.cens(-AJCC.7.i[imputation.i$SNstatus=="Positive"], S.Rec[imputation.i$SNstatus=="Positive",])
  cindex.AJCC.7.pos[i]<-rc.pos["C Index"]
  cindex.AJCC.7.se.pos[i]<-rc.pos["S.D."]/2

  # compute C-index for only negative sentinel node patients
  rc.neg <- Hmisc::rcorr.cens(-AJCC.7.i[imputation.i$SNstatus=="Negative"], S.Rec[imputation.i$SNstatus=="Negative",])
  cindex.AJCC.7.neg[i]<-rc.neg["C Index"]
  cindex.AJCC.7.se.neg[i]<-rc.neg["S.D."]/2
}
# combine C-index for each imputed data set
cindex.mi.AJCC.7<-Rubin.combine(cindex.AJCC.7,cindex.AJCC.7.se)
cindex.mi.AJCC.7.pos<-Rubin.combine(cindex.AJCC.7.pos,cindex.AJCC.7.se.pos)
cindex.mi.AJCC.7.neg<-Rubin.combine(cindex.AJCC.7.neg,cindex.AJCC.7.se.neg)

# save C-index and confidence interval
utils::write.table(rbind(c("All patients:", sprintf('%.2f', cindex.mi.AJCC.7$est),
              paste0("[", sprintf('%.2f', cindex.mi.AJCC.7$est+qnorm(.025)*cindex.mi.AJCC.7$se),
                      "; ", sprintf('%.2f', cindex.mi.AJCC.7$est+qnorm(.975)*cindex.mi.AJCC.7$se), "]")),
            c("Positive SN patients:", sprintf('%.2f', cindex.mi.AJCC.7.pos$est),
                  paste0("[", sprintf('%.2f', cindex.mi.AJCC.7.pos$est+qnorm(.025)*cindex.mi.AJCC.7.pos$se),
                         "; ", sprintf('%.2f', cindex.mi.AJCC.7.pos$est+qnorm(.975)*cindex.mi.AJCC.7.pos$se), "]")),
            c("Negative SN patients:", sprintf('%.2f', cindex.mi.AJCC.7.neg$est),
                  paste0("[", sprintf('%.2f', cindex.mi.AJCC.7.neg$est+qnorm(.025)*cindex.mi.AJCC.7.neg$se),
                         "; ", sprintf('%.2f', cindex.mi.AJCC.7.neg$est+qnorm(.975)*cindex.mi.AJCC.7.neg$se), "]"))),
            file='Z:/Project Melanoom/PaperMelanoma/Results/AJCC.7.Cindex.txt', row.names=FALSE, col.names=FALSE)

#####
##### COMPARE MODEL WITH AJCC.8 CLASIFCATION
#####
TNM.8 <- c()
AJCC.8 <- c()
cindex.AJCC.8 <- rep(0, m)
cindex.AJCC.8.se <- rep(0, m)
cindex.AJCC.8.pos <- rep(0, m)
cindex.AJCC.8.se.pos <- rep(0, m)
cindex.AJCC.8.neg <- rep(0, m)
cindex.AJCC.8.se.neg <- rep(0, m)
for (i in 1:m){
  imputation.i <- complete(imputed.data,i)
  TNM.8.i <- rep("", nrow(imputation.i))
  AJCC.8.i <- rep(0, nrow(imputation.i))
  for (patient in 1:nrow(imputation.i)){
    breslow.patient <- imputation.i[patient, "Breslow"]
    ulceration.patient <- as.character(imputation.i[patient, "Ulceration"])
    mitosis.patient <- as.character(imputation.i[patient, "Mitosis"])
    nodes.patient <- imputation.i[patient, "Tot_SNs_pos"]

    # T stage
    if (breslow.patient <= 1){
      if (ulceration.patient=="No" & breslow.patient < 0.8){
        TNM.8.i[patient] <- paste0(TNM.8.i[patient], "T1a")
      } else {
        TNM.8.i[patient] <- paste0(TNM.8.i[patient], "T1b")
      }
    } else if (breslow.patient>1 & breslow.patient <= 2){
      if (ulceration.patient=="No"){
        TNM.8.i[patient] <- paste0(TNM.8.i[patient], "T2a")
      } else if (ulceration.patient=="Yes"){
        TNM.8.i[patient] <- paste0(TNM.8.i[patient], "T2b")
      }
    } else if (breslow.patient>2 & breslow.patient <= 4){
      if (ulceration.patient=="No"){
        TNM.8.i[patient] <- paste0(TNM.8.i[patient], "T3a")
      } else if (ulceration.patient=="Yes"){
        TNM.8.i[patient] <- paste0(TNM.8.i[patient], "T3b")
      }
    } else if (breslow.patient>4){
      if (ulceration.patient=="No"){
        TNM.8.i[patient] <- paste0(TNM.8.i[patient], "T4a")
      } else if (ulceration.patient=="Yes"){
        TNM.8.i[patient] <- paste0(TNM.8.i[patient], "T4b")
      }
    }

    # N stage
    if (nodes.patient == 0){
      TNM.8.i[patient] <- paste0(TNM.8.i[patient], "N0")
    } else if (nodes.patient==1){
      TNM.8.i[patient] <- paste0(TNM.8.i[patient], "N1a")
    } else if (nodes.patient==2 | nodes.patient==3){
      TNM.8.i[patient] <- paste0(TNM.8.i[patient], "N2a")
    } else if (nodes.patient>3){
      TNM.8.i[patient] <- paste0(TNM.8.i[patient], "N3a")
    }

    # M stage for all patients in our data set
    TNM.8.i[patient] <- paste0(TNM.8.i[patient], "M0")

    # determine AJCC.8.i
    if (TNM.8.i[patient]=="T1aN0M0"){
      AJCC.8.i[patient] <- 1.1
    } else if (TNM.8.i[patient]=="T1bN0M0" | TNM.8.i[patient]=="T2aN0M0"){
      AJCC.8.i[patient] <- 1.2
    } else if (TNM.8.i[patient]=="T2bN0M0" | TNM.8.i[patient]=="T3aN0M0"){
      AJCC.8.i[patient] <- 2.1
    } else if (TNM.8.i[patient]=="T3bN0M0" | TNM.8.i[patient]=="T4aN0M0"){
      AJCC.8.i[patient] <- 2.2
    } else if (TNM.8.i[patient]=="T4bN0M0"){
      AJCC.8.i[patient] <- 2.3
    } else if (TNM.8.i[patient]=="T1aN1aM0" | TNM.8.i[patient]=="T1bN1aM0" | TNM.8.i[patient]=="T2aN1aM0"
               | TNM.8.i[patient]=="T1aN2aM0" | TNM.8.i[patient]=="T1bN2aM0" | TNM.8.i[patient]=="T2aN2aM0"){
      AJCC.8.i[patient] <- 3.1
    } else if (TNM.8.i[patient]=="T2bN1aM0" | TNM.8.i[patient]=="T3aN1aM0"){
      AJCC.8.i[patient] <- 3.2
    } else if (TNM.8.i[patient]=="T4bN3aM0"){
      AJCC.8.i[patient] <- 3.4
    } else {
      AJCC.8.i[patient] <- 3.3
    }

  }

  # save TNM.8 and AJCC.8 for each imputated data set
  TNM.8 <- cbind(TNM.8, TNM.8.i)
  AJCC.8 <- cbind(AJCC.8, AJCC.8.i)

  # compute C-index
  rc <- Hmisc::rcorr.cens(-AJCC.8.i, S.Rec) # add a minus to revert class order: from worst to best
  cindex.AJCC.8[i]<-rc["C Index"]
  cindex.AJCC.8.se[i]<-rc["S.D."]/2

  # compute C-index for only positive sentinel node patients
  rc.pos <- Hmisc::rcorr.cens(-AJCC.8.i[imputation.i$SNstatus=="Positive"], S.Rec[imputation.i$SNstatus=="Positive",])
  cindex.AJCC.8.pos[i]<-rc.pos["C Index"]
  cindex.AJCC.8.se.pos[i]<-rc.pos["S.D."]/2

  # compute C-index for only negative sentinel node patients
  rc.neg <- Hmisc::rcorr.cens(-AJCC.8.i[imputation.i$SNstatus=="Negative"], S.Rec[imputation.i$SNstatus=="Negative",])
  cindex.AJCC.8.neg[i]<-rc.neg["C Index"]
  cindex.AJCC.8.se.neg[i]<-rc.neg["S.D."]/2
}
# combine C-index for each imputed data set
cindex.mi.AJCC.8<-Rubin.combine(cindex.AJCC.8,cindex.AJCC.8.se)
cindex.mi.AJCC.8.pos<-Rubin.combine(cindex.AJCC.8.pos,cindex.AJCC.8.se.pos)
cindex.mi.AJCC.8.neg<-Rubin.combine(cindex.AJCC.8.neg,cindex.AJCC.8.se.neg)

# save C-index and confidence interval
utils::write.table(rbind(c("All patients:", sprintf('%.2f', cindex.mi.AJCC.8$est),
                        paste0("[", sprintf('%.2f', cindex.mi.AJCC.8$est+qnorm(.025)*cindex.mi.AJCC.8$se),
                               "; ", sprintf('%.2f', cindex.mi.AJCC.8$est+qnorm(.975)*cindex.mi.AJCC.8$se), "]")),
                  c("Positive SN patients:", sprintf('%.2f', cindex.mi.AJCC.8.pos$est),
                        paste0("[", sprintf('%.2f', cindex.mi.AJCC.8.pos$est+qnorm(.025)*cindex.mi.AJCC.8.pos$se),
                               "; ", sprintf('%.2f', cindex.mi.AJCC.8.pos$est+qnorm(.975)*cindex.mi.AJCC.8.pos$se), "]")),
                  c("Negative SN patients", sprintf('%.2f', cindex.mi.AJCC.8.neg$est),
                        paste0("[", sprintf('%.2f', cindex.mi.AJCC.8.neg$est+qnorm(.025)*cindex.mi.AJCC.8.neg$se),
                               "; ", sprintf('%.2f', cindex.mi.AJCC.8.neg$est+qnorm(.975)*cindex.mi.AJCC.8.neg$se), "]"))),
            file='Z:/Project Melanoom/PaperMelanoma/Results/AJCC.8.Cindex.txt', row.names=FALSE, col.names=FALSE)

# table of number of AJCC.7 and AJCC.8
table.AJCC.7 <- c()
table.AJCC.8 <- c()
for (i in 1:m){
  table.AJCC.7 <- cbind(table.AJCC.7, as.numeric(table(AJCC.7[,i])))
  table.AJCC.8 <- cbind(table.AJCC.8, as.numeric(table(AJCC.8[,i])))
}
table.AJCC.7.8 <- cbind(as.numeric(names(table(AJCC.8[,1]))),
                        c(round(rowMeans(table.AJCC.7), 0), 0),
                        round(rowMeans(table.AJCC.8), 0))
utils::write.table(table.AJCC.7.8,
            file='Z:/Project Melanoom/PaperMelanoma/Results/table.AJCC.7.and.8.txt', row.names=FALSE, col.names=FALSE)
table.AJCC.7.8
```

# Full model for recurrence
```{r, eval=TRUE}
form.full.Rec <- S.Rec ~
  SNstatus +
  Sex +
  log(Age.SN) +
  Ulceration +
  Loc_CAT +
  Histology +
  log(Breslow) +
  multiple.fields +
  Tot_SNs_neg +
  Tot_SNs_pos +
  log(Rdamcrit) +
  Dewar +
  SNstatus:Sex +
  SNstatus:log(Age.SN) +
  SNstatus:Ulceration +
  SNstatus:Loc_CAT +
  SNstatus:Histology +
  SNstatus:log(Breslow) +
  SNstatus:multiple.fields
f.mi.full.Rec <- Hmisc::fit.mult.impute(form.full.Rec, cph, xtrans=imputed.data,
                                 data=data.to.be.imputed,n.impute=m, pr=FALSE,
                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
f.mi.full.Rec
stats::anova(f.mi.full.Rec)
summary(f.mi.full.Rec)
```

### Test proportional hazards assumption
```{r, eval=TRUE}
test.prop<-survival::cox.zph(f.mi.full.Rec$fits[[1]])
test.prop
plot(test.prop)
```

# Final model for recurrence selected by backward selection with p = 0.01
```{r, eval=TRUE}
c.Breslow <- mean(log(complete.imputed.data$Breslow))
form.BS.Rec <- S.Rec ~
  SNstatus +
  log(Age.SN) +
  Ulceration +
  Loc_CAT +
  I(log(Breslow)-c.Breslow) +
  log(Rdamcrit) +
  SNstatus:I(log(Breslow)-c.Breslow)
f.mi.BS.Rec <- Hmisc::fit.mult.impute(form.BS.Rec,cph,xtrans=imputed.data,data=data.to.be.imputed,n.impute=m,
                           pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
round(stats::anova(f.mi.BS.Rec)[order(stats::anova(f.mi.BS.Rec)[, 'P']),], 3)
```

# Refitted full model for melanoma specific mortality
```{r, eval=TRUE}
form.full.MSM.refit <- S.MSM ~
  SNstatus +
  Sex +
  log(Age.SN) +
  Ulceration +
  Loc_CAT +
  Histology +
  log(Breslow) +
  multiple.fields +
  Tot_SNs_neg +
  Tot_SNs_pos +
  log(Rdamcrit) +
  Dewar +
  SNstatus:Sex +
  SNstatus:log(Age.SN) +
  SNstatus:Ulceration +
  SNstatus:Loc_CAT +
  SNstatus:Histology +
  SNstatus:log(Breslow) +
  SNstatus:multiple.fields
f.mi.full.MSM.refit <- Hmisc::fit.mult.impute(form.full.MSM.refit,cph,xtrans=imputed.data,data=data.to.be.imputed,n.impute=m,
                                 pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
f.mi.full.MSM.refit
summary(f.mi.full.MSM.refit)
stats::anova(f.mi.full.MSM.refit)
```

# Refitted final model for melanoma specific mortality
```{r, eval=TRUE}
form.BS.MSM.refit <- S.MSM ~
  SNstatus +
  log(Age.SN) +
  Ulceration +
  Loc_CAT +
  I(log(Breslow)-c.Breslow) +
  log(Rdamcrit) +
  SNstatus:I(log(Breslow)-c.Breslow)
f.mi.BS.MSM.refit <- Hmisc::fit.mult.impute(form.BS.MSM.refit,cph,xtrans=imputed.data,data=data.to.be.imputed,n.impute=m,
                                  pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
round(anova(f.mi.BS.MSM.refit)[order(anova(f.mi.BS.MSM.refit)[, 'P']),], 3)
f.mi.BS.MSM.refit
summary(f.mi.BS.MSM.refit)
stats::anova(f.mi.BS.MSM.refit)
```

# Calibrated final model for melanoma specific mortality
Fit disease specific mortality to the linear predictor of recurrence
```{r, eval=TRUE}
lp.BS<-f.mi.BS.Rec$linear
f.mi.BS.MSM<-rms::cph(S.MSM~lp.BS,y=TRUE,x=TRUE)
utils::write.table(paste0(sprintf("%.2f", stats::coef(f.mi.BS.MSM)), " [",
            sprintf("%.2f", as.numeric(stats::confint(f.mi.BS.MSM))[1]), "; ",
            sprintf("%.2f", as.numeric(stats::confint(f.mi.BS.MSM))[2]), "]"),
            file='Z:/Project Melanoom/PaperMelanoma/Results/calibration.MSS.txt', row.names=FALSE, col.names=FALSE)
MSM.cal.fact <- stats::coef(f.mi.BS.MSM)
stats::coef(f.mi.BS.MSM)
stats::confint(f.mi.BS.MSM)
```

### Create variables with results
##### Full model
```{r, eval=TRUE}
for (model in c('Rec', 'MSM.refit')){
  f.mi.full <- eval(parse(text=paste0('f.mi.full.', model)))

  CI.lower.neg <- exp(stats::confint(f.mi.full)[,1])
  CI.upper.neg <- exp(stats::confint(f.mi.full)[,2])
  CI.neg <- paste0("[", sprintf("%.2f", CI.lower.neg), "; ", sprintf("%.2f", CI.upper.neg), "]")
  CI.lower.pos <- c(CI.lower.neg[2:12]*CI.lower.neg[20:30], CI.lower.neg[14:19])
  CI.upper.pos <- c(CI.upper.neg[2:12]*CI.upper.neg[20:30], CI.upper.neg[14:19])
  CI.pos <- paste0("[", sprintf("%.2f", CI.lower.pos), "; ", sprintf("%.2f", CI.upper.pos), "]")

  anova.full <- stats::anova(f.mi.full)[,1]

  full.table <- cbind(c(NA, sprintf("%.2f", exp(f.mi.full$coefficients[2:13])), rep(NA, 5)), # COEF NEG
                      c(NA, CI.neg[2:13], rep(NA, 5)), # CI NEG
                      sprintf("%.2f", c(exp(f.mi.full$coefficients[1]), # COEF POS
                        exp(f.mi.full$coefficients[2:12])*exp(f.mi.full$coefficients[20:30]),
                        exp(f.mi.full$coefficients[14]), exp(f.mi.full$coefficients[15:19]))),
                      c(CI.neg[1], CI.pos), # CI POS
                      c(sprintf("%.1f", anova.full[c(1, 3, 5, 7, 9)]), rep(NA, 2), # CHI2 NEG
                            sprintf("%.1f", anova.full[11]), rep(NA, 2),
                            sprintf("%.1f", anova.full[c(13, 15, 17, 19, 20)]), rep(NA, 3)),
                      c(sprintf("%.1f", anova.full[c(2, 21, 22, 23, 24)]), rep(NA, 2), # CHI2 POS
                            sprintf("%.1f", anova.full[25]), rep(NA, 2),
                            sprintf("%.1f", anova.full[c(26:27, 18)]), rep(NA, 5)))

  ncols <- 6
  final.table <- rbind(full.table[1:4,],
                       rep(NA, ncols), # location
                       c("(ref)", NA, "(ref)", rep(NA, 3)), # ref arm
                       full.table[5:7,], # other locations
                       rep(NA, ncols), # histology
                       c("(ref)", NA, "(ref)", rep(NA, 3)), # ref SMM
                       full.table[8:14,],
                       rep(NA, ncols), # Dewar
                       c(rep(NA, 2), "(ref)", rep(NA, 3)), # ref subcap
                       full.table[15:18,]) # other Dewar

  colnames(final.table) <- c("HR.neg", "CI.neg", "HR.pos", "CI.pos", "Chi2.neg", "Chi2.pos")
  assign(paste0('final.full.table.', model), as.data.frame(final.table))
}
```

##### Final model
```{r, eval=TRUE}
for (model in c('Rec', 'MSM', 'MSM.refit')){
  if (model=="MSM"){
    mult.factor <- MSM.cal.fact # if MSM, multiply by MSM.cal.fact
    f.mi.BS <- eval(parse(text=paste0('f.mi.BS.Rec'))) # use recurrence model if calibrated MSM
  }
  else{
    mult.factor <- 1 # if Rec or MSM.refit, don't multiply
    f.mi.BS <- eval(parse(text=paste0('f.mi.BS.', model)))
  }

  CI.lower.neg <- exp(mult.factor*confint(f.mi.BS)[,1])
  CI.upper.neg <- exp(mult.factor*confint(f.mi.BS)[,2])
  CI.neg <- paste0("[", sprintf("%.2f", CI.lower.neg), "; ", sprintf("%.2f", CI.upper.neg), "]")

  anova.BS <- anova(f.mi.BS)[,1]

  CI.pos <- paste0("[", sprintf("%.2f", exp(mult.factor*confint(f.mi.BS)[7,1])*exp(mult.factor*confint(f.mi.BS)[9,1])),
                   "; ", sprintf("%.2f", exp(mult.factor*confint(f.mi.BS)[7,2])*exp(mult.factor*confint(f.mi.BS)[9,2])), "]")

  BS.table <- cbind(c(NA, sprintf("%.2f", exp(mult.factor*f.mi.BS$coefficients[2:7])), NA), # COEF NEG
                    c(NA, CI.neg[2:7], NA), # CI NEG
                    sprintf("%.2f", exp(mult.factor*f.mi.BS$coefficients[1:8])*c(rep(1, 6), exp(mult.factor*f.mi.BS$coefficients[9]), 1)), #COEF POS
                    c(CI.neg[1:6], CI.pos, CI.neg[8]), # CI POS
                    c(NA, sprintf("%.1f", anova.BS[3:5]), rep(NA, 2), sprintf("%.1f", anova.BS[c(6, 8)])), # CHI2 NEG
                    c(sprintf("%.1f", anova.BS[1]), rep(NA, 5), # CHI2 POS
                      sprintf("%.1f", anova.BS[9]), NA))

  BS.table <- rbind(BS.table[1:3,], rep(NA, 6), c("(ref)", NA, "(ref)", rep(NA, 3)), BS.table[4:8,])

  colnames(BS.table) <- c("HR.neg", "CI.neg", "HR.pos", "CI.pos", "Chi2.neg", "Chi2.pos")

  assign(paste0('final.BS.table.', model), as.data.frame(BS.table))
}
```

# Plot calibration of all centers combined and all centers seperate for recurrence and melanoma specific mortality refit and write results to Excel
```{r, eval=TRUE}
horizon <- 5
for (model in list('Rec', 'MSM.refit')){
  if (model == "Rec"){
    # plot predict
    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/plot.predict.Rec.png"))
    plot.predict <- plot(Predict(f.mi.full.Rec))
    show(plot.predict)
    grDevices::dev.off()

    # plot predict Breslow
    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/plot.predict.Breslow.Rec.png"))
    plot.Breslow <- plot(Predict(f.mi.full.Rec, Breslow=0:20, SNstatus), ~ Breslow, groups="SNstatus",
                         adj.subtitle=FALSE, lwd=2, labels=FALSE,
                         par.settings=list(superpose.line=list(lty=c(1, 2), col=c("red", "blue"))),
                         key=list(cex=.8,corner = c(0.1, .9),lines=list(col=c("red","blue"), lty=c(1,2), lwd=2),text=list(c("Negative","Positive"))))
    show(plot.Breslow)
    grDevices::dev.off()
  }

  # discrimination full and backward selected model
  for (type in c("full", "BS")){
    f.mi <- eval(parse(text=paste0('f.mi.', type, '.', model)))
    if (model=="Rec"){
      S <- S.Rec
    }
    else if (model=="MSM.refit"){
      S <- S.MSM
    }
    p <- matrix(NA, nrow(sel.data), m)
    for (i in 1:m){
      # predicted lp for imputation i
      lp.i <- predict(f.mi,newdata=complete(imputed.data,i), type="lp")
      # baseline hazard
      f.basehaz <- survival::basehaz(f.mi,TRUE)
      h0.i <- f.basehaz$hazard[f.basehaz$time==max(f.basehaz$time[f.basehaz$time<=horizon])]
      # probabilities
      p.i <- fun.event(lp=lp.i, h0=h0.i)
      p[, i] <- p.i
    }

    # omit the observation with predicted probability = 1
    omit <- unique(c(which(p[,1]==1), which(p[,2]==1), which(p[,3]==1), which(p[,4]==1), which(p[,5]==1)))
    if (length(omit)>0){
      cat('Omit:', omit, '\n')
      p <- p[-omit,]
      S <- S[-omit,]
    }

    out <- val.surv.mi(p,S,g=5,main="All centers",
                       file_path=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/all.centers.", model, '.', type, ".png"),
                       save_plots=TRUE,time=NULL,lim=c(0,1),dist=TRUE)

    final.full.table <- eval(parse(text=paste0('final.', type, '.table.', model)))
    final.full.table["C-index", "HR.neg"] <- sprintf('%.2f', out$cindex)
    final.full.table["C-index", "CI.neg"] <- paste0("[", sprintf('%.2f', out$cindex.lower),
                                                   "; ", sprintf('%.2f', out$cindex.upper), "]")

    assign(paste0('final.', type, '.table.', model), final.full.table)
  }
}

# MSM recalibration
for (type in c("full", "BS")){
  f.mi <- eval(parse(text=paste0('f.mi.', type, '.Rec')))
  cindex.MSM.refit <- rep(0, m)
  cindex.se.MSM.refit <- rep(0, m)
  for (i in 1:m){
    rc <- Hmisc::rcorr.cens(-predict(f.mi,newdata=complete(imputed.data,i), type="lp"), S.MSM)
    cindex.MSM.refit[i]<-rc["C Index"]
    cindex.se.MSM.refit[i]<-rc["S.D."]/2
  }
  cindex.mi.MSM.refit<-Rubin.combine(cindex.MSM.refit,cindex.se.MSM.refit)

  final.full.table <- eval(parse(text=paste0('final.', type, '.table.Rec')))
  final.full.table["C-index MSM", "HR.neg"] <- sprintf('%.2f', cindex.mi.MSM.refit$est)
  final.full.table["C-index MSM", "CI.neg"] <- paste0("[", sprintf('%.2f', cindex.mi.MSM.refit$est+qnorm(.025)*cindex.mi.MSM.refit$se),
                                                 "; ", sprintf('%.2f', cindex.mi.MSM.refit$est+qnorm(.975)*cindex.mi.MSM.refit$se), "]")

  assign(paste0('final.', type, '.table.Rec'), final.full.table)
}

# replace NA by blank
final.full.table.Rec[is.na(final.full.table.Rec)] <- ""
final.full.table.MSM.refit[is.na(final.full.table.MSM.refit)] <- ""
final.BS.table.Rec[is.na(final.BS.table.Rec)] <- ""
final.BS.table.MSM[is.na(final.BS.table.MSM)] <- ""
final.BS.table.MSM["C-index MSM",] <- final.BS.table.Rec[nrow(final.BS.table.Rec),]
final.BS.table.MSM.refit[is.na(final.BS.table.MSM.refit)] <- ""

# save tables
write.table(final.full.table.Rec,file='Z:/Project Melanoom/PaperMelanoma/Results/final.full.table.Rec.txt')
write.table(final.full.table.MSM.refit,file='Z:/Project Melanoom/PaperMelanoma/Results/final.full.table.MSM.refit.txt')
write.table(final.BS.table.Rec,file='Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.Rec.txt')
write.table(final.BS.table.MSM,file='Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.MSM.txt')
write.table(final.BS.table.MSM.refit,file='Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.MSM.refit.txt')
final.full.table.Rec
final.full.table.MSM.refit
final.BS.table.Rec
final.BS.table.MSM
final.BS.table.MSM.refit
```

# Cross-validation recurrence 5-years
```{r, eval=TRUE}
Centers <- sort(as.character(unique(sel.data$Center)))
Centers.anonimized <- c("A", "B", "C", "D")
cuts <- 5
for (model in c('Rec', 'MSM')){
  if (model == 'Rec'){
    S.model <- S.Rec
  }
  if (model == 'MSM'){
    S.model <- S.MSM
  }

  for (type in c('full', 'BS')){
    for (j in 1:4){
      S.j<-S.model[data.to.be.imputed$Center==Centers[j],]
      S.notj<-S.model[data.to.be.imputed$Center!=Centers[j],]
      S.Rec.notj<-S.Rec[data.to.be.imputed$Center!=Centers[j],]

      form.model<-eval(parse(text=paste0('form.', type, '.Rec')))
      form.notj<-update(form.model, S.Rec.notj  ~ . )
      f.mi.Rec.notj<-Hmisc::fit.mult.impute(form.notj,cph,xtrans=imputed.data,data=data.to.be.imputed,
                                 n.impute=m,pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE,
                                 sub=data.to.be.imputed$Center!=Centers[j])

      if (model=="Rec"){
        f.basehaz.notj<-survival::basehaz(f.mi.Rec.notj,TRUE)
        h0.notj<-f.basehaz.notj$hazard[f.basehaz.notj$time==max(f.basehaz.notj$time[f.basehaz.notj$time<=horizon])]
      } else if (model=="MSM"){
        lp.notj<-f.mi.Rec.notj$linear
        f.MSM.notj<-cph(S.notj~lp.notj,y=TRUE,x=TRUE)

        f.basehaz.notj<-basehaz(f.MSM.notj,TRUE)
        h0.notj<-f.basehaz.notj$hazard[f.basehaz.notj$time==max(f.basehaz.notj$time[f.basehaz.notj$time<=horizon])]
      }

      for (select.value in c("Positive", "Negative", "All")){
        # identify those patients that are selected
        if (select.value == "All"){
          select.patients <- TRUE
        } else if (select.value != "All"){
          select.patients <- data.to.be.imputed[data.to.be.imputed$Center==Centers[j],]$SNstatus==select.value
        }

        # calibration with metrics
        S <- S.j[select.patients,]
        p <- matrix(NA, nrow(S), m)
        for (i in 1:m){
          lp.i<-predict(f.mi.Rec.notj,newdata=complete(imputed.data,i)[data.to.be.imputed$Center==Centers[j],], type="lp")

          if (model=="MSM"){
            lp.i<-predict(f.MSM.notj,newdata=lp.i)
          }

          if (select.value != "All"){
            lp.i <- lp.i[select.patients]
          }

          # baseline hazard
          if (model=="Rec"){
            f.basehaz<-basehaz(f.mi.Rec.notj,TRUE)
          }
          else if (model=="MSM"){
            f.basehaz<-basehaz(f.MSM.notj,TRUE)
          }
          h0.i <- f.basehaz$hazard[f.basehaz$time==max(f.basehaz$time[f.basehaz$time<=horizon])]

          # probabilities
          p.i <- fun.event(lp=lp.i, h0=h0.i)
          p[, i] <- p.i
        }

        # omit the observation with predicted probability = 1
        omit <- unique(c(which(p[,1]==1), which(p[,2]==1), which(p[,3]==1), which(p[,4]==1), which(p[,5]==1)))
        if (length(omit)>0){
          cat('Omit:', omit, '\n')
          p <- p[-omit,]
          S <- S[-omit,]
        }

        out <- val.surv.mi(p, S, g=5, main=paste("Center", Centers.anonimized[j]),
                           file_path=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", model, '.', select.value, '.', type, '.', Centers[j], ".png"),
                           save_plots=TRUE,time=NULL,lim=c(0,1),dist=TRUE)
      }
    }
  }
}

# combine calibration plots into one
for (select.value in c("All", "Positive", "Negative")){
  for (y in c("Rec", "MSM")){
    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/", y, ".", select.value, ".png"), width=16, height=16, units="cm", res=300)
    par(mar=rep(0, 4))
    layout(matrix(1:4, ncol=2, byrow=TRUE))
    Centers.val <- rep(1:4, 3)
    for (i in 1:4){
      plot(NA, xlim=0:1, ylim=0:1, xaxt="n", yaxt="n", bty="n")
      img <- readPNG(paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", y, ".", select.value, ".BS.", Centers[Centers.val[i]], ".png"))
      rasterImage(img, 0, 0, 1, 1)
    }
    grDevices::dev.off()
  }
}
```

# Nomogram
```{r, eval=TRUE}
# Limits of Age, Breslow, Rdamcrit
Age.class <- seq(20, 80, 20)
Breslow.class <- c(seq(0.1, 0.9, 0.2), 1:7)
Rdamcrit.class <- c(seq(1, 10, 2)/100, seq(1, 10, 2)/10, seq(1, 10, 2))

# initial nomogram
nom.0 <- nomogram(f.mi.BS.Rec,maxscale=10,
                Breslow=Breslow.class, Rdamcrit=Rdamcrit.class, Age.SN=Age.class)

# Baseline hazard
rc.Rec <- (1/attributes(nom.0)$info$sc)
int.Rec <- attributes(nom.0)$info$Intercept
f.Rec.basehaz <- basehaz(f.mi.BS.Rec,TRUE)
h0.Rec <- f.Rec.basehaz$hazard[f.Rec.basehaz$time==max(f.Rec.basehaz$time[f.Rec.basehaz$time<=horizon])]
lp.range <- int.Rec+rc.Rec*(0:10)*2
par(mar = rep(0, 4))
nom <- nomogram(f.mi.BS.Rec, maxscale=10,
              Breslow=Breslow.class, Rdamcrit=Rdamcrit.class, Age.SN=Age.class,
              lp=TRUE, fun.at=fun.event(lp=lp.range, h0=h0.Rec))
attributes(nom)$names[attributes(nom)$names=="Age.SN"]<-"Age"
attributes(nom)$names[attributes(nom)$names=="Loc_CAT"]<-"Location"
attributes(nom)$names[attributes(nom)$names=="Rdamcrit"]<-"Tumor burden"
attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Positive)"]<-"Breslow + positive SN"
attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Negative)"]<-"Breslow + negative SN"
plot(nom, total.sep.page=TRUE, col.grid=gray(c(0.8, 0.95)))
nom.0
nom

# Slope and intercept of recurrence and MSM
int.Rec
# check that int is same as this:
log(20)*f.mi.BS.Rec$coef["Age.SN"]+log(0.1)*f.mi.BS.Rec$coef["Breslow"]+log(0.01)*f.mi.BS.Rec$coef["Rdamcrit"]-f.mi.BS.Rec$center
```

# Risk score distribution
```{r, eval=TRUE}
rc.MSM <- f.mi.BS.MSM$coef[1]*rc.Rec
int.MSM <- f.mi.BS.MSM$coef[1]*int.Rec

max.range <- 16
points.range <- 0:max.range
lp.points.Rec <- int.Rec+rc.Rec*points.range
lp.points.MSM <- int.MSM+rc.MSM*points.range

f.MSM.basehaz <- basehaz(f.mi.BS.MSM,TRUE)
h0.MSM <- f.MSM.basehaz$hazard[f.MSM.basehaz$time==max(f.MSM.basehaz$time[f.MSM.basehaz$time<=horizon])]
lp.MSM <- f.mi.BS.MSM$linear

Breslow.trunc <- complete.imputed.data$Breslow
Breslow.trunc[complete.imputed.data$Breslow<0.1] <- .1
Breslow.trunc[complete.imputed.data$Breslow>7] <- 7
lp.score.Rec <- predict(f.mi.BS.Rec,newdata=data.frame(SNstatus=complete.imputed.data$SNstatus,
                                                Age.SN=complete.imputed.data$Age.SN,
                                                Ulceration=complete.imputed.data$Ulceration,
                                                Loc_CAT=complete.imputed.data$Loc_CAT,
                                                Breslow=Breslow.trunc,
                                                Rdamcrit=complete.imputed.data$Rdamcrit))
mean(lp.score.Rec)
score.Rec <- round((lp.score.Rec-int.Rec)/rc.Rec,0)
max.breaks <- max.range+1
h.Rec <- hist(score.Rec,plot=FALSE,breaks=0:max.breaks,right=FALSE)

# score table
data.frame(table(score.Rec)/length(score.Rec))
```

# Check mean event rates of risk distribution
```{r, eval=TRUE}
# Recurrence
sum(h.Rec$density*fun.event(lp=lp.points.Rec, h0=h0.Rec))
# equal to the 5-year survival probability, kaplanmeier op 5 jaar, survfit op 5 jaar
SF<-survfit(S.Rec~1,data=sel.data)
100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
mean(fun.event(lp=lp.score.Rec, h0=h0.Rec))

# MSM
sum(h.Rec$density*fun.event(lp=lp.points.MSM, h0=h0.MSM))
SF<-survfit(S.MSM~1,data=sel.data)
100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
mean(fun.event(lp=lp.MSM, h0=h0.MSM))
```

# Plot of risk score distribution
```{r, eval=TRUE, message=FALSE}
h.plot <- h.Rec
h.plot$density <- 100*h.Rec$density
p.Rec <- fun.event(lp=lp.points.Rec, h0=h0.Rec)
p.MSM <- fun.event(lp=lp.points.MSM, h0=h0.MSM)
x.lim <- c(0, max.range+1)
y.lim <- c(0, 100)
y.lim.hist <- c(0, 25)
png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Risk.distribution.png"))
par(mar = c(5,5,2,5))
plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col='white')
abline(h=(0:30)*1,col="light grey",lty=2)
axis(side = 4)
mtext(side = 4, line = 3, 'Risk score distribution (%)')
par(new = TRUE)
plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col="white")
par(new = TRUE)
plot(points.range,p.Rec*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=15, xlab="Risk score",ylab="Predicted 5-year risk (%)")
points(points.range,p.MSM*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=16)
legend(x=0,y=100,legend=c("Recurrence","Melanoma-specific mortality"),pch=c(15,16),cex=.9,bg='white')
grDevices::dev.off()
# table of risk distribution
cbind(points.range, round(p.Rec*100, 1), round(p.MSM*100, 1))[c(1, 9, 10, 12, 13),]

# save for Rshiny app
coef.Rec <- coef(f.mi.BS.Rec)
center.Rec <- f.mi.BS.Rec$center
save(h0.Rec, h0.MSM, coef.Rec, c.Breslow, center.Rec,
     file=paste0('Z:/Project Melanoom/PaperMelanoma/Results/h0.Rec.MSM.Rdata'), compress=TRUE)
load(file=paste0('Z:/Project Melanoom/PaperMelanoma/Results/h0.Rec.MSM.Rdata'))
```

# Example to check probabilities with app
```{r, eval=TRUE}
patient.nr <- 115 # 395 and 396 = Negative, 397 and 398 = Positive, 30 = Head & neck, 115 has large Breslow
data.to.be.imputed[patient.nr, c("SNstatus", "Age.SN", "Ulceration", "Loc_CAT", "Breslow", "Rdamcrit")]
lp.score.Rec.true <- predict(f.mi.BS.Rec)[patient.nr]
lp.score.Rec.try <- predict(f.mi.BS.Rec, newdata=data.frame(SNstatus=data.to.be.imputed[patient.nr,"SNstatus"],
                                                            Age.SN=data.to.be.imputed[patient.nr,"Age.SN"],
                                                            Ulceration=data.to.be.imputed[patient.nr,"Ulceration"],
                                                            Loc_CAT=data.to.be.imputed[patient.nr,"Loc_CAT"],
                                                            Breslow=data.to.be.imputed[patient.nr,"Breslow"],
                                                            Rdamcrit=data.to.be.imputed[patient.nr,"Rdamcrit"]),
                                                            type="lp")
p.Rec.example <- fun.event(lp=lp.score.Rec.try, h0=h0.Rec)
p.MSM.example <- fun.event(lp=lp.score.Rec.try, h0=h0.MSM)

# score
round((lp.score.Rec.try-int.Rec)/rc.Rec,0)

# manual calculation
coef.Rec <- coef(f.mi.BS.Rec)
lp.manual <- coef.Rec["SNstatus=Positive"]*(as.numeric(data.to.be.imputed[patient.nr,"SNstatus"])-1)+
              coef.Rec["Age.SN"]*log(data.to.be.imputed[patient.nr,"Age.SN"])+
              coef.Rec["Ulceration=Yes"]*(as.numeric(data.to.be.imputed[patient.nr,"Ulceration"])-1)+
              coef.Rec["Loc_CAT=leg"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="leg")+
              coef.Rec["Loc_CAT=trunk"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="trunk")+
              coef.Rec["Loc_CAT=head & neck"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="head & neck")+
              coef.Rec["Breslow"]*(log(data.to.be.imputed[patient.nr,"Breslow"])-c.Breslow)+
              coef.Rec["Rdamcrit"]*log(data.to.be.imputed[patient.nr,"Rdamcrit"])+
              coef.Rec["SNstatus=Positive * Breslow"]*(log(data.to.be.imputed[patient.nr,"Breslow"])-c.Breslow)*(as.numeric(data.to.be.imputed[patient.nr,"SNstatus"])-1)
center.Rec <- f.mi.BS.Rec$center

# lp
lp.score.Rec.true
lp.score.Rec.try
lp.manual-center.Rec

# probability of recurrence
fun.event(lp=lp.score.Rec.try, h0=h0.Rec)
1-exp(-h0.Rec*exp(lp.manual-center.Rec))

# check MSM
lp.score.MSM.true <- predict(f.mi.BS.MSM)[patient.nr]
lp.score.MSM.true
lp.MSM.try <- MSM.cal.fact*(lp.manual-center.Rec)
lp.MSM.try
fun.event(lp=lp.MSM.try, h0=h0.MSM)
fun.event(lp=lp.MSM.try, h0=h0.MSM)
1-exp(-h0.MSM*exp(MSM.cal.fact*(lp.manual-center.Rec)))
```

### 2022-06-08 Code/descr.table.Rmd
Initial

---
output: word_document
---

```{r setup, include=FALSE}
# clear all
rm(list = ls(all.names = TRUE))
if(!is.null(dev.list())) dev.off()
cat('\014')

library(arsenal)
library(knitr)

load('Z:/Project Melanoom/Code/PaperMelanoma/Data/descr.data.Rdata')
data.to.be.imputed$MSM <- factor(data.to.be.imputed$MSM, labels=c("Other", "Death by melanoma"))
data.to.be.imputed$multiple.fields <- factor(data.to.be.imputed$multiple.fields, labels=c("Other", "Multiple SN fields"))
descr.table <- arsenal::tableby(Center ~ SNstatus+Sex
                                +Age.SN+includeNA(Ulceration)
                                +includeNA(Loc_CAT)+includeNA(Histology)
                                +Breslow+multiple.fields
                                +Tot_SNs_neg+Tot_SNs_pos
                                +Rdamcrit+includeNA(Dewar)+includeNA(Mitosis)
                                +Recurrence+dead
                                +MSM+logRec.time+logFU.time, data=data.to.be.imputed, 
                                total.pos="before",
                                digits = 1,
                                test = F,
                                total = T,
                                numeric.test = "kwt", cat.test = "chisq",
                                numeric.stats = c("meansd", "medianq1q3", "range", "Nmiss2"),
                                cat.stats = c("countpct"),
                                stats.labels = list(
                                  meansd = "Mean (SD)",
                                  medianq1q3 = "Median (Q1, Q3)",
                                  range = "Min - Max",
                                  Nmiss2 = "Missing"
                                ))
my_labels <- list(
  SNstatus = "SN status",
  Age.SN = "Age",
  Loc_CAT = "Location",
  multiple.fields = "Multiple fields",
  Tot_SNs_neg = "Total number of negative SN",
  Tot_SNs_pos = "Total number of positive SN",
  Rdamcrit = "SN tumour burden",
  Dewar = "Location metastasis in lymph",
  dead = "Deaths",
  MSM = "Melanoma specific mortality"
)
```

Table 1. Descriptive statistics.
```{r echo=FALSE}
knitr::kable(summary(descr.table, labelTranslations=my_labels))
```

### 2022-06-08 Code/check_probabilities.R

@@ -1,0 +1,28 @@
+h0.MSM <- 0.1438848
+MSM.cal.fact <- 1.0973015
+
+rscript <- function(status, age, ulceration, location, breslow, tumor_burden) {
+  h0.Rec <- 0.2534864
+  coef.Rec <- c(1.0264235, 0.3514440, 0.4914014, 0.2621760, 0.3803103,
+                0.7062802, 0.8212041, 0.1866868, -0.3838754)
+  names(coef.Rec) <- c("SNstatus=Positive", "Age.SN", "Ulceration=Yes",
+                       "Loc_CAT=leg", "Loc_CAT=trunk", "Loc_CAT=headneck",
+                       "Breslow", "Rdamcrit", "SNstatus=Positive * Breslow")
+  c.Breslow <- 0.7154091
+  center.Rec <- 2.031039
+  lp <- coef.Rec["SNstatus=Positive"]*status+
+    coef.Rec["Age.SN"]*log(age)+
+    coef.Rec["Ulceration=Yes"]*ulceration+
+    coef.Rec["Loc_CAT=leg"]*I(location == 1)+
+    coef.Rec["Loc_CAT=trunk"]*I(location == 2)+
+    coef.Rec["Loc_CAT=headneck"]*I(location == 3)+
+    coef.Rec["Breslow"]*(log(breslow)-c.Breslow)+
+    coef.Rec["Rdamcrit"]*log(tumor_burden)+
+    coef.Rec["SNstatus=Positive * Breslow"]*(log(breslow)-c.Breslow)*status-center.Rec
+  print(as.numeric(lp))
+  p.Rec <- 1-exp(-h0.Rec*exp(lp))
+  print(as.numeric(p.Rec))
+  S <- 100 - 100*h0.Rec^exp(lp)
+  return(as.numeric(S))
+}
+rscript(0, 55, 0, 0, 7, 1)

### 2022-06-08 Code/melanoma.R

@@ -1,0 +1,1089 @@
+# clear all
+rm(list = ls(all.names = TRUE))
+if(!is.null(dev.list())) grDevices::dev.off()
+cat('\014')
+
+#####
+##### Author: C. Maas
+##### Project Melanoom
+#####
+
+# load libraries
+library(foreign)   # Reading external (spps) data
+library(mice)      # Multipe imputation
+library(survminer) # Kaplan-Meier plot
+library(prodlim)   # median follow-up time
+library(glmnet)    # LASSO
+library(rms)       # rcs
+library(Hmisc)     # rcorr.cens and fit.mult.impute
+library(mitools)   # MIcombine
+library(arsenal)   # descriptives table
+library(png)       # reading PNG images
+set.seed(100)
+
+source("./Code/table.missing.values.R") # import function to create table of missing values
+source("./Code/val_surv_mi_v3.R")
+
+#####
+##### LOAD DATA
+#####
+original.data <- foreign::read.spss(file="./Data/Dataset Predictiemodel Stassen.sav",
+                                    to.data.frame=TRUE, origin="01-01-1970")
+
+# Select necessary variables
+selected.vars <- c("EXCLUSIE", "DatabaseNR", "Center", "Sex", "Age", "DOB",
+                   "Ulceration", "Loc_CAT",            # OLD data set: Loc_CAT2
+                   "Histology", "Breslow", "Tot_nr_SNs",
+                   "nr_SNs_Field_1", "nr_SNs_Field_2", "nr_SNs_Field_3",
+                   "Mitosis",                          # needed for AJCC.7.i
+                   "SNdate", "Recurrence", "Date_1st_Rec", "Last_FU",
+                   "dead", "Status",
+                   "SNstatus",                         # distinguish between positive and negative SN patients
+                   "Tot_SNs_pos", "Rdamcrit", "Dewar") # additional for positive SN patients
+
+# Exclude based on duplicates, urogenital melanoma, etc.
+sel.data <- original.data[original.data$EXCLUSIE!="Exclusion", selected.vars] # 2 = Exclusion
[…]

### 2022-06-08 Code/table.missing.values.R

@@ -1,0 +1,8 @@
+missings.table <- function(data){
+  missing.values.table <- data.frame(m=rep(0, ncol(data)))
+  rownames(missing.values.table) <- colnames(data)
+  for (variable in colnames(data)){
+    missing.values.table[variable,] <- sum(is.na(data[,variable]))
+  }
+  print(missing.values.table[order(missing.values.table$m, decreasing=TRUE), , drop=FALSE])  
+}

### 2022-06-08 Code/val_surv_mi_v3.R

@@ -1,0 +1,199 @@
+# # Input for val.surv.mi
+# p<-as.matrix(prediction)   # matrix with predicted probabilities for imputation i in columns (complete case analyis: one column)
+# y<-Surv(time,status)  # Time to event outcome as Surv object (time,status)
+# g<-4 # number of risk groups; default=5
+# main<-"Plot label" # Default = ""
+# save_plots<-FALSE # save plots to file? default=FALSE
+# time<-30 #time point at which to evaluate the predicted probabilities; default=NULL (not entered), the maximum time point will be taken
+#
+# val.surv.mi(p=p,y=y,g=g,main=main,save_plots = save_plots,time=time)
+
+Rubin.combine<-function(est,se){
+  m<-length(est)
+  est.mi<-mean(est)
+  var.w<-mean(se^2)
+  var.b<-0
+  if (m>1) var.b<-sum((est - est.mi)^2)/(m - 1)
+  se.mi<-sqrt(var.w+(1+1/m)*var.b)
+  return(list(est=est.mi,se=se.mi))
+}
+
+val.surv.mi<-function(p,y,g=5,main="",file_path="",save_plots=FALSE,time=NULL,
+                      lim=c(0,1),dist=TRUE,CI=FALSE, df=3){
+  if (!is.null(time)){
+    y[y[,1]>time,2]<-0
+    y[y[,1]>time,1]<-time
+  }
+
+  lp<-log(-log(1-p))
+
+  n<-length(y)
+  m.imp.val<-ncol(lp)
+
+  cindex<-rep(0,m.imp.val)
+  cindex.se<-rep(0,m.imp.val)
+  slope<-rep(0,m.imp.val)
+  slope.se<-rep(0,m.imp.val)
+  int<-rep(0,m.imp.val)
+  int.se<-rep(0,m.imp.val)
+  cindex<-rep(0,m.imp.val)
+  cindex.se<-rep(0,m.imp.val)
+
+  p.groups<-array(rep(0,g*m.imp.val),dim=c(m.imp.val,g),dimnames=list(1:m.imp.val,1:g))
+  y.groups<-array(rep(0,2*g*m.imp.val),dim=c(m.imp.val,g,2),dimnames=list(1:m.imp.val,1:g,c("obs","se")))
+  lp.range<-min(lp)+0:100*(max(lp)-min(lp))/100
[…]

### 2022-06-08 Code/PaperMelanoma.Rmd
Package dependencies

@@
 # load libraries
-library(foreign)   # Reading external (spps) data
-library(mice)      # Multipe imputation
-library(survminer) # Kaplan-Meier plot
-library(prodlim)   # median follow-up time
-library(glmnet)    # LASSO
-library(rms)       # rcs
-library(Hmisc)     # rcorr.cens and fit.mult.impute
-library(mitools)   # MIcombine
-library(arsenal)   # descriptives table
-library(png)       # reading PNG images
+library(foreign)
+library(mice)
+library(survminer)
+library(prodlim)
+library(rms)
+library(Hmisc)
+library(png)
 set.seed(100)
@@
 ```{r, eval=TRUE}
-median.FU.time.total <- stats::quantile(prodlim::prodlim(Hist(exp(logFU.time), MSM)~1, data=complete.imputed.data, reverse=TRUE))
+median.FU.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM)~1, data=complete.imputed.data, reverse=TRUE))
 tab.FU.time <- c("total", paste(sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "quantile"]),
@@
                         sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "upper"]), "]")))
-median.FU.time <- stats::quantile(prodlim::prodlim(Hist(exp(logFU.time), MSM)~Center, data=complete.imputed.data, reverse=TRUE))
+median.FU.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM)~Center, data=complete.imputed.data, reverse=TRUE))
 for (center in sort(as.character(unique(sel.data$Center)))){
@@
 Rec.table <- c("Total", sum(sel.data$Recurrence=="recurred"), sprintf("%.1f", sum(sel.data$Recurrence=="recurred")/nrow(sel.data)*100))
-Rec.table <- rbind(Rec.table, c("Positive", sum(sel.data$Recurrence=="recurred"&sel.data$SNstatus=="Positive"),
+Rec.table <- rbind(Rec.table, c("Positive", 
+                                sum(sel.data$Recurrence=="recurred"&sel.data$SNstatus=="Positive"),
                                 sprintf("%.1f", sum(sel.data$Recurrence=="recurred"&sel.data$SNstatus=="Positive")/sum(sel.data$SNstatus=="Positive")*100)))
-Rec.table <- rbind(Rec.table, c("Negative", sum(sel.data$Recurrence=="recurred"&sel.data$SNstatus=="Negative"),
+Rec.table <- rbind(Rec.table, c("Negative", 
+                                sum(sel.data$Recurrence=="recurred"&sel.data$SNstatus=="Negative"),
                                 sprintf("%.1f", sum(sel.data$Recurrence=="recurred"&sel.data$SNstatus=="Negative")/sum(sel.data$SNstatus=="Negative")*100)))
@@
 for (i in 1:m){
-  imputation.i <- complete(imputed.data,i)
+  imputation.i <- mice::complete(imputed.data,i)
   TNM.8.i <- rep("", nrow(imputation.i))
@@
       # predicted lp for imputation i
-      lp.i <- predict(f.mi,newdata=complete(imputed.data,i), type="lp")
+      lp.i <- predict(f.mi,newdata=mice::complete(imputed.data,i), type="lp")
       # baseline hazard
@@
   for (i in 1:m){
-    rc <- Hmisc::rcorr.cens(-predict(f.mi,newdata=complete(imputed.data,i), type="lp"), S.MSM)
+    rc <- Hmisc::rcorr.cens(-predict(f.mi,newdata=mice::complete(imputed.data,i), type="lp"), S.MSM)
     cindex.MSM.refit[i]<-rc["C Index"]
@@
         lp.notj<-f.mi.Rec.notj$linear
-        f.MSM.notj<-cph(S.notj~lp.notj,y=TRUE,x=TRUE)
+        f.MSM.notj<-rms::cph(S.notj~lp.notj,y=TRUE,x=TRUE)
 
@@
         for (i in 1:m){
-          lp.i<-predict(f.mi.Rec.notj,newdata=complete(imputed.data,i)[data.to.be.imputed$Center==Centers[j],], type="lp")
+          lp.i<-predict(f.mi.Rec.notj,newdata=mice::complete(imputed.data,i)[data.to.be.imputed$Center==Centers[j],], type="lp")
 

### 2022-06-08 Code/melanoma.R

@@ -9,16 +9,13 @@ cat('\014')
 #####
 
 # load libraries
-library(foreign)   # Reading external (spps) data
-library(mice)      # Multipe imputation
-library(survminer) # Kaplan-Meier plot
-library(prodlim)   # median follow-up time
-library(glmnet)    # LASSO
-library(rms)       # rcs
-library(Hmisc)     # rcorr.cens and fit.mult.impute
-library(mitools)   # MIcombine
-library(arsenal)   # descriptives table
-library(png)       # reading PNG images
+library(foreign)
+library(mice)
+library(survminer)
+library(prodlim)
+library(rms)
+library(Hmisc)
+library(png)
 set.seed(100)
 
 source("./Code/table.missing.values.R") # import function to create table of missing values
@@ -153,11 +150,11 @@ survminer::ggsurvplot(center.MSM.KM, conf.int=TRUE, risk.table=TRUE)
 #####
 ##### MEDIAN FOLLOW-UP TIME
 #####
-median.FU.time.total <- stats::quantile(prodlim::prodlim(Hist(exp(logFU.time), MSM)~1, data=complete.imputed.data, reverse=TRUE))
+median.FU.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM)~1, data=complete.imputed.data, reverse=TRUE))
 tab.FU.time <- c("total", paste(sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "quantile"]),
                  paste0("[", sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "lower"]), "; ",
                         sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "upper"]), "]")))
-median.FU.time <- stats::quantile(prodlim::prodlim(Hist(exp(logFU.time), MSM)~Center, data=complete.imputed.data, reverse=TRUE))
+median.FU.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM)~Center, data=complete.imputed.data, reverse=TRUE))
 median.FU.time
 for (center in sort(as.character(unique(sel.data$Center)))){
[…]

### 2022-06-09 Code/PaperMelanoma.Rmd
Added tests

@@
 ### Recode
-Set Rdamcrit of negative SN patients equal to 0
-```{r, eval=TRUE}
+Set Rdamcrit of negative SN patients equal to 0. 11 patients with positive SN have SN tumour burden of exactly 1
+```{r, eval=TRUE}
+sum(sel.data$Rdamcrit==1&sel.data$SNstatus=="Positive"&!is.na(sel.data$Rdamcrit))
 sel.data$Rdamcrit[sel.data$SNstatus=="Negative"] <- 1 # since log(1) = 0
@@
 load('Z:/Project Melanoom/PaperMelanoma/Data/imputed.data.RData')
-complete.imputed.data <- mice::complete(imputed.data, m)
-summary(complete.imputed.data)
-missings.table(complete.imputed.data)
+last.imputed.data.set <- mice::complete(imputed.data, m)
+summary(last.imputed.data.set)
+missings.table(last.imputed.data.set)
 ```
@@
 ```{r, eval=TRUE}
-S.Rec <- survival::Surv(exp(complete.imputed.data$logRec.time), sel.data$Recurrence=="recurred")
-S.MSM <- survival::Surv(exp(complete.imputed.data$logFU.time), sel.data$MSM)
+S.Rec <- survival::Surv(exp(last.imputed.data.set$logRec.time), sel.data$Recurrence=="recurred")
+S.MSM <- survival::Surv(exp(last.imputed.data.set$logFU.time), sel.data$MSM)
 ```
@@
 ```{r, eval=TRUE}
-median.FU.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM)~1, data=complete.imputed.data, reverse=TRUE))
+median.FU.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM)~1, data=last.imputed.data.set, reverse=TRUE))
 tab.FU.time <- c("total", paste(sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "quantile"]),
@@
                         sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "upper"]), "]")))
-median.FU.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM)~Center, data=complete.imputed.data, reverse=TRUE))
+median.FU.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM)~Center, data=last.imputed.data.set, reverse=TRUE))
 for (center in sort(as.character(unique(sel.data$Center)))){
@@
 
-### Test proportional hazards assumption
-```{r, eval=TRUE}
-test.prop<-survival::cox.zph(f.mi.full.Rec$fits[[1]])
-test.prop
-plot(test.prop)
-```
-
 # Final model for recurrence selected by backward selection with p = 0.01
 ```{r, eval=TRUE}
-c.Breslow <- mean(log(complete.imputed.data$Breslow))
+c.Breslow <- mean(log(last.imputed.data.set$Breslow))
 form.BS.Rec <- S.Rec ~
@@
                            pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
-round(stats::anova(f.mi.BS.Rec)[order(stats::anova(f.mi.BS.Rec)[, 'P']),], 3)
+# round(stats::anova(f.mi.BS.Rec)[order(stats::anova(f.mi.BS.Rec)[, 'P']),], 3)
+stats::anova(f.mi.BS.Rec)
 ```
@@
                                   pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
-round(anova(f.mi.BS.MSM.refit)[order(anova(f.mi.BS.MSM.refit)[, 'P']),], 3)
+# round(anova(f.mi.BS.MSM.refit)[order(anova(f.mi.BS.MSM.refit)[, 'P']),], 3)
 f.mi.BS.MSM.refit
@@
 ```{r, eval=TRUE}
-for (model in c('Rec', 'MSM', 'MSM.refit')){
-  if (model=="MSM"){
-    mult.factor <- MSM.cal.fact # if MSM, multiply by MSM.cal.fact
-    f.mi.BS <- eval(parse(text=paste0('f.mi.BS.Rec'))) # use recurrence model if calibrated MSM
-  }
-  else{
-    mult.factor <- 1 # if Rec or MSM.refit, don't multiply
+for (model in c('Rec', 'MSM.refit')){
   f.mi.BS <- eval(parse(text=paste0('f.mi.BS.', model)))
-  }
-
-  CI.lower.neg <- exp(mult.factor*confint(f.mi.BS)[,1])
-  CI.upper.neg <- exp(mult.factor*confint(f.mi.BS)[,2])
+
+  CI.lower.neg <- exp(confint(f.mi.BS)[,1])
+  CI.upper.neg <- exp(confint(f.mi.BS)[,2])
   CI.neg <- paste0("[", sprintf("%.2f", CI.lower.neg), "; ", sprintf("%.2f", CI.upper.neg), "]")
@@
 
-  CI.pos <- paste0("[", sprintf("%.2f", exp(mult.factor*confint(f.mi.BS)[7,1])*exp(mult.factor*confint(f.mi.BS)[9,1])),
-                   "; ", sprintf("%.2f", exp(mult.factor*confint(f.mi.BS)[7,2])*exp(mult.factor*confint(f.mi.BS)[9,2])), "]")
-
-  BS.table <- cbind(c(NA, sprintf("%.2f", exp(mult.factor*f.mi.BS$coefficients[2:7])), NA), # COEF NEG
+  CI.pos <- paste0("[", sprintf("%.2f", exp(confint(f.mi.BS)[7,1])*exp(confint(f.mi.BS)[9,1])),
+                   "; ", sprintf("%.2f", exp(confint(f.mi.BS)[7,2])*exp(confint(f.mi.BS)[9,2])), "]")
+
+  BS.table <- cbind(c(NA, sprintf("%.2f", exp(f.mi.BS$coefficients[2:7])), NA), # COEF NEG
                     c(NA, CI.neg[2:7], NA), # CI NEG
-                    sprintf("%.2f", exp(mult.factor*f.mi.BS$coefficients[1:8])*c(rep(1, 6), exp(mult.factor*f.mi.BS$coefficients[9]), 1)), #COEF POS
+                    sprintf("%.2f", exp(f.mi.BS$coefficients[1:8])*c(rep(1, 6), exp(f.mi.BS$coefficients[9]), 1)), #COEF POS
                     c(CI.neg[1:6], CI.pos, CI.neg[8]), # CI POS
@@
 
-# Plot calibration of all centers combined and all centers seperate for recurrence and melanoma specific mortality refit and write results to Excel
+# Plot calibration of all centers combined and all centers seperate 
+for recurrence and melanoma specific mortality refit 
+write results to Excel
 ```{r, eval=TRUE}
@@
 final.BS.table.Rec[is.na(final.BS.table.Rec)] <- ""
-final.BS.table.MSM[is.na(final.BS.table.MSM)] <- ""
-final.BS.table.MSM["C-index MSM",] <- final.BS.table.Rec[nrow(final.BS.table.Rec),]
 final.BS.table.MSM.refit[is.na(final.BS.table.MSM.refit)] <- ""
 
+# set rownames
+final.full.table.Rec <- as.data.frame(final.full.table.Rec)
+row.names(final.full.table.Rec) <- c("Positive SN status", "Male",
+                                    "Age", "Ulceration", "Location",
+                                    "Arm", "Leg", "Trunk", "Head and neck",
+                                    "Histology", "SSM", "NM", "ALM", "Other",
+                                    "Breslow", "Multiple fields", 
+                                    "Total number of nodes",
+                                    "SN tumour burden", 
+                                    "Location metastasis in lymph node",
+                                    "Subcap", "Combined", "Parenchymal",
+                                    "Multifocal", "Extensive",
+                                    rownames(final.full.table.Rec)[25:26])
+final.full.table.MSM.refit <- as.data.frame(final.full.table.MSM.refit)
+row.names(final.full.table.MSM.refit) <- c("Positive SN status", "Male",
+                                          "Age", "Ulceration", "Location",
+                                          "Arm", "Leg", "Trunk", "Head and neck",
+                                          "Histology", "SSM", "NM", "ALM", "Other",
+                                          "Breslow", "Multiple fields", 
+                                          "Total number of nodes",
+                                          "SN tumour burden", 
+                                          "Location metastasis in lymph node",
+                                          "Subcap", "Combined", "Parenchymal",
+                                          "Multifocal", "Extensive",
+                                          row.names(final.full.table.Rec)[25])
+final.BS.table.Rec <- as.data.frame(final.BS.table.Rec)
+row.names(final.BS.table.Rec) <- c("Positive SN status", "Age", "Ulceration",
+                                   "Location", "Arm", "Leg", "Trunk", 
+                                   "Head and neck",
+                                   "Breslow", "SN tumour burden", 
+                                   row.names(final.BS.table.Rec)[11:12])
+final.BS.table.MSM.refit <- as.data.frame(final.BS.table.MSM.refit)
+row.names(final.BS.table.MSM.refit) <- c("Positive SN status", "Age", "Ulceration",
+                                   "Location", "Arm", "Leg", "Trunk", 
+                                   "Head and neck",
+                                   "Breslow", "SN tumour burden", 
+                                   row.names(final.BS.table.MSM.refit)[11])
+
 # save tables
-write.table(final.full.table.Rec,file='Z:/Project Melanoom/PaperMelanoma/Results/final.full.table.Rec.txt')
-write.table(final.full.table.MSM.refit,file='Z:/Project Melanoom/PaperMelanoma/Results/final.full.table.MSM.refit.txt')
-write.table(final.BS.table.Rec,file='Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.Rec.txt')
-write.table(final.BS.table.MSM,file='Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.MSM.txt')
-write.table(final.BS.table.MSM.refit,file='Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.MSM.refit.txt')
+write.table(final.full.table.Rec,
+            file='Z:/Project Melanoom/PaperMelanoma/Results/final.full.table.Rec.txt', sep=",")
+write.table(final.full.table.MSM.refit,
+            file='Z:/Project Melanoom/PaperMelanoma/Results/final.full.table.MSM.refit.txt', sep=",")
+write.table(final.BS.table.Rec,
+            file='Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.Rec.txt', sep=",")
+write.table(final.BS.table.MSM.refit,
+            file='Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.MSM.refit.txt', sep=",")
+
+# show tables
 final.full.table.Rec
@@
 final.BS.table.Rec
-final.BS.table.MSM
 final.BS.table.MSM.refit
@@
 nom
-
-# Slope and intercept of recurrence and MSM
-int.Rec
-# check that int is same as this:
-log(20)*f.mi.BS.Rec$coef["Age.SN"]+log(0.1)*f.mi.BS.Rec$coef["Breslow"]+log(0.01)*f.mi.BS.Rec$coef["Rdamcrit"]-f.mi.BS.Rec$center
 ```
@@
 
-Breslow.trunc <- complete.imputed.data$Breslow
-Breslow.trunc[complete.imputed.data$Breslow<0.1] <- .1
-Breslow.trunc[complete.imputed.data$Breslow>7] <- 7
-lp.score.Rec <- predict(f.mi.BS.Rec,newdata=data.frame(SNstatus=complete.imputed.data$SNstatus,
-                                                Age.SN=complete.imputed.data$Age.SN,
-                                                Ulceration=complete.imputed.data$Ulceration,
-                                                Loc_CAT=complete.imputed.data$Loc_CAT,
+Breslow.trunc <- last.imputed.data.set$Breslow
+Breslow.trunc[last.imputed.data.set$Breslow<0.1] <- .1
+Breslow.trunc[last.imputed.data.set$Breslow>7] <- 7
+lp.score.Rec <- predict(f.mi.BS.Rec,newdata=data.frame(SNstatus=last.imputed.data.set$SNstatus,
+                                                Age.SN=last.imputed.data.set$Age.SN,
+                                                Ulceration=last.imputed.data.set$Ulceration,
+                                                Loc_CAT=last.imputed.data.set$Loc_CAT,
                                                 Breslow=Breslow.trunc,
-                                                Rdamcrit=complete.imputed.data$Rdamcrit))
+                                                Rdamcrit=last.imputed.data.set$Rdamcrit))
 mean(lp.score.Rec)
@@
 data.frame(table(score.Rec)/length(score.Rec))
-```
-
-# Check mean event rates of risk distribution
-```{r, eval=TRUE}
-# Recurrence
-sum(h.Rec$density*fun.event(lp=lp.points.Rec, h0=h0.Rec))
-# equal to the 5-year survival probability, kaplanmeier op 5 jaar, survfit op 5 jaar
-SF<-survfit(S.Rec~1,data=sel.data)
-100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
-mean(fun.event(lp=lp.score.Rec, h0=h0.Rec))
-
-# MSM
-sum(h.Rec$density*fun.event(lp=lp.points.MSM, h0=h0.MSM))
-SF<-survfit(S.MSM~1,data=sel.data)
-100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
-mean(fun.event(lp=lp.MSM, h0=h0.MSM))
 ```
@@
 grDevices::dev.off()
+```
+
+```{r, eval=TRUE, echo=FALSE}
+par(mar = c(5,5,2,5))
+plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col='white')
+abline(h=(0:30)*1,col="light grey",lty=2)
+axis(side = 4)
+mtext(side = 4, line = 3, 'Risk score distribution (%)')
+par(new = TRUE)
+plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col="white")
+
+par(new = TRUE)
+plot(points.range,p.Rec*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=15, xlab="Risk score",ylab="Predicted 5-year risk (%)")
+points(points.range,p.MSM*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=16)
+legend(x=0,y=100,legend=c("Recurrence","Melanoma-specific mortality"),pch=c(15,16),cex=.9,bg='white')
+```
+
+```{r, eval=TRUE}
+# reset margins
+par(mar = c(5,5,2,5))
+
 # table of risk distribution
@@
 # save for Rshiny app
-coef.Rec <- coef(f.mi.BS.Rec)
+coef.Rec <- stats::coef(f.mi.BS.Rec)
 center.Rec <- f.mi.BS.Rec$center
@@
 
-# Example to check probabilities with app
+# Test if model is correct
+### Check calibration
+```{r, eval=TRUE}
+# Slope and intercept of recurrence and MSM
+int.Rec
+# check that int is same as this:
+log(20)*f.mi.BS.Rec$coef["Age.SN"]+log(0.1)*f.mi.BS.Rec$coef["Breslow"]+log(0.01)*f.mi.BS.Rec$coef["Rdamcrit"]-f.mi.BS.Rec$center
+```
+
+### Check mean event rates of risk distribution
+```{r, eval=TRUE}
+# Recurrence
+sum(h.Rec$density*fun.event(lp=lp.points.Rec, h0=h0.Rec))
+# equal to the 5-year survival probability, kaplanmeier op 5 jaar, survfit op 5 jaar
+SF<-survfit(S.Rec~1,data=sel.data)
+100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
+mean(fun.event(lp=lp.score.Rec, h0=h0.Rec))
+
+# MSM
+sum(h.Rec$density*fun.event(lp=lp.points.MSM, h0=h0.MSM))
+SF<-survfit(S.MSM~1,data=sel.data)
+100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
+mean(fun.event(lp=lp.MSM, h0=h0.MSM))
+```
+
+### Example to check probabilities with app
 ```{r, eval=TRUE}
@@
 # manual calculation
-coef.Rec <- coef(f.mi.BS.Rec)
+coef.Rec <- stats::coef(f.mi.BS.Rec)
 lp.manual <- coef.Rec["SNstatus=Positive"]*(as.numeric(data.to.be.imputed[patient.nr,"SNstatus"])-1)+
@@
 ```
+
+### Test with old tool predictions recurrence for negative SN status patients
+```{r, eval=TRUE}
+rscript.negative <- function(breslow, ulceration, location) {
+  S0 <- 0.85837329
+  lp <- 0.79351989*log(breslow) +
+    0.61912477*I(ulceration == 1) +
+    0.29594843*I(location == 1) +
+    0.42802046*I(location == 2) +
+    0.86053044*I(location == 3) - 0.98860941
+  S <- 100 - 100*S0^exp(lp)
+  return(S)
+}
+old.negativeSN.tool <- rscript.negative(breslow=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Breslow"],
+        ulceration=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Ulceration"])-1,
+        location=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Loc_CAT"])-1)
+lp.negative <- predict(f.mi.BS.Rec, newdata=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative",], type="lp")
+new.negativeSN.tool <- 1-exp(-h0.Rec*exp(lp.negative))
+plot(x=fun.event(lp=new.negativeSN.tool, h0=h0.Rec)*100, y=old.negativeSN.tool,
+     xlim=c(0, 100), ylim=c(0, 100),
+     xlab="New predictions", ylab="Old predictions",
+     main="Predicted recurrence of patients with negative SN status")
+abline(a=0, b=1, col="red")
+```
+
+### Test with old tool predictions recurrence for positive SN status patients
+```{r, eval=TRUE}
+rscript.positive <- function(age, ulceration, tumor_burden, breslow) {
+  S0 <- 0.51762362
+  lp <- 0.56109945 * log(age) +
+    0.23383770 * log(tumor_burden) +
+    0.35489235 * log(breslow) +
+    0.36562098 * I(ulceration == 1) - 2.709225
+  S <- 100 - 100 * S0^exp(lp)
+  return(S)
+}
+old.positiveSN.tool <- rscript.positive(age=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Age.SN"],
+                               ulceration=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Ulceration"])-1,
+                               breslow=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Breslow"],
+                               tumor_burden=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Rdamcrit"]))
+lp.positive <- predict(f.mi.BS.Rec, newdata=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive",], type="lp")
+new.positiveSN.tool <- 1-exp(-h0.Rec*exp(lp.positive))
+plot(x=fun.event(lp=new.positiveSN.tool, h0=h0.Rec)*100, y=old.positiveSN.tool,
+     xlim=c(0, 100), ylim=c(0, 100),
+     xlab="New predictions", ylab="Old predictions",
+     main="Predicted recurrence of patients with positive SN status")
+abline(a=0, b=1, col="red")
+```
+
+### Histogram for negative SN status patients
+```{r, eval=TRUE}
+Breslow.trunc.neg <- last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Breslow"]
+Breslow.trunc.neg[last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Breslow"]<0.1] <- .1
+Breslow.trunc.neg[last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Breslow"]>7] <- 7
+lp.score.Rec.negative <- predict(f.mi.BS.Rec,newdata=data.frame(SNstatus=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "SNstatus"],
+                                                       Age.SN=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Age.SN"],
+                                                       Ulceration=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Ulceration"],
+                                                       Loc_CAT=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Loc_CAT"],
+                                                       Breslow=Breslow.trunc.neg,
+                                                       Rdamcrit=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Rdamcrit"]))
+mean(lp.score.Rec.negative)
+# OK to use int.Rec and rc.Rec based on all patients?
+score.Rec.neg <- round((lp.score.Rec.negative-int.Rec)/rc.Rec,0)
+h.Rec.neg <- hist(score.Rec.neg,plot=FALSE,breaks=0:max.breaks,right=FALSE)
+h.plot.neg <- h.Rec.neg
+h.plot.neg$density <- 100*h.Rec.neg$density
+```
+
+### Histogram for positive SN status patients
+```{r, eval=TRUE}
+Breslow.trunc.pos <- last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Breslow"]
+Breslow.trunc.pos[last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Breslow"]<0.1] <- .1
+Breslow.trunc.pos[last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Breslow"]>7] <- 7
+lp.score.Rec.positive <- predict(f.mi.BS.Rec,newdata=data.frame(SNstatus=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "SNstatus"],
+                                                                Age.SN=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Age.SN"],
+                                                                Ulceration=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Ulceration"],
+                                                                Loc_CAT=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Loc_CAT"],
+                                                                Breslow=Breslow.trunc.pos,
+                                                                Rdamcrit=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Rdamcrit"]))
+mean(lp.score.Rec.positive)
+# OK to use int.Rec and rc.Rec based on all patients?
+score.Rec.pos <- round((lp.score.Rec.positive-int.Rec)/rc.Rec,0)
+h.Rec.pos <- hist(score.Rec.pos,plot=FALSE,breaks=0:max.breaks,right=FALSE)
+h.plot.pos <- h.Rec.pos
+h.plot.pos$density <- 100*h.Rec.pos$density
+
+par(mfrow=c(1, 3))
+plot(h.plot,freq=FALSE,axes=TRUE, xlab="Risk score", xlim=x.lim,ylim=y.lim.hist,
+     ylab="Risk score distribution (%)", main="All patients",col="white")
+plot(h.plot.neg,freq=FALSE,axes=TRUE, xlab="Risk score", xlim=x.lim,
+     ylab="Risk score distribution (%)",ylim=y.lim.hist,main="Negative SN status",col="white")
+plot(h.plot.pos,freq=FALSE,axes=TRUE, xlab="Risk score", xlim=x.lim,
+     ylab="Risk score distribution (%)",ylim=y.lim.hist,main="Positive SN status",col="white")
+```
+
+### Test proportional hazards assumption
+```{r, eval=TRUE}
+test.prop<-survival::cox.zph(f.mi.full.Rec$fits[[1]])
+test.prop
+plot(test.prop)
+```

### 2022-06-09 Code/PaperMelanoma.R

@@ -47,6 +47,8 @@ cat("Number of patients excluded:", nrow(original.data)-nrow(sel.data), "\n")
 # Omit patient A0523 - no last follow-up date
 sel.data <- sel.data[sel.data$DatabaseNR!="A0523               ",]
 
+# TODO: 11 patients with positive SN have SN tumour burden of exactly 1
+sum(sel.data$Rdamcrit==1&sel.data$SNstatus=="Positive"&!is.na(sel.data$Rdamcrit))
 # Set Rdamcrit of negative SN patients equal to 0
 sel.data$Rdamcrit[sel.data$SNstatus=="Negative"] <- 1 # since log(1) = 0
 
@@ -129,15 +131,15 @@ m <- 5
 
 # OR LOAD
 load('./Data/imputed.data.RData')
-complete.imputed.data <- mice::complete(imputed.data, m)
-summary(complete.imputed.data)
-missings.table(complete.imputed.data)
+last.imputed.data.set <- mice::complete(imputed.data, m)
+summary(last.imputed.data.set)
+missings.table(last.imputed.data.set)
 
 #####
 ##### SET SURVIVAL
 #####
-S.Rec <- survival::Surv(exp(complete.imputed.data$logRec.time), sel.data$Recurrence=="recurred")
-S.MSM <- survival::Surv(exp(complete.imputed.data$logFU.time), sel.data$MSM)
+S.Rec <- survival::Surv(exp(last.imputed.data.set$logRec.time), sel.data$Recurrence=="recurred")
+S.MSM <- survival::Surv(exp(last.imputed.data.set$logFU.time), sel.data$MSM)
 
 #####
 ##### KAPLAN MEIER PLOTS
@@ -150,11 +152,11 @@ survminer::ggsurvplot(center.MSM.KM, conf.int=TRUE, risk.table=TRUE)
 #####
 ##### MEDIAN FOLLOW-UP TIME
 #####
-median.FU.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM)~1, data=complete.imputed.data, reverse=TRUE))
+median.FU.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM)~1, data=last.imputed.data.set, reverse=TRUE))
 tab.FU.time <- c("total", paste(sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "quantile"]),
[…]

### 2022-06-09 Code/check_probabilities.R

@@ -19,10 +19,7 @@ rscript <- function(status, age, ulceration, location, breslow, tumor_burden) {
     coef.Rec["Breslow"]*(log(breslow)-c.Breslow)+
     coef.Rec["Rdamcrit"]*log(tumor_burden)+
     coef.Rec["SNstatus=Positive * Breslow"]*(log(breslow)-c.Breslow)*status-center.Rec
-  print(as.numeric(lp))
   p.Rec <- 1-exp(-h0.Rec*exp(lp))
-  print(as.numeric(p.Rec))
-  S <- 100 - 100*h0.Rec^exp(lp)
-  return(as.numeric(S))
+  return(as.numeric(p.Rec))
 }
 rscript(0, 55, 0, 0, 7, 1)

### 2022-06-14 Code/PaperMelanoma.Rmd
Final checks

@@
 ### Exclude patients
-Exclude when Exclusion = 2: based on duplicates, urogenital melanoma, etc.
-Leave the patients included that have EXCLUSIE = 1 and Dbstatus = 8,
-even though diagnosis and operation date are very close.
+Exclude when Exclusion = 2: based on duplicates, urogenital melanoma, etc. Note:
+Leave the patients included that have EXCLUSIE = 1 and Dbstatus = 8 (Dbstatus=0: 
+negative, Dbstatus=1: positive), even though diagnosis and operation date are 
+very close.
 ```{r, eval=TRUE}
@@
 ```{r, eval=TRUE}
+sel.data[which(is.na(sel.data$Last_FU)), "DatabaseNR"]
 sel.data <- sel.data[sel.data$DatabaseNR!="A0523               ",]
@@
 ### Recode
-Set Rdamcrit of negative SN patients equal to 0. 11 patients with positive SN have SN tumour burden of exactly 1
+#### Use age at SN instead of age at diagnosis
+```{r, eval=TRUE}
+sel.data$Age.SN <- floor((sel.data$SNdate-sel.data$DOB)/(24*60*60)/365.25)
+```
+
+#### Location
+Merge location other with trunk
+```{r, eval=TRUE}
+sel.data$Loc_CAT[sel.data$Loc_CAT=="other"] <- "trunk"
+```
+
+Merge LMM other with other
+```{r, eval=TRUE}
+sel.data$Histology[sel.data$Histology=="LMM"] <- "other"
+```
+
+#### Define multiple fields
+```{r, eval=TRUE}
+sel.data$multiple.fields <- ifelse(!is.na(sel.data$nr_SNs_Field_2)&sel.data$nr_SNs_Field_2>0,1,0)
+```
+
+#### Number of positive and negative nodes
+Check: there are no negative patients that have positive nodes
+```{r, eval=TRUE}
+sum(!is.na(sel.data$Tot_SNs_pos)&sel.data$Tot_SNs_pos>0&sel.data$SNstatus=="Negative")
+```
+
+Create number of negative nodes removed
+```{r, eval=TRUE}
+sel.data$Tot_SNs_neg <- sel.data$Tot_nr_SNs - sel.data$Tot_SNs_pos
+```
+
+467 positive SN patients have negative nodes
+```{r, eval=TRUE}
+sum(!is.na(sel.data$Tot_SNs_neg)&sel.data$Tot_SNs_neg>0&sel.data$SNstatus=="Positive")
+```
+
+#### SN tumour burden
+11 patients with positive SN have SN tumour burden of exactly 1
 ```{r, eval=TRUE}
 sum(sel.data$Rdamcrit==1&sel.data$SNstatus=="Positive"&!is.na(sel.data$Rdamcrit))
+```
+
+Set Rdamcrit of negative SN patients equal to 0
+```{r, eval=TRUE}
 sel.data$Rdamcrit[sel.data$SNstatus=="Negative"] <- 1 # since log(1) = 0
@@
 
-Merge single cells with subcap and set Dewar for negative patients to subcap
+#### Location of metastasis
+Set Dewar for negative patients to subcap
+```{r, eval=TRUE}
+sum(sel.data$SNstatus=="Positive"&sel.data$Dewar=="subcap"&!is.na(sel.data$Dewar))
+sel.data$Dewar[sel.data$SNstatus=="Negative"] <- "subcap"
+```
+
+Merge single cells with subcap 
 ```{r, eval=TRUE}
 sel.data$Dewar[sel.data$Dewar=="single cells"] <- "subcap"
-sel.data$Dewar[sel.data$SNstatus=="Negative"] <- "subcap"
-```
-
-Merge location other with trunk
-```{r, eval=TRUE}
-sel.data$Loc_CAT[sel.data$Loc_CAT=="other"] <- "trunk"
-```
-
-Merge LMM other with other
-```{r, eval=TRUE}
-sel.data$Histology[sel.data$Histology=="LMM"] <- "other"
-```
-
-Melanoma specific mortality (MSM)
-```{r, eval=TRUE}
-sel.data$MSM<-1*(sel.data$Status=="DOD")
-```
-
-Define multiple fields
-```{r, eval=TRUE}
-sel.data$multiple.fields <- ifelse(!is.na(sel.data$nr_SNs_Field_2)&sel.data$nr_SNs_Field_2>0,1,0)
-```
-
-Create number of negative nodes removed
-```{r, eval=TRUE}
-sel.data$Tot_SNs_neg <- sel.data$Tot_nr_SNs - sel.data$Tot_SNs_pos
-```
-
-Use age at SN instead of age at diagnosis
-```{r, eval=TRUE}
-sel.data$Age.SN <- floor((sel.data$SNdate-sel.data$DOB)/(24*60*60)/365.25)
-```
-
+```
+
+#### Recurrence 
 Define time since diagnosis in years (follow-up time)
@@
 
+#### Melanoma specific mortality (MSM)
+```{r, eval=TRUE}
+sel.data$MSM<-1*(sel.data$Status=="DOD")
+```
+
+#### Mitosis
 Set Mitosis to Unknown if missing
@@
 
-Replace "Unknown" by NA
+#### Replace "Unknown" by NA
 ```{r, eval=TRUE}
@@
 
-Delete unused levels
+#### Delete unused levels
 ```{r, eval=TRUE}
@@
 
-### Restrict survival at 5 years
+# Restrict survival at 5 years
 ```{r, eval=TRUE}
@@
 f.mi.BS.MSM<-rms::cph(S.MSM~lp.BS,y=TRUE,x=TRUE)
+MSM.cal.fact <- stats::coef(f.mi.BS.MSM)
+MSM.cal.fact
+stats::confint(f.mi.BS.MSM)
 utils::write.table(paste0(sprintf("%.2f", stats::coef(f.mi.BS.MSM)), " [",
@@
             file='Z:/Project Melanoom/PaperMelanoma/Results/calibration.MSS.txt', row.names=FALSE, col.names=FALSE)
-MSM.cal.fact <- stats::coef(f.mi.BS.MSM)
-stats::coef(f.mi.BS.MSM)
-stats::confint(f.mi.BS.MSM)
 ```
@@
 for recurrence and melanoma specific mortality refit 
-write results to Excel
 ```{r, eval=TRUE}
@@
 
-# Baseline hazard
+# baseline hazard of recurrence
+f.Rec.basehaz <- basehaz(f.mi.BS.Rec,TRUE)
+h0.Rec <- f.Rec.basehaz$hazard[f.Rec.basehaz$time==max(f.Rec.basehaz$time[f.Rec.basehaz$time<=horizon])]
+
+# lp range Recurrence
 rc.Rec <- (1/attributes(nom.0)$info$sc)
 int.Rec <- attributes(nom.0)$info$Intercept
-f.Rec.basehaz <- basehaz(f.mi.BS.Rec,TRUE)
-h0.Rec <- f.Rec.basehaz$hazard[f.Rec.basehaz$time==max(f.Rec.basehaz$time[f.Rec.basehaz$time<=horizon])]
 lp.range <- int.Rec+rc.Rec*(0:10)*2
+
+# nomogram
 par(mar = rep(0, 4))
@@
 ```{r, eval=TRUE}
+# baseline hazard of MSM
+f.MSM.basehaz <- basehaz(f.mi.BS.MSM,TRUE)
+h0.MSM <- f.MSM.basehaz$hazard[f.MSM.basehaz$time==max(f.MSM.basehaz$time[f.MSM.basehaz$time<=horizon])]
+lp.MSM <- f.mi.BS.MSM$linear
+
+# lp range MSM
 rc.MSM <- f.mi.BS.MSM$coef[1]*rc.Rec
 int.MSM <- f.mi.BS.MSM$coef[1]*int.Rec
-
 max.range <- 16
@@
 
-f.MSM.basehaz <- basehaz(f.mi.BS.MSM,TRUE)
-h0.MSM <- f.MSM.basehaz$hazard[f.MSM.basehaz$time==max(f.MSM.basehaz$time[f.MSM.basehaz$time<=horizon])]
-lp.MSM <- f.mi.BS.MSM$linear
-
+# Recurrence lp scores
 Breslow.trunc <- last.imputed.data.set$Breslow
@@
 plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col="white")
-
 par(new = TRUE)
@@
 ### Check calibration
-```{r, eval=TRUE}
-# Slope and intercept of recurrence and MSM
+Intercept of recurrence should be the same as when setting covariates to 'zero' values minus the center
+```{r, eval=TRUE}
 int.Rec
-# check that int is same as this:
-log(20)*f.mi.BS.Rec$coef["Age.SN"]+log(0.1)*f.mi.BS.Rec$coef["Breslow"]+log(0.01)*f.mi.BS.Rec$coef["Rdamcrit"]-f.mi.BS.Rec$center
+log(20)*f.mi.BS.Rec$coef["Age.SN"]+
+  (log(0.1)-c.Breslow)*f.mi.BS.Rec$coef["Breslow"]+
+  log(0.01)*f.mi.BS.Rec$coef["Rdamcrit"]-
+  f.mi.BS.Rec$center
 ```
@@
 ### Check mean event rates of risk distribution
-```{r, eval=TRUE}
-# Recurrence
+#### Recurrence
+```{r, eval=TRUE}
 sum(h.Rec$density*fun.event(lp=lp.points.Rec, h0=h0.Rec))
-# equal to the 5-year survival probability, kaplanmeier op 5 jaar, survfit op 5 jaar
+# equal to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
 SF<-survfit(S.Rec~1,data=sel.data)
 100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
+# average probability of recurrence
 mean(fun.event(lp=lp.score.Rec, h0=h0.Rec))
-
-# MSM
+```
+
+#### MSM
+```{r, eval=TRUE}
 sum(h.Rec$density*fun.event(lp=lp.points.MSM, h0=h0.MSM))
+# equal to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
 SF<-survfit(S.MSM~1,data=sel.data)
 100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
+# average probability of MSM
 mean(fun.event(lp=lp.MSM, h0=h0.MSM))
@@
 
-# score
-round((lp.score.Rec.try-int.Rec)/rc.Rec,0)
-
 # manual calculation
-coef.Rec <- stats::coef(f.mi.BS.Rec)
-lp.manual <- coef.Rec["SNstatus=Positive"]*(as.numeric(data.to.be.imputed[patient.nr,"SNstatus"])-1)+
+lp.score.Rec.manual <- coef.Rec["SNstatus=Positive"]*(as.numeric(data.to.be.imputed[patient.nr,"SNstatus"])-1)+
               coef.Rec["Age.SN"]*log(data.to.be.imputed[patient.nr,"Age.SN"])+
@@
               coef.Rec["SNstatus=Positive * Breslow"]*(log(data.to.be.imputed[patient.nr,"Breslow"])-c.Breslow)*(as.numeric(data.to.be.imputed[patient.nr,"SNstatus"])-1)
-center.Rec <- f.mi.BS.Rec$center
-
-# lp
+
+# lp check
 lp.score.Rec.true
 lp.score.Rec.try
-lp.manual-center.Rec
-
-# probability of recurrence
+lp.score.Rec.manual-center.Rec
+
+# probability of recurrence with fun.event
 fun.event(lp=lp.score.Rec.try, h0=h0.Rec)
-1-exp(-h0.Rec*exp(lp.manual-center.Rec))
-
-# check MSM
+# probability of recurrence manually
+1-exp(-h0.Rec*exp(lp.score.Rec.manual-center.Rec))
+
+# check calculation of lp of MSM
 lp.score.MSM.true <- predict(f.mi.BS.MSM)[patient.nr]
 lp.score.MSM.true
-lp.MSM.try <- MSM.cal.fact*(lp.manual-center.Rec)
+lp.MSM.try <- MSM.cal.fact*(lp.score.Rec.manual-center.Rec)
 lp.MSM.try
+
+# probability of MSM with fun.event
 fun.event(lp=lp.MSM.try, h0=h0.MSM)
-fun.event(lp=lp.MSM.try, h0=h0.MSM)
-1-exp(-h0.MSM*exp(MSM.cal.fact*(lp.manual-center.Rec)))
+# probability of MSM manually
+1-exp(-h0.MSM*exp(MSM.cal.fact*(lp.score.Rec.manual-center.Rec)))
 ```
@@
 new.negativeSN.tool <- 1-exp(-h0.Rec*exp(lp.negative))
-plot(x=fun.event(lp=new.negativeSN.tool, h0=h0.Rec)*100, y=old.negativeSN.tool,
+png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/new.old.negatives.png"))
+plot(x=new.negativeSN.tool*100, y=old.negativeSN.tool,
+     xlim=c(0, 100), ylim=c(0, 100),
+     xlab="New predictions", ylab="Old predictions",
+     main="Predicted recurrence of patients with negative SN status")
+abline(a=0, b=1, col="red")
+grDevices::dev.off()
+```
+
+```{r, eval=TRUE, echo=FALSE}
+plot(x=new.negativeSN.tool*100, y=old.negativeSN.tool,
      xlim=c(0, 100), ylim=c(0, 100),
@@
 new.positiveSN.tool <- 1-exp(-h0.Rec*exp(lp.positive))
-plot(x=fun.event(lp=new.positiveSN.tool, h0=h0.Rec)*100, y=old.positiveSN.tool,
+png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/new.old.positives.png"))
+plot(x=new.positiveSN.tool*100, y=old.positiveSN.tool,
+     xlim=c(0, 100), ylim=c(0, 100),
+     xlab="New predictions", ylab="Old predictions",
+     main="Predicted recurrence of patients with positive SN status")
+abline(a=0, b=1, col="red")
+grDevices::dev.off()
+```
+
+```{r, eval=TRUE, echo=FALSE}
+plot(x=new.positiveSN.tool*100, y=old.positiveSN.tool,
      xlim=c(0, 100), ylim=c(0, 100),

### 2022-06-14 Code/PaperMelanoma.R

@@ -45,9 +45,10 @@ cat("Number of patients excluded:", nrow(original.data)-nrow(sel.data), "\n")
 # because diagnosis and operation date are very close but OK
 
 # Omit patient A0523 - no last follow-up date
+sel.data[which(is.na(sel.data$Last_FU)), "DatabaseNR"]
 sel.data <- sel.data[sel.data$DatabaseNR!="A0523               ",]
 
-# TODO: 11 patients with positive SN have SN tumour burden of exactly 1
+# 11 patients with positive SN have SN tumour burden of exactly 1
 sum(sel.data$Rdamcrit==1&sel.data$SNstatus=="Positive"&!is.na(sel.data$Rdamcrit))
 # Set Rdamcrit of negative SN patients equal to 0
 sel.data$Rdamcrit[sel.data$SNstatus=="Negative"] <- 1 # since log(1) = 0
@@ -68,8 +69,13 @@ sel.data$MSM<-1*(sel.data$Status=="DOD")
 # Define multiple fields
 sel.data$multiple.fields <- ifelse(!is.na(sel.data$nr_SNs_Field_2)&sel.data$nr_SNs_Field_2>0,1,0)
 
+# no negative patients have positive nodes
+sum(!is.na(sel.data$Tot_SNs_pos)&sel.data$Tot_SNs_pos>0&sel.data$SNstatus=="Negative")
+
 # Create number of negative nodes removed
 sel.data$Tot_SNs_neg <- sel.data$Tot_nr_SNs - sel.data$Tot_SNs_pos
+# 467 positive SN patients have negative nodes
+sum(!is.na(sel.data$Tot_SNs_neg)&sel.data$Tot_SNs_neg>0&sel.data$SNstatus=="Positive")
 
 # Use age at SN instead of age at diagnosis
 sel.data$Age.SN <- floor((sel.data$SNdate-sel.data$DOB)/(24*60*60)/365.25)
@@ -1038,15 +1044,13 @@ save(MSM.cal.fact, h0.Rec, h0.MSM, coef.Rec, c.Breslow, center.Rec,
 load(file=paste0('./Results/h0.Rec.MSM.Rdata'))
 
 # Test probabilities
-### Test proportional hazards assumption
-test.prop<-survival::cox.zph(f.mi.full.Rec$fits[[1]])
-test.prop
-plot(test.prop)
-
 # Slope and intercept of recurrence and MSM
 int.Rec
[…]

### 2022-06-20 Code/PaperMelanoma.Rmd
Limit follow-up to 5 years correctly

@@
 ```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
+# for reproducibility
+set.seed(100)
+
 # load libraries
@@
 library(png)
-set.seed(100)
 # import function to create table of missing values
@@
 
-### Exclude patients
+### Inclusion criteria
 Exclude when Exclusion = 2: based on duplicates, urogenital melanoma, etc. Note:
@@
 
-Omit patients with Rec date before SN date
+Omit patients with Rec date before SN date (TODO: or impute instead?)
 ```{r, eval=TRUE}
 old.n <- nrow(sel.data)
+sel.data[sel.data$Rec.time<0&!is.na(sel.data$Rec.time), "DatabaseNR"]
 sel.data <- sel.data[!(sel.data$Rec.time<0&!is.na(sel.data$Rec.time)),]
@@
 sel.data <- droplevels(sel.data)
+save(sel.data,file='Z:/Project Melanoom/PaperMelanoma/Data/sel.data.Rdata', compress=TRUE)
 ```
@@
 ```
+
 Save distributions of all variables, necessary for Harrell summaries
 ```{r, eval=TRUE}
-dd<-datadist(data.to.be.imputed)
+dd<-rms::datadist(data.to.be.imputed)
 options(datadist='dd')
@@
 write.table(as.data.frame(tab.FU.time),file='Z:/Project Melanoom/PaperMelanoma/Results/med.FU.table.txt',
-            row.names=FALSE)
+            row.names=FALSE, sep=",")
 tab.FU.time
@@
 ```{r, eval=TRUE}
-Rec.table <- c("Total", sum(sel.data$Recurrence=="recurred"), sprintf("%.1f", sum(sel.data$Recurrence=="recurred")/nrow(sel.data)*100))
-Rec.table <- rbind(Rec.table, c("Positive", 
-                                sum(sel.data$Recurrence=="recurred"&sel.data$SNstatus=="Positive"),
-                                sprintf("%.1f", sum(sel.data$Recurrence=="recurred"&sel.data$SNstatus=="Positive")/sum(sel.data$SNstatus=="Positive")*100)))
-Rec.table <- rbind(Rec.table, c("Negative", 
-                                sum(sel.data$Recurrence=="recurred"&sel.data$SNstatus=="Negative"),
-                                sprintf("%.1f", sum(sel.data$Recurrence=="recurred"&sel.data$SNstatus=="Negative")/sum(sel.data$SNstatus=="Negative")*100)))
-utils::write.table(as.data.frame(Rec.table),file='Z:/Project Melanoom/PaperMelanoma/Results/Rec.table.txt',
-                   row.names=FALSE)
-Rec.table
-```
-
-# Melanoma Specific Mortality at 5 years
-```{r, eval=TRUE}
-MSM.table <- c("Total", sum(sel.data$MSM==1), sprintf("%.1f", sum(sel.data$MSM==1)/nrow(sel.data)*100))
-MSM.table <- rbind(MSM.table, c("Positive", sum(sel.data$MSM==1&sel.data$SNstatus=="Positive"),
-                                sprintf("%.1f", sum(sel.data$MSM==1&sel.data$SNstatus=="Positive")/sum(sel.data$SNstatus=="Positive")*100)))
-MSM.table <- rbind(MSM.table, c("Negative", sum(sel.data$MSM==1&sel.data$SNstatus=="Negative"),
-                                sprintf("%.1f", sum(sel.data$MSM==1&sel.data$SNstatus=="Negative")/sum(sel.data$SNstatus=="Negative")*100)))
-utils::write.table(as.data.frame(MSM.table),file='Z:/Project Melanoom/PaperMelanoma/Results/MSM.table.txt',
-                   row.names=FALSE)
-MSM.table
+horizon <- 5
+SF.Rec.all<-survfit(S.Rec~1)
+Rec.all <- 100*(1-SF.Rec.all$surv[SF.Rec.all$time==max(SF.Rec.all$time[SF.Rec.all$time<=horizon])])
+
+SF.Rec.neg<-survfit(S.Rec[sel.data$SNstatus=="Negative"]~1)
+Rec.neg <- 100*(1-SF.Rec.neg$surv[SF.Rec.neg$time==max(SF.Rec.neg$time[SF.Rec.neg$time<=horizon])])
+
+SF.Rec.pos<-survfit(S.Rec[sel.data$SNstatus=="Positive"]~1)
+Rec.pos <- 100*(1-SF.Rec.pos$surv[SF.Rec.pos$time==max(SF.Rec.pos$time[SF.Rec.pos$time<=horizon])])
+
+SF.MSM.all<-survfit(S.MSM~1)
+MSM.all <- 100*(1-SF.MSM.all$surv[SF.MSM.all$time==max(SF.MSM.all$time[SF.MSM.all$time<=horizon])])
+
+SF.MSM.neg<-survfit(S.MSM[sel.data$SNstatus=="Negative"]~1)
+MSM.neg <- 100*(1-SF.MSM.neg$surv[SF.MSM.neg$time==max(SF.MSM.neg$time[SF.MSM.neg$time<=horizon])])
+
+SF.MSM.pos<-survfit(S.MSM[sel.data$SNstatus=="Positive"]~1)
+MSM.pos <- 100*(1-SF.MSM.pos$surv[SF.MSM.pos$time==max(SF.MSM.pos$time[SF.MSM.pos$time<=horizon])])
+
+Rec.MSM.table <- data.frame(n=c(nrow(sel.data), 
+                                sum(sel.data$SNstatus=="Negative"), 
+                                sum(sel.data$SNstatus=="Positive")),
+                            perc=c(100, 
+                                   sprintf("%.1f", sum(sel.data$SNstatus=="Negative")/nrow(sel.data)*100),
+                                   sprintf("%.1f", sum(sel.data$SNstatus=="Positive")/nrow(sel.data)*100)),
+                            Rec=sprintf("%.1f", c(Rec.all, Rec.neg, Rec.pos)), 
+                            MSM=sprintf("%.1f", c(MSM.all, MSM.neg, MSM.pos)))
+row.names(Rec.MSM.table) <- c("All", "Negative", "Positive")
+utils::write.table(as.data.frame(Rec.MSM.table),file='Z:/Project Melanoom/PaperMelanoma/Results/Rec.MSM.table.txt',
+                   row.names=TRUE, sep=",")
+Rec.MSM.table
 ```
@@
 ```{r, eval=TRUE}
-S.Rec[S.Rec[,1]>5, 1] <- 5 # set recurrence time to 5
-S.Rec[S.Rec[,1]>5, 2] <- 0 # set censoring
-S.MSM[S.MSM[,1]>5, 1] <- 5 # set follow-up time to 5
-S.MSM[S.MSM[,1]>5, 2] <- 0 # set censoring
+S.Rec.5 <- S.Rec
+S.Rec.5[S.Rec[,1]>5, 1] <- 5 # set recurrence time to 5
+S.Rec.5[S.Rec[,1]>5, 2] <- 0 # set censoring
+
+S.MSM.5 <- S.MSM
+S.MSM.5[S.MSM[,1]>5, 1] <- 5 # set follow-up time to 5
+S.MSM.5[S.MSM[,1]>5, 2] <- 0 # set censoring
 ```
@@
   # compute C-index
-  rc <- Hmisc::rcorr.cens(-AJCC.7.i, S.Rec) # add a minus to revert class order: from worst to best
+  rc <- Hmisc::rcorr.cens(-AJCC.7.i, S.Rec.5) # add a minus to revert class order: from worst to best
   cindex.AJCC.7[i]<-rc["C Index"]
@@
   # compute C-index for only positive sentinel node patients
-  rc.pos <- Hmisc::rcorr.cens(-AJCC.7.i[imputation.i$SNstatus=="Positive"], S.Rec[imputation.i$SNstatus=="Positive",])
+  rc.pos <- Hmisc::rcorr.cens(-AJCC.7.i[imputation.i$SNstatus=="Positive"], S.Rec.5[imputation.i$SNstatus=="Positive",])
   cindex.AJCC.7.pos[i]<-rc.pos["C Index"]
@@
   # compute C-index for only negative sentinel node patients
-  rc.neg <- Hmisc::rcorr.cens(-AJCC.7.i[imputation.i$SNstatus=="Negative"], S.Rec[imputation.i$SNstatus=="Negative",])
+  rc.neg <- Hmisc::rcorr.cens(-AJCC.7.i[imputation.i$SNstatus=="Negative"], S.Rec.5[imputation.i$SNstatus=="Negative",])
   cindex.AJCC.7.neg[i]<-rc.neg["C Index"]
@@
                          "; ", sprintf('%.2f', cindex.mi.AJCC.7.neg$est+qnorm(.975)*cindex.mi.AJCC.7.neg$se), "]"))),
-            file='Z:/Project Melanoom/PaperMelanoma/Results/AJCC.7.Cindex.txt', row.names=FALSE, col.names=FALSE)
+            file='Z:/Project Melanoom/PaperMelanoma/Results/AJCC.7.Cindex.txt', 
+            row.names=FALSE, col.names=FALSE, sep=",")
 
@@
   # compute C-index
-  rc <- Hmisc::rcorr.cens(-AJCC.8.i, S.Rec) # add a minus to revert class order: from worst to best
+  rc <- Hmisc::rcorr.cens(-AJCC.8.i, S.Rec.5) # add a minus to revert class order: from worst to best
   cindex.AJCC.8[i]<-rc["C Index"]
@@
   # compute C-index for only positive sentinel node patients
-  rc.pos <- Hmisc::rcorr.cens(-AJCC.8.i[imputation.i$SNstatus=="Positive"], S.Rec[imputation.i$SNstatus=="Positive",])
+  rc.pos <- Hmisc::rcorr.cens(-AJCC.8.i[imputation.i$SNstatus=="Positive"], S.Rec.5[imputation.i$SNstatus=="Positive",])
   cindex.AJCC.8.pos[i]<-rc.pos["C Index"]
@@
   # compute C-index for only negative sentinel node patients
-  rc.neg <- Hmisc::rcorr.cens(-AJCC.8.i[imputation.i$SNstatus=="Negative"], S.Rec[imputation.i$SNstatus=="Negative",])
+  rc.neg <- Hmisc::rcorr.cens(-AJCC.8.i[imputation.i$SNstatus=="Negative"], S.Rec.5[imputation.i$SNstatus=="Negative",])
   cindex.AJCC.8.neg[i]<-rc.neg["C Index"]
@@
                                "; ", sprintf('%.2f', cindex.mi.AJCC.8.neg$est+qnorm(.975)*cindex.mi.AJCC.8.neg$se), "]"))),
-            file='Z:/Project Melanoom/PaperMelanoma/Results/AJCC.8.Cindex.txt', row.names=FALSE, col.names=FALSE)
+            file='Z:/Project Melanoom/PaperMelanoma/Results/AJCC.8.Cindex.txt', 
+            row.names=FALSE, col.names=FALSE, sep=",")
 
@@
 utils::write.table(table.AJCC.7.8,
-            file='Z:/Project Melanoom/PaperMelanoma/Results/table.AJCC.7.and.8.txt', row.names=FALSE, col.names=FALSE)
+            file='Z:/Project Melanoom/PaperMelanoma/Results/table.AJCC.7.and.8.txt', 
+            row.names=FALSE, col.names=FALSE, sep=",")
 table.AJCC.7.8
@@
 
-# Full model for recurrence
-```{r, eval=TRUE}
-form.full.Rec <- S.Rec ~
+# Test non-lineairities
+### Test if Age should be modelled non-linearly
+```{r, eval=TRUE}
+form.linear.full.Rec <- S.Rec.5 ~
+  SNstatus +
+  Sex +
+  Age.SN +
+  Ulceration +
+  Loc_CAT +
+  Histology +
+  Breslow +
+  multiple.fields +
+  Tot_SNs_neg +
+  Tot_SNs_pos +
+  Rdamcrit +
+  Dewar +
+  SNstatus:Sex +
+  SNstatus:Age.SN +
+  SNstatus:Ulceration +
+  SNstatus:Loc_CAT +
+  SNstatus:Histology +
+  SNstatus:Breslow +
+  SNstatus:multiple.fields
+form.nonlinear.age.full.Rec <- S.Rec.5 ~
   SNstatus +
@@
   log(Age.SN) +
+  Ulceration +
+  Loc_CAT +
+  Histology +
+  Breslow +
+  multiple.fields +
+  Tot_SNs_neg +
+  Tot_SNs_pos +
+  Rdamcrit +
+  Dewar +
+  SNstatus:Sex +
+  SNstatus:log(Age.SN) +
+  SNstatus:Ulceration +
+  SNstatus:Loc_CAT +
+  SNstatus:Histology +
+  SNstatus:Breslow +
+  SNstatus:multiple.fields
+linear.full.Rec <- Hmisc::fit.mult.impute(form.linear.full.Rec, cph, xtrans=imputed.data,
+                                 data=data.to.be.imputed,n.impute=m, pr=FALSE,
+                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
+nonlinear.age.full.Rec <- Hmisc::fit.mult.impute(form.nonlinear.age.full.Rec, cph, xtrans=imputed.data,
+                                 data=data.to.be.imputed,n.impute=m, pr=FALSE,
+                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
+# distribution of Age
+hist(last.imputed.data.set$Age.SN)
+# linear model
+linear.full.Rec$loglik[2]
+# using log(Age) gives slightly better log likelihood
+nonlinear.age.full.Rec$loglik[2]
+```
+
+### Test if Breslow should be modelled non-linearly
+```{r, eval=TRUE}
+form.nonlinear.Breslow.full.Rec <- S.Rec.5 ~
+  SNstatus +
+  Sex +
+  Age.SN +
+  Ulceration +
+  Loc_CAT +
+  Histology +
+  log(Breslow) +
+  multiple.fields +
+  Tot_SNs_neg +
+  Tot_SNs_pos +
+  Rdamcrit +
+  Dewar +
+  SNstatus:Sex +
+  SNstatus:Age.SN +
+  SNstatus:Ulceration +
+  SNstatus:Loc_CAT +
+  SNstatus:Histology +
+  SNstatus:log(Breslow) +
+  SNstatus:multiple.fields
+nonlinear.Breslow.full.Rec <- Hmisc::fit.mult.impute(form.nonlinear.Breslow.full.Rec, cph, xtrans=imputed.data,
+                                 data=data.to.be.imputed,n.impute=m, pr=FALSE,
+                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
+# distribution of Breslow
+hist(last.imputed.data.set$Breslow)
+# linear model
+linear.full.Rec$loglik[2]
+# using log(Breslow) gives better log likelihood
+nonlinear.Breslow.full.Rec$loglik[2]
+```
+
+### Test if Rdamcrit should be modelled non-linearly
+```{r, eval=TRUE}
+form.nonlinear.Rdamcrit.full.Rec <- S.Rec.5 ~
+  SNstatus +
+  Sex +
+  Age.SN +
+  Ulceration +
+  Loc_CAT +
+  Histology +
+  Breslow +
+  multiple.fields +
+  Tot_SNs_neg +
+  Tot_SNs_pos +
+  log(Rdamcrit) +
+  Dewar +
+  SNstatus:Sex +
+  SNstatus:Age.SN +
+  SNstatus:Ulceration +
+  SNstatus:Loc_CAT +
+  SNstatus:Histology +
+  SNstatus:Breslow +
+  SNstatus:multiple.fields
+nonlinear.Rdamcrit.full.Rec <- Hmisc::fit.mult.impute(form.nonlinear.Rdamcrit.full.Rec, cph, xtrans=imputed.data,
+                                 data=data.to.be.imputed,n.impute=m, pr=FALSE,
+                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
+# distribution of Rdamcrit
+hist(last.imputed.data.set$Rdamcrit)
+# linear model
+linear.full.Rec$loglik[2]
+# using log(Rdamcrit) gives better log likelihood
+nonlinear.Rdamcrit.full.Rec$loglik[2]
+```
+
+# Full model for recurrence
+```{r, eval=TRUE}
+form.full.Rec.5 <- S.Rec.5 ~
+  SNstatus +
+  Sex +
+  Age.SN +
   Ulceration +
@@
   SNstatus:Sex +
-  SNstatus:log(Age.SN) +
+  SNstatus:Age.SN +
   SNstatus:Ulceration +
@@
   SNstatus:multiple.fields
-f.mi.full.Rec <- Hmisc::fit.mult.impute(form.full.Rec, cph, xtrans=imputed.data,
+f.mi.full.Rec.5 <- Hmisc::fit.mult.impute(form.full.Rec.5, cph, xtrans=imputed.data,
                                  data=data.to.be.imputed,n.impute=m, pr=FALSE,
                                  fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
-f.mi.full.Rec
-stats::anova(f.mi.full.Rec)
-summary(f.mi.full.Rec)
+# stats::anova(f.mi.full.Rec.5)
+summary(f.mi.full.Rec.5, Tot_SNs_pos=c(0, 1))
 ```
@@
 c.Breslow <- mean(log(last.imputed.data.set$Breslow))
-form.BS.Rec <- S.Rec ~
+form.BS.Rec.5 <- S.Rec.5 ~
   SNstatus +
-  log(Age.SN) +
+  Age.SN +
   Ulceration +
@@
   SNstatus:I(log(Breslow)-c.Breslow)
-f.mi.BS.Rec <- Hmisc::fit.mult.impute(form.BS.Rec,cph,xtrans=imputed.data,data=data.to.be.imputed,n.impute=m,
+f.mi.BS.Rec.5 <- Hmisc::fit.mult.impute(form.BS.Rec.5,cph,xtrans=imputed.data,data=data.to.be.imputed,n.impute=m,
                            pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
-# round(stats::anova(f.mi.BS.Rec)[order(stats::anova(f.mi.BS.Rec)[, 'P']),], 3)
-stats::anova(f.mi.BS.Rec)
+summary(f.mi.BS.Rec.5, Breslow=exp(c.Breslow))
+coef(f.mi.BS.Rec.5)
+# round(stats::anova(f.mi.BS.Rec.5)[order(stats::anova(f.mi.BS.Rec.5)[, 'P']),], 3)
+stats::anova(f.mi.BS.Rec.5)
 ```
@@
 ```{r, eval=TRUE}
-form.full.MSM.refit <- S.MSM ~
+form.full.MSM.5.refit <- S.MSM.5 ~
   SNstatus +
   Sex +
-  log(Age.SN) +
+  Age.SN +
   Ulceration +
@@
   SNstatus:multiple.fields
-f.mi.full.MSM.refit <- Hmisc::fit.mult.impute(form.full.MSM.refit,cph,xtrans=imputed.data,data=data.to.be.imputed,n.impute=m,
+f.mi.full.MSM.5.refit <- Hmisc::fit.mult.impute(form.full.MSM.5.refit,cph,xtrans=imputed.data,data=data.to.be.imputed,n.impute=m,
                                  pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
-f.mi.full.MSM.refit
-summary(f.mi.full.MSM.refit)
-stats::anova(f.mi.full.MSM.refit)
+f.mi.full.MSM.5.refit
+summary(f.mi.full.MSM.5.refit, Tot_SNs_pos=c(0, 1))
+stats::anova(f.mi.full.MSM.5.refit)
 ```
@@
 ```{r, eval=TRUE}
-form.BS.MSM.refit <- S.MSM ~
+form.BS.MSM.5.refit <- S.MSM.5 ~
   SNstatus +
-  log(Age.SN) +
+  Age.SN +
   Ulceration +
@@
   SNstatus:I(log(Breslow)-c.Breslow)
-f.mi.BS.MSM.refit <- Hmisc::fit.mult.impute(form.BS.MSM.refit,cph,xtrans=imputed.data,data=data.to.be.imputed,n.impute=m,
+f.mi.BS.MSM.5.refit <- Hmisc::fit.mult.impute(form.BS.MSM.5.refit,cph,xtrans=imputed.data,data=data.to.be.imputed,n.impute=m,
                                   pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
-# round(anova(f.mi.BS.MSM.refit)[order(anova(f.mi.BS.MSM.refit)[, 'P']),], 3)
-f.mi.BS.MSM.refit
-summary(f.mi.BS.MSM.refit)
-stats::anova(f.mi.BS.MSM.refit)
+# round(anova(f.mi.BS.MSM.5.refit)[order(anova(f.mi.BS.MSM.5.refit)[, 'P']),], 3)
+f.mi.BS.MSM.5.refit
+summary(f.mi.BS.MSM.5.refit, Breslow=exp(c.Breslow))
+stats::anova(f.mi.BS.MSM.5.refit)
 ```
@@
 ```{r, eval=TRUE}
-lp.BS<-f.mi.BS.Rec$linear
-f.mi.BS.MSM<-rms::cph(S.MSM~lp.BS,y=TRUE,x=TRUE)
-MSM.cal.fact <- stats::coef(f.mi.BS.MSM)
+lp.BS<-f.mi.BS.Rec.5$linear
+f.mi.BS.MSM.5<-rms::cph(S.MSM.5~lp.BS,y=TRUE,x=TRUE)
+MSM.cal.fact <- stats::coef(f.mi.BS.MSM.5)
 MSM.cal.fact
-stats::confint(f.mi.BS.MSM)
-utils::write.table(paste0(sprintf("%.2f", stats::coef(f.mi.BS.MSM)), " [",
-            sprintf("%.2f", as.numeric(stats::confint(f.mi.BS.MSM))[1]), "; ",
-            sprintf("%.2f", as.numeric(stats::confint(f.mi.BS.MSM))[2]), "]"),
-            file='Z:/Project Melanoom/PaperMelanoma/Results/calibration.MSS.txt', row.names=FALSE, col.names=FALSE)
+stats::confint(f.mi.BS.MSM.5)
+utils::write.table(paste0(sprintf("%.2f", stats::coef(f.mi.BS.MSM.5)), " [",
+            sprintf("%.2f", as.numeric(stats::confint(f.mi.BS.MSM.5))[1]), "; ",
+            sprintf("%.2f", as.numeric(stats::confint(f.mi.BS.MSM.5))[2]), "]"),
+            file='Z:/Project Melanoom/PaperMelanoma/Results/calibration.MSS.txt', 
+            row.names=FALSE, col.names=FALSE, sep=",")
 ```
@@
 ```{r, eval=TRUE}
-for (model in c('Rec', 'MSM.refit')){
+for (model in c('Rec.5', 'MSM.5.refit')){
+  # load model
   f.mi.full <- eval(parse(text=paste0('f.mi.full.', model)))
 
-  CI.lower.neg <- exp(stats::confint(f.mi.full)[,1])
-  CI.upper.neg <- exp(stats::confint(f.mi.full)[,2])
-  CI.neg <- paste0("[", sprintf("%.2f", CI.lower.neg), "; ", sprintf("%.2f", CI.upper.neg), "]")
-  CI.lower.pos <- c(CI.lower.neg[2:12]*CI.lower.neg[20:30], CI.lower.neg[14:19])
-  CI.upper.pos <- c(CI.upper.neg[2:12]*CI.upper.neg[20:30], CI.upper.neg[14:19])
-  CI.pos <- paste0("[", sprintf("%.2f", CI.lower.pos), "; ", sprintf("%.2f", CI.upper.pos), "]")
-
-  anova.full <- stats::anova(f.mi.full)[,1]
-
-  full.table <- cbind(c(NA, sprintf("%.2f", exp(f.mi.full$coefficients[2:13])), rep(NA, 5)), # COEF NEG
-                      c(NA, CI.neg[2:13], rep(NA, 5)), # CI NEG
-                      sprintf("%.2f", c(exp(f.mi.full$coefficients[1]), # COEF POS
-                        exp(f.mi.full$coefficients[2:12])*exp(f.mi.full$coefficients[20:30]),
-                        exp(f.mi.full$coefficients[14]), exp(f.mi.full$coefficients[15:19]))),
-                      c(CI.neg[1], CI.pos), # CI POS
-                      c(sprintf("%.1f", anova.full[c(1, 3, 5, 7, 9)]), rep(NA, 2), # CHI2 NEG
-                            sprintf("%.1f", anova.full[11]), rep(NA, 2),
-                            sprintf("%.1f", anova.full[c(13, 15, 17, 19, 20)]), rep(NA, 3)),
-                      c(sprintf("%.1f", anova.full[c(2, 21, 22, 23, 24)]), rep(NA, 2), # CHI2 POS
-                            sprintf("%.1f", anova.full[25]), rep(NA, 2),
-                            sprintf("%.1f", anova.full[c(26:27, 18)]), rep(NA, 5)))
-
-  ncols <- 6
+  # Coefficients with CI
+  # Negative results
+  neg.results <- summary(f.mi.full, SNstatus="Negative", Loc_CAT="arm")
+  neg.table <- rbind(rep(NA, 3), # SN status
+                   exp(neg.results[c("Sex - male:female",
+                        "Age.SN",
+                        "Ulceration - Yes:No",
+                        "Loc_CAT - leg:arm",
+                        "Loc_CAT - trunk:arm",
+                        "Loc_CAT - head & neck:arm",
+                        "Histology - NM:SSM",
+                        "Histology - ALM:SSM",
+                        "Histology - other:SSM",
+                        "Breslow",
+                        "multiple.fields",
+                        "Tot_SNs_neg"), c("Effect", "Lower 0.95", "Upper 0.95")]), 
+                   matrix(NA, nrow=6, ncol=3)) # Tot_SNs_pos, Rdamcrit, Combined, Parenchymal, Multifocal, Extensive
+  # add CI
+  neg.table <- cbind(sprintf("%.2f", neg.table[, 1]),
+                     paste0("[", sprintf("%.2f", neg.table[, 2]), "; ", sprintf("%.2f", neg.table[, 3]), "]"))
+
+  # Positive results
+  pos.results <- summary(f.mi.full, SNstatus="Positive", Loc_CAT="arm", Tot_SNs_pos=c(0, 1))
+  pos.table <- rbind(exp(neg.results["SNstatus - Positive:Negative", c("Effect", "Lower 0.95", "Upper 0.95")]),
+                     exp(pos.results[c("Sex - male:female",
+                        "Age.SN",
+                        "Ulceration - Yes:No",
+                        "Loc_CAT - leg:arm",
+                        "Loc_CAT - trunk:arm",
+                        "Loc_CAT - head & neck:arm",
+                        "Histology - NM:SSM",
+                        "Histology - ALM:SSM",
+                        "Histology - other:SSM",
+                        "Breslow",
+                        "multiple.fields",
+                        "Tot_SNs_neg",
+                        "Tot_SNs_pos",
+                        "Rdamcrit",
+                        "Dewar - combined:subcap",
+                        "Dewar - parenchymal:subcap",
+                        "Dewar - multifocal:subcap",
+                        "Dewar - extensive:subcap"), c("Effect", "Lower 0.95", "Upper 0.95")]))
+  # add CI
+  pos.table <- cbind(sprintf("%.2f", pos.table[, 1]), 
+                     paste0("[", sprintf("%.2f", pos.table[, 2]), "; ", sprintf("%.2f", pos.table[, 3]), "]"))
+
+  # combine postive with negative
+  full.table <- cbind(neg.table, pos.table)
+
+  # add reference rows
   final.table <- rbind(full.table[1:4,],
-                       rep(NA, ncols), # location
-                       c("(ref)", NA, "(ref)", rep(NA, 3)), # ref arm
+                       rep(NA, 4),     # location
+                       c("(ref)", NA, "(ref)", NA), # ref arm
                        full.table[5:7,],   # other locations
-                       rep(NA, ncols), # histology
-                       c("(ref)", NA, "(ref)", rep(NA, 3)), # ref SMM
-                       full.table[8:14,],
-                       rep(NA, ncols), # Dewar
-                       c(rep(NA, 2), "(ref)", rep(NA, 3)), # ref subcap
-                       full.table[15:18,]) # other Dewar
-
-  colnames(final.table) <- c("HR.neg", "CI.neg", "HR.pos", "CI.pos", "Chi2.neg", "Chi2.pos")
+                       rep(NA, 4),     # histology
+                       c("(ref)", NA, "(ref)", NA), # ref SMM
+                       full.table[8:15,],
+                       rep(NA, 4),     # Dewar
+                       c(rep(NA, 2), "(ref)", NA), # ref subcap
+                       full.table[16:nrow(full.table),]) # other Dewar
+
+  # add Chi-square statistics
+  anova.out <- anova(f.mi.full)[, "Chi-Square"]
+  anova.main <- c(anova.out[c("SNstatus  (Factor+Higher Order Factors)",
+                        "Sex  (Factor+Higher Order Factors)",
+                        "Age.SN  (Factor+Higher Order Factors)",
+                        "Ulceration  (Factor+Higher Order Factors)",
+                        "Loc_CAT  (Factor+Higher Order Factors)")],
+                  rep(NA, 4),
+                  anova.out["Histology  (Factor+Higher Order Factors)"],
+                  rep(NA, 4),
+                  anova.out[c("Breslow  (Factor+Higher Order Factors)",
+                        "multiple.fields  (Factor+Higher Order Factors)",
+                        "Tot_SNs_neg",
+                        "Tot_SNs_pos",
+                        "Rdamcrit",
+                        "Dewar")],
+                  rep(NA, 5))
+  anova.int <- c(anova.out[c("TOTAL INTERACTION",
+                       "SNstatus * Sex  (Factor+Higher Order Factors)",
+                       "SNstatus * Age.SN  (Factor+Higher Order Factors)",
+                       "SNstatus * Ulceration  (Factor+Higher Order Factors)",
+                       "SNstatus * Loc_CAT  (Factor+Higher Order Factors)")],
+                 rep(NA, 4),
+                 anova.out["SNstatus * Histology  (Factor+Higher Order Factors)"],
+                 rep(NA, 4),
+                 anova.out[c("SNstatus * Breslow  (Factor+Higher Order Factors)",
+                       "SNstatus * multiple.fields  (Factor+Higher Order Factors)")],
+                 rep(NA, 9))
+  final.table <- cbind(final.table, sprintf("%.0f", anova.main), sprintf("%.0f", anova.int))
+
+  colnames(final.table) <- c("HR.neg", "CI.neg", "HR.pos", "CI.pos", "Main", "Int")
   assign(paste0('final.full.table.', model), as.data.frame(final.table))
@@
 ```{r, eval=TRUE}
-for (model in c('Rec', 'MSM.refit')){
+for (model in c('Rec.5', 'MSM.5.refit')){
   f.mi.BS <- eval(parse(text=paste0('f.mi.BS.', model)))
 
-  CI.lower.neg <- exp(confint(f.mi.BS)[,1])
-  CI.upper.neg <- exp(confint(f.mi.BS)[,2])
-  CI.neg <- paste0("[", sprintf("%.2f", CI.lower.neg), "; ", sprintf("%.2f", CI.upper.neg), "]")
-
-  anova.BS <- anova(f.mi.BS)[,1]
-
-  CI.pos <- paste0("[", sprintf("%.2f", exp(confint(f.mi.BS)[7,1])*exp(confint(f.mi.BS)[9,1])),
-                   "; ", sprintf("%.2f", exp(confint(f.mi.BS)[7,2])*exp(confint(f.mi.BS)[9,2])), "]")
-
-  BS.table <- cbind(c(NA, sprintf("%.2f", exp(f.mi.BS$coefficients[2:7])), NA), # COEF NEG
-                    c(NA, CI.neg[2:7], NA), # CI NEG
-                    sprintf("%.2f", exp(f.mi.BS$coefficients[1:8])*c(rep(1, 6), exp(f.mi.BS$coefficients[9]), 1)), #COEF POS
-                    c(CI.neg[1:6], CI.pos, CI.neg[8]), # CI POS
-                    c(NA, sprintf("%.1f", anova.BS[3:5]), rep(NA, 2), sprintf("%.1f", anova.BS[c(6, 8)])), # CHI2 NEG
-                    c(sprintf("%.1f", anova.BS[1]), rep(NA, 5), # CHI2 POS
-                      sprintf("%.1f", anova.BS[9]), NA))
-
-  BS.table <- rbind(BS.table[1:3,], rep(NA, 6), c("(ref)", NA, "(ref)", rep(NA, 3)), BS.table[4:8,])
-
-  colnames(BS.table) <- c("HR.neg", "CI.neg", "HR.pos", "CI.pos", "Chi2.neg", "Chi2.pos")
-
+  # Coefficients with CI
+  # Negative results
+  neg.results.BS <- summary(f.mi.BS, SNstatus="Negative", Breslow=exp(c.Breslow), Loc_CAT="arm")
+  neg.table.BS <- rbind(rep(NA, 3), # SN status
+                   exp(neg.results.BS[c("Age.SN",
+                        "Ulceration - Yes:No",
+                        "Loc_CAT - leg:arm",
+                        "Loc_CAT - trunk:arm",
+                        "Loc_CAT - head & neck:arm",
+                        "Breslow"), c("Effect", "Lower 0.95", "Upper 0.95")]), 
+                   rep(NA, 3)) # Rdamcrit
+  # add CI
+  neg.table.BS <- cbind(sprintf("%.2f", neg.table.BS[, 1]), 
+                        paste0("[", sprintf("%.2f", neg.table.BS[, 2]), "; ", sprintf("%.2f", neg.table.BS[, 3]), "]"))
+
+  # Positive results
+  pos.results.BS <- summary(f.mi.BS, SNstatus="Positive", Breslow=exp(c.Breslow), Loc_CAT="arm")
+  pos.table.BS <- rbind(exp(neg.results.BS["SNstatus - Positive:Negative", c("Effect", "Lower 0.95", "Upper 0.95")]),
+                     exp(pos.results.BS[c("Age.SN",
+                        "Ulceration - Yes:No",
+                        "Loc_CAT - leg:arm",
+                        "Loc_CAT - trunk:arm",
+                        "Loc_CAT - head & neck:arm",
+                        "Breslow",
+                        "Rdamcrit"), c("Effect", "Lower 0.95", "Upper 0.95")]))
+  # add CI
+  pos.table.BS <- cbind(sprintf("%.2f", pos.table.BS[, 1]), 
+                        paste0("[", sprintf("%.2f", pos.table.BS[, 2]), "; ", sprintf("%.2f", pos.table.BS[, 3]), "]"))
+
+  # combine postive with negative
+  BS.table <- cbind(neg.table.BS, pos.table.BS)
+
+  # add reference rows
+  BS.table <- rbind(BS.table[1:3,], 
+                    rep(NA, 4), 
+                    c("(ref)", NA, "(ref)", NA), 
+                    BS.table[4:nrow(BS.table),])
+
+  # add Chi-square statistics
+  anova.out <- anova(f.mi.BS)[, "Chi-Square"]
+  anova.main <- c(anova.out[c("SNstatus  (Factor+Higher Order Factors)",
+                        "Age.SN",
+                        "Ulceration",
+                        "Loc_CAT")],
+                  rep(NA, 4),
+                  anova.out[c("Breslow  (Factor+Higher Order Factors)",
+                        "Rdamcrit")])
+  anova.int <- c(rep(NA, 8),
+                 anova.out["SNstatus * Breslow  (Factor+Higher Order Factors)"],
+                 NA)
+  BS.table <- cbind(BS.table, sprintf("%.0f", anova.main), sprintf("%.0f", anova.int))
+
+  colnames(BS.table) <- c("HR.neg", "CI.neg", "HR.pos", "CI.pos", "Main", "Int")
   assign(paste0('final.BS.table.', model), as.data.frame(BS.table))
@@
 ```{r, eval=TRUE}
-horizon <- 5
-for (model in list('Rec', 'MSM.refit')){
-  if (model == "Rec"){
+for (model in list('Rec.5', 'MSM.5.refit')){
+  if (model == 'Rec.5'){
     # plot predict
     png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/plot.predict.Rec.png"))
-    plot.predict <- plot(Predict(f.mi.full.Rec))
+    plot.predict <- plot(Predict(f.mi.full.Rec.5))
     show(plot.predict)
@@
     png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/plot.predict.Breslow.Rec.png"))
-    plot.Breslow <- plot(Predict(f.mi.full.Rec, Breslow=0:20, SNstatus), ~ Breslow, groups="SNstatus",
+    plot.Breslow <- plot(Predict(f.mi.full.Rec.5, Breslow=0:20, SNstatus), ~ Breslow, groups="SNstatus",
                          adj.subtitle=FALSE, lwd=2, labels=FALSE,
@@
     f.mi <- eval(parse(text=paste0('f.mi.', type, '.', model)))
-    if (model=="Rec"){
-      S <- S.Rec
+    if (model=='Rec.5'){
+      S <- S.Rec.5
     }
     else if (model=="MSM.refit"){
-      S <- S.MSM
+      S <- S.MSM.5
     }
@@
 for (type in c("full", "BS")){
-  f.mi <- eval(parse(text=paste0('f.mi.', type, '.Rec')))
+  f.mi <- eval(parse(text=paste0('f.mi.', type, '.Rec.5')))
   cindex.MSM.refit <- rep(0, m)
@@
   for (i in 1:m){
-    rc <- Hmisc::rcorr.cens(-predict(f.mi,newdata=mice::complete(imputed.data,i), type="lp"), S.MSM)
+    rc <- Hmisc::rcorr.cens(-predict(f.mi,newdata=mice::complete(imputed.data,i), type="lp"), S.MSM.5)
     cindex.MSM.refit[i]<-rc["C Index"]
@@
 
-  final.full.table <- eval(parse(text=paste0('final.', type, '.table.Rec')))
+  final.full.table <- eval(parse(text=paste0('final.', type, '.table.Rec.5')))
   final.full.table["C-index MSM", "HR.neg"] <- sprintf('%.2f', cindex.mi.MSM.refit$est)
@@
 
-  assign(paste0('final.', type, '.table.Rec'), final.full.table)
+  assign(paste0('final.', type, '.table.Rec.5'), final.full.table)
 }
@@
 # replace NA by blank
-final.full.table.Rec[is.na(final.full.table.Rec)] <- ""
-final.full.table.MSM.refit[is.na(final.full.table.MSM.refit)] <- ""
-final.BS.table.Rec[is.na(final.BS.table.Rec)] <- ""
-final.BS.table.MSM.refit[is.na(final.BS.table.MSM.refit)] <- ""
+for (df in c("final.full.table.Rec.5", "final.full.table.MSM.5.refit", "final.BS.table.Rec.5", "final.BS.table.MSM.5.refit")){
+  final.df <- eval(parse(text=df))
+  final.df[final.df=="[NA; NA]"] <- NA
+  final.df[final.df=="NA"] <- NA
+  final.df[is.na(final.df)] <- ""
+  assign(df, final.df)
+}
 
 # set rownames
-final.full.table.Rec <- as.data.frame(final.full.table.Rec)
-row.names(final.full.table.Rec) <- c("Positive SN status", "Male",
+final.full.table.Rec.5 <- as.data.frame(final.full.table.Rec.5)
+row.names(final.full.table.Rec.5) <- c("Positive SN status", "Male",
+                                    "Age iqr", "Ulceration", "Location",
+                                    "Arm", "Leg", "Trunk", "Head and neck",
+                                    "Histology", "SSM", "NM", "ALM", "Other",
+                                    "Breslow iqr", "Multiple fields", 
+                                    "Total number of negative nodes",
+                                    "Total number of positive nodes",
+                                    "SN tumour burden iqr", 
+                                    "Location metastasis in lymph node",
+                                    "Subcap", "Combined", "Parenchymal",
+                                    "Multifocal", "Extensive",
+                                    rownames(final.full.table.Rec.5)[(nrow(final.full.table.Rec.5)-1):nrow(final.full.table.Rec.5)])
+final.full.table.MSM.5.refit <- as.data.frame(final.full.table.MSM.5.refit)
+row.names(final.full.table.MSM.5.refit) <- c("Positive SN status", "Male",
                                           "Age", "Ulceration", "Location",
@@
                                           "Breslow", "Multiple fields", 
-                                    "Total number of nodes",
+                                          "Total number of negative nodes",
+                                          "Total number of positive nodes",
                                           "SN tumour burden", 
@@
                                           "Multifocal", "Extensive",
-                                    rownames(final.full.table.Rec)[25:26])
-final.full.table.MSM.refit <- as.data.frame(final.full.table.MSM.refit)
-row.names(final.full.table.MSM.refit) <- c("Positive SN status", "Male",
-                                          "Age", "Ulceration", "Location",
-                                          "Arm", "Leg", "Trunk", "Head and neck",
-                                          "Histology", "SSM", "NM", "ALM", "Other",
-                                          "Breslow", "Multiple fields", 
-                                          "Total number of nodes",
-                                          "SN tumour burden", 
-                                          "Location metastasis in lymph node",
-                                          "Subcap", "Combined", "Parenchymal",
-                                          "Multifocal", "Extensive",
-                                          row.names(final.full.table.Rec)[25])
-final.BS.table.Rec <- as.data.frame(final.BS.table.Rec)
-row.names(final.BS.table.Rec) <- c("Positive SN status", "Age", "Ulceration",
+                                          row.names(final.full.table.MSM.5.refit)[nrow(final.full.table.MSM.5.refit)])
+final.BS.table.Rec.5 <- as.data.frame(final.BS.table.Rec.5)
+row.names(final.BS.table.Rec.5) <- c("Positive SN status", "Age", "Ulceration",
                                    "Location", "Arm", "Leg", "Trunk", 
@@
                                    "Breslow", "SN tumour burden", 
-                                   row.names(final.BS.table.Rec)[11:12])
-final.BS.table.MSM.refit <- as.data.frame(final.BS.table.MSM.refit)
-row.names(final.BS.table.MSM.refit) <- c("Positive SN status", "Age", "Ulceration",
+                                   row.names(final.BS.table.Rec.5)[(nrow(final.BS.table.Rec.5)-1):nrow(final.BS.table.Rec.5)])
+final.BS.table.MSM.5.refit <- as.data.frame(final.BS.table.MSM.5.refit)
+row.names(final.BS.table.MSM.5.refit) <- c("Positive SN status", "Age", "Ulceration",
                                    "Location", "Arm", "Leg", "Trunk", 
@@
                                    "Breslow", "SN tumour burden", 
-                                   row.names(final.BS.table.MSM.refit)[11])
+                                   row.names(final.BS.table.MSM.5.refit)[nrow(final.BS.table.MSM.5.refit)])
 
 # save tables
-write.table(final.full.table.Rec,
-            file='Z:/Project Melanoom/PaperMelanoma/Results/final.full.table.Rec.txt', sep=",")
-write.table(final.full.table.MSM.refit,
-            file='Z:/Project Melanoom/PaperMelanoma/Results/final.full.table.MSM.refit.txt', sep=",")
-write.table(final.BS.table.Rec,
-            file='Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.Rec.txt', sep=",")
-write.table(final.BS.table.MSM.refit,
-            file='Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.MSM.refit.txt', sep=",")
+write.table(final.full.table.Rec.5,
+            file='Z:/Project Melanoom/PaperMelanoma/Results/final.full.table.Rec.txt', 
+            sep=",")
+write.table(final.full.table.MSM.5.refit,
+            file='Z:/Project Melanoom/PaperMelanoma/Results/final.full.table.MSM.refit.txt', 
+            sep=",")
+write.table(final.BS.table.Rec.5,
+            file='Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.Rec.txt', 
+            sep=",")
+write.table(final.BS.table.MSM.5.refit,
+            file='Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.MSM.refit.txt', 
+            sep=",")
 
 # show tables
-final.full.table.Rec
-final.full.table.MSM.refit
-final.BS.table.Rec
-final.BS.table.MSM.refit
+final.full.table.Rec.5
+final.full.table.MSM.5.refit
+final.BS.table.Rec.5
+final.BS.table.MSM.5.refit
 ```
@@
 cuts <- 5
-for (model in c('Rec', 'MSM')){
-  if (model == 'Rec'){
-    S.model <- S.Rec
-  }
-  if (model == 'MSM'){
-    S.model <- S.MSM
+for (model in c('Rec.5', 'MSM.5')){
+  if (model == 'Rec.5'){
+    S.model <- S.Rec.5
+  }
+  if (model == 'MSM.5'){
+    S.model <- S.MSM.5
   }
@@
       S.notj<-S.model[data.to.be.imputed$Center!=Centers[j],]
-      S.Rec.notj<-S.Rec[data.to.be.imputed$Center!=Centers[j],]
-
-      form.model<-eval(parse(text=paste0('form.', type, '.Rec')))
-      form.notj<-update(form.model, S.Rec.notj  ~ . )
+      S.Rec.5.notj<-S.Rec.5[data.to.be.imputed$Center!=Centers[j],]
+
+      form.model<-eval(parse(text=paste0('form.', type, '.Rec.5')))
+      form.notj<-update(form.model, S.Rec.5.notj  ~ . )
       f.mi.Rec.notj<-Hmisc::fit.mult.impute(form.notj,cph,xtrans=imputed.data,data=data.to.be.imputed,
@@
 
-      if (model=="Rec"){
+      if (model=='Rec.5'){
         f.basehaz.notj<-survival::basehaz(f.mi.Rec.notj,TRUE)
         h0.notj<-f.basehaz.notj$hazard[f.basehaz.notj$time==max(f.basehaz.notj$time[f.basehaz.notj$time<=horizon])]
-      } else if (model=="MSM"){
+      } else if (model=='MSM.5'){
         lp.notj<-f.mi.Rec.notj$linear
@@
 
-          if (model=="MSM"){
+          if (model=='MSM.5'){
             lp.i<-predict(f.MSM.notj,newdata=lp.i)
@@
           # baseline hazard
-          if (model=="Rec"){
+          if (model=='Rec.5'){
             f.basehaz<-basehaz(f.mi.Rec.notj,TRUE)
           }
-          else if (model=="MSM"){
+          else if (model=='MSM.5'){
             f.basehaz<-basehaz(f.MSM.notj,TRUE)
@@
 for (select.value in c("All", "Positive", "Negative")){
-  for (y in c("Rec", "MSM")){
+  for (y in c('Rec.5', 'MSM.5')){
     png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/", y, ".", select.value, ".png"), width=16, height=16, units="cm", res=300)
@@
 # initial nomogram
-nom.0 <- nomogram(f.mi.BS.Rec,maxscale=10,
+nom.0 <- nomogram(f.mi.BS.Rec.5,maxscale=10,
                 Breslow=Breslow.class, Rdamcrit=Rdamcrit.class, Age.SN=Age.class)
@@
 # baseline hazard of recurrence
-f.Rec.basehaz <- basehaz(f.mi.BS.Rec,TRUE)
+f.Rec.basehaz <- basehaz(f.mi.BS.Rec.5,TRUE)
 h0.Rec <- f.Rec.basehaz$hazard[f.Rec.basehaz$time==max(f.Rec.basehaz$time[f.Rec.basehaz$time<=horizon])]
@@
 # nomogram
-par(mar = rep(0, 4))
-nom <- nomogram(f.mi.BS.Rec, maxscale=10,
+nom <- nomogram(f.mi.BS.Rec.5, maxscale=10,
               Breslow=Breslow.class, Rdamcrit=Rdamcrit.class, Age.SN=Age.class,
@@
 attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Negative)"]<-"Breslow + negative SN"
-plot(nom, total.sep.page=TRUE, col.grid=gray(c(0.8, 0.95)))
+png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Nomogram.png"))
+par(mar = rep(0, 4))
+plot(nom, total.sep.page=FALSE, col.grid=gray(c(0.8, 0.95)))
+grDevices::dev.off()
 nom.0
@@
 # baseline hazard of MSM
-f.MSM.basehaz <- basehaz(f.mi.BS.MSM,TRUE)
+f.MSM.basehaz <- basehaz(f.mi.BS.MSM.5,TRUE)
 h0.MSM <- f.MSM.basehaz$hazard[f.MSM.basehaz$time==max(f.MSM.basehaz$time[f.MSM.basehaz$time<=horizon])]
-lp.MSM <- f.mi.BS.MSM$linear
+lp.MSM <- f.mi.BS.MSM.5$linear
 
 # lp range MSM
-rc.MSM <- f.mi.BS.MSM$coef[1]*rc.Rec
-int.MSM <- f.mi.BS.MSM$coef[1]*int.Rec
+rc.MSM <- f.mi.BS.MSM.5$coef[1]*rc.Rec
+int.MSM <- f.mi.BS.MSM.5$coef[1]*int.Rec
 max.range <- 16
 points.range <- 0:max.range
-lp.points.Rec <- int.Rec+rc.Rec*points.range
-lp.points.MSM <- int.MSM+rc.MSM*points.range
+lp.points.Rec.5 <- int.Rec+rc.Rec*points.range
+lp.points.MSM.5 <- int.MSM+rc.MSM*points.range
 
@@
 Breslow.trunc[last.imputed.data.set$Breslow>7] <- 7
-lp.score.Rec <- predict(f.mi.BS.Rec,newdata=data.frame(SNstatus=last.imputed.data.set$SNstatus,
+lp.score.Rec <- predict(f.mi.BS.Rec.5,newdata=data.frame(SNstatus=last.imputed.data.set$SNstatus,
                                                 Age.SN=last.imputed.data.set$Age.SN,
@@
 h.plot$density <- 100*h.Rec$density
-p.Rec <- fun.event(lp=lp.points.Rec, h0=h0.Rec)
-p.MSM <- fun.event(lp=lp.points.MSM, h0=h0.MSM)
+p.Rec <- fun.event(lp=lp.points.Rec.5, h0=h0.Rec)
+p.MSM <- fun.event(lp=lp.points.MSM.5, h0=h0.MSM)
 x.lim <- c(0, max.range+1)
@@
 # save for Rshiny app
-coef.Rec <- stats::coef(f.mi.BS.Rec)
-center.Rec <- f.mi.BS.Rec$center
+coef.Rec <- stats::coef(f.mi.BS.Rec.5)
+center.Rec <- f.mi.BS.Rec.5$center
 save(h0.Rec, h0.MSM, coef.Rec, c.Breslow, center.Rec,
@@
 int.Rec
-log(20)*f.mi.BS.Rec$coef["Age.SN"]+
-  (log(0.1)-c.Breslow)*f.mi.BS.Rec$coef["Breslow"]+
-  log(0.01)*f.mi.BS.Rec$coef["Rdamcrit"]-
-  f.mi.BS.Rec$center
+20*f.mi.BS.Rec.5$coef["Age.SN"]+
+  (log(0.1)-c.Breslow)*f.mi.BS.Rec.5$coef["Breslow"]+
+  log(0.01)*f.mi.BS.Rec.5$coef["Rdamcrit"]-
+  f.mi.BS.Rec.5$center
 ```
@@
 ```{r, eval=TRUE}
-sum(h.Rec$density*fun.event(lp=lp.points.Rec, h0=h0.Rec))
+sum(h.Rec$density*fun.event(lp=lp.points.Rec.5, h0=h0.Rec))
 # equal to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
-SF<-survfit(S.Rec~1,data=sel.data)
+SF<-survfit(S.Rec.5~1,data=sel.data)
 100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
@@
 ```{r, eval=TRUE}
-sum(h.Rec$density*fun.event(lp=lp.points.MSM, h0=h0.MSM))
+sum(h.Rec$density*fun.event(lp=lp.points.MSM.5, h0=h0.MSM))
 # equal to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
-SF<-survfit(S.MSM~1,data=sel.data)
+SF<-survfit(S.MSM.5~1,data=sel.data)
 100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
@@
 
-### Example to check probabilities with app
-```{r, eval=TRUE}
-patient.nr <- 115 # 395 and 396 = Negative, 397 and 398 = Positive, 30 = Head & neck, 115 has large Breslow
-data.to.be.imputed[patient.nr, c("SNstatus", "Age.SN", "Ulceration", "Loc_CAT", "Breslow", "Rdamcrit")]
-lp.score.Rec.true <- predict(f.mi.BS.Rec)[patient.nr]
-lp.score.Rec.try <- predict(f.mi.BS.Rec, newdata=data.frame(SNstatus=data.to.be.imputed[patient.nr,"SNstatus"],
-                                                            Age.SN=data.to.be.imputed[patient.nr,"Age.SN"],
-                                                            Ulceration=data.to.be.imputed[patient.nr,"Ulceration"],
-                                                            Loc_CAT=data.to.be.imputed[patient.nr,"Loc_CAT"],
-                                                            Breslow=data.to.be.imputed[patient.nr,"Breslow"],
-                                                            Rdamcrit=data.to.be.imputed[patient.nr,"Rdamcrit"]),
+# Check if predictions of new and old model align for negative SN patients
+## New model
+```{r, eval=TRUE}
+lp.negative <- predict(f.mi.BS.Rec.5,
+                       newdata=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative",],
                        type="lp")
-p.Rec.example <- fun.event(lp=lp.score.Rec.try, h0=h0.Rec)
-p.MSM.example <- fun.event(lp=lp.score.Rec.try, h0=h0.MSM)
-
-# manual calculation
-lp.score.Rec.manual <- coef.Rec["SNstatus=Positive"]*(as.numeric(data.to.be.imputed[patient.nr,"SNstatus"])-1)+
-              coef.Rec["Age.SN"]*log(data.to.be.imputed[patient.nr,"Age.SN"])+
-              coef.Rec["Ulceration=Yes"]*(as.numeric(data.to.be.imputed[patient.nr,"Ulceration"])-1)+
-              coef.Rec["Loc_CAT=leg"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="leg")+
-              coef.Rec["Loc_CAT=trunk"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="trunk")+
-              coef.Rec["Loc_CAT=head & neck"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="head & neck")+
-              coef.Rec["Breslow"]*(log(data.to.be.imputed[patient.nr,"Breslow"])-c.Breslow)+
-              coef.Rec["Rdamcrit"]*log(data.to.be.imputed[patient.nr,"Rdamcrit"])+
-              coef.Rec["SNstatus=Positive * Breslow"]*(log(data.to.be.imputed[patient.nr,"Breslow"])-c.Breslow)*(as.numeric(data.to.be.imputed[patient.nr,"SNstatus"])-1)
-
-# lp check
-lp.score.Rec.true
-lp.score.Rec.try
-lp.score.Rec.manual-center.Rec
-
-# probability of recurrence with fun.event
-fun.event(lp=lp.score.Rec.try, h0=h0.Rec)
-# probability of recurrence manually
-1-exp(-h0.Rec*exp(lp.score.Rec.manual-center.Rec))
-
-# check calculation of lp of MSM
-lp.score.MSM.true <- predict(f.mi.BS.MSM)[patient.nr]
-lp.score.MSM.true
-lp.MSM.try <- MSM.cal.fact*(lp.score.Rec.manual-center.Rec)
-lp.MSM.try
-
-# probability of MSM with fun.event
-fun.event(lp=lp.MSM.try, h0=h0.MSM)
-# probability of MSM manually
-1-exp(-h0.MSM*exp(MSM.cal.fact*(lp.score.Rec.manual-center.Rec)))
-```
-
-### Test with old tool predictions recurrence for negative SN status patients
+new.negativeSN.tool <- 1-exp(-h0.Rec*exp(lp.negative))
+```
+
+## Old model with new data
 ```{r, eval=TRUE}
@@
         location=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Loc_CAT"])-1)
-lp.negative <- predict(f.mi.BS.Rec, newdata=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative",], type="lp")
-new.negativeSN.tool <- 1-exp(-h0.Rec*exp(lp.negative))
-png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/new.old.negatives.png"))
+```
+
+## Compare
+```{r, eval=TRUE}
 plot(x=new.negativeSN.tool*100, y=old.negativeSN.tool,
@@
 abline(a=0, b=1, col="red")
-grDevices::dev.off()
-```
-
-```{r, eval=TRUE, echo=FALSE}
-plot(x=new.negativeSN.tool*100, y=old.negativeSN.tool,
+```
+
+## Refitted old model new data
+```{r, eval=TRUE}
+refitted.old.negative <- rms::cph(S.Rec.5 ~ log(Breslow)+Ulceration+Loc_CAT,
+                     data=last.imputed.data.set,
+                     y=TRUE, x=TRUE,
+                     subset=last.imputed.data.set$SNstatus=="Negative")
+refitted.old.negative
+f.Rec.basehaz.neg <- basehaz(refitted.old.negative,TRUE)
+h0.neg <- f.Rec.basehaz.neg$hazard[f.Rec.basehaz.neg$time==max(f.Rec.basehaz.neg$time[f.Rec.basehaz.neg$time<=horizon])]
+refit.old.negativeSN.tool <- fun.event(predict(refitted.old.negative),  h0.neg)
+```
+
+## Compare
+```{r, eval=TRUE}
+plot(x=new.negativeSN.tool*100, y=refit.old.negativeSN.tool*100,
      xlim=c(0, 100), ylim=c(0, 100),
@@
 
-### Test with old tool predictions recurrence for positive SN status patients
+## Check mean recurrence probabilities
+```{r, eval=TRUE}
+SF<-survfit(S.Rec.5[last.imputed.data.set$SNstatus=="Negative"]~1)
+100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
+mean(new.negativeSN.tool)
+mean(refit.old.negativeSN.tool)
+mean(old.negativeSN.tool)
+```
+
+# Check if predictions of new and old model align for positive SN patients
+## New model
+```{r, eval=TRUE}
+lp.positive <- predict(f.mi.BS.Rec.5,
+                       newdata=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive",],
+                       type="lp")
+new.positiveSN.tool <- 1-exp(-h0.Rec*exp(lp.positive))
+```
+
+## Old model with new data
 ```{r, eval=TRUE}
@@
                                         tumor_burden=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Rdamcrit"]))
-lp.positive <- predict(f.mi.BS.Rec, newdata=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive",], type="lp")
-new.positiveSN.tool <- 1-exp(-h0.Rec*exp(lp.positive))
-png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/new.old.positives.png"))
+```
+
+## Compare
+```{r, eval=TRUE}
 plot(x=new.positiveSN.tool*100, y=old.positiveSN.tool,
@@
 abline(a=0, b=1, col="red")
-grDevices::dev.off()
-```
-
-```{r, eval=TRUE, echo=FALSE}
-plot(x=new.positiveSN.tool*100, y=old.positiveSN.tool,
+```
+
+## Refitted old model new data
+```{r, eval=TRUE}
+refitted.old.positive <- rms::cph(S.Rec.5 ~ log(Breslow)+Ulceration+Loc_CAT, 
+                      data=last.imputed.data.set,
+                      y=TRUE, x=TRUE,
+                      subset=last.imputed.data.set$SNstatus=="Positive")
+refitted.old.positive
+f.Rec.basehaz.pos <- basehaz(refitted.old.positive,TRUE)
+h0.pos <- f.Rec.basehaz.pos$hazard[f.Rec.basehaz.pos$time==max(f.Rec.basehaz.pos$time[f.Rec.basehaz.pos$time<=horizon])]
+refit.old.positiveSN.tool <- fun.event(predict(refitted.old.positive),  h0.pos)
+```
+
+## Compare
+```{r, eval=TRUE}
+plot(x=new.positiveSN.tool*100, y=refit.old.positiveSN.tool*100,
      xlim=c(0, 100), ylim=c(0, 100),
@@
 
+## Check mean recurrence probabilities
+```{r, eval=TRUE}
+SF<-survfit(S.Rec.5[last.imputed.data.set$SNstatus=="Positive"]~1)
+100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
+mean(new.positiveSN.tool)
+mean(refit.old.positiveSN.tool)
+mean(old.positiveSN.tool)
+```
+
+#### Example to check probabilities with app
+```{r, eval=TRUE}
+patient.nr <- 115 # 395 and 396 = Negative, 397 and 398 = Positive, 30 = Head & neck, 115 has large Breslow
+data.to.be.imputed[patient.nr, c("SNstatus", "Age.SN", "Ulceration", "Loc_CAT", "Breslow", "Rdamcrit")]
+lp.score.Rec.true <- predict(f.mi.BS.Rec.5)[patient.nr]
+lp.score.Rec.try <- predict(f.mi.BS.Rec.5, newdata=data.frame(SNstatus=data.to.be.imputed[patient.nr,"SNstatus"],
+                                                            Age.SN=data.to.be.imputed[patient.nr,"Age.SN"],
+                                                            Ulceration=data.to.be.imputed[patient.nr,"Ulceration"],
+                                                            Loc_CAT=data.to.be.imputed[patient.nr,"Loc_CAT"],
+                                                            Breslow=data.to.be.imputed[patient.nr,"Breslow"],
+                                                            Rdamcrit=data.to.be.imputed[patient.nr,"Rdamcrit"]),
+                                                            type="lp")
+p.Rec.example <- fun.event(lp=lp.score.Rec.try, h0=h0.Rec)
+p.MSM.example <- fun.event(lp=lp.score.Rec.try, h0=h0.MSM)
+
+# manual calculation
+lp.score.Rec.manual <- coef.Rec["SNstatus=Positive"]*(as.numeric(data.to.be.imputed[patient.nr,"SNstatus"])-1)+
+              coef.Rec["Age.SN"]*data.to.be.imputed[patient.nr,"Age.SN"]+
+              coef.Rec["Ulceration=Yes"]*(as.numeric(data.to.be.imputed[patient.nr,"Ulceration"])-1)+
+              coef.Rec["Loc_CAT=leg"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="leg")+
+              coef.Rec["Loc_CAT=trunk"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="trunk")+
+              coef.Rec["Loc_CAT=head & neck"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="head & neck")+
+              coef.Rec["Breslow"]*(log(data.to.be.imputed[patient.nr,"Breslow"])-c.Breslow)+
+              coef.Rec["Rdamcrit"]*log(data.to.be.imputed[patient.nr,"Rdamcrit"])+
+              coef.Rec["SNstatus=Positive * Breslow"]*(log(data.to.be.imputed[patient.nr,"Breslow"])-c.Breslow)*(as.numeric(data.to.be.imputed[patient.nr,"SNstatus"])-1)
+
+# lp check
+lp.score.Rec.true
+lp.score.Rec.try
+lp.score.Rec.manual-center.Rec
+
+# probability of recurrence with fun.event
+fun.event(lp=lp.score.Rec.try, h0=h0.Rec)
+# probability of recurrence manually
+1-exp(-h0.Rec*exp(lp.score.Rec.manual-center.Rec))
+
+# check calculation of lp of MSM
+lp.score.MSM.true <- predict(f.mi.BS.MSM.5)[patient.nr]
+lp.score.MSM.true
+lp.MSM.try <- MSM.cal.fact*(lp.score.Rec.manual-center.Rec)
+lp.MSM.try
+
+# probability of MSM with fun.event
+fun.event(lp=lp.MSM.try, h0=h0.MSM)
+# probability of MSM manually
+1-exp(-h0.MSM*exp(MSM.cal.fact*(lp.score.Rec.manual-center.Rec)))
+```
+
 ### Histogram for negative SN status patients
@@
 Breslow.trunc.neg[last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Breslow"]>7] <- 7
-lp.score.Rec.negative <- predict(f.mi.BS.Rec,newdata=data.frame(SNstatus=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "SNstatus"],
+lp.score.Rec.negative <- predict(f.mi.BS.Rec.5,newdata=data.frame(SNstatus=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "SNstatus"],
                                                        Age.SN=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Age.SN"],
@@
 Breslow.trunc.pos[last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Breslow"]>7] <- 7
-lp.score.Rec.positive <- predict(f.mi.BS.Rec,newdata=data.frame(SNstatus=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "SNstatus"],
+lp.score.Rec.positive <- predict(f.mi.BS.Rec.5,newdata=data.frame(SNstatus=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "SNstatus"],
                                                                 Age.SN=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Age.SN"],
@@
 ```{r, eval=TRUE}
-test.prop<-survival::cox.zph(f.mi.full.Rec$fits[[1]])
+test.prop<-survival::cox.zph(f.mi.full.Rec.5$fits[[1]])
 test.prop

### 2022-06-20 Code/PaperMelanoma.R

@@ -18,6 +18,34 @@ library(Hmisc)
 library(png)
 set.seed(100)
 
+#####
+##### HELPER FUNCTIONS FOR CALIBRATION
+#####
+km <- function(sel, S.km, horizon){
+  S.km.sel <- S.km[sel,]
+  sf <- survfit(S.km.sel~1)
+  year1 <- max(sf$time[sf$time<=horizon])
+  1-sf$surv[sf$time==year1]
+}
+km.lower <- function(sel, S.km, horizon){
+  S.km.sel <- S.km[sel,]
+  sf <- survfit(S.km.sel~1)
+  year1 <- max(sf$time[sf$time<=horizon])
+  1-sf$lower[sf$time==year1]
+}
+km.upper <- function(sel, S.km, horizon){
+  S.km.sel <- S.km[sel,]
+  sf <- survfit(S.km.sel~1)
+  year1 <- max(sf$time[sf$time<=horizon])
+  1-sf$upper[sf$time==year1]
+}
+fun.event <- function(lp, h0)
+{
+  h <- h0*exp(lp)
+  p <- 1-exp(-h)
+  return(p)
+}
+
 source("./Code/table.missing.values.R") # import function to create table of missing values
 source("./Code/val_surv_mi_v3.R")
 
@@ -43,6 +71,10 @@ sel.data <- original.data[original.data$EXCLUSIE!="Exclusion", selected.vars] # 
 cat("Number of patients excluded:", nrow(original.data)-nrow(sel.data), "\n")
 # Leave the patients included that have EXCLUSIE = 1 and Dbstatus = 8,
 # because diagnosis and operation date are very close but OK
+
+# CHECK with previous data
+cat("Number of negative patients excluded:",
+    sum(original.data$SNstatus=="Negative")-sum(sel.data$SNstatus=="Negative"), "\n")
 
 # Omit patient A0523 - no last follow-up date
 sel.data[which(is.na(sel.data$Last_FU)), "DatabaseNR"]
@@ -202,10 +234,15 @@ MSM.table
 #####
 ##### RESTRICT SURVIVAL
 #####
-S.Rec[S.Rec[,1]>5, 1] <- 5 # set recurrence time to 5
-S.Rec[S.Rec[,1]>5, 2] <- 0 # set censoring
-S.MSM[S.MSM[,1]>5, 1] <- 5 # set follow-up time to 5
[…]

### 2022-06-20 Code/old_analysis.R

@@ -1,0 +1,1533 @@
+# De belangrijkste primaire factoren die we mee zouden willen nemen zijn (waarvan de laatste 2 iets minder dan de eerste 6):
+# -	Leeftijd (continu)
+# -	Gender (male/female)
+# -	Locatie melanoom (o.a. arm, hoofd/hals)
+# -	Breslow dikte (continu)
+# -	Histologie (o.a. NM, ALM)
+# -	Ulceratie (yes/no). (Er zijn slechts 128 unknown = 4.1%, ik was in de war met andere factor, dus eigenlijk weinig unknown!).
+# -	Clark level. Dit was altijd een zeer sterke prognostische factor in allerlei studies, mogelijk wordt deze factor in de toekomst iets minder van belang maar we denken dat hij misschien toch wel meegenomen zou moeten worden.
+# -	Hoeveelheid verwijderde sentinel nodes (=Tot_nr_SNs)
+#
+# Daarnaast mogelijk interessant:
+# -	Hoeveelheid SN fields (totaal aantal locaties waar de SN is afgenomen, er zijn er een aantal met 2 of zelfs 3, zoals beide oksels en/of ??n lies)? En/of locatie SN field. Maar wij denken dat deze beide factoren te maken hebben met de locatie van het melanoom (bijv. een melanoom op de rug geeft de kans op meerdere SN fields, terwijl een melanoom op het been altijd maar ??n SN field heeft namelijk in de lies aan dezelfde kant). Locatie melanoom staat reeds hierboven genoemd om mee te nemen.
[…]

### 2022-06-20 Code/PaperMelanoma.Rmd
Last results

@@
 attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Negative)"]<-"Breslow + negative SN"
-png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Nomogram.png"))
 par(mar = rep(0, 4))
-plot(nom, total.sep.page=FALSE, col.grid=gray(c(0.8, 0.95)))
-grDevices::dev.off()
+plot(nom, total.sep.page=TRUE, col.grid=gray(c(0.8, 0.95)))
 nom.0

### 2022-06-30 Code/PaperMelanoma.Rmd
C-index negative and positive seperate added

@@
 ```{r, eval=TRUE}
+# 504 patients died due to melanoma, 102 patients died from other causes
+table(sel.data$Status, sel.data$dead)
 sel.data$MSM<-1*(sel.data$Status=="DOD")
@@
   Sex +
-  log(Age.SN) +
+  rcs(Age.SN) +
   Ulceration +
@@
   SNstatus:Sex +
-  SNstatus:log(Age.SN) +
+  SNstatus:rcs(Age.SN) +
   SNstatus:Ulceration +
@@
   Histology +
-  log(Breslow) +
+  rcs(Breslow) +
   multiple.fields +
@@
   SNstatus:Histology +
-  SNstatus:log(Breslow) +
+  SNstatus:rcs(Breslow) +
   SNstatus:multiple.fields
@@
 nonlinear.Breslow.full.Rec$loglik[2]
+# test
+anova(nonlinear.Breslow.full.Rec)
 ```
@@
   Tot_SNs_pos +
-  log(Rdamcrit) +
+  rcs(Rdamcrit) +
   Dewar +
@@
                                  fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
-# stats::anova(f.mi.full.Rec.5)
+stats::anova(f.mi.full.Rec.5)
 summary(f.mi.full.Rec.5, Tot_SNs_pos=c(0, 1))
@@
                            pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
+# backward selection
+# round(stats::anova(f.mi.BS.Rec.5)[order(stats::anova(f.mi.BS.Rec.5)[, 'P']),], 3)
 summary(f.mi.BS.Rec.5, Breslow=exp(c.Breslow))
 coef(f.mi.BS.Rec.5)
-# round(stats::anova(f.mi.BS.Rec.5)[order(stats::anova(f.mi.BS.Rec.5)[, 'P']),], 3)
 stats::anova(f.mi.BS.Rec.5)
@@
                                  pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
-f.mi.full.MSM.5.refit
 summary(f.mi.full.MSM.5.refit, Tot_SNs_pos=c(0, 1))
@@
                                   pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
-# round(anova(f.mi.BS.MSM.5.refit)[order(anova(f.mi.BS.MSM.5.refit)[, 'P']),], 3)
-f.mi.BS.MSM.5.refit
 summary(f.mi.BS.MSM.5.refit, Breslow=exp(c.Breslow))
@@
 ```{r, eval=TRUE}
-lp.BS<-f.mi.BS.Rec.5$linear
-f.mi.BS.MSM.5<-rms::cph(S.MSM.5~lp.BS,y=TRUE,x=TRUE)
+lp.BS <- f.mi.BS.Rec.5$linear
+f.mi.BS.MSM.5 <- rms::cph(S.MSM.5~lp.BS,y=TRUE,x=TRUE)
 MSM.cal.fact <- stats::coef(f.mi.BS.MSM.5)
@@
 
+# C-index stratified by SN status
+```{r, eval=TRUE}
+for (outcome in c("Rec.5", "MSM.5.refit")){
+  for (arm in c("Positive", "Negative")){
+    f.mi.stratified <- eval(parse(text=paste0('f.mi.BS.', outcome)))
+    if (outcome=="Rec.5"){
+      S.5 <- S.Rec.5
+    } else if (outcome=="MSM.5.refit"){
+      S.5 <- S.MSM.5
+    }
+    S.5.subset <- S.5[last.imputed.data.set$SNstatus==arm,]
+
+    cindex.stratified <- rep(0, m)
+    cindex.se.stratified <- rep(0, m)
+    for (i in 1:m){
+      subset.imputed.data <- as.data.frame(mice::complete(imputed.data,i)[last.imputed.data.set$SNstatus==arm,])
+      rc <- Hmisc::rcorr.cens(-predict(f.mi.stratified,newdata=subset.imputed.data, type="lp"), S.5.subset)
+      cindex.stratified[i]<-rc["C Index"]
+      cindex.se.stratified[i]<-rc["S.D."]/2
+    }
+    cindex.mi.stratified<-Rubin.combine(cindex.stratified,cindex.se.stratified)
+
+    print(c(outcome, arm, sprintf('%.2f', cindex.mi.stratified$est),
+    paste0("[", sprintf('%.2f', cindex.mi.stratified$est+qnorm(.025)*cindex.mi.stratified$se), 
+           "; ", sprintf('%.2f', cindex.mi.stratified$est+qnorm(.975)*cindex.mi.stratified$se), "]")))
+  }
+}
+
+```
+
 # Cross-validation recurrence 5-years
@@
 center.Rec <- f.mi.BS.Rec.5$center
-save(h0.Rec, h0.MSM, coef.Rec, c.Breslow, center.Rec,
+save(h0.Rec, h0.MSM, coef.Rec, c.Breslow, center.Rec, MSM.cal.fact,
      file=paste0('Z:/Project Melanoom/PaperMelanoma/Results/h0.Rec.MSM.Rdata'), compress=TRUE)
@@
 patient.nr <- 115 # 395 and 396 = Negative, 397 and 398 = Positive, 30 = Head & neck, 115 has large Breslow
-data.to.be.imputed[patient.nr, c("SNstatus", "Age.SN", "Ulceration", "Loc_CAT", "Breslow", "Rdamcrit")]
+print(data.to.be.imputed[patient.nr, c("SNstatus", "Age.SN", "Ulceration", "Loc_CAT", "Breslow", "Rdamcrit")])
+
+# calculate linear predictor
 lp.score.Rec.true <- predict(f.mi.BS.Rec.5)[patient.nr]

### 2022-06-30 Code/PaperMelanoma.R

@@ -54,9 +54,12 @@ source("./Code/val_surv_mi_v3.R")
 #####
 original.data <- foreign::read.spss(file="./Data/Dataset Predictiemodel Stassen.sav",
                                     to.data.frame=TRUE, origin="01-01-1970")
+cat("No SN status:", sum(original.data$SNstatus=="No SN"))
+cat("Duplicates:", sum(original.data$COMMENTS=="double patient!                                                                                                                                                                                         "))
+unique(original.data[original.data$EXCLUSIE=="Exclusion", "COMMENTS"])
 
 # Select necessary variables
-selected.vars <- c("EXCLUSIE", "DatabaseNR", "Center", "Sex", "Age", "DOB",
+selected.vars <- c("EXCLUSIE", "COMMENTS", "DatabaseNR", "Center", "Sex", "Age", "DOB",
                    "Ulceration", "Loc_CAT",            # OLD data set: Loc_CAT2
                    "Histology", "Breslow", "Tot_nr_SNs",
                    "nr_SNs_Field_1", "nr_SNs_Field_2", "nr_SNs_Field_3",

### 2022-11-29 Code/0_PrepareTrainingData.Rmd
Validation data added

---
title: "Prepare Training Data"
author: "Carolien C.H.M. Maas"
date: "14-11-2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Training data preparation

### Load packages, functions, and data 
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# erase memory
rm(list = ls(all.names = TRUE))

# for reproducibility
set.seed(100)

# load libraries
library(foreign)
library(rms)
source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")

# load data file
original.train.data <- foreign::read.spss(file="Z:/Project Melanoom/PaperMelanoma/Data/training/Dataset Predictiemodel.sav",
                                    to.data.frame=TRUE, origin="01-01-1970")

# Select necessary variables
selected.vars <- c("EXCLUSIE", "DatabaseNR", "Center", "Sex", "Age", "DOB",
                   "Ulceration", "Loc_CAT",            # OLD data set: Loc_CAT2
                   "Histology", "Breslow", "Tot_nr_SNs",
                   "nr_SNs_Field_1", "nr_SNs_Field_2", "nr_SNs_Field_3",
                   "Mitosis",                          # needed for AJCC.7.i
                   "SNdate", "Recurrence", "Date_1st_Rec", "Last_FU",
                   "dead", "Status",
                   "SNstatus",                         # distinguish between positive and negative SN patients
                   "Tot_SNs_pos", "Rdamcrit", "Dewar") # additional for positive SN patients
```

### Inclusion criteria
Exclude when Exclusion = 2: based on duplicates, urogenital melanoma, etc. Note:
Leave the patients included that have EXCLUSIE = 1 and Dbstatus = 8 (Dbstatus=0: 
negative, Dbstatus=1: positive), even though diagnosis and operation date are 
very close.
```{r, eval=TRUE}
sel.train.data <- original.train.data[original.train.data$EXCLUSIE!="Exclusion", selected.vars]
cat("Number of patients excluded:", nrow(original.train.data)-nrow(sel.train.data), "\n")
```
Omit one patient that has no last follow-up date
```{r, eval=TRUE, echo=FALSE}
sel.train.data <- sel.train.data[!is.na(sel.train.data$Last_FU),]
```

### Recode
#### Use age at SN instead of age at diagnosis
```{r, eval=TRUE}
sel.train.data$Age.SN <- floor((sel.train.data$SNdate-sel.train.data$DOB)/(24*60*60)/365.25)
```

#### Location
Merge location other with trunk
```{r, eval=TRUE}
sel.train.data$Loc_CAT[sel.train.data$Loc_CAT=="other"] <- "trunk"
```

Merge LMM other with other
```{r, eval=TRUE}
sel.train.data$Histology[sel.train.data$Histology=="LMM"] <- "other"
```

#### Define multiple fields
```{r, eval=TRUE}
sel.train.data$multiple.fields <- ifelse(!is.na(sel.train.data$nr_SNs_Field_2)&sel.train.data$nr_SNs_Field_2>0,1,0)
```

#### Number of positive and negative nodes
Check: there are no negative patients that have positive nodes
```{r, eval=TRUE}
sum(!is.na(sel.train.data$Tot_SNs_pos)&sel.train.data$Tot_SNs_pos>0&sel.train.data$SNstatus=="Negative")
```

Create number of negative nodes removed
```{r, eval=TRUE}
sel.train.data$Tot_SNs_neg <- sel.train.data$Tot_nr_SNs - sel.train.data$Tot_SNs_pos
```

467 positive SN patients have negative nodes
```{r, eval=TRUE}
sum(!is.na(sel.train.data$Tot_SNs_neg)&sel.train.data$Tot_SNs_neg>0&sel.train.data$SNstatus=="Positive")
```

#### SN tumour burden
11 patients with positive SN have SN tumour burden of exactly 1
```{r, eval=TRUE}
sum(sel.train.data$Rdamcrit==1&sel.train.data$SNstatus=="Positive"&!is.na(sel.train.data$Rdamcrit))
```

Set Rdamcrit of negative SN patients equal to 0
```{r, eval=TRUE}
sel.train.data$Rdamcrit[sel.train.data$SNstatus=="Negative"] <- 1 # since log(1) = 0
```

#### Location of metastasis
Set Dewar for negative patients to subcap
```{r, eval=TRUE}
sum(sel.train.data$SNstatus=="Positive"&sel.train.data$Dewar=="subcap"&!is.na(sel.train.data$Dewar))
sel.train.data$Dewar[sel.train.data$SNstatus=="Negative"] <- "subcap"
```

Merge single cells with subcap 
```{r, eval=TRUE}
sel.train.data$Dewar[sel.train.data$Dewar=="single cells"] <- "subcap"
```

#### Recurrence 
Define time since diagnosis in years (follow-up time)
```{r, eval=TRUE}
sel.train.data$FU.time <- (sel.train.data$Last_FU-sel.train.data$SNdate)/(24*60*60)/365.25
```

Define recurrence time in years
```{r, eval=TRUE}
sel.train.data$Rec.time <- (sel.train.data$Date_1st_Rec-sel.train.data$SNdate)/(24*60*60)/365.25
```

Omit five patients with Rec date before SN date
```{r, eval=TRUE}
old.n <- nrow(sel.train.data)
sel.train.data <- sel.train.data[!(sel.train.data$Rec.time<0&!is.na(sel.train.data$Rec.time)),]
cat("Number of patients excluded:", old.n-nrow(sel.train.data), "\n")
```

Set recurrence time equal to follow-up time for those without recurrence
```{r, eval=TRUE}
sel.train.data$Rec.time[sel.train.data$Recurrence=="no recurrence"|is.na(sel.train.data$SNdate)]<-sel.train.data$FU.time[sel.train.data$Recurrence=="no recurrence"|is.na(sel.train.data$SNdate)]
table(sel.train.data$Status, sel.train.data$Recurrence)
```

Impute recurrence time for 45 patients with missing Rec.time but Recurrence=="recurred"
```{r, eval=TRUE}
cat('Number of patients dead with recurrence without Date_1st_Rec:', sum(is.na(sel.train.data$Date_1st_Rec)&sel.train.data$Recurrence=="recurred"&sel.train.data$dead=="Yes"), '\n')
cat('Number of patients alive with recurrence without Date_1st_Rec:', sum(is.na(sel.train.data$Date_1st_Rec)&sel.train.data$Recurrence=="recurred"&sel.train.data$dead=="No"))
```

#### Melanoma specific mortality (MSM)
```{r, eval=TRUE}
# 504 patients died due to melanoma, 102 patients died from other causes
table(sel.train.data$Status, sel.train.data$dead)
sel.train.data$MSM<-1*(sel.train.data$Status=="DOD")
```

#### Mitosis
Set Mitosis to Unknown if missing
```{r, eval=TRUE}
sel.train.data$Mitosis[is.na(sel.train.data$Mitosis)] <- "Unknown" # mitosis unknown for 80% of patients
```

#### Replace "Unknown" by NA
```{r, eval=TRUE}
sel.train.data$Ulceration[sel.train.data$Ulceration=="unknown"]<-NA
sel.train.data$Loc_CAT[sel.train.data$Loc_CAT=="unknown"]<-NA
sel.train.data$Histology[sel.train.data$Histology=="unknown"]<-NA
sel.train.data$Dewar[sel.train.data$Dewar=="unknown"] <- NA
sel.train.data$Mitosis[sel.train.data$Mitosis=="Unknown"] <- NA
```

#### Delete unused levels
```{r, eval=TRUE}
sel.train.data <- droplevels(sel.train.data)

# set levels for multiple.fields
sel.train.data$multiple.fields <- factor(sel.train.data$multiple.fields, labels=c("Other", "Multiple SN fields"))
sel.train.data$MSM <- factor(sel.train.data$MSM, labels=c("Other", "Death by melanoma"))
```

# Multiple imputation
```{r, eval=TRUE}
variables.to.be.imputed <- c("Center", "SNstatus", "Sex", "Age.SN", "Ulceration",
                             "Loc_CAT", "Histology", "Breslow", "multiple.fields",
                             "Tot_SNs_neg", "Tot_SNs_pos", "Rdamcrit", "Dewar",
                             "Mitosis",
                             "Recurrence", "dead", "MSM")
training.data.to.be.imputed <- sel.train.data[, variables.to.be.imputed]
```

Save distributions of all variables, necessary for Harrell summaries
```{r, eval=TRUE}
dd.training<-rms::datadist(training.data.to.be.imputed)
options(datadist='dd.training')
options(digits=8)
```

Relation between outcome and covariates
```{r, eval=TRUE}
training.data.to.be.imputed$logRec.time <- log(sel.train.data$Rec.time+1/365.25)
training.data.to.be.imputed$logFU.time <- log(sel.train.data$FU.time+1/365.25)
missings.table(training.data.to.be.imputed)
save(list=c("dd.training", "training.data.to.be.imputed"),file='Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata', compress=TRUE)
```

### 2022-11-29 Code/0_PrepareValidationData.Rmd
Validation data added

---
title: "Prepare Validation Data"
author: "Carolien C.H.M. Maas"
date: "14-11-2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Validation data preparation

### Load packages, functions, and data 
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# erase memory
rm(list = ls(all.names = TRUE))

# for reproducibility
set.seed(100)

# load libraries
library(foreign)
library(rms)
source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")

# load data file
original.validation.data <- foreign::read.spss(file="Z:/Project Melanoom/PaperMelanoma/Data/validation/Validation_cohort_MIA.sav",
                                               to.data.frame=TRUE, origin="01-01-1970")

# remove space after Negative # Bespreken met Robert
levels(original.validation.data$SNstatus)[levels(original.validation.data$SNstatus)=="Negative "] <- "Negative"

# all patients may be included
sel.val.data <- original.validation.data
```

### Recode
#### Use age at SN instead of age at diagnosis
```{r, eval=TRUE}
sel.val.data$Age.SN <- floor((sel.val.data$SNdate-sel.val.data$DOB)/(24*60*60)/365.25)
```

#### Location
Merge location other with trunk
```{r, eval=TRUE}
sel.val.data$Loc_CAT[sel.val.data$Loc_CAT=="other"] <- "trunk"
```

Merge LMM other with other
```{r, eval=TRUE}
sel.val.data$Histology[sel.val.data$Histology=="LLM"] <- "other" # Bespreken met Robert, is LLM and LMM the same?
```

#### Define multiple fields
```{r, eval=TRUE}
sel.val.data$multiple.fields <- as.numeric(!is.na(sel.val.data$Tot_nr_SNs)&sel.val.data$Tot_nr_SNs>1) # Bespreken met ask Robert
```

#### Number of positive and negative nodes
Check: there are no negative patients that have positive nodes
```{r, eval=TRUE}
sum(!is.na(sel.val.data$Tot_SNs_pos)&sel.val.data$Tot_SNs_pos>0&sel.val.data$SNstatus=="Negative")
```

Create number of negative nodes removed
```{r, eval=TRUE}
sel.val.data$Tot_nr_SNs <- sel.val.data$Tot_SNs_neg + sel.val.data$Tot_SNs_pos
```

607 positive SN patients have negative nodes
```{r, eval=TRUE}
sum(!is.na(sel.val.data$Tot_SNs_neg)&sel.val.data$Tot_SNs_neg>0&sel.val.data$SNstatus=="Positive")
```

#### SN tumour burden
43 patients with positive SN have SN tumour burden of exactly 1
```{r, eval=TRUE}
sum(sel.val.data$Rdamcrit==1&sel.val.data$SNstatus=="Positive"&!is.na(sel.val.data$Rdamcrit))
```

Set Rdamcrit of negative SN patients equal to 0, SNstatus will capture the effect of these patients
```{r, eval=TRUE}
sel.val.data$Rdamcrit[sel.val.data$SNstatus=="Negative"] <- 1 # since log(1) = 0
```

#### Location of metastasis
Set Dewar for negative patients to subcap to merge them with the reference category, SN status will capture the effect of Negative patients
```{r, eval=TRUE}
sel.val.data$Dewar <- sel.val.data$Subcap
sel.val.data$Dewar[sel.val.data$SNstatus=="Negative"] <- "subcap"

# 339 patients have subcap with positive SN
sum(sel.val.data$SNstatus=="Positive"&sel.val.data$Dewar=="subcap"&!is.na(sel.val.data$Dewar))
```
#### Recurrence 
Define time since diagnosis in years (follow-up time)
```{r, eval=TRUE}
sel.val.data$FU.time <- (sel.val.data$Last_FU-sel.val.data$SNdate)/(24*60*60)/365.25
```

Define recurrence time in years
```{r, eval=TRUE}
sel.val.data$Rec.time <- (sel.val.data$Date_1st_Rec-sel.val.data$SNdate)/(24*60*60)/365.25
```

Check: no patients with Rec date before SN date
```{r, eval=TRUE}
sum(!(!(sel.val.data$Rec.time<0&!is.na(sel.val.data$Rec.time))))
```

Impute recurrence time for 45 patients with missing Rec.time but Recurrence=="recurred"
```{r, eval=TRUE}
cat('Number of patients dead with recurrence without Date_1st_Rec:', sum(is.na(sel.val.data$Date_1st_Rec)&sel.val.data$Status=="Dead, melanoma       "), '\n')
cat('Number of patients alive with recurrence without Date_1st_Rec:', sum(is.na(sel.val.data$Date_1st_Rec)&sel.val.data$Status=="Alive with Melanoma  "))
```

Set recurrence indicator
```{r, eval=TRUE}
Rec.date <- as.numeric(!is.na(sel.val.data$Date_1st_Rec)) # n=1294
Rec.date <- factor(Rec.date, labels=c("no recurrence", "recurred"))
table(sel.val.data$Status, Rec.date) # Bespreken met Robert

cat("Number of alive patients with Date_1st_Rec, but status NSR:", sum(!is.na(sel.val.data$Date_1st_Rec)&sel.val.data$Status=="Alive NSR            "), "\n")
cat("Number of alive patients with Date_1st_Rec, but status unknown:", sum(!is.na(sel.val.data$Date_1st_Rec)&sel.val.data$Status=="Alive, status unknown"), "\n")
cat("Number of dead patients with Date_1st_Rec, but status unknown:", sum(!is.na(sel.val.data$Date_1st_Rec)&sel.val.data$Status=="Dead, cause unknown  "), "\n")
cat("Number of dead patients with Date_1st_Rec, but status not melanoma:", sum(!is.na(sel.val.data$Date_1st_Rec)&sel.val.data$Status=="Dead, not melanoma   "), "\n")

# Recurrence if patient has Date_1st_Rec, Alive with melanoma, or dead with melanoma
sel.val.data$Recurrence <- as.numeric(sel.val.data$Status=="Alive with Melanoma  " | sel.val.data$Status=="Dead, melanoma       " | !is.na(sel.val.data$Date_1st_Rec)) # n=1352
sum(sel.val.data$Recurrence)
```

#### Melanoma specific mortality (MSM)
```{r, eval=TRUE}
sel.val.data$dead <- as.numeric(stringr::str_detect(sel.val.data$Status, "Dead"))

# 504 patients died due to melanoma, 102 patients died from other causes
table(sel.val.data$Status, sel.val.data$dead)

sel.val.data$MSM <- as.numeric(sel.val.data$Status=="Dead, melanoma       ")
```
#### Mitosis
Define Mitosis as Yes/No
```{r, eval=TRUE}
sel.val.data$Mitosis <- as.numeric(!is.na(sel.val.data$Mitosis))
```

#### Delete unused levels
```{r, eval=TRUE}
sel.val.data <- droplevels(sel.val.data)
sel.val.data$Center <- "Australia"
```

#### Set levels
```{r, eval=TRUE}
sel.val.data$Ulceration <- factor(sel.val.data$Ulceration, labels=c("No", "Yes"))
sel.val.data$Mitosis <- factor(sel.val.data$Mitosis, labels=c("Yes", "No"))

sel.val.data$multiple.fields <- factor(sel.val.data$multiple.fields, labels=c("Multiple SN fields", "Other"))

sel.val.data$Recurrence <- factor(sel.val.data$Recurrence, labels=c("no recurrence", "recurred"))
sel.val.data$dead <- factor(sel.val.data$dead, labels=c("No", "Yes"))
sel.val.data$MSM <- factor(sel.val.data$MSM, labels=c("Other", "Death by melanoma"))
```

# Multiple imputation
```{r, eval=TRUE}
variables.to.be.imputed <- c("Center", "SNstatus", "Sex", "Age.SN", "Ulceration",
                             "Loc_CAT", "Histology", "Breslow", "multiple.fields",
                             "Tot_SNs_neg", "Tot_SNs_pos", "Rdamcrit", "Dewar",
                             "Mitosis",
                             "Recurrence", "dead", "MSM")
validation.data.to.be.imputed <- sel.val.data[, variables.to.be.imputed]
```

Save distributions of all variables, necessary for Harrell summaries
```{r, eval=TRUE}
dd.validation<-rms::datadist(validation.data.to.be.imputed)
options(datadist='dd.validation')
options(digits=8)
```

Relation between outcome and covariates
```{r, eval=TRUE}
validation.data.to.be.imputed$logRec.time <- log(sel.val.data$Rec.time+1/365.25)
validation.data.to.be.imputed$logFU.time <- log(sel.val.data$FU.time+1/365.25)
missings.table(validation.data.to.be.imputed)
save(list=c("dd.validation", "validation.data.to.be.imputed"),file='Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.Rdata', compress=TRUE)
```

### 2022-11-29 Code/10_CreateRiskScoreHistogram.Rmd
Validation data added

---
title: "Create histogram with risk scores"
author: "Carolien C.H.M. Maas"
date: "14-11-2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Create calibration plots

### Load packages, functions, and data 
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# erase memory
rm(list = ls(all.names = TRUE))

# for reproducibility
set.seed(100)

# load libraries
library(grDevices)
library(stats)
library(survival)
source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")
```

# Risk score distribution
```{r, eval=TRUE}
load('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData')
# baseline hazard of MSM
f.MSM.basehaz <- survival::basehaz(f.mi.BS.MSM.5,TRUE)
h0.MSM <- f.MSM.basehaz$hazard[f.MSM.basehaz$time==max(f.MSM.basehaz$time[f.MSM.basehaz$time<=horizon])]
lp.MSM <- f.mi.BS.MSM.5$linear

# lp range MSM
rc.MSM <- f.mi.BS.MSM.5$coef[1]*rc.Rec
int.MSM <- f.mi.BS.MSM.5$coef[1]*int.Rec
max.range <- 16
points.range <- 0:max.range
lp.points.Rec.5 <- int.Rec+rc.Rec*points.range
lp.points.MSM.5 <- int.MSM+rc.MSM*points.range

# Recurrence lp scores
Breslow.trunc <- last.imputed.training.data.set$Breslow
Breslow.trunc[last.imputed.training.data.set$Breslow<0.1] <- .1
Breslow.trunc[last.imputed.training.data.set$Breslow>7] <- 7
lp.score.Rec <- predict(f.mi.BS.Rec.5,newdata=data.frame(SNstatus=last.imputed.training.data.set$SNstatus,
                                                         Age.SN=last.imputed.training.data.set$Age.SN,
                                                         Ulceration=last.imputed.training.data.set$Ulceration,
                                                         Loc_CAT=last.imputed.training.data.set$Loc_CAT,
                                                         Breslow=Breslow.trunc,
                                                         Rdamcrit=last.imputed.training.data.set$Rdamcrit))
mean(lp.score.Rec)
score.Rec <- round((lp.score.Rec-int.Rec)/rc.Rec,0)
max.breaks <- max.range+1
h.Rec <- hist(score.Rec,plot=FALSE,breaks=0:max.breaks,right=FALSE)

# score table
data.frame(table(score.Rec)/length(score.Rec))
```

# Plot of risk score distribution
```{r, eval=TRUE, message=FALSE}
h.plot <- h.Rec
h.plot$density <- 100*h.Rec$density
p.Rec <- fun.event(lp=lp.points.Rec.5, h0=h0.Rec)
p.MSM <- fun.event(lp=lp.points.MSM.5, h0=h0.MSM)
x.lim <- c(0, max.range+1)
y.lim <- c(0, 100)
y.lim.hist <- c(0, 25)
png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Risk.distribution.png"))
par(mar = c(5,5,2,5))
plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col='white')
abline(h=(0:30)*1,col="light grey",lty=2)
axis(side = 4)
mtext(side = 4, line = 3, 'Risk score distribution (%)')
par(new = TRUE)
plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col="white")
par(new = TRUE)
plot(points.range,p.Rec*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=15, xlab="Risk score",ylab="Predicted 5-year risk (%)")
points(points.range,p.MSM*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=16)
legend(x=0,y=100,legend=c("Recurrence","Melanoma-specific mortality"),pch=c(15,16),cex=.9,bg='white')
grDevices::dev.off()
```

```{r, eval=TRUE, echo=FALSE}
par(mar = c(5,5,2,5))
plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col='white')
abline(h=(0:30)*1,col="light grey",lty=2)
axis(side = 4)
mtext(side = 4, line = 3, 'Risk score distribution (%)')
par(new = TRUE)
plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col="white")
par(new = TRUE)
plot(points.range,p.Rec*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=15, xlab="Risk score",ylab="Predicted 5-year risk (%)")
points(points.range,p.MSM*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=16)
legend(x=0,y=100,legend=c("Recurrence","Melanoma-specific mortality"),pch=c(15,16),cex=.9,bg='white')

# table of risk distribution
cbind(points.range, round(p.Rec*100, 1), round(p.MSM*100, 1))[c(1, 9, 10, 12, 13),]
```
# Save for Rshiny app
```{r, eval=TRUE}
coef.Rec <- stats::coef(f.mi.BS.Rec.5)
center.Rec <- f.mi.BS.Rec.5$center
save(h0.Rec, h0.MSM, coef.Rec, c.Breslow, center.Rec, MSM.cal.fact,
     file=paste0('Z:/Project Melanoom/PaperMelanoma/Results/h0.Rec.MSM.Rdata'), compress=TRUE)
save(list=c("dd.training", "training.data.to.be.imputed", "m", "imputed.training.data", "last.imputed.training.data.set", "horizon", "S.Rec.5.training", "S.MSM.5.training",
            "f.mi.full.Rec.5", "f.mi.BS.Rec.5", "f.mi.full.MSM.5.refit", "f.mi.BS.MSM.5.refit", "f.mi.BS.MSM.5", "MSM.cal.fact", "c.Breslow", 
            "form.full.Rec.5", "form.BS.Rec.5", "form.full.MSM.5.refit", "form.BS.MSM.5.refit", 
            "rc.Rec", "int.Rec", "h0.Rec", "h0.MSM", "coef.Rec", "center.Rec", "h.Rec", "lp.points.Rec.5", "lp.points.MSM.5", "lp.score.Rec", "lp.MSM", 
            "max.breaks", "h.plot", "y.lim.hist", "x.lim"),
     file='Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata', compress=TRUE)
```

### 2022-11-29 Code/11_TestModel.Rmd
Validation data added

---
title: "Create histogram with risk scores"
author: "Carolien C.H.M. Maas"
date: "14-11-2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Create calibration plots

### Load packages, functions, and data 
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# erase memory
rm(list = ls(all.names = TRUE))

# for reproducibility
set.seed(100)

# load libraries
library(rms)
library(survival)
source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")
```

# Test if model is correct
### Check calibration
Intercept of recurrence should be the same as when setting covariates to 'zero' values minus the center
```{r, eval=TRUE}
load('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData')

int.Rec
20*f.mi.BS.Rec.5$coef["Age.SN"]+
  (log(0.1)-c.Breslow)*f.mi.BS.Rec.5$coef["Breslow"]+
  log(0.01)*f.mi.BS.Rec.5$coef["Rdamcrit"]-
  f.mi.BS.Rec.5$center
```

### Check mean event rates of risk distribution
#### Recurrence
```{r, eval=TRUE}
sum(h.Rec$density*fun.event(lp=lp.points.Rec.5, h0=h0.Rec))
# equal to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
SF<-survival::survfit(S.Rec.5.training~1,data=training.data.to.be.imputed)
100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
# average probability of recurrence
mean(fun.event(lp=lp.score.Rec, h0=h0.Rec))
```

#### MSM
```{r, eval=TRUE}
sum(h.Rec$density*fun.event(lp=lp.points.MSM.5, h0=h0.MSM))
# equal to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
SF<-survfit(S.MSM.5.training~1,data=training.data.to.be.imputed)
100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
# average probability of MSM
mean(fun.event(lp=lp.MSM, h0=h0.MSM))
```

# Check if predictions of new and old model align for negative SN patients
## New model
```{r, eval=TRUE}
lp.negative <- predict(f.mi.BS.Rec.5,
                       newdata=last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Negative",],
                       type="lp")
new.negativeSN.tool <- 1-exp(-h0.Rec*exp(lp.negative))
```

## Old model with new data
```{r, eval=TRUE}
rscript.negative <- function(breslow, ulceration, location) {
  S0 <- 0.85837329
  lp <- 0.79351989*log(breslow) +
    0.61912477*I(ulceration == 1) +
    0.29594843*I(location == 1) +
    0.42802046*I(location == 2) +
    0.86053044*I(location == 3) - 0.98860941
  S <- 100 - 100*S0^exp(lp)
  return(S)
}
old.negativeSN.tool <- rscript.negative(breslow=last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Negative", "Breslow"],
                                        ulceration=as.numeric(last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Negative", "Ulceration"])-1,
                                        location=as.numeric(last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Negative", "Loc_CAT"])-1)
```

## Compare
```{r, eval=TRUE}
plot(x=new.negativeSN.tool*100, y=old.negativeSN.tool,
     xlim=c(0, 100), ylim=c(0, 100),
     xlab="New predictions", ylab="Old predictions",
     main="Predicted recurrence of patients with negative SN status")
abline(a=0, b=1, col="red")
```

## Refitted old model new data
```{r, eval=TRUE}
refitted.old.negative <- rms::cph(S.Rec.5.training ~ log(Breslow)+Ulceration+Loc_CAT,
                                  data=last.imputed.training.data.set,
                                  y=TRUE, x=TRUE,
                                  subset=last.imputed.training.data.set$SNstatus=="Negative")
refitted.old.negative
f.Rec.basehaz.neg <- basehaz(refitted.old.negative,TRUE)
h0.neg <- f.Rec.basehaz.neg$hazard[f.Rec.basehaz.neg$time==max(f.Rec.basehaz.neg$time[f.Rec.basehaz.neg$time<=horizon])]
refit.old.negativeSN.tool <- fun.event(predict(refitted.old.negative),  h0.neg)
```

## Compare
```{r, eval=TRUE}
plot(x=new.negativeSN.tool*100, y=refit.old.negativeSN.tool*100,
     xlim=c(0, 100), ylim=c(0, 100),
     xlab="New predictions", ylab="Old predictions",
     main="Predicted recurrence of patients with negative SN status")
abline(a=0, b=1, col="red")
```

## Check mean recurrence probabilities
```{r, eval=TRUE}
SF<-survfit(S.Rec.5.training[last.imputed.training.data.set$SNstatus=="Negative"]~1)
100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
mean(new.negativeSN.tool)
mean(refit.old.negativeSN.tool)
mean(old.negativeSN.tool)
```

# Check if predictions of new and old model align for positive SN patients
## New model
```{r, eval=TRUE}
lp.positive <- predict(f.mi.BS.Rec.5,
                       newdata=last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Positive",],
                       type="lp")
new.positiveSN.tool <- 1-exp(-h0.Rec*exp(lp.positive))
```

## Old model with new data
```{r, eval=TRUE}
rscript.positive <- function(age, ulceration, tumor_burden, breslow) {
  S0 <- 0.51762362
  lp <- 0.56109945 * log(age) +
    0.23383770 * log(tumor_burden) +
    0.35489235 * log(breslow) +
    0.36562098 * I(ulceration == 1) - 2.709225
  S <- 100 - 100 * S0^exp(lp)
  return(S)
}
old.positiveSN.tool <- rscript.positive(age=last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Positive", "Age.SN"],
                                        ulceration=as.numeric(last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Positive", "Ulceration"])-1,
                                        breslow=last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Positive", "Breslow"],
                                        tumor_burden=as.numeric(last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Positive", "Rdamcrit"]))
```

## Compare
```{r, eval=TRUE}
plot(x=new.positiveSN.tool*100, y=old.positiveSN.tool,
     xlim=c(0, 100), ylim=c(0, 100),
     xlab="New predictions", ylab="Old predictions",
     main="Predicted recurrence of patients with positive SN status")
abline(a=0, b=1, col="red")
```

## Refitted old model new data
```{r, eval=TRUE}
refitted.old.positive <- rms::cph(S.Rec.5.training ~ log(Breslow)+Ulceration+Loc_CAT,
                                  data=last.imputed.training.data.set,
                                  y=TRUE, x=TRUE,
                                  subset=last.imputed.training.data.set$SNstatus=="Positive")
refitted.old.positive
f.Rec.basehaz.pos <- basehaz(refitted.old.positive,TRUE)
h0.pos <- f.Rec.basehaz.pos$hazard[f.Rec.basehaz.pos$time==max(f.Rec.basehaz.pos$time[f.Rec.basehaz.pos$time<=horizon])]
refit.old.positiveSN.tool <- fun.event(predict(refitted.old.positive),  h0.pos)
```

## Compare
```{r, eval=TRUE}
plot(x=new.positiveSN.tool*100, y=refit.old.positiveSN.tool*100,
     xlim=c(0, 100), ylim=c(0, 100),
     xlab="New predictions", ylab="Old predictions",
     main="Predicted recurrence of patients with positive SN status")
abline(a=0, b=1, col="red")
```

## Check mean recurrence probabilities
```{r, eval=TRUE}
SF<-survfit(S.Rec.5.training[last.imputed.training.data.set$SNstatus=="Positive"]~1)
100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
mean(new.positiveSN.tool)
mean(refit.old.positiveSN.tool)
mean(old.positiveSN.tool)
```

#### Example to check probabilities with app
```{r, eval=TRUE}
patient.nr <- 115 # 395 and 396 = Negative, 397 and 398 = Positive, 30 = Head & neck, 115 has large Breslow
print(training.data.to.be.imputed[patient.nr, c("SNstatus", "Age.SN", "Ulceration", "Loc_CAT", "Breslow", "Rdamcrit")])

# calculate linear predictor
lp.score.Rec.true <- predict(f.mi.BS.Rec.5)[patient.nr]
lp.score.Rec.try <- predict(f.mi.BS.Rec.5, newdata=data.frame(SNstatus=training.data.to.be.imputed[patient.nr,"SNstatus"],
                                                              Age.SN=training.data.to.be.imputed[patient.nr,"Age.SN"],
                                                              Ulceration=training.data.to.be.imputed[patient.nr,"Ulceration"],
                                                              Loc_CAT=training.data.to.be.imputed[patient.nr,"Loc_CAT"],
                                                              Breslow=training.data.to.be.imputed[patient.nr,"Breslow"],
                                                              Rdamcrit=training.data.to.be.imputed[patient.nr,"Rdamcrit"]),
                            type="lp")
p.Rec.example <- fun.event(lp=lp.score.Rec.try, h0=h0.Rec)
p.MSM.example <- fun.event(lp=lp.score.Rec.try, h0=h0.MSM)

# manual calculation
lp.score.Rec.manual <- coef.Rec["SNstatus=Positive"]*(as.numeric(training.data.to.be.imputed[patient.nr,"SNstatus"])-1)+
  coef.Rec["Age.SN"]*training.data.to.be.imputed[patient.nr,"Age.SN"]+
  coef.Rec["Ulceration=Yes"]*(as.numeric(training.data.to.be.imputed[patient.nr,"Ulceration"])-1)+
  coef.Rec["Loc_CAT=leg"]*as.numeric(training.data.to.be.imputed[patient.nr,"Loc_CAT"]=="leg")+
  coef.Rec["Loc_CAT=trunk"]*as.numeric(training.data.to.be.imputed[patient.nr,"Loc_CAT"]=="trunk")+
  coef.Rec["Loc_CAT=head & neck"]*as.numeric(training.data.to.be.imputed[patient.nr,"Loc_CAT"]=="head & neck")+
  coef.Rec["Breslow"]*(log(training.data.to.be.imputed[patient.nr,"Breslow"])-c.Breslow)+
  coef.Rec["Rdamcrit"]*log(training.data.to.be.imputed[patient.nr,"Rdamcrit"])+
  coef.Rec["SNstatus=Positive * Breslow"]*(log(training.data.to.be.imputed[patient.nr,"Breslow"])-c.Breslow)*(as.numeric(training.data.to.be.imputed[patient.nr,"SNstatus"])-1)

# lp check
lp.score.Rec.true
lp.score.Rec.try
lp.score.Rec.manual-center.Rec

# probability of recurrence with fun.event
fun.event(lp=lp.score.Rec.try, h0=h0.Rec)
# probability of recurrence manually
1-exp(-h0.Rec*exp(lp.score.Rec.manual-center.Rec))

# check calculation of lp of MSM
lp.score.MSM.true <- predict(f.mi.BS.MSM.5)[patient.nr]
lp.score.MSM.true
lp.MSM.try <- MSM.cal.fact*(lp.score.Rec.manual-center.Rec)
lp.MSM.try

# probability of MSM with fun.event
fun.event(lp=lp.MSM.try, h0=h0.MSM)
# probability of MSM manually
1-exp(-h0.MSM*exp(MSM.cal.fact*(lp.score.Rec.manual-center.Rec)))
```

### Histogram for negative SN status patients
```{r, eval=TRUE}
Breslow.trunc.neg <- last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Negative", "Breslow"]
Breslow.trunc.neg[last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Negative", "Breslow"]<0.1] <- .1
Breslow.trunc.neg[last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Negative", "Breslow"]>7] <- 7
lp.score.Rec.negative <- predict(f.mi.BS.Rec.5,newdata=data.frame(SNstatus=last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Negative", "SNstatus"],
                                                                  Age.SN=last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Negative", "Age.SN"],
                                                                  Ulceration=last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Negative", "Ulceration"],
                                                                  Loc_CAT=last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Negative", "Loc_CAT"],
                                                                  Breslow=Breslow.trunc.neg,
                                                                  Rdamcrit=last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Negative", "Rdamcrit"]))
mean(lp.score.Rec.negative)
# OK to use int.Rec and rc.Rec based on all patients?
score.Rec.neg <- round((lp.score.Rec.negative-int.Rec)/rc.Rec,0)
h.Rec.neg <- hist(score.Rec.neg,plot=FALSE,breaks=0:max.breaks,right=FALSE)
h.plot.neg <- h.Rec.neg
h.plot.neg$density <- 100*h.Rec.neg$density
```

### Histogram for positive SN status patients
```{r, eval=TRUE}
Breslow.trunc.pos <- last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Positive", "Breslow"]
Breslow.trunc.pos[last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Positive", "Breslow"]<0.1] <- .1
Breslow.trunc.pos[last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Positive", "Breslow"]>7] <- 7
lp.score.Rec.positive <- predict(f.mi.BS.Rec.5,newdata=data.frame(SNstatus=last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Positive", "SNstatus"],
                                                                  Age.SN=last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Positive", "Age.SN"],
                                                                  Ulceration=last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Positive", "Ulceration"],
                                                                  Loc_CAT=last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Positive", "Loc_CAT"],
                                                                  Breslow=Breslow.trunc.pos,
                                                                  Rdamcrit=last.imputed.training.data.set[last.imputed.training.data.set$SNstatus=="Positive", "Rdamcrit"]))
mean(lp.score.Rec.positive)
# OK to use int.Rec and rc.Rec based on all patients?
score.Rec.pos <- round((lp.score.Rec.positive-int.Rec)/rc.Rec,0)
h.Rec.pos <- hist(score.Rec.pos,plot=FALSE,breaks=0:max.breaks,right=FALSE)
h.plot.pos <- h.Rec.pos
h.plot.pos$density <- 100*h.Rec.pos$density

par(mfrow=c(1, 3))
plot(h.plot,freq=FALSE,axes=TRUE, xlab="Risk score", xlim=x.lim,ylim=y.lim.hist,
     ylab="Risk score distribution (%)", main="All patients",col="white")
plot(h.plot.neg,freq=FALSE,axes=TRUE, xlab="Risk score", xlim=x.lim,
     ylab="Risk score distribution (%)",ylim=y.lim.hist,main="Negative SN status",col="white")
plot(h.plot.pos,freq=FALSE,axes=TRUE, xlab="Risk score", xlim=x.lim,
     ylab="Risk score distribution (%)",ylim=y.lim.hist,main="Positive SN status",col="white")
```

### Test proportional hazards assumption
```{r, eval=TRUE}
test.prop<-survival::cox.zph(f.mi.full.Rec.5$fits[[1]])
test.prop
plot(test.prop)
```

### 2022-11-29 Code/12_Validation.Rmd
Validation data added

---
title: "Create histogram with risk scores"
author: "Carolien C.H.M. Maas"
date: "14-11-2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Create calibration plots

### Load packages, functions, and data 
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# erase memory
rm(list = ls(all.names = TRUE))

# for reproducibility
set.seed(100)

# load libraries
library(mice)
library(survival)         # needed to use
# remotes::install_github("CHMMaas/PredictionTools")
library(PredictionTools)  # to plot calibration
# import helper functions
source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")
```

# Validation
# Compute predictions for each multiple imputation and make calibration plots with metrics
```{r, eval=TRUE}
# laod data
load('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData')
options(datadist='dd.training')
options(digits=8)
load('Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData')
# load model
load('Z:/Project Melanoom/PaperMelanoma/Results/h0.Rec.MSM.Rdata')
p.Rec.val <- matrix(NA, nrow(validation.data.to.be.imputed), m)
p.MSM.val <- matrix(NA, nrow(validation.data.to.be.imputed), m)
for (i in 1:m){
  # Linear predictor
  lp.val.i <- predict(f.mi.BS.Rec.5,
                      newdata=mice::complete(imputed.validation.data, i),
                      type="lp")

  # Recurrence
  p.Rec.val[, i] <- 1-exp(-h0.Rec*exp(lp.val.i))

  # MSM
  p.MSM.val[, i] <- 1-exp(-h0.MSM*exp(MSM.cal.fact*lp.val.i))
}

# calibration curve for recurrence, TODO: based on dd.training?
png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/val.Rec.png"))
PredictionTools::val.surv.mi(p.Rec.val,S.Rec.5.validation,g=5,main="Calibration plot of recurrence for all patients",
                             time=NULL,lim=c(0,1),dist=TRUE)
grDevices::dev.off()

# calibration curve for MSM
png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/val.MSM.png"))
PredictionTools::val.surv.mi(p.MSM.val,S.MSM.5.validation,g=5,main="Calibration plot of MSM for all patients",
                             time=NULL,lim=c(0,1),dist=TRUE)
grDevices::dev.off()

```

### 2022-11-29 Code/1_DescriptivesTable.Rmd
Validation data added

---
output: word_document
---

```{r setup, include=FALSE}
# clear all
rm(list = ls(all.names = TRUE))
if(!is.null(dev.list())) dev.off()
cat('\014')

library(arsenal)
library(knitr)

load('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata')
load('Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.Rdata')
descr.table <- arsenal::tableby(Center ~ SNstatus
                                +Sex
                                +Age.SN
                                +arsenal::includeNA(Ulceration)
                                +arsenal::includeNA(Loc_CAT)
                                +arsenal::includeNA(Histology)
                                +Breslow
                                +multiple.fields
                                +Tot_SNs_neg
                                +Tot_SNs_pos
                                +Rdamcrit
                                +arsenal::includeNA(Dewar)
                                +arsenal::includeNA(Mitosis)
                                +Recurrence
                                +dead
                                +MSM
                                +logRec.time
                                +logFU.time, data=rbind(training.data.to.be.imputed, validation.data.to.be.imputed), 
                                total.pos="before",
                                digits = 1,
                                test = F,
                                total = T,
                                numeric.test = "kwt", cat.test = "chisq",
                                numeric.stats = c("meansd", "medianq1q3", "range", "Nmiss2"),
                                cat.stats = c("countpct"),
                                stats.labels = list(
                                  meansd = "Mean (SD)",
                                  medianq1q3 = "Median (Q1, Q3)",
                                  range = "Min - Max",
                                  Nmiss2 = "Missing"
                                ))
my_labels <- list(
  SNstatus = "SN status",
  Age.SN = "Age",
  Loc_CAT = "Location",
  multiple.fields = "Multiple fields",
  Tot_SNs_neg = "Total number of negative SN",
  Tot_SNs_pos = "Total number of positive SN",
  Rdamcrit = "SN tumour burden",
  Dewar = "Location metastasis in lymph",
  dead = "Deaths",
  MSM = "Melanoma specific mortality"
)
```

Table 1. Descriptive statistics.
```{r echo=FALSE}
knitr::kable(summary(descr.table, labelTranslations=my_labels))
```

### 2022-11-29 Code/2_ImputeData.Rmd
Validation data added

---
title: "Impute missing values of training and validation data using multiple imputation"
author: "Carolien C.H.M. Maas"
date: "14-11-2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Multiple imputation

### Load packages, functions, and data 
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# erase memory
rm(list = ls(all.names = TRUE))

# for reproducibility
set.seed(100)

# load libraries
library(mice)
```

Function to perform multiple imputation for training or validation data
```{r, eval=TRUE}
impute.missing.values <- function(type=NULL, m=1, new.imputation=FALSE){
  # load data
  load(paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/', type, '.data.Rdata'))

  if (new.imputation){
    # impute data
    data.to.be.imputed <- eval(parse(text=paste0(type, '.data.to.be.imputed')))
    imputed.data <- mice::mice(data.to.be.imputed, m=m, maxit=5, print=FALSE, seed=500)
    assign(paste0('imputed.', type, '.data'), imputed.data)
    save(list=c(paste0('imputed.', type, '.data')),
         file=paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/imputed.', type, '.data.Rdata'), compress=TRUE)
  }
  load(paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/imputed.', type, '.data.Rdata'))
  save(list=c(paste0('dd.', type), paste0(type, '.data.to.be.imputed'), 'm', paste0('imputed.', type, '.data')),
       file=paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/', type, '.data.Rdata'), compress=TRUE)
}
```

Perform multiple imputation
```{r, eval=TRUE}
# erase memory
m <- 5
new.imputation <- FALSE
impute.missing.values(type='training', m=m, new.imputation=new.imputation)
impute.missing.values(type='validation', m=m, new.imputation=new.imputation)

# training data gives warning messages for Mitosis
# validation data gives multicolinearity of logRectime and recurrence
# print(imputed.data$loggedEvents)
```

### 2022-11-29 Code/3_Descriptives.Rmd
Validation data added

---
title: "Multiple imputation"
author: "Carolien C.H.M. Maas"
date: "14-11-2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Describe data

### Load packages, functions, and data 
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# erase memory
rm(list = ls(all.names = TRUE))

# for reproducibility
set.seed(100)

# load libraries
library(mice)
library(survival)
library(survminer)
library(stats)
library(prodlim)
library(utils)
source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")
```

```{r, eval=TRUE}
prepare.data <- function(type){
  # Load imputed data set 
  load(paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/', type, '.data.RData'))
  imputed.data <- eval(parse(text=paste0("imputed.", type, ".data")))
  last.imputed.data.set <- mice::complete(imputed.data, m)

  assign(paste0('S.Rec.', type), survival::Surv(exp(last.imputed.data.set$logRec.time), last.imputed.data.set$Recurrence=="recurred"))
  assign(paste0('S.MSM.', type), survival::Surv(exp(last.imputed.data.set$logFU.time), last.imputed.data.set$MSM=="Death by melanoma"))

  assign(paste0("imputed.", type, ".data"), imputed.data)
  assign(paste0("last.imputed.", type, ".data.set"), last.imputed.data.set)

  save(list=c(paste0("dd.", type), "m", 
              paste0(type, ".data.to.be.imputed"), 
              paste0("imputed.", type, ".data"), 
              paste0("last.imputed.", type, ".data.set"), 
              paste0("S.Rec.", type), 
              paste0("S.MSM.", type)),
       file=paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/', type, '.data.Rdata'), compress=TRUE)
}
prepare.data(type='training')
prepare.data(type='validation')
```

### Kaplan-Meier plots
```{r, eval=TRUE}
load(paste0('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData'))
survminer::ggsurvplot(survival::survfit(S.Rec.training ~ Center, data=last.imputed.training.data.set), conf.int=TRUE, risk.table=TRUE)
survminer::ggsurvplot(survival::survfit(S.MSM.training ~ Center, data=last.imputed.training.data.set), conf.int=TRUE, risk.table=TRUE)
load(paste0('Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData'))
survminer::ggsurvplot(survival::survfit(S.Rec.validation ~ Center, data=last.imputed.validation.data.set), conf.int=TRUE, risk.table=TRUE)
survminer::ggsurvplot(survival::survfit(S.MSM.validation ~ Center, data=last.imputed.validation.data.set), conf.int=TRUE, risk.table=TRUE)
```

# Median follow-up time
```{r, eval=TRUE}
median.FU.time <- function(data=NA){
  median.FU.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM=="Death by melanoma")~1, data=data[data$Center!="Australia",], reverse=TRUE))
  tab.FU.time <- c("Total Development", paste(sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "quantile"]),
                   paste0("[", sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "lower"]), "; ",
                          sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "upper"]), "]")))
  median.FU.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM=="Death by melanoma")~Center, data=data, reverse=TRUE))
  for (center in c("Amsterdam", "Berlin", "Rotterdam", "Warsaw", "Australia")){
    med.FU.table <- eval(parse(text=paste0("median.FU.time$`Center=", center, "`")))
    med.FU <- med.FU.table[3,"quantile"]
    lower.FU <- med.FU.table[3,"lower"]
    upper.FU <- med.FU.table[3,"upper"]
    tab.FU.time <- rbind(tab.FU.time, c(center, paste(sprintf("%.2f", med.FU), paste0("[", sprintf("%.2f", lower.FU), "; ", sprintf("%.2f", upper.FU), "]"))))
  }
  write.table(as.data.frame(tab.FU.time),
              file=paste0('Z:/Project Melanoom/PaperMelanoma/Results/med.FU.table.txt'),
              row.names=FALSE, sep=",")
  return(tab.FU.time)
}
median.FU.time(data=rbind(training.data.to.be.imputed[, c("logFU.time", "MSM", "Center")], 
                          validation.data.to.be.imputed[, c("logFU.time", "MSM", "Center")]))
```

# Recurrence at 5 years
```{r, eval=TRUE}
Rec.MSM.table <- function(horizon=5, type=NULL, data.to.be.imputed=NULL){
  # all patients, Recurrence
  S.Rec <- eval(parse(text=paste0('S.Rec.', type)))
  SF.Rec.all<-survfit(S.Rec~1, data=data.to.be.imputed)
  Rec.all <- 100*(1-SF.Rec.all$surv[SF.Rec.all$time==max(SF.Rec.all$time[SF.Rec.all$time<=horizon])])

  # only negative patients, Recurrence
  SF.Rec.neg<-survfit(S.Rec[data.to.be.imputed$SNstatus=="Negative"]~1, data=data.to.be.imputed)
  Rec.neg <- 100*(1-SF.Rec.neg$surv[SF.Rec.neg$time==max(SF.Rec.neg$time[SF.Rec.neg$time<=horizon])])

  # only positive patients, Recurrence
  SF.Rec.pos<-survfit(S.Rec[data.to.be.imputed$SNstatus=="Positive"]~1, data=data.to.be.imputed)
  Rec.pos <- 100*(1-SF.Rec.pos$surv[SF.Rec.pos$time==max(SF.Rec.pos$time[SF.Rec.pos$time<=horizon])])

  # all patients, MSM
  S.MSM <- eval(parse(text=paste0('S.MSM.', type)))
  SF.MSM.all<-survfit(S.MSM~1, data=data.to.be.imputed)
  MSM.all <- 100*(1-SF.MSM.all$surv[SF.MSM.all$time==max(SF.MSM.all$time[SF.MSM.all$time<=horizon])])

  # negative patients, MSM
  SF.MSM.neg<-survfit(S.MSM[data.to.be.imputed$SNstatus=="Negative"]~1, data=data.to.be.imputed)
  MSM.neg <- 100*(1-SF.MSM.neg$surv[SF.MSM.neg$time==max(SF.MSM.neg$time[SF.MSM.neg$time<=horizon])])

  # positive patients, MSM
  SF.MSM.pos<-survfit(S.MSM[data.to.be.imputed$SNstatus=="Positive"]~1, data=data.to.be.imputed)
  MSM.pos <- 100*(1-SF.MSM.pos$surv[SF.MSM.pos$time==max(SF.MSM.pos$time[SF.MSM.pos$time<=horizon])])

  # merge in one table
  Rec.MSM.table <- data.frame(n=c(nrow(data.to.be.imputed),
                                  sum(data.to.be.imputed$SNstatus=="Negative"),
                                  sum(data.to.be.imputed$SNstatus=="Positive")),
                              perc=c(100,
                                     sprintf("%.1f", sum(data.to.be.imputed$SNstatus=="Negative")/nrow(data.to.be.imputed)*100),
                                     sprintf("%.1f", sum(data.to.be.imputed$SNstatus=="Positive")/nrow(data.to.be.imputed)*100)),
                              Rec=sprintf("%.1f", c(Rec.all, Rec.neg, Rec.pos)),
                              MSM=sprintf("%.1f", c(MSM.all, MSM.neg, MSM.pos)))
  row.names(Rec.MSM.table) <- c("All", "Negative", "Positive")
  utils::write.table(as.data.frame(Rec.MSM.table),file=paste0('Z:/Project Melanoom/PaperMelanoma/Results/Rec.MSM.table.', type, '.txt'),
                     row.names=TRUE, sep=",")
  return(Rec.MSM.table)
}

horizon <- 5
Rec.MSM.table(horizon=horizon, type='training', data.to.be.imputed=training.data.to.be.imputed)
Rec.MSM.table(horizon=horizon, type='validation', data.to.be.imputed=validation.data.to.be.imputed)
```

# Restrict survival at 5 years
```{r, eval=TRUE}
for (type in c('training',  'validation')){
  load(paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/', type, '.data.Rdata'))
  for (outcome in c('Rec', 'MSM')){
    S <- eval(parse(text=paste0('S.', outcome, '.', type)))
    S.5 <- S
    S.5[S[,1]>5, 1] <- 5 # set time to 5
    S.5[S[,1]>5, 2] <- 0 # set censoring

    assign(paste0('S.', outcome, '.5.', type), S.5)
  }
  save(list=c(paste0("dd.", type), "m", "horizon", 
              paste0(type, ".data.to.be.imputed"),
              paste0("imputed.", type, ".data"), 
              paste0("last.imputed.", type, ".data.set"),
              paste0("S.Rec.", type), 
              paste0("S.MSM.", type),
              paste0("S.Rec.5.", type), 
              paste0("S.MSM.5.", type)),
     file=paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/', type, '.data.Rdata'), compress=TRUE)
}
```

### 2022-11-29 Code/4_AJCC.Rmd
Validation data added

--- 2022-06-20 Code/PaperMelanoma.Rmd
@@
 ---
-title: "PaperMelanoma"
+title: "Calculate AJCC"
 author: "Carolien C.H.M. Maas"
-date: "3-6-2022"
+date: "14-11-2022"
 output: html_document
@@
 
-## Development and validation of a clinial prediction model to predict recurrence and disease-specific mortality in patients with melanoma after sentinel lymph node biopsy
+## Calculate AJCC
 
-This file includes the code and results for the above mentioned paper.
-
-# Data preparation
 ### Load packages, functions, and data 
 ```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
+# erase memory
+rm(list = ls(all.names = TRUE))
+
 # for reproducibility
@@
 # load libraries
-library(foreign)
-library(mice)
-library(survminer)
-library(prodlim)
-library(rms)
-library(Hmisc)
-library(png)
-# import function to create table of missing values
-source("Z:/Project Melanoom/PaperMelanoma/Code/table.missing.values.R") 
-# import function to plot calibration
-source("Z:/Project Melanoom/PaperMelanoma/Code/val_surv_mi_v3.R")
-# helper functions for calibration plots
-km <- function(sel, S.km, horizon){
-  S.km.sel <- S.km[sel,]
-  sf <- survfit(S.km.sel~1)
-  year1 <- max(sf$time[sf$time<=horizon])
-  1-sf$surv[sf$time==year1]
-}
-km.lower <- function(sel, S.km, horizon){
-  S.km.sel <- S.km[sel,]
-  sf <- survfit(S.km.sel~1)
-  year1 <- max(sf$time[sf$time<=horizon])
-  1-sf$lower[sf$time==year1]
-}
-km.upper <- function(sel, S.km, horizon){
-  S.km.sel <- S.km[sel,]
-  sf <- survfit(S.km.sel~1)
-  year1 <- max(sf$time[sf$time<=horizon])
-  1-sf$upper[sf$time==year1]
-}
-fun.event <- function(lp, h0)
-{
-  h <- h0*exp(lp)
-  p <- 1-exp(-h)
-  return(p)
-}
-
-# load data file
-original.data <- foreign::read.spss(file="Z:/Project Melanoom/PaperMelanoma/Data/Dataset Predictiemodel Stassen.sav",
-                                    to.data.frame=TRUE, origin="01-01-1970")
-
-# Select necessary variables
-selected.vars <- c("EXCLUSIE", "DatabaseNR", "Center", "Sex", "Age", "DOB",
-                   "Ulceration", "Loc_CAT",            # OLD data set: Loc_CAT2
-                   "Histology", "Breslow", "Tot_nr_SNs",
-                   "nr_SNs_Field_1", "nr_SNs_Field_2", "nr_SNs_Field_3",
-                   "Mitosis",                          # needed for AJCC.7.i
-                   "SNdate", "Recurrence", "Date_1st_Rec", "Last_FU",
-                   "dead", "Status",
-                   "SNstatus",                         # distinguish between positive and negative SN patients
-                   "Tot_SNs_pos", "Rdamcrit", "Dewar") # additional for positive SN patients
+# remotes::install_github("CHMMaas/PredictionTools")
+library(PredictionTools)  # to plot calibration
 ```
-
-### Inclusion criteria
-Exclude when Exclusion = 2: based on duplicates, urogenital melanoma, etc. Note:
-Leave the patients included that have EXCLUSIE = 1 and Dbstatus = 8 (Dbstatus=0: 
-negative, Dbstatus=1: positive), even though diagnosis and operation date are 
-very close.
-```{r, eval=TRUE}
-sel.data <- original.data[original.data$EXCLUSIE!="Exclusion", selected.vars]
-cat("Number of patients excluded:", nrow(original.data)-nrow(sel.data), "\n")
-```
-Omit patient A0523 - no last follow-up date
-```{r, eval=TRUE}
-sel.data[which(is.na(sel.data$Last_FU)), "DatabaseNR"]
-sel.data <- sel.data[sel.data$DatabaseNR!="A0523               ",]
-```
-
-### Recode
-#### Use age at SN instead of age at diagnosis
-```{r, eval=TRUE}
-sel.data$Age.SN <- floor((sel.data$SNdate-sel.data$DOB)/(24*60*60)/365.25)
-```
-
-#### Location
-Merge location other with trunk
-```{r, eval=TRUE}
-sel.data$Loc_CAT[sel.data$Loc_CAT=="other"] <- "trunk"
-```
-
-Merge LMM other with other
-```{r, eval=TRUE}
-sel.data$Histology[sel.data$Histology=="LMM"] <- "other"
-```
-
-#### Define multiple fields
-```{r, eval=TRUE}
-sel.data$multiple.fields <- ifelse(!is.na(sel.data$nr_SNs_Field_2)&sel.data$nr_SNs_Field_2>0,1,0)
-```
-
-#### Number of positive and negative nodes
-Check: there are no negative patients that have positive nodes
-```{r, eval=TRUE}
-sum(!is.na(sel.data$Tot_SNs_pos)&sel.data$Tot_SNs_pos>0&sel.data$SNstatus=="Negative")
-```
-
-Create number of negative nodes removed
-```{r, eval=TRUE}
-sel.data$Tot_SNs_neg <- sel.data$Tot_nr_SNs - sel.data$Tot_SNs_pos
-```
-
-467 positive SN patients have negative nodes
-```{r, eval=TRUE}
-sum(!is.na(sel.data$Tot_SNs_neg)&sel.data$Tot_SNs_neg>0&sel.data$SNstatus=="Positive")
-```
-
-#### SN tumour burden
-11 patients with positive SN have SN tumour burden of exactly 1
-```{r, eval=TRUE}
-sum(sel.data$Rdamcrit==1&sel.data$SNstatus=="Positive"&!is.na(sel.data$Rdamcrit))
-```
-
-Set Rdamcrit of negative SN patients equal to 0
-```{r, eval=TRUE}
-sel.data$Rdamcrit[sel.data$SNstatus=="Negative"] <- 1 # since log(1) = 0
-```
-
-#### Location of metastasis
-Set Dewar for negative patients to subcap
-```{r, eval=TRUE}
-sum(sel.data$SNstatus=="Positive"&sel.data$Dewar=="subcap"&!is.na(sel.data$Dewar))
-sel.data$Dewar[sel.data$SNstatus=="Negative"] <- "subcap"
-```
-
-Merge single cells with subcap 
-```{r, eval=TRUE}
-sel.data$Dewar[sel.data$Dewar=="single cells"] <- "subcap"
-```
-
-#### Recurrence 
-Define time since diagnosis in years (follow-up time)
-```{r, eval=TRUE}
-sel.data$FU.time <- (sel.data$Last_FU-sel.data$SNdate)/(24*60*60)/365.25
-```
-
-Define recurrence time in years
-```{r, eval=TRUE}
-sel.data$Rec.time<- (sel.data$Date_1st_Rec-sel.data$SNdate)/(24*60*60)/365.25
-```
-
-Omit patients with Rec date before SN date (TODO: or impute instead?)
-```{r, eval=TRUE}
-old.n <- nrow(sel.data)
-sel.data[sel.data$Rec.time<0&!is.na(sel.data$Rec.time), "DatabaseNR"]
-sel.data <- sel.data[!(sel.data$Rec.time<0&!is.na(sel.data$Rec.time)),]
-cat("Number of patients excluded:", old.n-nrow(sel.data), "\n")
-```
-
-Set recurrence time equal to follow-up time for those without recurrence
-```{r, eval=TRUE}
-sel.data$Rec.time[sel.data$Recurrence=="no recurrence"|is.na(sel.data$SNdate)]<-sel.data$FU.time[sel.data$Recurrence=="no recurrence"|is.na(sel.data$SNdate)]
-```
-
-#### Melanoma specific mortality (MSM)
-```{r, eval=TRUE}
-sel.data$MSM<-1*(sel.data$Status=="DOD")
-```
-
-#### Mitosis
-Set Mitosis to Unknown if missing
-```{r, eval=TRUE}
-sel.data$Mitosis[is.na(sel.data$Mitosis)] <- "Unknown" # mitosis unknown for 80% of patients
-```
-
-#### Replace "Unknown" by NA
-```{r, eval=TRUE}
-sel.data$Ulceration[sel.data$Ulceration=="unknown"]<-NA
-sel.data$Loc_CAT[sel.data$Loc_CAT=="unknown"]<-NA
-sel.data$Histology[sel.data$Histology=="unknown"]<-NA
-sel.data$Dewar[sel.data$Dewar=="unknown"] <- NA
-sel.data$Mitosis[sel.data$Mitosis=="Unknown"] <- NA
-```
-
-#### Delete unused levels
-```{r, eval=TRUE}
-sel.data <- droplevels(sel.data)
-save(sel.data,file='Z:/Project Melanoom/PaperMelanoma/Data/sel.data.Rdata', compress=TRUE)
-```
-
-# Multiple imputation
-```{r, eval=TRUE}
-variables.to.be.imputed <- c("Center", "SNstatus", "Sex", "Age.SN", "Ulceration",
-                             "Loc_CAT", "Histology", "Breslow", "multiple.fields",
-                             "Tot_SNs_neg", "Tot_SNs_pos", "Rdamcrit", "Dewar",
-                             "Mitosis",
-                             "Recurrence", "dead", "MSM")
-data.to.be.imputed <- sel.data[,variables.to.be.imputed]
-```
-
-Save distributions of all variables, necessary for Harrell summaries
-```{r, eval=TRUE}
-dd<-rms::datadist(data.to.be.imputed)
-options(datadist='dd')
-options(digits=8)
-```
-
-Relation between outcome and covariates
-```{r, eval=TRUE}
-data.to.be.imputed$logRec.time <- log(sel.data$Rec.time+1/365.25)
-data.to.be.imputed$logFU.time <- log(sel.data$FU.time+1/365.25)
-missings.table(data.to.be.imputed)
-save(data.to.be.imputed,file='Z:/Project Melanoom/PaperMelanoma/Data/descr.data.Rdata', compress=TRUE)
-```
-
-Number of imputations
-```{r, eval=TRUE}
-m <- 5
-```
-
-Perform multiple imputation
-```{r, eval=FALSE}
-imputed.data <- mice::mice(data.to.be.imputed, m=m, maxit=5, print=FALSE, seed=500)
-print(imputed.data$loggedEvents)
-save(imputed.data,file='Z:/Project Melanoom/PaperMelanoma/Data/imputed.data.RData', compress=TRUE)
-```
-
-Load imputed data set
-```{r, eval=TRUE}
-load('Z:/Project Melanoom/PaperMelanoma/Data/imputed.data.RData')
-last.imputed.data.set <- mice::complete(imputed.data, m)
-summary(last.imputed.data.set)
-missings.table(last.imputed.data.set)
-```
-
-# Set as survival data
-```{r, eval=TRUE}
-S.Rec <- survival::Surv(exp(last.imputed.data.set$logRec.time), sel.data$Recurrence=="recurred")
-S.MSM <- survival::Surv(exp(last.imputed.data.set$logFU.time), sel.data$MSM)
-```
-
-### Kaplan-Meier plots
-```{r, eval=TRUE}
-center.Rec.KM <- survival::survfit(S.Rec ~ Center, data=sel.data)
-survminer::ggsurvplot(center.Rec.KM, conf.int=TRUE, risk.table=TRUE)
-center.MSM.KM <- survival::survfit(S.MSM ~ Center, data=sel.data)
-survminer::ggsurvplot(center.MSM.KM, conf.int=TRUE, risk.table=TRUE)
-```
-
-# Median follow-up time
-```{r, eval=TRUE}
-median.FU.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM)~1, data=last.imputed.data.set, reverse=TRUE))
-tab.FU.time <- c("total", paste(sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "quantile"]),
-                 paste0("[", sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "lower"]), "; ",
-                        sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "upper"]), "]")))
-median.FU.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM)~Center, data=last.imputed.data.set, reverse=TRUE))
-for (center in sort(as.character(unique(sel.data$Center)))){
-  med.FU.table <- eval(parse(text=paste0("median.FU.time$`Center=", center, "`")))
-  med.FU <- med.FU.table[3,"quantile"]
-  lower.FU <- med.FU.table[3,"lower"]
-  upper.FU <- med.FU.table[3,"upper"]
-  tab.FU.time <- rbind(tab.FU.time, c(center, paste(sprintf("%.2f", med.FU), paste0("[", sprintf("%.2f", lower.FU), "; ", sprintf("%.2f", upper.FU), "]"))))
-}
-write.table(as.data.frame(tab.FU.time),file='Z:/Project Melanoom/PaperMelanoma/Results/med.FU.table.txt',
-            row.names=FALSE, sep=",")
-tab.FU.time
-```
-
-# Recurrence at 5 years
-```{r, eval=TRUE}
-horizon <- 5
-SF.Rec.all<-survfit(S.Rec~1)
-Rec.all <- 100*(1-SF.Rec.all$surv[SF.Rec.all$time==max(SF.Rec.all$time[SF.Rec.all$time<=horizon])])
-
-SF.Rec.neg<-survfit(S.Rec[sel.data$SNstatus=="Negative"]~1)
-Rec.neg <- 100*(1-SF.Rec.neg$surv[SF.Rec.neg$time==max(SF.Rec.neg$time[SF.Rec.neg$time<=horizon])])
-
-SF.Rec.pos<-survfit(S.Rec[sel.data$SNstatus=="Positive"]~1)
-Rec.pos <- 100*(1-SF.Rec.pos$surv[SF.Rec.pos$time==max(SF.Rec.pos$time[SF.Rec.pos$time<=horizon])])
-
-SF.MSM.all<-survfit(S.MSM~1)
-MSM.all <- 100*(1-SF.MSM.all$surv[SF.MSM.all$time==max(SF.MSM.all$time[SF.MSM.all$time<=horizon])])
-
-SF.MSM.neg<-survfit(S.MSM[sel.data$SNstatus=="Negative"]~1)
-MSM.neg <- 100*(1-SF.MSM.neg$surv[SF.MSM.neg$time==max(SF.MSM.neg$time[SF.MSM.neg$time<=horizon])])
-
-SF.MSM.pos<-survfit(S.MSM[sel.data$SNstatus=="Positive"]~1)
-MSM.pos <- 100*(1-SF.MSM.pos$surv[SF.MSM.pos$time==max(SF.MSM.pos$time[SF.MSM.pos$time<=horizon])])
-
-Rec.MSM.table <- data.frame(n=c(nrow(sel.data), 
-                                sum(sel.data$SNstatus=="Negative"), 
-                                sum(sel.data$SNstatus=="Positive")),
-                            perc=c(100, 
-                                   sprintf("%.1f", sum(sel.data$SNstatus=="Negative")/nrow(sel.data)*100),
-                                   sprintf("%.1f", sum(sel.data$SNstatus=="Positive")/nrow(sel.data)*100)),
-                            Rec=sprintf("%.1f", c(Rec.all, Rec.neg, Rec.pos)), 
-                            MSM=sprintf("%.1f", c(MSM.all, MSM.neg, MSM.pos)))
-row.names(Rec.MSM.table) <- c("All", "Negative", "Positive")
-utils::write.table(as.data.frame(Rec.MSM.table),file='Z:/Project Melanoom/PaperMelanoma/Results/Rec.MSM.table.txt',
-                   row.names=TRUE, sep=",")
-Rec.MSM.table
-```
-
-# Restrict survival at 5 years
-```{r, eval=TRUE}
-S.Rec.5 <- S.Rec
-S.Rec.5[S.Rec[,1]>5, 1] <- 5 # set recurrence time to 5
-S.Rec.5[S.Rec[,1]>5, 2] <- 0 # set censoring
-
-S.MSM.5 <- S.MSM
-S.MSM.5[S.MSM[,1]>5, 1] <- 5 # set follow-up time to 5
-S.MSM.5[S.MSM[,1]>5, 2] <- 0 # set censoring
-```
-
 # Obtain C-index of AJCC7 and AJCC8
 ```{r, eval=TRUE}
+calculate.AJCC.7.8 <- function(imputed.data=NULL, type=NULL, S.Rec.5=NULL, S.MSM.5=NULL){
+  ####
+  #### AJCC 7
+  ####
   TNM.7 <- c()
@@
                            "; ", sprintf('%.2f', cindex.mi.AJCC.7.neg$est+qnorm(.975)*cindex.mi.AJCC.7.neg$se), "]"))),
-            file='Z:/Project Melanoom/PaperMelanoma/Results/AJCC.7.Cindex.txt', 
+              file=paste0('Z:/Project Melanoom/PaperMelanoma/Results/AJCC.7.Cindex.', type, '.txt'), 
               row.names=FALSE, col.names=FALSE, sep=",")
 
-#####
-##### COMPARE MODEL WITH AJCC.8 CLASIFCATION
-#####
+  ####
+  #### AJCC8
+  ####
   TNM.8 <- c()
@@
                                  "; ", sprintf('%.2f', cindex.mi.AJCC.8.neg$est+qnorm(.975)*cindex.mi.AJCC.8.neg$se), "]"))),
-            file='Z:/Project Melanoom/PaperMelanoma/Results/AJCC.8.Cindex.txt', 
+              file=paste0('Z:/Project Melanoom/PaperMelanoma/Results/AJCC.8.Cindex.', type, '.txt'), 
               row.names=FALSE, col.names=FALSE, sep=",")
@@
   }
-table.AJCC.7.8 <- cbind(as.numeric(names(table(AJCC.8[,1]))),
+
+  # print(sort(unique(AJCC.7[,1])))
+  # print(sort(unique(AJCC.8[,1])))
+
+  if (type=="training"){
+    table.AJCC.7.8 <- cbind(c('1.1', '1.2', '2.1', '2.2', '2.3', '3.1', '3.2', '3.3', '3.4'),
                           c(round(rowMeans(table.AJCC.7), 0), 0),
-                        round(rowMeans(table.AJCC.8), 0))
+                          round(rowMeans(table.AJCC.8)))
+  } else if (type=="validation"){
+    table.AJCC.7.8 <- cbind(c('1.1', '1.2', '2.1', '2.2', '2.3', '3.1', '3.2', '3.3', '3.4'),
+                          c(round(rowMeans(table.AJCC.7), 0), 0),
+                          c(0, round(rowMeans(table.AJCC.8), 0)))
+  }
   utils::write.table(table.AJCC.7.8,
-            file='Z:/Project Melanoom/PaperMelanoma/Results/table.AJCC.7.and.8.txt', 
+              file=paste0('Z:/Project Melanoom/PaperMelanoma/Results/table.AJCC.7.and.8.', type, '.txt'),
               row.names=FALSE, col.names=FALSE, sep=",")
-table.AJCC.7.8
+}
+load('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData')
+calculate.AJCC.7.8(imputed.data=imputed.training.data, S.Rec.5=S.Rec.5.training, S.MSM.5=S.MSM.5.training, type='training')
+load('Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData')
+calculate.AJCC.7.8(imputed.data=imputed.validation.data, S.Rec.5=S.Rec.5.validation, S.MSM.5=S.MSM.5.validation, type='validation')
 ```
-
-# Test non-lineairities
-### Test if Age should be modelled non-linearly
-```{r, eval=TRUE}
-form.linear.full.Rec <- S.Rec.5 ~
-  SNstatus +
-  Sex +
-  Age.SN +
-  Ulceration +
-  Loc_CAT +
-  Histology +
-  Breslow +
-  multiple.fields +
-  Tot_SNs_neg +
-  Tot_SNs_pos +
-  Rdamcrit +
-  Dewar +
-  SNstatus:Sex +
-  SNstatus:Age.SN +
-  SNstatus:Ulceration +
-  SNstatus:Loc_CAT +
-  SNstatus:Histology +
-  SNstatus:Breslow +
-  SNstatus:multiple.fields
-form.nonlinear.age.full.Rec <- S.Rec.5 ~
-  SNstatus +
-  Sex +
-  log(Age.SN) +
-  Ulceration +
-  Loc_CAT +
-  Histology +
-  Breslow +
-  multiple.fields +
-  Tot_SNs_neg +
-  Tot_SNs_pos +
-  Rdamcrit +
-  Dewar +
-  SNstatus:Sex +
-  SNstatus:log(Age.SN) +
-  SNstatus:Ulceration +
-  SNstatus:Loc_CAT +
-  SNstatus:Histology +
-  SNstatus:Breslow +
-  SNstatus:multiple.fields
-linear.full.Rec <- Hmisc::fit.mult.impute(form.linear.full.Rec, cph, xtrans=imputed.data,
-                                 data=data.to.be.imputed,n.impute=m, pr=FALSE,
-                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
-nonlinear.age.full.Rec <- Hmisc::fit.mult.impute(form.nonlinear.age.full.Rec, cph, xtrans=imputed.data,
-                                 data=data.to.be.imputed,n.impute=m, pr=FALSE,
-                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
-# distribution of Age
-hist(last.imputed.data.set$Age.SN)
-# linear model
-linear.full.Rec$loglik[2]
-# using log(Age) gives slightly better log likelihood
-nonlinear.age.full.Rec$loglik[2]
-```
-
-### Test if Breslow should be modelled non-linearly
-```{r, eval=TRUE}
-form.nonlinear.Breslow.full.Rec <- S.Rec.5 ~
-  SNstatus +
-  Sex +
-  Age.SN +
-  Ulceration +
-  Loc_CAT +
-  Histology +
-  log(Breslow) +
-  multiple.fields +
-  Tot_SNs_neg +
-  Tot_SNs_pos +
-  Rdamcrit +
-  Dewar +
-  SNstatus:Sex +
-  SNstatus:Age.SN +
-  SNstatus:Ulceration +
-  SNstatus:Loc_CAT +
-  SNstatus:Histology +
-  SNstatus:log(Breslow) +
-  SNstatus:multiple.fields
-nonlinear.Breslow.full.Rec <- Hmisc::fit.mult.impute(form.nonlinear.Breslow.full.Rec, cph, xtrans=imputed.data,
-                                 data=data.to.be.imputed,n.impute=m, pr=FALSE,
-                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
-# distribution of Breslow
-hist(last.imputed.data.set$Breslow)
-# linear model
-linear.full.Rec$loglik[2]
-# using log(Breslow) gives better log likelihood
-nonlinear.Breslow.full.Rec$loglik[2]
-```
-
-### Test if Rdamcrit should be modelled non-linearly
-```{r, eval=TRUE}
-form.nonlinear.Rdamcrit.full.Rec <- S.Rec.5 ~
-  SNstatus +
-  Sex +
-  Age.SN +
-  Ulceration +
-  Loc_CAT +
-  Histology +
-  Breslow +
-  multiple.fields +
-  Tot_SNs_neg +
-  Tot_SNs_pos +
-  log(Rdamcrit) +
-  Dewar +
-  SNstatus:Sex +
-  SNstatus:Age.SN +
-  SNstatus:Ulceration +
-  SNstatus:Loc_CAT +
-  SNstatus:Histology +
-  SNstatus:Breslow +
-  SNstatus:multiple.fields
-nonlinear.Rdamcrit.full.Rec <- Hmisc::fit.mult.impute(form.nonlinear.Rdamcrit.full.Rec, cph, xtrans=imputed.data,
-                                 data=data.to.be.imputed,n.impute=m, pr=FALSE,
-                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
-# distribution of Rdamcrit
-hist(last.imputed.data.set$Rdamcrit)
-# linear model
-linear.full.Rec$loglik[2]
-# using log(Rdamcrit) gives better log likelihood
-nonlinear.Rdamcrit.full.Rec$loglik[2]
-```
-
-# Full model for recurrence
-```{r, eval=TRUE}
-form.full.Rec.5 <- S.Rec.5 ~
-  SNstatus +
-  Sex +
-  Age.SN +
-  Ulceration +
-  Loc_CAT +
-  Histology +
-  log(Breslow) +
-  multiple.fields +
-  Tot_SNs_neg +
-  Tot_SNs_pos +
-  log(Rdamcrit) +
-  Dewar +
-  SNstatus:Sex +
-  SNstatus:Age.SN +
-  SNstatus:Ulceration +
-  SNstatus:Loc_CAT +
-  SNstatus:Histology +
-  SNstatus:log(Breslow) +
-  SNstatus:multiple.fields
-f.mi.full.Rec.5 <- Hmisc::fit.mult.impute(form.full.Rec.5, cph, xtrans=imputed.data,
-                                 data=data.to.be.imputed,n.impute=m, pr=FALSE,
-                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
-# stats::anova(f.mi.full.Rec.5)
-summary(f.mi.full.Rec.5, Tot_SNs_pos=c(0, 1))
-```
-
-# Final model for recurrence selected by backward selection with p = 0.01
-```{r, eval=TRUE}
-c.Breslow <- mean(log(last.imputed.data.set$Breslow))
-form.BS.Rec.5 <- S.Rec.5 ~
-  SNstatus +
-  Age.SN +
-  Ulceration +
-  Loc_CAT +
-  I(log(Breslow)-c.Breslow) +
-  log(Rdamcrit) +
-  SNstatus:I(log(Breslow)-c.Breslow)
-f.mi.BS.Rec.5 <- Hmisc::fit.mult.impute(form.BS.Rec.5,cph,xtrans=imputed.data,data=data.to.be.imputed,n.impute=m,
-                           pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
-summary(f.mi.BS.Rec.5, Breslow=exp(c.Breslow))
-coef(f.mi.BS.Rec.5)
-# round(stats::anova(f.mi.BS.Rec.5)[order(stats::anova(f.mi.BS.Rec.5)[, 'P']),], 3)
-stats::anova(f.mi.BS.Rec.5)
-```
-
-# Refitted full model for melanoma specific mortality
-```{r, eval=TRUE}
-form.full.MSM.5.refit <- S.MSM.5 ~
-  SNstatus +
-  Sex +
-  Age.SN +
-  Ulceration +
-  Loc_CAT +
-  Histology +
-  log(Breslow) +
-  multiple.fields +
-  Tot_SNs_neg +
-  Tot_SNs_pos +
-  log(Rdamcrit) +
-  Dewar +
-  SNstatus:Sex +
-  SNstatus:Age.SN +
-  SNstatus:Ulceration +
-  SNstatus:Loc_CAT +
-  SNstatus:Histology +
-  SNstatus:log(Breslow) +
-  SNstatus:multiple.fields
-f.mi.full.MSM.5.refit <- Hmisc::fit.mult.impute(form.full.MSM.5.refit,cph,xtrans=imputed.data,data=data.to.be.imputed,n.impute=m,
-                                 pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
-f.mi.full.MSM.5.refit
-summary(f.mi.full.MSM.5.refit, Tot_SNs_pos=c(0, 1))
-stats::anova(f.mi.full.MSM.5.refit)
-```
-
-# Refitted final model for melanoma specific mortality
-```{r, eval=TRUE}
-form.BS.MSM.5.refit <- S.MSM.5 ~
-  SNstatus +
-  Age.SN +
-  Ulceration +
-  Loc_CAT +
-  I(log(Breslow)-c.Breslow) +
-  log(Rdamcrit) +
-  SNstatus:I(log(Breslow)-c.Breslow)
-f.mi.BS.MSM.5.refit <- Hmisc::fit.mult.impute(form.BS.MSM.5.refit,cph,xtrans=imputed.data,data=data.to.be.imputed,n.impute=m,
-                                  pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
-# round(anova(f.mi.BS.MSM.5.refit)[order(anova(f.mi.BS.MSM.5.refit)[, 'P']),], 3)
-f.mi.BS.MSM.5.refit
-summary(f.mi.BS.MSM.5.refit, Breslow=exp(c.Breslow))
-stats::anova(f.mi.BS.MSM.5.refit)
-```
-
-# Calibrated final model for melanoma specific mortality
-Fit disease specific mortality to the linear predictor of recurrence
-```{r, eval=TRUE}
-lp.BS<-f.mi.BS.Rec.5$linear
-f.mi.BS.MSM.5<-rms::cph(S.MSM.5~lp.BS,y=TRUE,x=TRUE)
-MSM.cal.fact <- stats::coef(f.mi.BS.MSM.5)
-MSM.cal.fact
-stats::confint(f.mi.BS.MSM.5)
-utils::write.table(paste0(sprintf("%.2f", stats::coef(f.mi.BS.MSM.5)), " [",
-            sprintf("%.2f", as.numeric(stats::confint(f.mi.BS.MSM.5))[1]), "; ",
-            sprintf("%.2f", as.numeric(stats::confint(f.mi.BS.MSM.5))[2]), "]"),
-            file='Z:/Project Melanoom/PaperMelanoma/Results/calibration.MSS.txt', 
-            row.names=FALSE, col.names=FALSE, sep=",")
-```
-
-### Create variables with results
-##### Full model
-```{r, eval=TRUE}
-for (model in c('Rec.5', 'MSM.5.refit')){
-  # load model
-  f.mi.full <- eval(parse(text=paste0('f.mi.full.', model)))
-
-  # Coefficients with CI
-  # Negative results
-  neg.results <- summary(f.mi.full, SNstatus="Negative", Loc_CAT="arm")
-  neg.table <- rbind(rep(NA, 3), # SN status
-                   exp(neg.results[c("Sex - male:female",
-                        "Age.SN",
-                        "Ulceration - Yes:No",
-                        "Loc_CAT - leg:arm",
-                        "Loc_CAT - trunk:arm",
-                        "Loc_CAT - head & neck:arm",
-                        "Histology - NM:SSM",
-                        "Histology - ALM:SSM",
-                        "Histology - other:SSM",
-                        "Breslow",
-                        "multiple.fields",
-                        "Tot_SNs_neg"), c("Effect", "Lower 0.95", "Upper 0.95")]), 
-                   matrix(NA, nrow=6, ncol=3)) # Tot_SNs_pos, Rdamcrit, Combined, Parenchymal, Multifocal, Extensive
-  # add CI
-  neg.table <- cbind(sprintf("%.2f", neg.table[, 1]),
-                     paste0("[", sprintf("%.2f", neg.table[, 2]), "; ", sprintf("%.2f", neg.table[, 3]), "]"))
-
-  # Positive results
-  pos.results <- summary(f.mi.full, SNstatus="Positive", Loc_CAT="arm", Tot_SNs_pos=c(0, 1))
-  pos.table <- rbind(exp(neg.results["SNstatus - Positive:Negative", c("Effect", "Lower 0.95", "Upper 0.95")]),
-                     exp(pos.results[c("Sex - male:female",
-                        "Age.SN",
-                        "Ulceration - Yes:No",
-                        "Loc_CAT - leg:arm",
-                        "Loc_CAT - trunk:arm",
-                        "Loc_CAT - head & neck:arm",
-                        "Histology - NM:SSM",
-                        "Histology - ALM:SSM",
-                        "Histology - other:SSM",
-                        "Breslow",
-                        "multiple.fields",
-                        "Tot_SNs_neg",
-                        "Tot_SNs_pos",
-                        "Rdamcrit",
-                        "Dewar - combined:subcap",
-                        "Dewar - parenchymal:subcap",
-                        "Dewar - multifocal:subcap",
-                        "Dewar - extensive:subcap"), c("Effect", "Lower 0.95", "Upper 0.95")]))
-  # add CI
-  pos.table <- cbind(sprintf("%.2f", pos.table[, 1]), 
-                     paste0("[", sprintf("%.2f", pos.table[, 2]), "; ", sprintf("%.2f", pos.table[, 3]), "]"))
-
-  # combine postive with negative
-  full.table <- cbind(neg.table, pos.table)
-
-  # add reference rows
-  final.table <- rbind(full.table[1:4,],
-                       rep(NA, 4),     # location
-                       c("(ref)", NA, "(ref)", NA), # ref arm
-                       full.table[5:7,],   # other locations
-                       rep(NA, 4),     # histology
-                       c("(ref)", NA, "(ref)", NA), # ref SMM
-                       full.table[8:15,],
-                       rep(NA, 4),     # Dewar
-                       c(rep(NA, 2), "(ref)", NA), # ref subcap
-                       full.table[16:nrow(full.table),]) # other Dewar
-
-  # add Chi-square statistics
-  anova.out <- anova(f.mi.full)[, "Chi-Square"]
-  anova.main <- c(anova.out[c("SNstatus  (Factor+Higher Order Factors)",
-                        "Sex  (Factor+Higher Order Factors)",
-                        "Age.SN  (Factor+Higher Order Factors)",
-                        "Ulceration  (Factor+Higher Order Factors)",
-                        "Loc_CAT  (Factor+Higher Order Factors)")],
-                  rep(NA, 4),
-                  anova.out["Histology  (Factor+Higher Order Factors)"],
-                  rep(NA, 4),
-                  anova.out[c("Breslow  (Factor+Higher Order Factors)",
-                        "multiple.fields  (Factor+Higher Order Factors)",
-                        "Tot_SNs_neg",
-                        "Tot_SNs_pos",
-                        "Rdamcrit",
-                        "Dewar")],
-                  rep(NA, 5))
-  anova.int <- c(anova.out[c("TOTAL INTERACTION",
-                       "SNstatus * Sex  (Factor+Higher Order Factors)",
-                       "SNstatus * Age.SN  (Factor+Higher Order Factors)",
-                       "SNstatus * Ulceration  (Factor+Higher Order Factors)",
-                       "SNstatus * Loc_CAT  (Factor+Higher Order Factors)")],
-                 rep(NA, 4),
-                 anova.out["SNstatus * Histology  (Factor+Higher Order Factors)"],
-                 rep(NA, 4),
-                 anova.out[c("SNstatus * Breslow  (Factor+Higher Order Factors)",
-                       "SNstatus * multiple.fields  (Factor+Higher Order Factors)")],
-                 rep(NA, 9))
-  final.table <- cbind(final.table, sprintf("%.0f", anova.main), sprintf("%.0f", anova.int))
-
-  colnames(final.table) <- c("HR.neg", "CI.neg", "HR.pos", "CI.pos", "Main", "Int")
-  assign(paste0('final.full.table.', model), as.data.frame(final.table))
-}
-```
-
-##### Final model
-```{r, eval=TRUE}
-for (model in c('Rec.5', 'MSM.5.refit')){
-  f.mi.BS <- eval(parse(text=paste0('f.mi.BS.', model)))
-
-  # Coefficients with CI
-  # Negative results
-  neg.results.BS <- summary(f.mi.BS, SNstatus="Negative", Breslow=exp(c.Breslow), Loc_CAT="arm")
-  neg.table.BS <- rbind(rep(NA, 3), # SN status
-                   exp(neg.results.BS[c("Age.SN",
-                        "Ulceration - Yes:No",
-                        "Loc_CAT - leg:arm",
-                        "Loc_CAT - trunk:arm",
-                        "Loc_CAT - head & neck:arm",
-                        "Breslow"), c("Effect", "Lower 0.95", "Upper 0.95")]), 
-                   rep(NA, 3)) # Rdamcrit
-  # add CI
-  neg.table.BS <- cbind(sprintf("%.2f", neg.table.BS[, 1]), 
-                        paste0("[", sprintf("%.2f", neg.table.BS[, 2]), "; ", sprintf("%.2f", neg.table.BS[, 3]), "]"))
-
-  # Positive results
-  pos.results.BS <- summary(f.mi.BS, SNstatus="Positive", Breslow=exp(c.Breslow), Loc_CAT="arm")
-  pos.table.BS <- rbind(exp(neg.results.BS["SNstatus - Positive:Negative", c("Effect", "Lower 0.95", "Upper 0.95")]),
-                     exp(pos.results.BS[c("Age.SN",
-                        "Ulceration - Yes:No",
-                        "Loc_CAT - leg:arm",
-                        "Loc_CAT - trunk:arm",
-                        "Loc_CAT - head & neck:arm",
-                        "Breslow",
-                        "Rdamcrit"), c("Effect", "Lower 0.95", "Upper 0.95")]))
-  # add CI
-  pos.table.BS <- cbind(sprintf("%.2f", pos.table.BS[, 1]), 
-                        paste0("[", sprintf("%.2f", pos.table.BS[, 2]), "; ", sprintf("%.2f", pos.table.BS[, 3]), "]"))
-
-  # combine postive with negative
-  BS.table <- cbind(neg.table.BS, pos.table.BS)
-
-  # add reference rows
-  BS.table <- rbind(BS.table[1:3,], 
-                    rep(NA, 4), 
-                    c("(ref)", NA, "(ref)", NA), 
-                    BS.table[4:nrow(BS.table),])
-
-  # add Chi-square statistics
-  anova.out <- anova(f.mi.BS)[, "Chi-Square"]
-  anova.main <- c(anova.out[c("SNstatus  (Factor+Higher Order Factors)",
-                        "Age.SN",
-                        "Ulceration",
-                        "Loc_CAT")],
-                  rep(NA, 4),
-                  anova.out[c("Breslow  (Factor+Higher Order Factors)",
-                        "Rdamcrit")])
-  anova.int <- c(rep(NA, 8),
-                 anova.out["SNstatus * Breslow  (Factor+Higher Order Factors)"],
-                 NA)
-  BS.table <- cbind(BS.table, sprintf("%.0f", anova.main), sprintf("%.0f", anova.int))
-
-  colnames(BS.table) <- c("HR.neg", "CI.neg", "HR.pos", "CI.pos", "Main", "Int")
-  assign(paste0('final.BS.table.', model), as.data.frame(BS.table))
-}
-```
-
-# Plot calibration of all centers combined and all centers seperate 
-for recurrence and melanoma specific mortality refit 
-```{r, eval=TRUE}
-for (model in list('Rec.5', 'MSM.5.refit')){
-  if (model == 'Rec.5'){
-    # plot predict
-    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/plot.predict.Rec.png"))
-    plot.predict <- plot(Predict(f.mi.full.Rec.5))
-    show(plot.predict)
-    grDevices::dev.off()
-
-    # plot predict Breslow
-    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/plot.predict.Breslow.Rec.png"))
-    plot.Breslow <- plot(Predict(f.mi.full.Rec.5, Breslow=0:20, SNstatus), ~ Breslow, groups="SNstatus",
-                         adj.subtitle=FALSE, lwd=2, labels=FALSE,
-                         par.settings=list(superpose.line=list(lty=c(1, 2), col=c("red", "blue"))),
-                         key=list(cex=.8,corner = c(0.1, .9),lines=list(col=c("red","blue"), lty=c(1,2), lwd=2),text=list(c("Negative","Positive"))))
-    show(plot.Breslow)
-    grDevices::dev.off()
-  }
-
-  # discrimination full and backward selected model
-  for (type in c("full", "BS")){
-    f.mi <- eval(parse(text=paste0('f.mi.', type, '.', model)))
-    if (model=='Rec.5'){
-      S <- S.Rec.5
-    }
-    else if (model=="MSM.refit"){
-      S <- S.MSM.5
-    }
-    p <- matrix(NA, nrow(sel.data), m)
-    for (i in 1:m){
-      # predicted lp for imputation i
-      lp.i <- predict(f.mi,newdata=mice::complete(imputed.data,i), type="lp")
-      # baseline hazard
-      f.basehaz <- survival::basehaz(f.mi,TRUE)
-      h0.i <- f.basehaz$hazard[f.basehaz$time==max(f.basehaz$time[f.basehaz$time<=horizon])]
-      # probabilities
-      p.i <- fun.event(lp=lp.i, h0=h0.i)
-      p[, i] <- p.i
-    }
-
-    # omit the observation with predicted probability = 1
-    omit <- unique(c(which(p[,1]==1), which(p[,2]==1), which(p[,3]==1), which(p[,4]==1), which(p[,5]==1)))
-    if (length(omit)>0){
-      cat('Omit:', omit, '\n')
-      p <- p[-omit,]
-      S <- S[-omit,]
-    }
-
-    out <- val.surv.mi(p,S,g=5,main="All centers",
-                       file_path=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/all.centers.", model, '.', type, ".png"),
-                       save_plots=TRUE,time=NULL,lim=c(0,1),dist=TRUE)
-
-    final.full.table <- eval(parse(text=paste0('final.', type, '.table.', model)))
-    final.full.table["C-index", "HR.neg"] <- sprintf('%.2f', out$cindex)
-    final.full.table["C-index", "CI.neg"] <- paste0("[", sprintf('%.2f', out$cindex.lower),
-                                                   "; ", sprintf('%.2f', out$cindex.upper), "]")
-
-    assign(paste0('final.', type, '.table.', model), final.full.table)
-  }
-}
-
-# MSM recalibration
-for (type in c("full", "BS")){
-  f.mi <- eval(parse(text=paste0('f.mi.', type, '.Rec.5')))
-  cindex.MSM.refit <- rep(0, m)
-  cindex.se.MSM.refit <- rep(0, m)
-  for (i in 1:m){
-    rc <- Hmisc::rcorr.cens(-predict(f.mi,newdata=mice::complete(imputed.data,i), type="lp"), S.MSM.5)
-    cindex.MSM.refit[i]<-rc["C Index"]
-    cindex.se.MSM.refit[i]<-rc["S.D."]/2
-  }
-  cindex.mi.MSM.refit<-Rubin.combine(cindex.MSM.refit,cindex.se.MSM.refit)
-
-  final.full.table <- eval(parse(text=paste0('final.', type, '.table.Rec.5')))
-  final.full.table["C-index MSM", "HR.neg"] <- sprintf('%.2f', cindex.mi.MSM.refit$est)
-  final.full.table["C-index MSM", "CI.neg"] <- paste0("[", sprintf('%.2f', cindex.mi.MSM.refit$est+qnorm(.025)*cindex.mi.MSM.refit$se),
-                                                 "; ", sprintf('%.2f', cindex.mi.MSM.refit$est+qnorm(.975)*cindex.mi.MSM.refit$se), "]")
-
-  assign(paste0('final.', type, '.table.Rec.5'), final.full.table)
-}
-
-# replace NA by blank
-for (df in c("final.full.table.Rec.5", "final.full.table.MSM.5.refit", "final.BS.table.Rec.5", "final.BS.table.MSM.5.refit")){
-  final.df <- eval(parse(text=df))
-  final.df[final.df=="[NA; NA]"] <- NA
-  final.df[final.df=="NA"] <- NA
-  final.df[is.na(final.df)] <- ""
-  assign(df, final.df)
-}
-
-# set rownames
-final.full.table.Rec.5 <- as.data.frame(final.full.table.Rec.5)
-row.names(final.full.table.Rec.5) <- c("Positive SN status", "Male",
-                                    "Age iqr", "Ulceration", "Location",
-                                    "Arm", "Leg", "Trunk", "Head and neck",
-                                    "Histology", "SSM", "NM", "ALM", "Other",
-                                    "Breslow iqr", "Multiple fields", 
-                                    "Total number of negative nodes",
-                                    "Total number of positive nodes",
-                                    "SN tumour burden iqr", 
-                                    "Location metastasis in lymph node",
-                                    "Subcap", "Combined", "Parenchymal",
-                                    "Multifocal", "Extensive",
-                                    rownames(final.full.table.Rec.5)[(nrow(final.full.table.Rec.5)-1):nrow(final.full.table.Rec.5)])
-final.full.table.MSM.5.refit <- as.data.frame(final.full.table.MSM.5.refit)
-row.names(final.full.table.MSM.5.refit) <- c("Positive SN status", "Male",
-                                          "Age", "Ulceration", "Location",
-                                          "Arm", "Leg", "Trunk", "Head and neck",
-                                          "Histology", "SSM", "NM", "ALM", "Other",
-                                          "Breslow", "Multiple fields", 
-                                          "Total number of negative nodes",
-                                          "Total number of positive nodes",
-                                          "SN tumour burden", 
-                                          "Location metastasis in lymph node",
-                                          "Subcap", "Combined", "Parenchymal",
-                                          "Multifocal", "Extensive",
-                                          row.names(final.full.table.MSM.5.refit)[nrow(final.full.table.MSM.5.refit)])
-final.BS.table.Rec.5 <- as.data.frame(final.BS.table.Rec.5)
-row.names(final.BS.table.Rec.5) <- c("Positive SN status", "Age", "Ulceration",
-                                   "Location", "Arm", "Leg", "Trunk", 
-                                   "Head and neck",
-                                   "Breslow", "SN tumour burden", 
-                                   row.names(final.BS.table.Rec.5)[(nrow(final.BS.table.Rec.5)-1):nrow(final.BS.table.Rec.5)])
-final.BS.table.MSM.5.refit <- as.data.frame(final.BS.table.MSM.5.refit)
-row.names(final.BS.table.MSM.5.refit) <- c("Positive SN status", "Age", "Ulceration",
-                                   "Location", "Arm", "Leg", "Trunk", 
-                                   "Head and neck",
-                                   "Breslow", "SN tumour burden", 
-                                   row.names(final.BS.table.MSM.5.refit)[nrow(final.BS.table.MSM.5.refit)])
-
-# save tables
-write.table(final.full.table.Rec.5,
-            file='Z:/Project Melanoom/PaperMelanoma/Results/final.full.table.Rec.txt', 
-            sep=",")
-write.table(final.full.table.MSM.5.refit,
-            file='Z:/Project Melanoom/PaperMelanoma/Results/final.full.table.MSM.refit.txt', 
-            sep=",")
-write.table(final.BS.table.Rec.5,
-            file='Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.Rec.txt', 
-            sep=",")
-write.table(final.BS.table.MSM.5.refit,
-            file='Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.MSM.refit.txt', 
-            sep=",")
-
-# show tables
-final.full.table.Rec.5
-final.full.table.MSM.5.refit
-final.BS.table.Rec.5
-final.BS.table.MSM.5.refit
-```
-
-# Cross-validation recurrence 5-years
-```{r, eval=TRUE}
-Centers <- sort(as.character(unique(sel.data$Center)))
-Centers.anonimized <- c("A", "B", "C", "D")
-cuts <- 5
-for (model in c('Rec.5', 'MSM.5')){
-  if (model == 'Rec.5'){
-    S.model <- S.Rec.5
-  }
-  if (model == 'MSM.5'){
-    S.model <- S.MSM.5
-  }
-
-  for (type in c('full', 'BS')){
-    for (j in 1:4){
-      S.j<-S.model[data.to.be.imputed$Center==Centers[j],]
-      S.notj<-S.model[data.to.be.imputed$Center!=Centers[j],]
-      S.Rec.5.notj<-S.Rec.5[data.to.be.imputed$Center!=Centers[j],]
-
-      form.model<-eval(parse(text=paste0('form.', type, '.Rec.5')))
-      form.notj<-update(form.model, S.Rec.5.notj  ~ . )
-      f.mi.Rec.notj<-Hmisc::fit.mult.impute(form.notj,cph,xtrans=imputed.data,data=data.to.be.imputed,
-                                 n.impute=m,pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE,
-                                 sub=data.to.be.imputed$Center!=Centers[j])
-
-      if (model=='Rec.5'){
-        f.basehaz.notj<-survival::basehaz(f.mi.Rec.notj,TRUE)
-        h0.notj<-f.basehaz.notj$hazard[f.basehaz.notj$time==max(f.basehaz.notj$time[f.basehaz.notj$time<=horizon])]
-      } else if (model=='MSM.5'){
-        lp.notj<-f.mi.Rec.notj$linear
-        f.MSM.notj<-rms::cph(S.notj~lp.notj,y=TRUE,x=TRUE)
-
-        f.basehaz.notj<-basehaz(f.MSM.notj,TRUE)
-        h0.notj<-f.basehaz.notj$hazard[f.basehaz.notj$time==max(f.basehaz.notj$time[f.basehaz.notj$time<=horizon])]
-      }
-
-      for (select.value in c("Positive", "Negative", "All")){
-        # identify those patients that are selected
-        if (select.value == "All"){
-          select.patients <- TRUE
-        } else if (select.value != "All"){
-          select.patients <- data.to.be.imputed[data.to.be.imputed$Center==Centers[j],]$SNstatus==select.value
-        }
-
-        # calibration with metrics
-        S <- S.j[select.patients,]
-        p <- matrix(NA, nrow(S), m)
-        for (i in 1:m){
-          lp.i<-predict(f.mi.Rec.notj,newdata=mice::complete(imputed.data,i)[data.to.be.imputed$Center==Centers[j],], type="lp")
-
-          if (model=='MSM.5'){
-            lp.i<-predict(f.MSM.notj,newdata=lp.i)
-          }
-
-          if (select.value != "All"){
-            lp.i <- lp.i[select.patients]
-          }
-
-          # baseline hazard
-          if (model=='Rec.5'){
-            f.basehaz<-basehaz(f.mi.Rec.notj,TRUE)
-          }
-          else if (model=='MSM.5'){
-            f.basehaz<-basehaz(f.MSM.notj,TRUE)
-          }
-          h0.i <- f.basehaz$hazard[f.basehaz$time==max(f.basehaz$time[f.basehaz$time<=horizon])]
-
-          # probabilities
-          p.i <- fun.event(lp=lp.i, h0=h0.i)
-          p[, i] <- p.i
-        }
-
-        # omit the observation with predicted probability = 1
-        omit <- unique(c(which(p[,1]==1), which(p[,2]==1), which(p[,3]==1), which(p[,4]==1), which(p[,5]==1)))
-        if (length(omit)>0){
-          cat('Omit:', omit, '\n')
-          p <- p[-omit,]
-          S <- S[-omit,]
-        }
-
-        out <- val.surv.mi(p, S, g=5, main=paste("Center", Centers.anonimized[j]),
-                           file_path=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", model, '.', select.value, '.', type, '.', Centers[j], ".png"),
-                           save_plots=TRUE,time=NULL,lim=c(0,1),dist=TRUE)
-      }
-    }
-  }
-}
-
-# combine calibration plots into one
-for (select.value in c("All", "Positive", "Negative")){
-  for (y in c('Rec.5', 'MSM.5')){
-    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/", y, ".", select.value, ".png"), width=16, height=16, units="cm", res=300)
-    par(mar=rep(0, 4))
-    layout(matrix(1:4, ncol=2, byrow=TRUE))
-    Centers.val <- rep(1:4, 3)
-    for (i in 1:4){
-      plot(NA, xlim=0:1, ylim=0:1, xaxt="n", yaxt="n", bty="n")
-      img <- readPNG(paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", y, ".", select.value, ".BS.", Centers[Centers.val[i]], ".png"))
-      rasterImage(img, 0, 0, 1, 1)
-    }
-    grDevices::dev.off()
-  }
-}
-```
-
-# Nomogram
-```{r, eval=TRUE}
-# Limits of Age, Breslow, Rdamcrit
-Age.class <- seq(20, 80, 20)
-Breslow.class <- c(seq(0.1, 0.9, 0.2), 1:7)
-Rdamcrit.class <- c(seq(1, 10, 2)/100, seq(1, 10, 2)/10, seq(1, 10, 2))
-
-# initial nomogram
-nom.0 <- nomogram(f.mi.BS.Rec.5,maxscale=10,
-                Breslow=Breslow.class, Rdamcrit=Rdamcrit.class, Age.SN=Age.class)
-
-# baseline hazard of recurrence
-f.Rec.basehaz <- basehaz(f.mi.BS.Rec.5,TRUE)
-h0.Rec <- f.Rec.basehaz$hazard[f.Rec.basehaz$time==max(f.Rec.basehaz$time[f.Rec.basehaz$time<=horizon])]
-
-# lp range Recurrence
-rc.Rec <- (1/attributes(nom.0)$info$sc)
-int.Rec <- attributes(nom.0)$info$Intercept
-lp.range <- int.Rec+rc.Rec*(0:10)*2
-
-# nomogram
-nom <- nomogram(f.mi.BS.Rec.5, maxscale=10,
-              Breslow=Breslow.class, Rdamcrit=Rdamcrit.class, Age.SN=Age.class,
-              lp=TRUE, fun.at=fun.event(lp=lp.range, h0=h0.Rec))
-attributes(nom)$names[attributes(nom)$names=="Age.SN"]<-"Age"
-attributes(nom)$names[attributes(nom)$names=="Loc_CAT"]<-"Location"
-attributes(nom)$names[attributes(nom)$names=="Rdamcrit"]<-"Tumor burden"
-attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Positive)"]<-"Breslow + positive SN"
-attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Negative)"]<-"Breslow + negative SN"
-par(mar = rep(0, 4))
-plot(nom, total.sep.page=TRUE, col.grid=gray(c(0.8, 0.95)))
-nom.0
-nom
-```
-
-# Risk score distribution
-```{r, eval=TRUE}
-# baseline hazard of MSM
-f.MSM.basehaz <- basehaz(f.mi.BS.MSM.5,TRUE)
-h0.MSM <- f.MSM.basehaz$hazard[f.MSM.basehaz$time==max(f.MSM.basehaz$time[f.MSM.basehaz$time<=horizon])]
-lp.MSM <- f.mi.BS.MSM.5$linear
-
-# lp range MSM
-rc.MSM <- f.mi.BS.MSM.5$coef[1]*rc.Rec
-int.MSM <- f.mi.BS.MSM.5$coef[1]*int.Rec
-max.range <- 16
-points.range <- 0:max.range
-lp.points.Rec.5 <- int.Rec+rc.Rec*points.range
-lp.points.MSM.5 <- int.MSM+rc.MSM*points.range
-
-# Recurrence lp scores
-Breslow.trunc <- last.imputed.data.set$Breslow
-Breslow.trunc[last.imputed.data.set$Breslow<0.1] <- .1
-Breslow.trunc[last.imputed.data.set$Breslow>7] <- 7
-lp.score.Rec <- predict(f.mi.BS.Rec.5,newdata=data.frame(SNstatus=last.imputed.data.set$SNstatus,
-                                                Age.SN=last.imputed.data.set$Age.SN,
-                                                Ulceration=last.imputed.data.set$Ulceration,
-                                                Loc_CAT=last.imputed.data.set$Loc_CAT,
-                                                Breslow=Breslow.trunc,
-                                                Rdamcrit=last.imputed.data.set$Rdamcrit))
-mean(lp.score.Rec)
-score.Rec <- round((lp.score.Rec-int.Rec)/rc.Rec,0)
-max.breaks <- max.range+1
-h.Rec <- hist(score.Rec,plot=FALSE,breaks=0:max.breaks,right=FALSE)
-
-# score table
-data.frame(table(score.Rec)/length(score.Rec))
-```
-
-# Plot of risk score distribution
-```{r, eval=TRUE, message=FALSE}
-h.plot <- h.Rec
-h.plot$density <- 100*h.Rec$density
-p.Rec <- fun.event(lp=lp.points.Rec.5, h0=h0.Rec)
-p.MSM <- fun.event(lp=lp.points.MSM.5, h0=h0.MSM)
-x.lim <- c(0, max.range+1)
-y.lim <- c(0, 100)
-y.lim.hist <- c(0, 25)
-png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Risk.distribution.png"))
-par(mar = c(5,5,2,5))
-plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col='white')
-abline(h=(0:30)*1,col="light grey",lty=2)
-axis(side = 4)
-mtext(side = 4, line = 3, 'Risk score distribution (%)')
-par(new = TRUE)
-plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col="white")
-par(new = TRUE)
-plot(points.range,p.Rec*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=15, xlab="Risk score",ylab="Predicted 5-year risk (%)")
-points(points.range,p.MSM*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=16)
-legend(x=0,y=100,legend=c("Recurrence","Melanoma-specific mortality"),pch=c(15,16),cex=.9,bg='white')
-grDevices::dev.off()
-```
-
-```{r, eval=TRUE, echo=FALSE}
-par(mar = c(5,5,2,5))
-plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col='white')
-abline(h=(0:30)*1,col="light grey",lty=2)
-axis(side = 4)
-mtext(side = 4, line = 3, 'Risk score distribution (%)')
-par(new = TRUE)
-plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col="white")
-par(new = TRUE)
-plot(points.range,p.Rec*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=15, xlab="Risk score",ylab="Predicted 5-year risk (%)")
-points(points.range,p.MSM*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=16)
-legend(x=0,y=100,legend=c("Recurrence","Melanoma-specific mortality"),pch=c(15,16),cex=.9,bg='white')
-```
-
-```{r, eval=TRUE}
-# reset margins
-par(mar = c(5,5,2,5))
-
-# table of risk distribution
-cbind(points.range, round(p.Rec*100, 1), round(p.MSM*100, 1))[c(1, 9, 10, 12, 13),]
-
-# save for Rshiny app
-coef.Rec <- stats::coef(f.mi.BS.Rec.5)
-center.Rec <- f.mi.BS.Rec.5$center
-save(h0.Rec, h0.MSM, coef.Rec, c.Breslow, center.Rec,
-     file=paste0('Z:/Project Melanoom/PaperMelanoma/Results/h0.Rec.MSM.Rdata'), compress=TRUE)
-load(file=paste0('Z:/Project Melanoom/PaperMelanoma/Results/h0.Rec.MSM.Rdata'))
-```
-
-# Test if model is correct
-### Check calibration
-Intercept of recurrence should be the same as when setting covariates to 'zero' values minus the center
-```{r, eval=TRUE}
-int.Rec
-20*f.mi.BS.Rec.5$coef["Age.SN"]+
-  (log(0.1)-c.Breslow)*f.mi.BS.Rec.5$coef["Breslow"]+
-  log(0.01)*f.mi.BS.Rec.5$coef["Rdamcrit"]-
-  f.mi.BS.Rec.5$center
-```
-
-### Check mean event rates of risk distribution
-#### Recurrence
-```{r, eval=TRUE}
-sum(h.Rec$density*fun.event(lp=lp.points.Rec.5, h0=h0.Rec))
-# equal to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
-SF<-survfit(S.Rec.5~1,data=sel.data)
-100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
-# average probability of recurrence
-mean(fun.event(lp=lp.score.Rec, h0=h0.Rec))
-```
-
-#### MSM
-```{r, eval=TRUE}
-sum(h.Rec$density*fun.event(lp=lp.points.MSM.5, h0=h0.MSM))
-# equal to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
-SF<-survfit(S.MSM.5~1,data=sel.data)
-100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
-# average probability of MSM
-mean(fun.event(lp=lp.MSM, h0=h0.MSM))
-```
-
-# Check if predictions of new and old model align for negative SN patients
-## New model
-```{r, eval=TRUE}
-lp.negative <- predict(f.mi.BS.Rec.5,
-                       newdata=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative",],
-                       type="lp")
-new.negativeSN.tool <- 1-exp(-h0.Rec*exp(lp.negative))
-```
-
-## Old model with new data
-```{r, eval=TRUE}
-rscript.negative <- function(breslow, ulceration, location) {
-  S0 <- 0.85837329
-  lp <- 0.79351989*log(breslow) +
-    0.61912477*I(ulceration == 1) +
-    0.29594843*I(location == 1) +
-    0.42802046*I(location == 2) +
-    0.86053044*I(location == 3) - 0.98860941
-  S <- 100 - 100*S0^exp(lp)
-  return(S)
-}
-old.negativeSN.tool <- rscript.negative(breslow=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Breslow"],
-        ulceration=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Ulceration"])-1,
-        location=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Loc_CAT"])-1)
-```
-
-## Compare
-```{r, eval=TRUE}
-plot(x=new.negativeSN.tool*100, y=old.negativeSN.tool,
-     xlim=c(0, 100), ylim=c(0, 100),
-     xlab="New predictions", ylab="Old predictions",
-     main="Predicted recurrence of patients with negative SN status")
-abline(a=0, b=1, col="red")
-```
-
-## Refitted old model new data
-```{r, eval=TRUE}
-refitted.old.negative <- rms::cph(S.Rec.5 ~ log(Breslow)+Ulceration+Loc_CAT,
-                     data=last.imputed.data.set,
-                     y=TRUE, x=TRUE,
-                     subset=last.imputed.data.set$SNstatus=="Negative")
-refitted.old.negative
-f.Rec.basehaz.neg <- basehaz(refitted.old.negative,TRUE)
-h0.neg <- f.Rec.basehaz.neg$hazard[f.Rec.basehaz.neg$time==max(f.Rec.basehaz.neg$time[f.Rec.basehaz.neg$time<=horizon])]
-refit.old.negativeSN.tool <- fun.event(predict(refitted.old.negative),  h0.neg)
-```
-
-## Compare
-```{r, eval=TRUE}
-plot(x=new.negativeSN.tool*100, y=refit.old.negativeSN.tool*100,
-     xlim=c(0, 100), ylim=c(0, 100),
-     xlab="New predictions", ylab="Old predictions",
-     main="Predicted recurrence of patients with negative SN status")
-abline(a=0, b=1, col="red")
-```
-
-## Check mean recurrence probabilities
-```{r, eval=TRUE}
-SF<-survfit(S.Rec.5[last.imputed.data.set$SNstatus=="Negative"]~1)
-100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
-mean(new.negativeSN.tool)
-mean(refit.old.negativeSN.tool)
-mean(old.negativeSN.tool)
-```
-
-# Check if predictions of new and old model align for positive SN patients
-## New model
-```{r, eval=TRUE}
-lp.positive <- predict(f.mi.BS.Rec.5,
-                       newdata=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive",],
-                       type="lp")
-new.positiveSN.tool <- 1-exp(-h0.Rec*exp(lp.positive))
-```
-
-## Old model with new data
-```{r, eval=TRUE}
-rscript.positive <- function(age, ulceration, tumor_burden, breslow) {
-  S0 <- 0.51762362
-  lp <- 0.56109945 * log(age) +
-    0.23383770 * log(tumor_burden) +
-    0.35489235 * log(breslow) +
-    0.36562098 * I(ulceration == 1) - 2.709225
-  S <- 100 - 100 * S0^exp(lp)
-  return(S)
-}
-old.positiveSN.tool <- rscript.positive(age=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Age.SN"],
-                                        ulceration=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Ulceration"])-1,
-                                        breslow=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Breslow"],
-                                        tumor_burden=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Rdamcrit"]))
-```
-
-## Compare
-```{r, eval=TRUE}
-plot(x=new.positiveSN.tool*100, y=old.positiveSN.tool,
-     xlim=c(0, 100), ylim=c(0, 100),
-     xlab="New predictions", ylab="Old predictions",
-     main="Predicted recurrence of patients with positive SN status")
-abline(a=0, b=1, col="red")
-```
-
-## Refitted old model new data
-```{r, eval=TRUE}
-refitted.old.positive <- rms::cph(S.Rec.5 ~ log(Breslow)+Ulceration+Loc_CAT, 
-                      data=last.imputed.data.set,
-                      y=TRUE, x=TRUE,
-                      subset=last.imputed.data.set$SNstatus=="Positive")
-refitted.old.positive
-f.Rec.basehaz.pos <- basehaz(refitted.old.positive,TRUE)
-h0.pos <- f.Rec.basehaz.pos$hazard[f.Rec.basehaz.pos$time==max(f.Rec.basehaz.pos$time[f.Rec.basehaz.pos$time<=horizon])]
-refit.old.positiveSN.tool <- fun.event(predict(refitted.old.positive),  h0.pos)
-```
-
-## Compare
-```{r, eval=TRUE}
-plot(x=new.positiveSN.tool*100, y=refit.old.positiveSN.tool*100,
-     xlim=c(0, 100), ylim=c(0, 100),
-     xlab="New predictions", ylab="Old predictions",
-     main="Predicted recurrence of patients with positive SN status")
-abline(a=0, b=1, col="red")
-```
-
-## Check mean recurrence probabilities
-```{r, eval=TRUE}
-SF<-survfit(S.Rec.5[last.imputed.data.set$SNstatus=="Positive"]~1)
-100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
-mean(new.positiveSN.tool)
-mean(refit.old.positiveSN.tool)
-mean(old.positiveSN.tool)
-```
-
-#### Example to check probabilities with app
-```{r, eval=TRUE}
-patient.nr <- 115 # 395 and 396 = Negative, 397 and 398 = Positive, 30 = Head & neck, 115 has large Breslow
-data.to.be.imputed[patient.nr, c("SNstatus", "Age.SN", "Ulceration", "Loc_CAT", "Breslow", "Rdamcrit")]
-lp.score.Rec.true <- predict(f.mi.BS.Rec.5)[patient.nr]
-lp.score.Rec.try <- predict(f.mi.BS.Rec.5, newdata=data.frame(SNstatus=data.to.be.imputed[patient.nr,"SNstatus"],
-                                                            Age.SN=data.to.be.imputed[patient.nr,"Age.SN"],
-                                                            Ulceration=data.to.be.imputed[patient.nr,"Ulceration"],
-                                                            Loc_CAT=data.to.be.imputed[patient.nr,"Loc_CAT"],
-                                                            Breslow=data.to.be.imputed[patient.nr,"Breslow"],
-                                                            Rdamcrit=data.to.be.imputed[patient.nr,"Rdamcrit"]),
-                                                            type="lp")
-p.Rec.example <- fun.event(lp=lp.score.Rec.try, h0=h0.Rec)
-p.MSM.example <- fun.event(lp=lp.score.Rec.try, h0=h0.MSM)
-
-# manual calculation
-lp.score.Rec.manual <- coef.Rec["SNstatus=Positive"]*(as.numeric(data.to.be.imputed[patient.nr,"SNstatus"])-1)+
-              coef.Rec["Age.SN"]*data.to.be.imputed[patient.nr,"Age.SN"]+
-              coef.Rec["Ulceration=Yes"]*(as.numeric(data.to.be.imputed[patient.nr,"Ulceration"])-1)+
-              coef.Rec["Loc_CAT=leg"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="leg")+
-              coef.Rec["Loc_CAT=trunk"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="trunk")+
-              coef.Rec["Loc_CAT=head & neck"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="head & neck")+
-              coef.Rec["Breslow"]*(log(data.to.be.imputed[patient.nr,"Breslow"])-c.Breslow)+
-              coef.Rec["Rdamcrit"]*log(data.to.be.imputed[patient.nr,"Rdamcrit"])+
-              coef.Rec["SNstatus=Positive * Breslow"]*(log(data.to.be.imputed[patient.nr,"Breslow"])-c.Breslow)*(as.numeric(data.to.be.imputed[patient.nr,"SNstatus"])-1)
-
-# lp check
-lp.score.Rec.true
-lp.score.Rec.try
-lp.score.Rec.manual-center.Rec
-
-# probability of recurrence with fun.event
-fun.event(lp=lp.score.Rec.try, h0=h0.Rec)
-# probability of recurrence manually
-1-exp(-h0.Rec*exp(lp.score.Rec.manual-center.Rec))
-
-# check calculation of lp of MSM
-lp.score.MSM.true <- predict(f.mi.BS.MSM.5)[patient.nr]
-lp.score.MSM.true
-lp.MSM.try <- MSM.cal.fact*(lp.score.Rec.manual-center.Rec)
-lp.MSM.try
-
-# probability of MSM with fun.event
-fun.event(lp=lp.MSM.try, h0=h0.MSM)
-# probability of MSM manually
-1-exp(-h0.MSM*exp(MSM.cal.fact*(lp.score.Rec.manual-center.Rec)))
-```
-
-### Histogram for negative SN status patients
-```{r, eval=TRUE}
-Breslow.trunc.neg <- last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Breslow"]
-Breslow.trunc.neg[last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Breslow"]<0.1] <- .1
-Breslow.trunc.neg[last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Breslow"]>7] <- 7
-lp.score.Rec.negative <- predict(f.mi.BS.Rec.5,newdata=data.frame(SNstatus=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "SNstatus"],
-                                                       Age.SN=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Age.SN"],
-                                                       Ulceration=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Ulceration"],
-                                                       Loc_CAT=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Loc_CAT"],
-                                                       Breslow=Breslow.trunc.neg,
-                                                       Rdamcrit=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Rdamcrit"]))
-mean(lp.score.Rec.negative)
-# OK to use int.Rec and rc.Rec based on all patients?
-score.Rec.neg <- round((lp.score.Rec.negative-int.Rec)/rc.Rec,0)
-h.Rec.neg <- hist(score.Rec.neg,plot=FALSE,breaks=0:max.breaks,right=FALSE)
-h.plot.neg <- h.Rec.neg
-h.plot.neg$density <- 100*h.Rec.neg$density
-```
-
-### Histogram for positive SN status patients
-```{r, eval=TRUE}
-Breslow.trunc.pos <- last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Breslow"]
-Breslow.trunc.pos[last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Breslow"]<0.1] <- .1
-Breslow.trunc.pos[last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Breslow"]>7] <- 7
-lp.score.Rec.positive <- predict(f.mi.BS.Rec.5,newdata=data.frame(SNstatus=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "SNstatus"],
-                                                                Age.SN=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Age.SN"],
-                                                                Ulceration=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Ulceration"],
-                                                                Loc_CAT=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Loc_CAT"],
-                                                                Breslow=Breslow.trunc.pos,
-                                                                Rdamcrit=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Rdamcrit"]))
-mean(lp.score.Rec.positive)
-# OK to use int.Rec and rc.Rec based on all patients?
-score.Rec.pos <- round((lp.score.Rec.positive-int.Rec)/rc.Rec,0)
-h.Rec.pos <- hist(score.Rec.pos,plot=FALSE,breaks=0:max.breaks,right=FALSE)
-h.plot.pos <- h.Rec.pos
-h.plot.pos$density <- 100*h.Rec.pos$density
-
-par(mfrow=c(1, 3))
-plot(h.plot,freq=FALSE,axes=TRUE, xlab="Risk score", xlim=x.lim,ylim=y.lim.hist,
-     ylab="Risk score distribution (%)", main="All patients",col="white")
-plot(h.plot.neg,freq=FALSE,axes=TRUE, xlab="Risk score", xlim=x.lim,
-     ylab="Risk score distribution (%)",ylim=y.lim.hist,main="Negative SN status",col="white")
-plot(h.plot.pos,freq=FALSE,axes=TRUE, xlab="Risk score", xlim=x.lim,
-     ylab="Risk score distribution (%)",ylim=y.lim.hist,main="Positive SN status",col="white")
-```
-
-### Test proportional hazards assumption
-```{r, eval=TRUE}
-test.prop<-survival::cox.zph(f.mi.full.Rec.5$fits[[1]])
-test.prop
-plot(test.prop)
-```

### 2022-11-29 Code/5_TestNonlinearity.Rmd
Validation data added

---
title: "Test non-linearity"
author: "Carolien C.H.M. Maas"
date: "14-11-2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Test non-linearity

### Load packages, functions, and data 
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# erase memory
rm(list = ls(all.names = TRUE))

# for reproducibility
set.seed(100)

# load libraries
library(foreign)
library(mice)
library(survminer)
library(prodlim)
library(rms)
library(Hmisc)
library(png)
# remotes::install_github("CHMMaas/PredictionTools")
library(PredictionTools)  # to plot calibration
# import helper functions
source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")
```

# Test non-lineairities
### Test if Age should be modelled non-linearly
```{r, eval=TRUE}
load('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData')
options(datadist='dd.training')
options(digits=8)

form.linear.full.Rec <- S.Rec.5.training ~
  SNstatus +
  Sex +
  Age.SN +
  Ulceration +
  Loc_CAT +
  Histology +
  Breslow +
  multiple.fields +
  Tot_SNs_neg +
  Tot_SNs_pos +
  Rdamcrit +
  Dewar +
  SNstatus:Sex +
  SNstatus:Age.SN +
  SNstatus:Ulceration +
  SNstatus:Loc_CAT +
  SNstatus:Histology +
  SNstatus:Breslow +
  SNstatus:multiple.fields
form.nonlinear.age.full.Rec <- S.Rec.5.training ~
  SNstatus +
  Sex +
  rcs(Age.SN) +
  Ulceration +
  Loc_CAT +
  Histology +
  Breslow +
  multiple.fields +
  Tot_SNs_neg +
  Tot_SNs_pos +
  Rdamcrit +
  Dewar +
  SNstatus:Sex +
  SNstatus:rcs(Age.SN) +
  SNstatus:Ulceration +
  SNstatus:Loc_CAT +
  SNstatus:Histology +
  SNstatus:Breslow +
  SNstatus:multiple.fields
linear.full.Rec <- Hmisc::fit.mult.impute(form.linear.full.Rec, cph, xtrans=imputed.training.data,
                                          data=training.data.to.be.imputed,n.impute=m, pr=FALSE,
                                          fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
nonlinear.age.full.Rec <- Hmisc::fit.mult.impute(form.nonlinear.age.full.Rec, cph, xtrans=imputed.training.data,
                                                 data=training.data.to.be.imputed,n.impute=m, pr=FALSE,
                                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
# distribution of Age
hist(last.imputed.training.data.set$Age.SN)
# linear model
linear.full.Rec$loglik[2]
# using log(Age) gives slightly better log likelihood
nonlinear.age.full.Rec$loglik[2]
```

### Test if Breslow should be modelled non-linearly
```{r, eval=TRUE}
form.nonlinear.Breslow.full.Rec <- S.Rec.5.training ~
  SNstatus +
  Sex +
  Age.SN +
  Ulceration +
  Loc_CAT +
  Histology +
  rcs(Breslow) +
  multiple.fields +
  Tot_SNs_neg +
  Tot_SNs_pos +
  Rdamcrit +
  Dewar +
  SNstatus:Sex +
  SNstatus:Age.SN +
  SNstatus:Ulceration +
  SNstatus:Loc_CAT +
  SNstatus:Histology +
  SNstatus:rcs(Breslow) +
  SNstatus:multiple.fields
nonlinear.Breslow.full.Rec <- Hmisc::fit.mult.impute(form.nonlinear.Breslow.full.Rec, cph, xtrans=imputed.training.data,
                                                     data=training.data.to.be.imputed,n.impute=m, pr=FALSE,
                                                     fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
# distribution of Breslow
hist(last.imputed.training.data.set$Breslow)
# linear model
linear.full.Rec$loglik[2]
# using log(Breslow) gives better log likelihood
nonlinear.Breslow.full.Rec$loglik[2]
# test
anova(nonlinear.Breslow.full.Rec)
```

### Test if Rdamcrit should be modelled non-linearly
```{r, eval=TRUE}
form.nonlinear.Rdamcrit.full.Rec <- S.Rec.5.training ~
  SNstatus +
  Sex +
  Age.SN +
  Ulceration +
  Loc_CAT +
  Histology +
  Breslow +
  multiple.fields +
  Tot_SNs_neg +
  Tot_SNs_pos +
  rcs(Rdamcrit) +
  Dewar +
  SNstatus:Sex +
  SNstatus:Age.SN +
  SNstatus:Ulceration +
  SNstatus:Loc_CAT +
  SNstatus:Histology +
  SNstatus:Breslow +
  SNstatus:multiple.fields
nonlinear.Rdamcrit.full.Rec <- Hmisc::fit.mult.impute(form.nonlinear.Rdamcrit.full.Rec, cph, xtrans=imputed.training.data,
                                                      data=training.data.to.be.imputed,n.impute=m, pr=FALSE,
                                                      fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
# distribution of Rdamcrit
hist(last.imputed.training.data.set$Rdamcrit)
# linear model
linear.full.Rec$loglik[2]
# using log(Rdamcrit) gives better log likelihood
nonlinear.Rdamcrit.full.Rec$loglik[2]
```

### 2022-11-29 Code/6_FitModels.Rmd
Validation data added

---
title: "Fit models"
author: "Carolien C.H.M. Maas"
date: "14-11-2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Fit models

### Load packages, functions, and data 
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# erase memory
rm(list = ls(all.names = TRUE))

# for reproducibility
set.seed(100)

# load libraries
library(Hmisc)
library(stats)
library(rms)
library(utils)
```

# Full model for recurrence
```{r, eval=TRUE}
load('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData')
options(datadist='dd.training')
options(digits=8)

form.full.Rec.5 <- S.Rec.5.training ~
  SNstatus +
  Sex +
  Age.SN +
  Ulceration +
  Loc_CAT +
  Histology +
  log(Breslow) +
  multiple.fields +
  Tot_SNs_neg +
  Tot_SNs_pos +
  log(Rdamcrit) +
  Dewar +
  SNstatus:Sex +
  SNstatus:Age.SN +
  SNstatus:Ulceration +
  SNstatus:Loc_CAT +
  SNstatus:Histology +
  SNstatus:log(Breslow) +
  SNstatus:multiple.fields
f.mi.full.Rec.5 <- Hmisc::fit.mult.impute(form.full.Rec.5, cph, xtrans=imputed.training.data,
                                 data=training.data.to.be.imputed,n.impute=m, pr=FALSE,
                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
stats::anova(f.mi.full.Rec.5)
summary(f.mi.full.Rec.5, Tot_SNs_pos=c(0, 1))
```

# Final model for recurrence selected by backward selection with p = 0.01
```{r, eval=TRUE}
c.Breslow <- mean(log(last.imputed.training.data.set$Breslow))
form.BS.Rec.5 <- S.Rec.5.training ~
  SNstatus +
  Age.SN +
  Ulceration +
  Loc_CAT +
  I(log(Breslow)-c.Breslow) +
  log(Rdamcrit) +
  SNstatus:I(log(Breslow)-c.Breslow)
f.mi.BS.Rec.5 <- Hmisc::fit.mult.impute(form.BS.Rec.5,cph,xtrans=imputed.training.data,data=training.data.to.be.imputed,n.impute=m,
                           pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
# backward selection
# round(stats::anova(f.mi.BS.Rec.5)[order(stats::anova(f.mi.BS.Rec.5)[, 'P']),], 3)
summary(f.mi.BS.Rec.5, Breslow=exp(c.Breslow))
coef(f.mi.BS.Rec.5)
stats::anova(f.mi.BS.Rec.5)
```

# Refitted full model for melanoma specific mortality
```{r, eval=TRUE}
form.full.MSM.5.refit <- S.MSM.5.training ~
  SNstatus +
  Sex +
  Age.SN +
  Ulceration +
  Loc_CAT +
  Histology +
  log(Breslow) +
  multiple.fields +
  Tot_SNs_neg +
  Tot_SNs_pos +
  log(Rdamcrit) +
  Dewar +
  SNstatus:Sex +
  SNstatus:Age.SN +
  SNstatus:Ulceration +
  SNstatus:Loc_CAT +
  SNstatus:Histology +
  SNstatus:log(Breslow) +
  SNstatus:multiple.fields
f.mi.full.MSM.5.refit <- Hmisc::fit.mult.impute(form.full.MSM.5.refit,cph,xtrans=imputed.training.data,data=training.data.to.be.imputed,n.impute=m,
                                 pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
summary(f.mi.full.MSM.5.refit, Tot_SNs_pos=c(0, 1))
stats::anova(f.mi.full.MSM.5.refit)
```

# Refitted final model for melanoma specific mortality
```{r, eval=TRUE}
form.BS.MSM.5.refit <- S.MSM.5.training ~
  SNstatus +
  Age.SN +
  Ulceration +
  Loc_CAT +
  I(log(Breslow)-c.Breslow) +
  log(Rdamcrit) +
  SNstatus:I(log(Breslow)-c.Breslow)
f.mi.BS.MSM.5.refit <- Hmisc::fit.mult.impute(form.BS.MSM.5.refit,cph,xtrans=imputed.training.data,data=training.data.to.be.imputed,n.impute=m,
                                  pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
summary(f.mi.BS.MSM.5.refit, Breslow=exp(c.Breslow))
stats::anova(f.mi.BS.MSM.5.refit)
```

# Calibrated final model for melanoma specific mortality
Fit disease specific mortality to the linear predictor of recurrence
```{r, eval=TRUE}
lp.BS <- f.mi.BS.Rec.5$linear
f.mi.BS.MSM.5 <- rms::cph(S.MSM.5.training~lp.BS,y=TRUE,x=TRUE)
MSM.cal.fact <- stats::coef(f.mi.BS.MSM.5)
MSM.cal.fact
stats::confint(f.mi.BS.MSM.5)
utils::write.table(paste0(sprintf("%.2f", stats::coef(f.mi.BS.MSM.5)), " [",
            sprintf("%.2f", as.numeric(stats::confint(f.mi.BS.MSM.5))[1]), "; ",
            sprintf("%.2f", as.numeric(stats::confint(f.mi.BS.MSM.5))[2]), "]"),
            file='Z:/Project Melanoom/PaperMelanoma/Results/calibration.MSS.txt', 
            row.names=FALSE, col.names=FALSE, sep=",")

save(list=c("dd.training", "training.data.to.be.imputed", "m", "imputed.training.data", "last.imputed.training.data.set", 
            "horizon", "S.Rec.training", "S.MSM.training", "S.Rec.5.training", "S.MSM.5.training",
            "f.mi.full.Rec.5", "f.mi.BS.Rec.5", "f.mi.full.MSM.5.refit", "f.mi.BS.MSM.5.refit", "f.mi.BS.MSM.5", "MSM.cal.fact", "c.Breslow", 
            "form.full.Rec.5", "form.BS.Rec.5", "form.full.MSM.5.refit", "form.BS.MSM.5.refit"),
     file='Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata', compress=TRUE)
load('Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData')
save(list=c("dd.validation", "validation.data.to.be.imputed", "m", "imputed.validation.data", "last.imputed.validation.data.set", "horizon", 
            "S.Rec.5.validation", "S.MSM.5.validation", "f.mi.BS.Rec.5"),
     file=paste0('Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.Rdata'), compress=TRUE)
```

### 2022-11-29 Code/7_CreateTablesandPlots.Rmd
Validation data added

--- 2022-06-20 Code/PaperMelanoma.Rmd
@@
 ---
-title: "PaperMelanoma"
+title: "Create Tables"
 author: "Carolien C.H.M. Maas"
-date: "3-6-2022"
+date: "14-11-2022"
 output: html_document
@@
 
-## Development and validation of a clinial prediction model to predict recurrence and disease-specific mortality in patients with melanoma after sentinel lymph node biopsy
+## Create Tables
 
-This file includes the code and results for the above mentioned paper.
-
-# Data preparation
 ### Load packages, functions, and data 
 ```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
+# erase memory
+rm(list = ls(all.names = TRUE))
+
 # for reproducibility
@@
 library(png)
-# import function to create table of missing values
-source("Z:/Project Melanoom/PaperMelanoma/Code/table.missing.values.R") 
-# import function to plot calibration
-source("Z:/Project Melanoom/PaperMelanoma/Code/val_surv_mi_v3.R")
-# helper functions for calibration plots
-km <- function(sel, S.km, horizon){
-  S.km.sel <- S.km[sel,]
-  sf <- survfit(S.km.sel~1)
-  year1 <- max(sf$time[sf$time<=horizon])
-  1-sf$surv[sf$time==year1]
-}
-km.lower <- function(sel, S.km, horizon){
-  S.km.sel <- S.km[sel,]
-  sf <- survfit(S.km.sel~1)
-  year1 <- max(sf$time[sf$time<=horizon])
-  1-sf$lower[sf$time==year1]
-}
-km.upper <- function(sel, S.km, horizon){
-  S.km.sel <- S.km[sel,]
-  sf <- survfit(S.km.sel~1)
-  year1 <- max(sf$time[sf$time<=horizon])
-  1-sf$upper[sf$time==year1]
-}
-fun.event <- function(lp, h0)
-{
-  h <- h0*exp(lp)
-  p <- 1-exp(-h)
-  return(p)
-}
-
-# load data file
-original.data <- foreign::read.spss(file="Z:/Project Melanoom/PaperMelanoma/Data/Dataset Predictiemodel Stassen.sav",
-                                    to.data.frame=TRUE, origin="01-01-1970")
-
-# Select necessary variables
-selected.vars <- c("EXCLUSIE", "DatabaseNR", "Center", "Sex", "Age", "DOB",
-                   "Ulceration", "Loc_CAT",            # OLD data set: Loc_CAT2
-                   "Histology", "Breslow", "Tot_nr_SNs",
-                   "nr_SNs_Field_1", "nr_SNs_Field_2", "nr_SNs_Field_3",
-                   "Mitosis",                          # needed for AJCC.7.i
-                   "SNdate", "Recurrence", "Date_1st_Rec", "Last_FU",
-                   "dead", "Status",
-                   "SNstatus",                         # distinguish between positive and negative SN patients
-                   "Tot_SNs_pos", "Rdamcrit", "Dewar") # additional for positive SN patients
-```
-
-### Inclusion criteria
-Exclude when Exclusion = 2: based on duplicates, urogenital melanoma, etc. Note:
-Leave the patients included that have EXCLUSIE = 1 and Dbstatus = 8 (Dbstatus=0: 
-negative, Dbstatus=1: positive), even though diagnosis and operation date are 
-very close.
-```{r, eval=TRUE}
-sel.data <- original.data[original.data$EXCLUSIE!="Exclusion", selected.vars]
-cat("Number of patients excluded:", nrow(original.data)-nrow(sel.data), "\n")
-```
-Omit patient A0523 - no last follow-up date
-```{r, eval=TRUE}
-sel.data[which(is.na(sel.data$Last_FU)), "DatabaseNR"]
-sel.data <- sel.data[sel.data$DatabaseNR!="A0523               ",]
-```
-
-### Recode
-#### Use age at SN instead of age at diagnosis
-```{r, eval=TRUE}
-sel.data$Age.SN <- floor((sel.data$SNdate-sel.data$DOB)/(24*60*60)/365.25)
-```
-
-#### Location
-Merge location other with trunk
-```{r, eval=TRUE}
-sel.data$Loc_CAT[sel.data$Loc_CAT=="other"] <- "trunk"
-```
-
-Merge LMM other with other
-```{r, eval=TRUE}
-sel.data$Histology[sel.data$Histology=="LMM"] <- "other"
-```
-
-#### Define multiple fields
-```{r, eval=TRUE}
-sel.data$multiple.fields <- ifelse(!is.na(sel.data$nr_SNs_Field_2)&sel.data$nr_SNs_Field_2>0,1,0)
-```
-
-#### Number of positive and negative nodes
-Check: there are no negative patients that have positive nodes
-```{r, eval=TRUE}
-sum(!is.na(sel.data$Tot_SNs_pos)&sel.data$Tot_SNs_pos>0&sel.data$SNstatus=="Negative")
-```
-
-Create number of negative nodes removed
-```{r, eval=TRUE}
-sel.data$Tot_SNs_neg <- sel.data$Tot_nr_SNs - sel.data$Tot_SNs_pos
-```
-
-467 positive SN patients have negative nodes
-```{r, eval=TRUE}
-sum(!is.na(sel.data$Tot_SNs_neg)&sel.data$Tot_SNs_neg>0&sel.data$SNstatus=="Positive")
-```
-
-#### SN tumour burden
-11 patients with positive SN have SN tumour burden of exactly 1
-```{r, eval=TRUE}
-sum(sel.data$Rdamcrit==1&sel.data$SNstatus=="Positive"&!is.na(sel.data$Rdamcrit))
-```
-
-Set Rdamcrit of negative SN patients equal to 0
-```{r, eval=TRUE}
-sel.data$Rdamcrit[sel.data$SNstatus=="Negative"] <- 1 # since log(1) = 0
-```
-
-#### Location of metastasis
-Set Dewar for negative patients to subcap
-```{r, eval=TRUE}
-sum(sel.data$SNstatus=="Positive"&sel.data$Dewar=="subcap"&!is.na(sel.data$Dewar))
-sel.data$Dewar[sel.data$SNstatus=="Negative"] <- "subcap"
-```
-
-Merge single cells with subcap 
-```{r, eval=TRUE}
-sel.data$Dewar[sel.data$Dewar=="single cells"] <- "subcap"
-```
-
-#### Recurrence 
-Define time since diagnosis in years (follow-up time)
-```{r, eval=TRUE}
-sel.data$FU.time <- (sel.data$Last_FU-sel.data$SNdate)/(24*60*60)/365.25
-```
-
-Define recurrence time in years
-```{r, eval=TRUE}
-sel.data$Rec.time<- (sel.data$Date_1st_Rec-sel.data$SNdate)/(24*60*60)/365.25
-```
-
-Omit patients with Rec date before SN date (TODO: or impute instead?)
-```{r, eval=TRUE}
-old.n <- nrow(sel.data)
-sel.data[sel.data$Rec.time<0&!is.na(sel.data$Rec.time), "DatabaseNR"]
-sel.data <- sel.data[!(sel.data$Rec.time<0&!is.na(sel.data$Rec.time)),]
-cat("Number of patients excluded:", old.n-nrow(sel.data), "\n")
-```
-
-Set recurrence time equal to follow-up time for those without recurrence
-```{r, eval=TRUE}
-sel.data$Rec.time[sel.data$Recurrence=="no recurrence"|is.na(sel.data$SNdate)]<-sel.data$FU.time[sel.data$Recurrence=="no recurrence"|is.na(sel.data$SNdate)]
-```
-
-#### Melanoma specific mortality (MSM)
-```{r, eval=TRUE}
-sel.data$MSM<-1*(sel.data$Status=="DOD")
-```
-
-#### Mitosis
-Set Mitosis to Unknown if missing
-```{r, eval=TRUE}
-sel.data$Mitosis[is.na(sel.data$Mitosis)] <- "Unknown" # mitosis unknown for 80% of patients
-```
-
-#### Replace "Unknown" by NA
-```{r, eval=TRUE}
-sel.data$Ulceration[sel.data$Ulceration=="unknown"]<-NA
-sel.data$Loc_CAT[sel.data$Loc_CAT=="unknown"]<-NA
-sel.data$Histology[sel.data$Histology=="unknown"]<-NA
-sel.data$Dewar[sel.data$Dewar=="unknown"] <- NA
-sel.data$Mitosis[sel.data$Mitosis=="Unknown"] <- NA
-```
-
-#### Delete unused levels
-```{r, eval=TRUE}
-sel.data <- droplevels(sel.data)
-save(sel.data,file='Z:/Project Melanoom/PaperMelanoma/Data/sel.data.Rdata', compress=TRUE)
-```
-
-# Multiple imputation
-```{r, eval=TRUE}
-variables.to.be.imputed <- c("Center", "SNstatus", "Sex", "Age.SN", "Ulceration",
-                             "Loc_CAT", "Histology", "Breslow", "multiple.fields",
-                             "Tot_SNs_neg", "Tot_SNs_pos", "Rdamcrit", "Dewar",
-                             "Mitosis",
-                             "Recurrence", "dead", "MSM")
-data.to.be.imputed <- sel.data[,variables.to.be.imputed]
-```
-
-Save distributions of all variables, necessary for Harrell summaries
-```{r, eval=TRUE}
-dd<-rms::datadist(data.to.be.imputed)
-options(datadist='dd')
-options(digits=8)
-```
-
-Relation between outcome and covariates
-```{r, eval=TRUE}
-data.to.be.imputed$logRec.time <- log(sel.data$Rec.time+1/365.25)
-data.to.be.imputed$logFU.time <- log(sel.data$FU.time+1/365.25)
-missings.table(data.to.be.imputed)
-save(data.to.be.imputed,file='Z:/Project Melanoom/PaperMelanoma/Data/descr.data.Rdata', compress=TRUE)
-```
-
-Number of imputations
-```{r, eval=TRUE}
-m <- 5
-```
-
-Perform multiple imputation
-```{r, eval=FALSE}
-imputed.data <- mice::mice(data.to.be.imputed, m=m, maxit=5, print=FALSE, seed=500)
-print(imputed.data$loggedEvents)
-save(imputed.data,file='Z:/Project Melanoom/PaperMelanoma/Data/imputed.data.RData', compress=TRUE)
-```
-
-Load imputed data set
-```{r, eval=TRUE}
-load('Z:/Project Melanoom/PaperMelanoma/Data/imputed.data.RData')
-last.imputed.data.set <- mice::complete(imputed.data, m)
-summary(last.imputed.data.set)
-missings.table(last.imputed.data.set)
-```
-
-# Set as survival data
-```{r, eval=TRUE}
-S.Rec <- survival::Surv(exp(last.imputed.data.set$logRec.time), sel.data$Recurrence=="recurred")
-S.MSM <- survival::Surv(exp(last.imputed.data.set$logFU.time), sel.data$MSM)
-```
-
-### Kaplan-Meier plots
-```{r, eval=TRUE}
-center.Rec.KM <- survival::survfit(S.Rec ~ Center, data=sel.data)
-survminer::ggsurvplot(center.Rec.KM, conf.int=TRUE, risk.table=TRUE)
-center.MSM.KM <- survival::survfit(S.MSM ~ Center, data=sel.data)
-survminer::ggsurvplot(center.MSM.KM, conf.int=TRUE, risk.table=TRUE)
-```
-
-# Median follow-up time
-```{r, eval=TRUE}
-median.FU.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM)~1, data=last.imputed.data.set, reverse=TRUE))
-tab.FU.time <- c("total", paste(sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "quantile"]),
-                 paste0("[", sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "lower"]), "; ",
-                        sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "upper"]), "]")))
-median.FU.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM)~Center, data=last.imputed.data.set, reverse=TRUE))
-for (center in sort(as.character(unique(sel.data$Center)))){
-  med.FU.table <- eval(parse(text=paste0("median.FU.time$`Center=", center, "`")))
-  med.FU <- med.FU.table[3,"quantile"]
-  lower.FU <- med.FU.table[3,"lower"]
-  upper.FU <- med.FU.table[3,"upper"]
-  tab.FU.time <- rbind(tab.FU.time, c(center, paste(sprintf("%.2f", med.FU), paste0("[", sprintf("%.2f", lower.FU), "; ", sprintf("%.2f", upper.FU), "]"))))
-}
-write.table(as.data.frame(tab.FU.time),file='Z:/Project Melanoom/PaperMelanoma/Results/med.FU.table.txt',
-            row.names=FALSE, sep=",")
-tab.FU.time
-```
-
-# Recurrence at 5 years
-```{r, eval=TRUE}
-horizon <- 5
-SF.Rec.all<-survfit(S.Rec~1)
-Rec.all <- 100*(1-SF.Rec.all$surv[SF.Rec.all$time==max(SF.Rec.all$time[SF.Rec.all$time<=horizon])])
-
-SF.Rec.neg<-survfit(S.Rec[sel.data$SNstatus=="Negative"]~1)
-Rec.neg <- 100*(1-SF.Rec.neg$surv[SF.Rec.neg$time==max(SF.Rec.neg$time[SF.Rec.neg$time<=horizon])])
-
-SF.Rec.pos<-survfit(S.Rec[sel.data$SNstatus=="Positive"]~1)
-Rec.pos <- 100*(1-SF.Rec.pos$surv[SF.Rec.pos$time==max(SF.Rec.pos$time[SF.Rec.pos$time<=horizon])])
-
-SF.MSM.all<-survfit(S.MSM~1)
-MSM.all <- 100*(1-SF.MSM.all$surv[SF.MSM.all$time==max(SF.MSM.all$time[SF.MSM.all$time<=horizon])])
-
-SF.MSM.neg<-survfit(S.MSM[sel.data$SNstatus=="Negative"]~1)
-MSM.neg <- 100*(1-SF.MSM.neg$surv[SF.MSM.neg$time==max(SF.MSM.neg$time[SF.MSM.neg$time<=horizon])])
-
-SF.MSM.pos<-survfit(S.MSM[sel.data$SNstatus=="Positive"]~1)
-MSM.pos <- 100*(1-SF.MSM.pos$surv[SF.MSM.pos$time==max(SF.MSM.pos$time[SF.MSM.pos$time<=horizon])])
-
-Rec.MSM.table <- data.frame(n=c(nrow(sel.data), 
-                                sum(sel.data$SNstatus=="Negative"), 
-                                sum(sel.data$SNstatus=="Positive")),
-                            perc=c(100, 
-                                   sprintf("%.1f", sum(sel.data$SNstatus=="Negative")/nrow(sel.data)*100),
-                                   sprintf("%.1f", sum(sel.data$SNstatus=="Positive")/nrow(sel.data)*100)),
-                            Rec=sprintf("%.1f", c(Rec.all, Rec.neg, Rec.pos)), 
-                            MSM=sprintf("%.1f", c(MSM.all, MSM.neg, MSM.pos)))
-row.names(Rec.MSM.table) <- c("All", "Negative", "Positive")
-utils::write.table(as.data.frame(Rec.MSM.table),file='Z:/Project Melanoom/PaperMelanoma/Results/Rec.MSM.table.txt',
-                   row.names=TRUE, sep=",")
-Rec.MSM.table
-```
-
-# Restrict survival at 5 years
-```{r, eval=TRUE}
-S.Rec.5 <- S.Rec
-S.Rec.5[S.Rec[,1]>5, 1] <- 5 # set recurrence time to 5
-S.Rec.5[S.Rec[,1]>5, 2] <- 0 # set censoring
-
-S.MSM.5 <- S.MSM
-S.MSM.5[S.MSM[,1]>5, 1] <- 5 # set follow-up time to 5
-S.MSM.5[S.MSM[,1]>5, 2] <- 0 # set censoring
-```
-
-# Obtain C-index of AJCC7 and AJCC8
-```{r, eval=TRUE}
-TNM.7 <- c()
-AJCC.7 <- c()
-cindex.AJCC.7 <- rep(0, m)
-cindex.AJCC.7.se <- rep(0, m)
-cindex.AJCC.7.pos <- rep(0, m)
-cindex.AJCC.7.se.pos <- rep(0, m)
-cindex.AJCC.7.neg <- rep(0, m)
-cindex.AJCC.7.se.neg <- rep(0, m)
-for (i in 1:m){
-  imputation.i <- mice::complete(imputed.data,i)
-  TNM.7.i <- rep("", nrow(imputation.i))
-  AJCC.7.i <- rep(0, nrow(imputation.i))
-  for (patient in 1:nrow(imputation.i)){
-    breslow.patient <- imputation.i[patient, "Breslow"]
-    ulceration.patient <- as.character(imputation.i[patient, "Ulceration"])
-    mitosis.patient <- as.character(imputation.i[patient, "Mitosis"])
-    nodes.patient <- imputation.i[patient, "Tot_SNs_pos"]
-
-    snstatus.patient <- imputation.i[patient, "SNstatus"]
-
-    # T stage
-    if (breslow.patient <= 1){
-      if (ulceration.patient=="No" & mitosis.patient=="No"){
-        TNM.7.i[patient] <- paste0(TNM.7.i[patient], "T1a")
-      } else if (ulceration.patient=="Yes" | mitosis.patient=="Yes"){
-        TNM.7.i[patient] <- paste0(TNM.7.i[patient], "T1b")
-      }
-    } else if (breslow.patient>1 & breslow.patient <= 2){
-      if (ulceration.patient=="No"){
-        TNM.7.i[patient] <- paste0(TNM.7.i[patient], "T2a")
-      } else if (ulceration.patient=="Yes"){
-        TNM.7.i[patient] <- paste0(TNM.7.i[patient], "T2b")
-      }
-    } else if (breslow.patient>2 & breslow.patient <= 4){
-      if (ulceration.patient=="No"){
-        TNM.7.i[patient] <- paste0(TNM.7.i[patient], "T3a")
-      } else if (ulceration.patient=="Yes"){
-        TNM.7.i[patient] <- paste0(TNM.7.i[patient], "T3b")
-      }
-    } else if (breslow.patient>4){
-      if (ulceration.patient=="No"){
-        TNM.7.i[patient] <- paste0(TNM.7.i[patient], "T4a")
-      } else if (ulceration.patient=="Yes"){
-        TNM.7.i[patient] <- paste0(TNM.7.i[patient], "T4b")
-      }
-    }
-
-    # N stage
-    if (nodes.patient == 0){
-      TNM.7.i[patient] <- paste0(TNM.7.i[patient], "N0")
-    } else if (nodes.patient==1){
-      TNM.7.i[patient] <- paste0(TNM.7.i[patient], "N1a")
-    } else if (nodes.patient==2 | nodes.patient==3){
-      TNM.7.i[patient] <- paste0(TNM.7.i[patient], "N2a")
-    } else if (nodes.patient>3){
-      TNM.7.i[patient] <- paste0(TNM.7.i[patient], "N3a")
-    }
-
-    # M stage for all patients in our data set
-    TNM.7.i[patient] <- paste0(TNM.7.i[patient], "M0")
-
-    # determine AJCC.7.i
-    if (TNM.7.i[patient]=="T1aN0M0"){
-      AJCC.7.i[patient] <- 1.1
-    } else if (TNM.7.i[patient]=="T1bN0M0" | TNM.7.i[patient]=="T2aN0M0"){
-      AJCC.7.i[patient] <- 1.2
-    } else if (TNM.7.i[patient]=="T2bN0M0" | TNM.7.i[patient]=="T3aN0M0"){
-      AJCC.7.i[patient] <- 2.1
-    } else if (TNM.7.i[patient]=="T3bN0M0" | TNM.7.i[patient]=="T4aN0M0"){
-      AJCC.7.i[patient] <- 2.2
-    } else if (TNM.7.i[patient]=="T4bN0M0"){
-      AJCC.7.i[patient] <- 2.3
-    } else if (TNM.7.i[patient]=="T1aN1aM0" | TNM.7.i[patient]=="T2aN1aM0" | # N1a
-               TNM.7.i[patient]=="T3aN1aM0" | TNM.7.i[patient]=="T4aN1aM0" |
-               TNM.7.i[patient]=="T1aN2aM0" | TNM.7.i[patient]=="T2aN2aM0" | # N2a
-               TNM.7.i[patient]=="T3aN2aM0" | TNM.7.i[patient]=="T4aN2aM0"){
-      AJCC.7.i[patient] <- 3.1
-    } else if (TNM.7.i[patient]=="T1bN1aM0" | TNM.7.i[patient]=="T2bN1aM0" | # N1a
-               TNM.7.i[patient]=="T3bN1aM0" | TNM.7.i[patient]=="T4bN1aM0" |
-               TNM.7.i[patient]=="T1bN2aM0" | TNM.7.i[patient]=="T2bN2aM0" | # N2a
-               TNM.7.i[patient]=="T3bN2aM0" | TNM.7.i[patient]=="T4bN2aM0" |
-               TNM.7.i[patient]=="T1aN1bM0" | TNM.7.i[patient]=="T2aN1bM0" | # N1b
-               TNM.7.i[patient]=="T3aN1bM0" | TNM.7.i[patient]=="T4aN1bM0" |
-               TNM.7.i[patient]=="T1aN2bM0" | TNM.7.i[patient]=="T2aN2bM0" | # N2b
-               TNM.7.i[patient]=="T3aN2bM0" | TNM.7.i[patient]=="T4aN2bM0" |
-               TNM.7.i[patient]=="T1aN2cM0" | TNM.7.i[patient]=="T2aN2cM0" | # N2c
-               TNM.7.i[patient]=="T3aN2cM0" | TNM.7.i[patient]=="T4aN2cM0"){
-      AJCC.7.i[patient] <- 3.2
-    } else {
-      AJCC.7.i[patient] <- 3.3
-    }
-  }
-
-  # save TNM.7 and AJCC.7 for each imputated data set
-  TNM.7 <- cbind(TNM.7, TNM.7.i)
-  AJCC.7 <- cbind(AJCC.7, AJCC.7.i)
-
-  # compute C-index
-  rc <- Hmisc::rcorr.cens(-AJCC.7.i, S.Rec.5) # add a minus to revert class order: from worst to best
-  cindex.AJCC.7[i]<-rc["C Index"]
-  cindex.AJCC.7.se[i]<-rc["S.D."]/2
-
-  # compute C-index for only positive sentinel node patients
-  rc.pos <- Hmisc::rcorr.cens(-AJCC.7.i[imputation.i$SNstatus=="Positive"], S.Rec.5[imputation.i$SNstatus=="Positive",])
-  cindex.AJCC.7.pos[i]<-rc.pos["C Index"]
-  cindex.AJCC.7.se.pos[i]<-rc.pos["S.D."]/2
-
-  # compute C-index for only negative sentinel node patients
-  rc.neg <- Hmisc::rcorr.cens(-AJCC.7.i[imputation.i$SNstatus=="Negative"], S.Rec.5[imputation.i$SNstatus=="Negative",])
-  cindex.AJCC.7.neg[i]<-rc.neg["C Index"]
-  cindex.AJCC.7.se.neg[i]<-rc.neg["S.D."]/2
-}
-# combine C-index for each imputed data set
-cindex.mi.AJCC.7<-Rubin.combine(cindex.AJCC.7,cindex.AJCC.7.se)
-cindex.mi.AJCC.7.pos<-Rubin.combine(cindex.AJCC.7.pos,cindex.AJCC.7.se.pos)
-cindex.mi.AJCC.7.neg<-Rubin.combine(cindex.AJCC.7.neg,cindex.AJCC.7.se.neg)
-
-# save C-index and confidence interval
-utils::write.table(rbind(c("All patients:", sprintf('%.2f', cindex.mi.AJCC.7$est),
-              paste0("[", sprintf('%.2f', cindex.mi.AJCC.7$est+qnorm(.025)*cindex.mi.AJCC.7$se),
-                      "; ", sprintf('%.2f', cindex.mi.AJCC.7$est+qnorm(.975)*cindex.mi.AJCC.7$se), "]")),
-            c("Positive SN patients:", sprintf('%.2f', cindex.mi.AJCC.7.pos$est),
-                  paste0("[", sprintf('%.2f', cindex.mi.AJCC.7.pos$est+qnorm(.025)*cindex.mi.AJCC.7.pos$se),
-                         "; ", sprintf('%.2f', cindex.mi.AJCC.7.pos$est+qnorm(.975)*cindex.mi.AJCC.7.pos$se), "]")),
-            c("Negative SN patients:", sprintf('%.2f', cindex.mi.AJCC.7.neg$est),
-                  paste0("[", sprintf('%.2f', cindex.mi.AJCC.7.neg$est+qnorm(.025)*cindex.mi.AJCC.7.neg$se),
-                         "; ", sprintf('%.2f', cindex.mi.AJCC.7.neg$est+qnorm(.975)*cindex.mi.AJCC.7.neg$se), "]"))),
-            file='Z:/Project Melanoom/PaperMelanoma/Results/AJCC.7.Cindex.txt', 
-            row.names=FALSE, col.names=FALSE, sep=",")
-
-#####
-##### COMPARE MODEL WITH AJCC.8 CLASIFCATION
-#####
-TNM.8 <- c()
-AJCC.8 <- c()
-cindex.AJCC.8 <- rep(0, m)
-cindex.AJCC.8.se <- rep(0, m)
-cindex.AJCC.8.pos <- rep(0, m)
-cindex.AJCC.8.se.pos <- rep(0, m)
-cindex.AJCC.8.neg <- rep(0, m)
-cindex.AJCC.8.se.neg <- rep(0, m)
-for (i in 1:m){
-  imputation.i <- mice::complete(imputed.data,i)
-  TNM.8.i <- rep("", nrow(imputation.i))
-  AJCC.8.i <- rep(0, nrow(imputation.i))
-  for (patient in 1:nrow(imputation.i)){
-    breslow.patient <- imputation.i[patient, "Breslow"]
-    ulceration.patient <- as.character(imputation.i[patient, "Ulceration"])
-    mitosis.patient <- as.character(imputation.i[patient, "Mitosis"])
-    nodes.patient <- imputation.i[patient, "Tot_SNs_pos"]
-
-    # T stage
-    if (breslow.patient <= 1){
-      if (ulceration.patient=="No" & breslow.patient < 0.8){
-        TNM.8.i[patient] <- paste0(TNM.8.i[patient], "T1a")
-      } else {
-        TNM.8.i[patient] <- paste0(TNM.8.i[patient], "T1b")
-      }
-    } else if (breslow.patient>1 & breslow.patient <= 2){
-      if (ulceration.patient=="No"){
-        TNM.8.i[patient] <- paste0(TNM.8.i[patient], "T2a")
-      } else if (ulceration.patient=="Yes"){
-        TNM.8.i[patient] <- paste0(TNM.8.i[patient], "T2b")
-      }
-    } else if (breslow.patient>2 & breslow.patient <= 4){
-      if (ulceration.patient=="No"){
-        TNM.8.i[patient] <- paste0(TNM.8.i[patient], "T3a")
-      } else if (ulceration.patient=="Yes"){
-        TNM.8.i[patient] <- paste0(TNM.8.i[patient], "T3b")
-      }
-    } else if (breslow.patient>4){
-      if (ulceration.patient=="No"){
-        TNM.8.i[patient] <- paste0(TNM.8.i[patient], "T4a")
-      } else if (ulceration.patient=="Yes"){
-        TNM.8.i[patient] <- paste0(TNM.8.i[patient], "T4b")
-      }
-    }
-
-    # N stage
-    if (nodes.patient == 0){
-      TNM.8.i[patient] <- paste0(TNM.8.i[patient], "N0")
-    } else if (nodes.patient==1){
-      TNM.8.i[patient] <- paste0(TNM.8.i[patient], "N1a")
-    } else if (nodes.patient==2 | nodes.patient==3){
-      TNM.8.i[patient] <- paste0(TNM.8.i[patient], "N2a")
-    } else if (nodes.patient>3){
-      TNM.8.i[patient] <- paste0(TNM.8.i[patient], "N3a")
-    }
-
-    # M stage for all patients in our data set
-    TNM.8.i[patient] <- paste0(TNM.8.i[patient], "M0")
-
-    # determine AJCC.8.i
-    if (TNM.8.i[patient]=="T1aN0M0"){
-      AJCC.8.i[patient] <- 1.1
-    } else if (TNM.8.i[patient]=="T1bN0M0" | TNM.8.i[patient]=="T2aN0M0"){
-      AJCC.8.i[patient] <- 1.2
-    } else if (TNM.8.i[patient]=="T2bN0M0" | TNM.8.i[patient]=="T3aN0M0"){
-      AJCC.8.i[patient] <- 2.1
-    } else if (TNM.8.i[patient]=="T3bN0M0" | TNM.8.i[patient]=="T4aN0M0"){
-      AJCC.8.i[patient] <- 2.2
-    } else if (TNM.8.i[patient]=="T4bN0M0"){
-      AJCC.8.i[patient] <- 2.3
-    } else if (TNM.8.i[patient]=="T1aN1aM0" | TNM.8.i[patient]=="T1bN1aM0" | TNM.8.i[patient]=="T2aN1aM0"
-               | TNM.8.i[patient]=="T1aN2aM0" | TNM.8.i[patient]=="T1bN2aM0" | TNM.8.i[patient]=="T2aN2aM0"){
-      AJCC.8.i[patient] <- 3.1
-    } else if (TNM.8.i[patient]=="T2bN1aM0" | TNM.8.i[patient]=="T3aN1aM0"){
-      AJCC.8.i[patient] <- 3.2
-    } else if (TNM.8.i[patient]=="T4bN3aM0"){
-      AJCC.8.i[patient] <- 3.4
-    } else {
-      AJCC.8.i[patient] <- 3.3
-    }
-
-  }
-
-  # save TNM.8 and AJCC.8 for each imputated data set
-  TNM.8 <- cbind(TNM.8, TNM.8.i)
-  AJCC.8 <- cbind(AJCC.8, AJCC.8.i)
-
-  # compute C-index
-  rc <- Hmisc::rcorr.cens(-AJCC.8.i, S.Rec.5) # add a minus to revert class order: from worst to best
-  cindex.AJCC.8[i]<-rc["C Index"]
-  cindex.AJCC.8.se[i]<-rc["S.D."]/2
-
-  # compute C-index for only positive sentinel node patients
-  rc.pos <- Hmisc::rcorr.cens(-AJCC.8.i[imputation.i$SNstatus=="Positive"], S.Rec.5[imputation.i$SNstatus=="Positive",])
-  cindex.AJCC.8.pos[i]<-rc.pos["C Index"]
-  cindex.AJCC.8.se.pos[i]<-rc.pos["S.D."]/2
-
-  # compute C-index for only negative sentinel node patients
-  rc.neg <- Hmisc::rcorr.cens(-AJCC.8.i[imputation.i$SNstatus=="Negative"], S.Rec.5[imputation.i$SNstatus=="Negative",])
-  cindex.AJCC.8.neg[i]<-rc.neg["C Index"]
-  cindex.AJCC.8.se.neg[i]<-rc.neg["S.D."]/2
-}
-# combine C-index for each imputed data set
-cindex.mi.AJCC.8<-Rubin.combine(cindex.AJCC.8,cindex.AJCC.8.se)
-cindex.mi.AJCC.8.pos<-Rubin.combine(cindex.AJCC.8.pos,cindex.AJCC.8.se.pos)
-cindex.mi.AJCC.8.neg<-Rubin.combine(cindex.AJCC.8.neg,cindex.AJCC.8.se.neg)
-
-# save C-index and confidence interval
-utils::write.table(rbind(c("All patients:", sprintf('%.2f', cindex.mi.AJCC.8$est),
-                        paste0("[", sprintf('%.2f', cindex.mi.AJCC.8$est+qnorm(.025)*cindex.mi.AJCC.8$se),
-                               "; ", sprintf('%.2f', cindex.mi.AJCC.8$est+qnorm(.975)*cindex.mi.AJCC.8$se), "]")),
-                  c("Positive SN patients:", sprintf('%.2f', cindex.mi.AJCC.8.pos$est),
-                        paste0("[", sprintf('%.2f', cindex.mi.AJCC.8.pos$est+qnorm(.025)*cindex.mi.AJCC.8.pos$se),
-                               "; ", sprintf('%.2f', cindex.mi.AJCC.8.pos$est+qnorm(.975)*cindex.mi.AJCC.8.pos$se), "]")),
-                  c("Negative SN patients", sprintf('%.2f', cindex.mi.AJCC.8.neg$est),
-                        paste0("[", sprintf('%.2f', cindex.mi.AJCC.8.neg$est+qnorm(.025)*cindex.mi.AJCC.8.neg$se),
-                               "; ", sprintf('%.2f', cindex.mi.AJCC.8.neg$est+qnorm(.975)*cindex.mi.AJCC.8.neg$se), "]"))),
-            file='Z:/Project Melanoom/PaperMelanoma/Results/AJCC.8.Cindex.txt', 
-            row.names=FALSE, col.names=FALSE, sep=",")
-
-# table of number of AJCC.7 and AJCC.8
-table.AJCC.7 <- c()
-table.AJCC.8 <- c()
-for (i in 1:m){
-  table.AJCC.7 <- cbind(table.AJCC.7, as.numeric(table(AJCC.7[,i])))
-  table.AJCC.8 <- cbind(table.AJCC.8, as.numeric(table(AJCC.8[,i])))
-}
-table.AJCC.7.8 <- cbind(as.numeric(names(table(AJCC.8[,1]))),
-                        c(round(rowMeans(table.AJCC.7), 0), 0),
-                        round(rowMeans(table.AJCC.8), 0))
-utils::write.table(table.AJCC.7.8,
-            file='Z:/Project Melanoom/PaperMelanoma/Results/table.AJCC.7.and.8.txt', 
-            row.names=FALSE, col.names=FALSE, sep=",")
-table.AJCC.7.8
-```
-
-# Test non-lineairities
-### Test if Age should be modelled non-linearly
-```{r, eval=TRUE}
-form.linear.full.Rec <- S.Rec.5 ~
-  SNstatus +
-  Sex +
-  Age.SN +
-  Ulceration +
-  Loc_CAT +
-  Histology +
-  Breslow +
-  multiple.fields +
-  Tot_SNs_neg +
-  Tot_SNs_pos +
-  Rdamcrit +
-  Dewar +
-  SNstatus:Sex +
-  SNstatus:Age.SN +
-  SNstatus:Ulceration +
-  SNstatus:Loc_CAT +
-  SNstatus:Histology +
-  SNstatus:Breslow +
-  SNstatus:multiple.fields
-form.nonlinear.age.full.Rec <- S.Rec.5 ~
-  SNstatus +
-  Sex +
-  log(Age.SN) +
-  Ulceration +
-  Loc_CAT +
-  Histology +
-  Breslow +
-  multiple.fields +
-  Tot_SNs_neg +
-  Tot_SNs_pos +
-  Rdamcrit +
-  Dewar +
-  SNstatus:Sex +
-  SNstatus:log(Age.SN) +
-  SNstatus:Ulceration +
-  SNstatus:Loc_CAT +
-  SNstatus:Histology +
-  SNstatus:Breslow +
-  SNstatus:multiple.fields
-linear.full.Rec <- Hmisc::fit.mult.impute(form.linear.full.Rec, cph, xtrans=imputed.data,
-                                 data=data.to.be.imputed,n.impute=m, pr=FALSE,
-                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
-nonlinear.age.full.Rec <- Hmisc::fit.mult.impute(form.nonlinear.age.full.Rec, cph, xtrans=imputed.data,
-                                 data=data.to.be.imputed,n.impute=m, pr=FALSE,
-                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
-# distribution of Age
-hist(last.imputed.data.set$Age.SN)
-# linear model
-linear.full.Rec$loglik[2]
-# using log(Age) gives slightly better log likelihood
-nonlinear.age.full.Rec$loglik[2]
-```
-
-### Test if Breslow should be modelled non-linearly
-```{r, eval=TRUE}
-form.nonlinear.Breslow.full.Rec <- S.Rec.5 ~
-  SNstatus +
-  Sex +
-  Age.SN +
-  Ulceration +
-  Loc_CAT +
-  Histology +
-  log(Breslow) +
-  multiple.fields +
-  Tot_SNs_neg +
-  Tot_SNs_pos +
-  Rdamcrit +
-  Dewar +
-  SNstatus:Sex +
-  SNstatus:Age.SN +
-  SNstatus:Ulceration +
-  SNstatus:Loc_CAT +
-  SNstatus:Histology +
-  SNstatus:log(Breslow) +
-  SNstatus:multiple.fields
-nonlinear.Breslow.full.Rec <- Hmisc::fit.mult.impute(form.nonlinear.Breslow.full.Rec, cph, xtrans=imputed.data,
-                                 data=data.to.be.imputed,n.impute=m, pr=FALSE,
-                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
-# distribution of Breslow
-hist(last.imputed.data.set$Breslow)
-# linear model
-linear.full.Rec$loglik[2]
-# using log(Breslow) gives better log likelihood
-nonlinear.Breslow.full.Rec$loglik[2]
-```
-
-### Test if Rdamcrit should be modelled non-linearly
-```{r, eval=TRUE}
-form.nonlinear.Rdamcrit.full.Rec <- S.Rec.5 ~
-  SNstatus +
-  Sex +
-  Age.SN +
-  Ulceration +
-  Loc_CAT +
-  Histology +
-  Breslow +
-  multiple.fields +
-  Tot_SNs_neg +
-  Tot_SNs_pos +
-  log(Rdamcrit) +
-  Dewar +
-  SNstatus:Sex +
-  SNstatus:Age.SN +
-  SNstatus:Ulceration +
-  SNstatus:Loc_CAT +
-  SNstatus:Histology +
-  SNstatus:Breslow +
-  SNstatus:multiple.fields
-nonlinear.Rdamcrit.full.Rec <- Hmisc::fit.mult.impute(form.nonlinear.Rdamcrit.full.Rec, cph, xtrans=imputed.data,
-                                 data=data.to.be.imputed,n.impute=m, pr=FALSE,
-                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
-# distribution of Rdamcrit
-hist(last.imputed.data.set$Rdamcrit)
-# linear model
-linear.full.Rec$loglik[2]
-# using log(Rdamcrit) gives better log likelihood
-nonlinear.Rdamcrit.full.Rec$loglik[2]
-```
-
-# Full model for recurrence
-```{r, eval=TRUE}
-form.full.Rec.5 <- S.Rec.5 ~
-  SNstatus +
-  Sex +
-  Age.SN +
-  Ulceration +
-  Loc_CAT +
-  Histology +
-  log(Breslow) +
-  multiple.fields +
-  Tot_SNs_neg +
-  Tot_SNs_pos +
-  log(Rdamcrit) +
-  Dewar +
-  SNstatus:Sex +
-  SNstatus:Age.SN +
-  SNstatus:Ulceration +
-  SNstatus:Loc_CAT +
-  SNstatus:Histology +
-  SNstatus:log(Breslow) +
-  SNstatus:multiple.fields
-f.mi.full.Rec.5 <- Hmisc::fit.mult.impute(form.full.Rec.5, cph, xtrans=imputed.data,
-                                 data=data.to.be.imputed,n.impute=m, pr=FALSE,
-                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
-# stats::anova(f.mi.full.Rec.5)
-summary(f.mi.full.Rec.5, Tot_SNs_pos=c(0, 1))
-```
-
-# Final model for recurrence selected by backward selection with p = 0.01
-```{r, eval=TRUE}
-c.Breslow <- mean(log(last.imputed.data.set$Breslow))
-form.BS.Rec.5 <- S.Rec.5 ~
-  SNstatus +
-  Age.SN +
-  Ulceration +
-  Loc_CAT +
-  I(log(Breslow)-c.Breslow) +
-  log(Rdamcrit) +
-  SNstatus:I(log(Breslow)-c.Breslow)
-f.mi.BS.Rec.5 <- Hmisc::fit.mult.impute(form.BS.Rec.5,cph,xtrans=imputed.data,data=data.to.be.imputed,n.impute=m,
-                           pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
-summary(f.mi.BS.Rec.5, Breslow=exp(c.Breslow))
-coef(f.mi.BS.Rec.5)
-# round(stats::anova(f.mi.BS.Rec.5)[order(stats::anova(f.mi.BS.Rec.5)[, 'P']),], 3)
-stats::anova(f.mi.BS.Rec.5)
-```
-
-# Refitted full model for melanoma specific mortality
-```{r, eval=TRUE}
-form.full.MSM.5.refit <- S.MSM.5 ~
-  SNstatus +
-  Sex +
-  Age.SN +
-  Ulceration +
-  Loc_CAT +
-  Histology +
-  log(Breslow) +
-  multiple.fields +
-  Tot_SNs_neg +
-  Tot_SNs_pos +
-  log(Rdamcrit) +
-  Dewar +
-  SNstatus:Sex +
-  SNstatus:Age.SN +
-  SNstatus:Ulceration +
-  SNstatus:Loc_CAT +
-  SNstatus:Histology +
-  SNstatus:log(Breslow) +
-  SNstatus:multiple.fields
-f.mi.full.MSM.5.refit <- Hmisc::fit.mult.impute(form.full.MSM.5.refit,cph,xtrans=imputed.data,data=data.to.be.imputed,n.impute=m,
-                                 pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
-f.mi.full.MSM.5.refit
-summary(f.mi.full.MSM.5.refit, Tot_SNs_pos=c(0, 1))
-stats::anova(f.mi.full.MSM.5.refit)
-```
-
-# Refitted final model for melanoma specific mortality
-```{r, eval=TRUE}
-form.BS.MSM.5.refit <- S.MSM.5 ~
-  SNstatus +
-  Age.SN +
-  Ulceration +
-  Loc_CAT +
-  I(log(Breslow)-c.Breslow) +
-  log(Rdamcrit) +
-  SNstatus:I(log(Breslow)-c.Breslow)
-f.mi.BS.MSM.5.refit <- Hmisc::fit.mult.impute(form.BS.MSM.5.refit,cph,xtrans=imputed.data,data=data.to.be.imputed,n.impute=m,
-                                  pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
-# round(anova(f.mi.BS.MSM.5.refit)[order(anova(f.mi.BS.MSM.5.refit)[, 'P']),], 3)
-f.mi.BS.MSM.5.refit
-summary(f.mi.BS.MSM.5.refit, Breslow=exp(c.Breslow))
-stats::anova(f.mi.BS.MSM.5.refit)
-```
-
-# Calibrated final model for melanoma specific mortality
-Fit disease specific mortality to the linear predictor of recurrence
-```{r, eval=TRUE}
-lp.BS<-f.mi.BS.Rec.5$linear
-f.mi.BS.MSM.5<-rms::cph(S.MSM.5~lp.BS,y=TRUE,x=TRUE)
-MSM.cal.fact <- stats::coef(f.mi.BS.MSM.5)
-MSM.cal.fact
-stats::confint(f.mi.BS.MSM.5)
-utils::write.table(paste0(sprintf("%.2f", stats::coef(f.mi.BS.MSM.5)), " [",
-            sprintf("%.2f", as.numeric(stats::confint(f.mi.BS.MSM.5))[1]), "; ",
-            sprintf("%.2f", as.numeric(stats::confint(f.mi.BS.MSM.5))[2]), "]"),
-            file='Z:/Project Melanoom/PaperMelanoma/Results/calibration.MSS.txt', 
-            row.names=FALSE, col.names=FALSE, sep=",")
+# remotes::install_github("CHMMaas/PredictionTools")
+library(PredictionTools)  # to plot calibration
+# import helper functions
+source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")
 ```
@@
 ### Create variables with results
-##### Full model
+##### Put hazard ratio's and anova Chi-square statistics of positive and negative SN patients in Table
 ```{r, eval=TRUE}
+load('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData')
 for (model in c('Rec.5', 'MSM.5.refit')){
@@
                                        "Breslow",
-                        "multiple.fields",
+                                       "multiple.fields - Multiple SN fields:Other",
                                        "Tot_SNs_neg"), c("Effect", "Lower 0.95", "Upper 0.95")]),
@@
                                        "Breslow",
-                        "multiple.fields",
+                                       "multiple.fields - Multiple SN fields:Other",
                                        "Tot_SNs_neg",
@@
 ##### Final model
+##### Put hazard ratio's and anova Chi-square statistics of positive and negative SN patients in Table
 ```{r, eval=TRUE}
@@
 
-# Plot calibration of all centers combined and all centers seperate 
-for recurrence and melanoma specific mortality refit 
+# Calculate optimism using bootstrap
+```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
+set.seed(100)
+slope<-NULL
+cindex.orig<-NULL
+opt<-NULL
+cindex.B<-NULL
+for (i in 1:m){
+  v<-validate(f.mi.full.Rec.5$fits[[i]],method='boot',bw=TRUE,rule='p',sls=0.05,B=100,pr=FALSE,type='individual')
+  slope<-c(slope,v["Slope","test"]) # slope of test
+  cindex.orig<-c(cindex.orig,(v["Dxy","index.orig"]+1)/2) # original C-index
+  opt<-c(opt,v["Dxy","optimism"]/2) # optimism C-index
+  cindex.B<-c(cindex.B,(v["Dxy","index.corrected"]+1)/2) # index corrected C-index
+}
+```
+```{r, eval=TRUE}
+mean(cindex.orig)
+optimism <- mean(opt)
+optimism
+mean(cindex.B)
+mean(cindex.orig)-optimism
+```
+# Plot calibration of all centers combined and all centers seperate for recurrence and melanoma specific mortality refit 
+#### Add C-index to Recurrence and MSM refit Table
 ```{r, eval=TRUE}
@@
   for (type in c("full", "BS")){
+    # load model
     f.mi <- eval(parse(text=paste0('f.mi.', type, '.', model)))
+
+    # load outcome
     if (model=='Rec.5'){
-      S <- S.Rec.5
+      S <- S.Rec.5.training
     }
     else if (model=="MSM.refit"){
-      S <- S.MSM.5
+      S <- S.MSM.5.training
     }
-    p <- matrix(NA, nrow(sel.data), m)
+
+    # obtain predictions for each imputed data set
+    p <- matrix(NA, nrow(training.data.to.be.imputed), m)
     for (i in 1:m){
       # predicted lp for imputation i
-      lp.i <- predict(f.mi,newdata=mice::complete(imputed.data,i), type="lp")
+      lp.i <- predict(f.mi,newdata=mice::complete(imputed.training.data,i), type="lp")
       # baseline hazard
@@
 
-    # omit the observation with predicted probability = 1
-    omit <- unique(c(which(p[,1]==1), which(p[,2]==1), which(p[,3]==1), which(p[,4]==1), which(p[,5]==1)))
-    if (length(omit)>0){
-      cat('Omit:', omit, '\n')
-      p <- p[-omit,]
-      S <- S[-omit,]
-    }
-
-    out <- val.surv.mi(p,S,g=5,main="All centers",
-                       file_path=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/all.centers.", model, '.', type, ".png"),
-                       save_plots=TRUE,time=NULL,lim=c(0,1),dist=TRUE)
+    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/all.centers.", model, '.', type, ".png"))
+    out <- PredictionTools::val.surv.mi(p,S,g=5,main="All centers",
+                       time=NULL,lim=c(0,1),dist=TRUE)
+    grDevices::dev.off()
 
     final.full.table <- eval(parse(text=paste0('final.', type, '.table.', model)))
-    final.full.table["C-index", "HR.neg"] <- sprintf('%.2f', out$cindex)
-    final.full.table["C-index", "CI.neg"] <- paste0("[", sprintf('%.2f', out$cindex.lower),
-                                                   "; ", sprintf('%.2f', out$cindex.upper), "]")
+    final.full.table["C-index", "HR.neg"] <- sprintf('%.2f', out$cindex-optimism)
+    final.full.table["C-index", "CI.neg"] <- paste0("[", sprintf('%.2f', out$cindex.lower-optimism),
+                                                   "; ", sprintf('%.2f', out$cindex.upper-optimism), "]")
 
@@
 }
+```
 
-# MSM recalibration
+Add the C-index for MSM for model of recurrence
+```{r, eval=TRUE}
 for (type in c("full", "BS")){
@@
   for (i in 1:m){
-    rc <- Hmisc::rcorr.cens(-predict(f.mi,newdata=mice::complete(imputed.data,i), type="lp"), S.MSM.5)
+    rc <- Hmisc::rcorr.cens(-predict(f.mi,newdata=mice::complete(imputed.training.data,i), type="lp"), S.MSM.5.training)
     cindex.MSM.refit[i]<-rc["C Index"]
@@
   final.full.table <- eval(parse(text=paste0('final.', type, '.table.Rec.5')))
-  final.full.table["C-index MSM", "HR.neg"] <- sprintf('%.2f', cindex.mi.MSM.refit$est)
-  final.full.table["C-index MSM", "CI.neg"] <- paste0("[", sprintf('%.2f', cindex.mi.MSM.refit$est+qnorm(.025)*cindex.mi.MSM.refit$se),
-                                                 "; ", sprintf('%.2f', cindex.mi.MSM.refit$est+qnorm(.975)*cindex.mi.MSM.refit$se), "]")
+  final.full.table["C-index MSM", "HR.neg"] <- sprintf('%.2f', cindex.mi.MSM.refit$est-optimism)
+  final.full.table["C-index MSM", "CI.neg"] <- paste0("[", sprintf('%.2f', cindex.mi.MSM.refit$est+qnorm(.025)*cindex.mi.MSM.refit$se-optimism),
+                                                 "; ", sprintf('%.2f', cindex.mi.MSM.refit$est+qnorm(.975)*cindex.mi.MSM.refit$se-optimism), "]")
 
@@
 }
+```
 
+# Save dataframes of result Tables
+```{r, eval=TRUE}
 # replace NA by blank
@@
 
-# Cross-validation recurrence 5-years
+# C-index stratified by SN status
 ```{r, eval=TRUE}
-Centers <- sort(as.character(unique(sel.data$Center)))
-Centers.anonimized <- c("A", "B", "C", "D")
-cuts <- 5
-for (model in c('Rec.5', 'MSM.5')){
-  if (model == 'Rec.5'){
-    S.model <- S.Rec.5
+for (outcome in c("Rec.5", "MSM.5.refit")){
+  for (arm in c("Positive", "Negative")){
+    f.mi.stratified <- eval(parse(text=paste0('f.mi.BS.', outcome)))
+    if (outcome=="Rec.5"){
+      S.5 <- S.Rec.5.training
+    } else if (outcome=="MSM.5.refit"){
+      S.5 <- S.MSM.5.training
     }
-  if (model == 'MSM.5'){
-    S.model <- S.MSM.5
+    S.5.subset <- S.5[last.imputed.training.data.set$SNstatus==arm,]
+
+    cindex.stratified <- rep(0, m)
+    cindex.se.stratified <- rep(0, m)
+    for (i in 1:m){
+      subset.imputed.training.data <- as.data.frame(mice::complete(imputed.training.data,i)[last.imputed.training.data.set$SNstatus==arm,])
+      rc <- Hmisc::rcorr.cens(-predict(f.mi.stratified,newdata=subset.imputed.training.data, type="lp"), S.5.subset)
+      cindex.stratified[i]<-rc["C Index"]
+      cindex.se.stratified[i]<-rc["S.D."]/2
     }
+    cindex.mi.stratified<-Rubin.combine(cindex.stratified,cindex.se.stratified)
 
-  for (type in c('full', 'BS')){
-    for (j in 1:4){
-      S.j<-S.model[data.to.be.imputed$Center==Centers[j],]
-      S.notj<-S.model[data.to.be.imputed$Center!=Centers[j],]
-      S.Rec.5.notj<-S.Rec.5[data.to.be.imputed$Center!=Centers[j],]
-
-      form.model<-eval(parse(text=paste0('form.', type, '.Rec.5')))
-      form.notj<-update(form.model, S.Rec.5.notj  ~ . )
-      f.mi.Rec.notj<-Hmisc::fit.mult.impute(form.notj,cph,xtrans=imputed.data,data=data.to.be.imputed,
-                                 n.impute=m,pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE,
-                                 sub=data.to.be.imputed$Center!=Centers[j])
-
-      if (model=='Rec.5'){
-        f.basehaz.notj<-survival::basehaz(f.mi.Rec.notj,TRUE)
-        h0.notj<-f.basehaz.notj$hazard[f.basehaz.notj$time==max(f.basehaz.notj$time[f.basehaz.notj$time<=horizon])]
-      } else if (model=='MSM.5'){
-        lp.notj<-f.mi.Rec.notj$linear
-        f.MSM.notj<-rms::cph(S.notj~lp.notj,y=TRUE,x=TRUE)
-
-        f.basehaz.notj<-basehaz(f.MSM.notj,TRUE)
-        h0.notj<-f.basehaz.notj$hazard[f.basehaz.notj$time==max(f.basehaz.notj$time[f.basehaz.notj$time<=horizon])]
-      }
-
-      for (select.value in c("Positive", "Negative", "All")){
-        # identify those patients that are selected
-        if (select.value == "All"){
-          select.patients <- TRUE
-        } else if (select.value != "All"){
-          select.patients <- data.to.be.imputed[data.to.be.imputed$Center==Centers[j],]$SNstatus==select.value
-        }
-
-        # calibration with metrics
-        S <- S.j[select.patients,]
-        p <- matrix(NA, nrow(S), m)
-        for (i in 1:m){
-          lp.i<-predict(f.mi.Rec.notj,newdata=mice::complete(imputed.data,i)[data.to.be.imputed$Center==Centers[j],], type="lp")
-
-          if (model=='MSM.5'){
-            lp.i<-predict(f.MSM.notj,newdata=lp.i)
-          }
-
-          if (select.value != "All"){
-            lp.i <- lp.i[select.patients]
-          }
-
-          # baseline hazard
-          if (model=='Rec.5'){
-            f.basehaz<-basehaz(f.mi.Rec.notj,TRUE)
-          }
-          else if (model=='MSM.5'){
-            f.basehaz<-basehaz(f.MSM.notj,TRUE)
-          }
-          h0.i <- f.basehaz$hazard[f.basehaz$time==max(f.basehaz$time[f.basehaz$time<=horizon])]
-
-          # probabilities
-          p.i <- fun.event(lp=lp.i, h0=h0.i)
-          p[, i] <- p.i
-        }
-
-        # omit the observation with predicted probability = 1
-        omit <- unique(c(which(p[,1]==1), which(p[,2]==1), which(p[,3]==1), which(p[,4]==1), which(p[,5]==1)))
-        if (length(omit)>0){
-          cat('Omit:', omit, '\n')
-          p <- p[-omit,]
-          S <- S[-omit,]
-        }
-
-        out <- val.surv.mi(p, S, g=5, main=paste("Center", Centers.anonimized[j]),
-                           file_path=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", model, '.', select.value, '.', type, '.', Centers[j], ".png"),
-                           save_plots=TRUE,time=NULL,lim=c(0,1),dist=TRUE)
-      }
-    }
+    print(c(outcome, arm, sprintf('%.2f', cindex.mi.stratified$est),
+    paste0("[", sprintf('%.2f', cindex.mi.stratified$est+qnorm(.025)*cindex.mi.stratified$se), 
+           "; ", sprintf('%.2f', cindex.mi.stratified$est+qnorm(.975)*cindex.mi.stratified$se), "]")))
   }
@@
 
-# combine calibration plots into one
-for (select.value in c("All", "Positive", "Negative")){
-  for (y in c('Rec.5', 'MSM.5')){
-    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/", y, ".", select.value, ".png"), width=16, height=16, units="cm", res=300)
-    par(mar=rep(0, 4))
-    layout(matrix(1:4, ncol=2, byrow=TRUE))
-    Centers.val <- rep(1:4, 3)
-    for (i in 1:4){
-      plot(NA, xlim=0:1, ylim=0:1, xaxt="n", yaxt="n", bty="n")
-      img <- readPNG(paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", y, ".", select.value, ".BS.", Centers[Centers.val[i]], ".png"))
-      rasterImage(img, 0, 0, 1, 1)
-    }
-    grDevices::dev.off()
-  }
-}
 ```
-
-# Nomogram
-```{r, eval=TRUE}
-# Limits of Age, Breslow, Rdamcrit
-Age.class <- seq(20, 80, 20)
-Breslow.class <- c(seq(0.1, 0.9, 0.2), 1:7)
-Rdamcrit.class <- c(seq(1, 10, 2)/100, seq(1, 10, 2)/10, seq(1, 10, 2))
-
-# initial nomogram
-nom.0 <- nomogram(f.mi.BS.Rec.5,maxscale=10,
-                Breslow=Breslow.class, Rdamcrit=Rdamcrit.class, Age.SN=Age.class)
-
-# baseline hazard of recurrence
-f.Rec.basehaz <- basehaz(f.mi.BS.Rec.5,TRUE)
-h0.Rec <- f.Rec.basehaz$hazard[f.Rec.basehaz$time==max(f.Rec.basehaz$time[f.Rec.basehaz$time<=horizon])]
-
-# lp range Recurrence
-rc.Rec <- (1/attributes(nom.0)$info$sc)
-int.Rec <- attributes(nom.0)$info$Intercept
-lp.range <- int.Rec+rc.Rec*(0:10)*2
-
-# nomogram
-nom <- nomogram(f.mi.BS.Rec.5, maxscale=10,
-              Breslow=Breslow.class, Rdamcrit=Rdamcrit.class, Age.SN=Age.class,
-              lp=TRUE, fun.at=fun.event(lp=lp.range, h0=h0.Rec))
-attributes(nom)$names[attributes(nom)$names=="Age.SN"]<-"Age"
-attributes(nom)$names[attributes(nom)$names=="Loc_CAT"]<-"Location"
-attributes(nom)$names[attributes(nom)$names=="Rdamcrit"]<-"Tumor burden"
-attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Positive)"]<-"Breslow + positive SN"
-attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Negative)"]<-"Breslow + negative SN"
-par(mar = rep(0, 4))
-plot(nom, total.sep.page=TRUE, col.grid=gray(c(0.8, 0.95)))
-nom.0
-nom
-```
-
-# Risk score distribution
-```{r, eval=TRUE}
-# baseline hazard of MSM
-f.MSM.basehaz <- basehaz(f.mi.BS.MSM.5,TRUE)
-h0.MSM <- f.MSM.basehaz$hazard[f.MSM.basehaz$time==max(f.MSM.basehaz$time[f.MSM.basehaz$time<=horizon])]
-lp.MSM <- f.mi.BS.MSM.5$linear
-
-# lp range MSM
-rc.MSM <- f.mi.BS.MSM.5$coef[1]*rc.Rec
-int.MSM <- f.mi.BS.MSM.5$coef[1]*int.Rec
-max.range <- 16
-points.range <- 0:max.range
-lp.points.Rec.5 <- int.Rec+rc.Rec*points.range
-lp.points.MSM.5 <- int.MSM+rc.MSM*points.range
-
-# Recurrence lp scores
-Breslow.trunc <- last.imputed.data.set$Breslow
-Breslow.trunc[last.imputed.data.set$Breslow<0.1] <- .1
-Breslow.trunc[last.imputed.data.set$Breslow>7] <- 7
-lp.score.Rec <- predict(f.mi.BS.Rec.5,newdata=data.frame(SNstatus=last.imputed.data.set$SNstatus,
-                                                Age.SN=last.imputed.data.set$Age.SN,
-                                                Ulceration=last.imputed.data.set$Ulceration,
-                                                Loc_CAT=last.imputed.data.set$Loc_CAT,
-                                                Breslow=Breslow.trunc,
-                                                Rdamcrit=last.imputed.data.set$Rdamcrit))
-mean(lp.score.Rec)
-score.Rec <- round((lp.score.Rec-int.Rec)/rc.Rec,0)
-max.breaks <- max.range+1
-h.Rec <- hist(score.Rec,plot=FALSE,breaks=0:max.breaks,right=FALSE)
-
-# score table
-data.frame(table(score.Rec)/length(score.Rec))
-```
-
-# Plot of risk score distribution
-```{r, eval=TRUE, message=FALSE}
-h.plot <- h.Rec
-h.plot$density <- 100*h.Rec$density
-p.Rec <- fun.event(lp=lp.points.Rec.5, h0=h0.Rec)
-p.MSM <- fun.event(lp=lp.points.MSM.5, h0=h0.MSM)
-x.lim <- c(0, max.range+1)
-y.lim <- c(0, 100)
-y.lim.hist <- c(0, 25)
-png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Risk.distribution.png"))
-par(mar = c(5,5,2,5))
-plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col='white')
-abline(h=(0:30)*1,col="light grey",lty=2)
-axis(side = 4)
-mtext(side = 4, line = 3, 'Risk score distribution (%)')
-par(new = TRUE)
-plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col="white")
-par(new = TRUE)
-plot(points.range,p.Rec*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=15, xlab="Risk score",ylab="Predicted 5-year risk (%)")
-points(points.range,p.MSM*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=16)
-legend(x=0,y=100,legend=c("Recurrence","Melanoma-specific mortality"),pch=c(15,16),cex=.9,bg='white')
-grDevices::dev.off()
-```
-
-```{r, eval=TRUE, echo=FALSE}
-par(mar = c(5,5,2,5))
-plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col='white')
-abline(h=(0:30)*1,col="light grey",lty=2)
-axis(side = 4)
-mtext(side = 4, line = 3, 'Risk score distribution (%)')
-par(new = TRUE)
-plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col="white")
-par(new = TRUE)
-plot(points.range,p.Rec*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=15, xlab="Risk score",ylab="Predicted 5-year risk (%)")
-points(points.range,p.MSM*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=16)
-legend(x=0,y=100,legend=c("Recurrence","Melanoma-specific mortality"),pch=c(15,16),cex=.9,bg='white')
-```
-
-```{r, eval=TRUE}
-# reset margins
-par(mar = c(5,5,2,5))
-
-# table of risk distribution
-cbind(points.range, round(p.Rec*100, 1), round(p.MSM*100, 1))[c(1, 9, 10, 12, 13),]
-
-# save for Rshiny app
-coef.Rec <- stats::coef(f.mi.BS.Rec.5)
-center.Rec <- f.mi.BS.Rec.5$center
-save(h0.Rec, h0.MSM, coef.Rec, c.Breslow, center.Rec,
-     file=paste0('Z:/Project Melanoom/PaperMelanoma/Results/h0.Rec.MSM.Rdata'), compress=TRUE)
-load(file=paste0('Z:/Project Melanoom/PaperMelanoma/Results/h0.Rec.MSM.Rdata'))
-```
-
-# Test if model is correct
-### Check calibration
-Intercept of recurrence should be the same as when setting covariates to 'zero' values minus the center
-```{r, eval=TRUE}
-int.Rec
-20*f.mi.BS.Rec.5$coef["Age.SN"]+
-  (log(0.1)-c.Breslow)*f.mi.BS.Rec.5$coef["Breslow"]+
-  log(0.01)*f.mi.BS.Rec.5$coef["Rdamcrit"]-
-  f.mi.BS.Rec.5$center
-```
-
-### Check mean event rates of risk distribution
-#### Recurrence
-```{r, eval=TRUE}
-sum(h.Rec$density*fun.event(lp=lp.points.Rec.5, h0=h0.Rec))
-# equal to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
-SF<-survfit(S.Rec.5~1,data=sel.data)
-100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
-# average probability of recurrence
-mean(fun.event(lp=lp.score.Rec, h0=h0.Rec))
-```
-
-#### MSM
-```{r, eval=TRUE}
-sum(h.Rec$density*fun.event(lp=lp.points.MSM.5, h0=h0.MSM))
-# equal to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
-SF<-survfit(S.MSM.5~1,data=sel.data)
-100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
-# average probability of MSM
-mean(fun.event(lp=lp.MSM, h0=h0.MSM))
-```
-
-# Check if predictions of new and old model align for negative SN patients
-## New model
-```{r, eval=TRUE}
-lp.negative <- predict(f.mi.BS.Rec.5,
-                       newdata=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative",],
-                       type="lp")
-new.negativeSN.tool <- 1-exp(-h0.Rec*exp(lp.negative))
-```
-
-## Old model with new data
-```{r, eval=TRUE}
-rscript.negative <- function(breslow, ulceration, location) {
-  S0 <- 0.85837329
-  lp <- 0.79351989*log(breslow) +
-    0.61912477*I(ulceration == 1) +
-    0.29594843*I(location == 1) +
-    0.42802046*I(location == 2) +
-    0.86053044*I(location == 3) - 0.98860941
-  S <- 100 - 100*S0^exp(lp)
-  return(S)
-}
-old.negativeSN.tool <- rscript.negative(breslow=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Breslow"],
-        ulceration=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Ulceration"])-1,
-        location=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Loc_CAT"])-1)
-```
-
-## Compare
-```{r, eval=TRUE}
-plot(x=new.negativeSN.tool*100, y=old.negativeSN.tool,
-     xlim=c(0, 100), ylim=c(0, 100),
-     xlab="New predictions", ylab="Old predictions",
-     main="Predicted recurrence of patients with negative SN status")
-abline(a=0, b=1, col="red")
-```
-
-## Refitted old model new data
-```{r, eval=TRUE}
-refitted.old.negative <- rms::cph(S.Rec.5 ~ log(Breslow)+Ulceration+Loc_CAT,
-                     data=last.imputed.data.set,
-                     y=TRUE, x=TRUE,
-                     subset=last.imputed.data.set$SNstatus=="Negative")
-refitted.old.negative
-f.Rec.basehaz.neg <- basehaz(refitted.old.negative,TRUE)
-h0.neg <- f.Rec.basehaz.neg$hazard[f.Rec.basehaz.neg$time==max(f.Rec.basehaz.neg$time[f.Rec.basehaz.neg$time<=horizon])]
-refit.old.negativeSN.tool <- fun.event(predict(refitted.old.negative),  h0.neg)
-```
-
-## Compare
-```{r, eval=TRUE}
-plot(x=new.negativeSN.tool*100, y=refit.old.negativeSN.tool*100,
-     xlim=c(0, 100), ylim=c(0, 100),
-     xlab="New predictions", ylab="Old predictions",
-     main="Predicted recurrence of patients with negative SN status")
-abline(a=0, b=1, col="red")
-```
-
-## Check mean recurrence probabilities
-```{r, eval=TRUE}
-SF<-survfit(S.Rec.5[last.imputed.data.set$SNstatus=="Negative"]~1)
-100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
-mean(new.negativeSN.tool)
-mean(refit.old.negativeSN.tool)
-mean(old.negativeSN.tool)
-```
-
-# Check if predictions of new and old model align for positive SN patients
-## New model
-```{r, eval=TRUE}
-lp.positive <- predict(f.mi.BS.Rec.5,
-                       newdata=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive",],
-                       type="lp")
-new.positiveSN.tool <- 1-exp(-h0.Rec*exp(lp.positive))
-```
-
-## Old model with new data
-```{r, eval=TRUE}
-rscript.positive <- function(age, ulceration, tumor_burden, breslow) {
-  S0 <- 0.51762362
-  lp <- 0.56109945 * log(age) +
-    0.23383770 * log(tumor_burden) +
-    0.35489235 * log(breslow) +
-    0.36562098 * I(ulceration == 1) - 2.709225
-  S <- 100 - 100 * S0^exp(lp)
-  return(S)
-}
-old.positiveSN.tool <- rscript.positive(age=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Age.SN"],
-                                        ulceration=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Ulceration"])-1,
-                                        breslow=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Breslow"],
-                                        tumor_burden=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Rdamcrit"]))
-```
-
-## Compare
-```{r, eval=TRUE}
-plot(x=new.positiveSN.tool*100, y=old.positiveSN.tool,
-     xlim=c(0, 100), ylim=c(0, 100),
-     xlab="New predictions", ylab="Old predictions",
-     main="Predicted recurrence of patients with positive SN status")
-abline(a=0, b=1, col="red")
-```
-
-## Refitted old model new data
-```{r, eval=TRUE}
-refitted.old.positive <- rms::cph(S.Rec.5 ~ log(Breslow)+Ulceration+Loc_CAT, 
-                      data=last.imputed.data.set,
-                      y=TRUE, x=TRUE,
-                      subset=last.imputed.data.set$SNstatus=="Positive")
-refitted.old.positive
-f.Rec.basehaz.pos <- basehaz(refitted.old.positive,TRUE)
-h0.pos <- f.Rec.basehaz.pos$hazard[f.Rec.basehaz.pos$time==max(f.Rec.basehaz.pos$time[f.Rec.basehaz.pos$time<=horizon])]
-refit.old.positiveSN.tool <- fun.event(predict(refitted.old.positive),  h0.pos)
-```
-
-## Compare
-```{r, eval=TRUE}
-plot(x=new.positiveSN.tool*100, y=refit.old.positiveSN.tool*100,
-     xlim=c(0, 100), ylim=c(0, 100),
-     xlab="New predictions", ylab="Old predictions",
-     main="Predicted recurrence of patients with positive SN status")
-abline(a=0, b=1, col="red")
-```
-
-## Check mean recurrence probabilities
-```{r, eval=TRUE}
-SF<-survfit(S.Rec.5[last.imputed.data.set$SNstatus=="Positive"]~1)
-100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
-mean(new.positiveSN.tool)
-mean(refit.old.positiveSN.tool)
-mean(old.positiveSN.tool)
-```
-
-#### Example to check probabilities with app
-```{r, eval=TRUE}
-patient.nr <- 115 # 395 and 396 = Negative, 397 and 398 = Positive, 30 = Head & neck, 115 has large Breslow
-data.to.be.imputed[patient.nr, c("SNstatus", "Age.SN", "Ulceration", "Loc_CAT", "Breslow", "Rdamcrit")]
-lp.score.Rec.true <- predict(f.mi.BS.Rec.5)[patient.nr]
-lp.score.Rec.try <- predict(f.mi.BS.Rec.5, newdata=data.frame(SNstatus=data.to.be.imputed[patient.nr,"SNstatus"],
-                                                            Age.SN=data.to.be.imputed[patient.nr,"Age.SN"],
-                                                            Ulceration=data.to.be.imputed[patient.nr,"Ulceration"],
-                                                            Loc_CAT=data.to.be.imputed[patient.nr,"Loc_CAT"],
-                                                            Breslow=data.to.be.imputed[patient.nr,"Breslow"],
-                                                            Rdamcrit=data.to.be.imputed[patient.nr,"Rdamcrit"]),
-                                                            type="lp")
-p.Rec.example <- fun.event(lp=lp.score.Rec.try, h0=h0.Rec)
-p.MSM.example <- fun.event(lp=lp.score.Rec.try, h0=h0.MSM)
-
-# manual calculation
-lp.score.Rec.manual <- coef.Rec["SNstatus=Positive"]*(as.numeric(data.to.be.imputed[patient.nr,"SNstatus"])-1)+
-              coef.Rec["Age.SN"]*data.to.be.imputed[patient.nr,"Age.SN"]+
-              coef.Rec["Ulceration=Yes"]*(as.numeric(data.to.be.imputed[patient.nr,"Ulceration"])-1)+
-              coef.Rec["Loc_CAT=leg"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="leg")+
-              coef.Rec["Loc_CAT=trunk"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="trunk")+
-              coef.Rec["Loc_CAT=head & neck"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="head & neck")+
-              coef.Rec["Breslow"]*(log(data.to.be.imputed[patient.nr,"Breslow"])-c.Breslow)+
-              coef.Rec["Rdamcrit"]*log(data.to.be.imputed[patient.nr,"Rdamcrit"])+
-              coef.Rec["SNstatus=Positive * Breslow"]*(log(data.to.be.imputed[patient.nr,"Breslow"])-c.Breslow)*(as.numeric(data.to.be.imputed[patient.nr,"SNstatus"])-1)
-
-# lp check
-lp.score.Rec.true
-lp.score.Rec.try
-lp.score.Rec.manual-center.Rec
-
-# probability of recurrence with fun.event
-fun.event(lp=lp.score.Rec.try, h0=h0.Rec)
-# probability of recurrence manually
-1-exp(-h0.Rec*exp(lp.score.Rec.manual-center.Rec))
-
-# check calculation of lp of MSM
-lp.score.MSM.true <- predict(f.mi.BS.MSM.5)[patient.nr]
-lp.score.MSM.true
-lp.MSM.try <- MSM.cal.fact*(lp.score.Rec.manual-center.Rec)
-lp.MSM.try
-
-# probability of MSM with fun.event
-fun.event(lp=lp.MSM.try, h0=h0.MSM)
-# probability of MSM manually
-1-exp(-h0.MSM*exp(MSM.cal.fact*(lp.score.Rec.manual-center.Rec)))
-```
-
-### Histogram for negative SN status patients
-```{r, eval=TRUE}
-Breslow.trunc.neg <- last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Breslow"]
-Breslow.trunc.neg[last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Breslow"]<0.1] <- .1
-Breslow.trunc.neg[last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Breslow"]>7] <- 7
-lp.score.Rec.negative <- predict(f.mi.BS.Rec.5,newdata=data.frame(SNstatus=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "SNstatus"],
-                                                       Age.SN=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Age.SN"],
-                                                       Ulceration=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Ulceration"],
-                                                       Loc_CAT=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Loc_CAT"],
-                                                       Breslow=Breslow.trunc.neg,
-                                                       Rdamcrit=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Rdamcrit"]))
-mean(lp.score.Rec.negative)
-# OK to use int.Rec and rc.Rec based on all patients?
-score.Rec.neg <- round((lp.score.Rec.negative-int.Rec)/rc.Rec,0)
-h.Rec.neg <- hist(score.Rec.neg,plot=FALSE,breaks=0:max.breaks,right=FALSE)
-h.plot.neg <- h.Rec.neg
-h.plot.neg$density <- 100*h.Rec.neg$density
-```
-
-### Histogram for positive SN status patients
-```{r, eval=TRUE}
-Breslow.trunc.pos <- last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Breslow"]
-Breslow.trunc.pos[last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Breslow"]<0.1] <- .1
-Breslow.trunc.pos[last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Breslow"]>7] <- 7
-lp.score.Rec.positive <- predict(f.mi.BS.Rec.5,newdata=data.frame(SNstatus=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "SNstatus"],
-                                                                Age.SN=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Age.SN"],
-                                                                Ulceration=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Ulceration"],
-                                                                Loc_CAT=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Loc_CAT"],
-                                                                Breslow=Breslow.trunc.pos,
-                                                                Rdamcrit=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Rdamcrit"]))
-mean(lp.score.Rec.positive)
-# OK to use int.Rec and rc.Rec based on all patients?
-score.Rec.pos <- round((lp.score.Rec.positive-int.Rec)/rc.Rec,0)
-h.Rec.pos <- hist(score.Rec.pos,plot=FALSE,breaks=0:max.breaks,right=FALSE)
-h.plot.pos <- h.Rec.pos
-h.plot.pos$density <- 100*h.Rec.pos$density
-
-par(mfrow=c(1, 3))
-plot(h.plot,freq=FALSE,axes=TRUE, xlab="Risk score", xlim=x.lim,ylim=y.lim.hist,
-     ylab="Risk score distribution (%)", main="All patients",col="white")
-plot(h.plot.neg,freq=FALSE,axes=TRUE, xlab="Risk score", xlim=x.lim,
-     ylab="Risk score distribution (%)",ylim=y.lim.hist,main="Negative SN status",col="white")
-plot(h.plot.pos,freq=FALSE,axes=TRUE, xlab="Risk score", xlim=x.lim,
-     ylab="Risk score distribution (%)",ylim=y.lim.hist,main="Positive SN status",col="white")
-```
-
-### Test proportional hazards assumption
-```{r, eval=TRUE}
-test.prop<-survival::cox.zph(f.mi.full.Rec.5$fits[[1]])
-test.prop
-plot(test.prop)
-```

### 2022-11-29 Code/8_CreateCrossValidatedCalibrationPlots.Rmd
Validation data added

---
title: "Create calibration plots"
author: "Carolien C.H.M. Maas"
date: "14-11-2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Create calibration plots

### Load packages, functions, and data 
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# erase memory
rm(list = ls(all.names = TRUE))

# for reproducibility
set.seed(100)

# load libraries
library(Hmisc)
library(survival)
library(rms)
library(mice)
library(grDevices)
source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")
# remotes::install_github("CHMMaas/PredictionTools")
library(PredictionTools)  # to plot calibration
```

# Cross-validation recurrence 5-years
```{r, eval=TRUE}
load('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData')
Centers <- sort(as.character(unique(training.data.to.be.imputed$Center)))
Centers.anonimized <- c("A", "B", "C", "D")
cuts <- 5
for (model in c('Rec.5', 'MSM.5')){
  if (model == 'Rec.5'){
    S.model <- S.Rec.5.training
  }
  if (model == 'MSM.5'){
    S.model <- S.MSM.5.training
  }

  for (type in c('full', 'BS')){
    for (j in 1:4){
      S.j<-S.model[training.data.to.be.imputed$Center==Centers[j],]
      S.notj<-S.model[training.data.to.be.imputed$Center!=Centers[j],]
      S.Rec.5.notj<-S.Rec.5.training[training.data.to.be.imputed$Center!=Centers[j],]

      form.model<-eval(parse(text=paste0('form.', type, '.Rec.5')))
      form.notj<-update(form.model, S.Rec.5.notj  ~ . )
      f.mi.Rec.notj<-Hmisc::fit.mult.impute(form.notj,cph,xtrans=imputed.training.data,data=training.data.to.be.imputed,
                                            n.impute=m,pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE,
                                            sub=training.data.to.be.imputed$Center!=Centers[j])

      if (model=='Rec.5'){
        f.basehaz.notj<-survival::basehaz(f.mi.Rec.notj,TRUE)
        h0.notj<-f.basehaz.notj$hazard[f.basehaz.notj$time==max(f.basehaz.notj$time[f.basehaz.notj$time<=horizon])]
      } else if (model=='MSM.5'){
        lp.notj<-f.mi.Rec.notj$linear
        f.MSM.notj<-rms::cph(S.notj~lp.notj,y=TRUE,x=TRUE)

        f.basehaz.notj<-basehaz(f.MSM.notj,TRUE)
        h0.notj<-f.basehaz.notj$hazard[f.basehaz.notj$time==max(f.basehaz.notj$time[f.basehaz.notj$time<=horizon])]
      }

      for (select.value in c("Positive", "Negative", "All")){
        # identify those patients that are selected
        if (select.value == "All"){
          select.patients <- TRUE
        } else if (select.value != "All"){
          select.patients <- training.data.to.be.imputed[training.data.to.be.imputed$Center==Centers[j],]$SNstatus==select.value
        }

        # calibration with metrics
        S <- S.j[select.patients,]
        p <- matrix(NA, nrow(S), m)
        for (i in 1:m){
          lp.i<-predict(f.mi.Rec.notj,newdata=mice::complete(imputed.training.data,i)[training.data.to.be.imputed$Center==Centers[j],], type="lp")

          if (model=='MSM.5'){
            lp.i<-predict(f.MSM.notj,newdata=lp.i)
          }

          if (select.value != "All"){
            lp.i <- lp.i[select.patients]
          }

          # baseline hazard
          if (model=='Rec.5'){
            f.basehaz<-basehaz(f.mi.Rec.notj,TRUE)
          }
          else if (model=='MSM.5'){
            f.basehaz<-basehaz(f.MSM.notj,TRUE)
          }
          h0.i <- f.basehaz$hazard[f.basehaz$time==max(f.basehaz$time[f.basehaz$time<=horizon])]

          # probabilities
          p.i <- fun.event(lp=lp.i, h0=h0.i)
          p[, i] <- p.i
        }

        # omit the observation with predicted probability = 1
        omit <- unique(c(which(p[,1]==1), which(p[,2]==1), which(p[,3]==1), which(p[,4]==1), which(p[,5]==1)))
        if (length(omit)>0){
          cat('Omit:', omit, '\n')
          p <- p[-omit,]
          S <- S[-omit,]
        }

        png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", model, '.', select.value, '.', type, '.', Centers[j], ".png"))
        out <- PredictionTools::val.surv.mi(p, S, g=5, main=paste("Center", Centers.anonimized[j]),
                                            time=NULL,lim=c(0,1),dist=TRUE)
        grDevices::dev.off()
      }
    }
  }
}

# combine calibration plots into one
for (select.value in c("All", "Positive", "Negative")){
  for (y in c('Rec.5', 'MSM.5')){
    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/", y, ".", select.value, ".png"), width=16, height=16, units="cm", res=300)
    par(mar=rep(0, 4))
    layout(matrix(1:4, ncol=2, byrow=TRUE))
    Centers.val <- rep(1:4, 3)
    for (i in 1:4){
      plot(NA, xlim=0:1, ylim=0:1, xaxt="n", yaxt="n", bty="n")
      img <- png::readPNG(paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", y, ".", select.value, ".BS.", Centers[Centers.val[i]], ".png"))
      rasterImage(img, 0, 0, 1, 1)
    }
    grDevices::dev.off()
  }
}
```

### 2022-11-29 Code/9_CreateNomogram.Rmd
Validation data added

---
title: "Create Nomogram"
author: "Carolien C.H.M. Maas"
date: "14-11-2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Create Nomogram

### Load packages, functions, and data 
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# erase memory
rm(list = ls(all.names = TRUE))

# for reproducibility
set.seed(100)

# load libraries
library(rms)
library(survival)
source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")
```

# Nomogram
```{r, eval=TRUE}
load('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData')
# Limits of Age, Breslow, Rdamcrit
Age.class <- seq(20, 80, 20)
Breslow.class <- c(seq(0.1, 0.9, 0.2), 1:7)
Rdamcrit.class <- c(seq(1, 10, 2)/100, seq(1, 10, 2)/10, seq(1, 10, 2))

# initial nomogram
nom.0 <- rms::nomogram(f.mi.BS.Rec.5,maxscale=10,
                  Breslow=Breslow.class, Rdamcrit=Rdamcrit.class, Age.SN=Age.class)

# baseline hazard of recurrence
f.Rec.basehaz <- survival::basehaz(f.mi.BS.Rec.5,TRUE)
h0.Rec <- f.Rec.basehaz$hazard[f.Rec.basehaz$time==max(f.Rec.basehaz$time[f.Rec.basehaz$time<=horizon])]

# lp range Recurrence
rc.Rec <- (1/attributes(nom.0)$info$sc)
int.Rec <- attributes(nom.0)$info$Intercept
lp.range <- int.Rec+rc.Rec*(0:10)*2

# nomogram
nom <- rms::nomogram(f.mi.BS.Rec.5, maxscale=10,
                Breslow=Breslow.class, Rdamcrit=Rdamcrit.class, Age.SN=Age.class,
                lp=TRUE, fun.at=fun.event(lp=lp.range, h0=h0.Rec))
attributes(nom)$names[attributes(nom)$names=="Age.SN"]<-"Age"
attributes(nom)$names[attributes(nom)$names=="Loc_CAT"]<-"Location"
attributes(nom)$names[attributes(nom)$names=="Rdamcrit"]<-"Tumor burden"
attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Positive)"]<-"Breslow + positive SN"
attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Negative)"]<-"Breslow + negative SN"
par(mar = rep(0, 4))
plot(nom, total.sep.page=TRUE, col.grid=gray(c(0.8, 0.95)))
nom.0
nom

save(list=c("dd.training", "training.data.to.be.imputed", "m", "imputed.training.data", "last.imputed.training.data.set", "horizon", "S.Rec.5.training", "S.MSM.5.training",
            "f.mi.full.Rec.5", "f.mi.BS.Rec.5", "f.mi.full.MSM.5.refit", "f.mi.BS.MSM.5.refit", "f.mi.BS.MSM.5", "MSM.cal.fact", "c.Breslow", 
            "form.full.Rec.5", "form.BS.Rec.5", "form.full.MSM.5.refit", "form.BS.MSM.5.refit", 
            "rc.Rec", "int.Rec", "h0.Rec"),
     file='Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata', compress=TRUE)
```

### 2022-11-29 Code/PaperMelanoma.R

@@ -1,1281 +1,0 @@
-# clear all
-rm(list = ls(all.names = TRUE))
-if(!is.null(dev.list())) grDevices::dev.off()
-cat('\014')
-
-#####
-##### Author: C. Maas
-##### Project Melanoom
-#####
-
-# load libraries
-library(foreign)
-library(mice)
-library(survminer)
-library(prodlim)
-library(rms)
-library(Hmisc)
-library(png)
-set.seed(100)
-
-#####
-##### HELPER FUNCTIONS FOR CALIBRATION
-#####
-km <- function(sel, S.km, horizon){
-  S.km.sel <- S.km[sel,]
-  sf <- survfit(S.km.sel~1)
-  year1 <- max(sf$time[sf$time<=horizon])
-  1-sf$surv[sf$time==year1]
-}
-km.lower <- function(sel, S.km, horizon){
-  S.km.sel <- S.km[sel,]
-  sf <- survfit(S.km.sel~1)
-  year1 <- max(sf$time[sf$time<=horizon])
-  1-sf$lower[sf$time==year1]
-}
-km.upper <- function(sel, S.km, horizon){
-  S.km.sel <- S.km[sel,]
-  sf <- survfit(S.km.sel~1)
-  year1 <- max(sf$time[sf$time<=horizon])
-  1-sf$upper[sf$time==year1]
-}
-fun.event <- function(lp, h0)
-{
-  h <- h0*exp(lp)
-  p <- 1-exp(-h)
-  return(p)
-}
-
-source("./Code/table.missing.values.R") # import function to create table of missing values
-source("./Code/val_surv_mi_v3.R")
-
-#####
-##### LOAD DATA
-#####
-original.data <- foreign::read.spss(file="./Data/Dataset Predictiemodel Stassen.sav",
-                                    to.data.frame=TRUE, origin="01-01-1970")
-cat("No SN status:", sum(original.data$SNstatus=="No SN"))
-cat("Duplicates:", sum(original.data$COMMENTS=="double patient!                                                                                                                                                                                         "))
-unique(original.data[original.data$EXCLUSIE=="Exclusion", "COMMENTS"])
-
-# Select necessary variables
-selected.vars <- c("EXCLUSIE", "COMMENTS", "DatabaseNR", "Center", "Sex", "Age", "DOB",
[…]

### 2022-11-29 Code/PaperMelanomaFunctions.R

@@ -1,0 +1,33 @@
+# helper functions for calibration plots
+km <- function(sel, S.km, horizon){
+  S.km.sel <- S.km[sel,]
+  sf <- survfit(S.km.sel~1)
+  year1 <- max(sf$time[sf$time<=horizon])
+  1-sf$surv[sf$time==year1]
+}
+km.lower <- function(sel, S.km, horizon){
+  S.km.sel <- S.km[sel,]
+  sf <- survfit(S.km.sel~1)
+  year1 <- max(sf$time[sf$time<=horizon])
+  1-sf$lower[sf$time==year1]
+}
+km.upper <- function(sel, S.km, horizon){
+  S.km.sel <- S.km[sel,]
+  sf <- survfit(S.km.sel~1)
+  year1 <- max(sf$time[sf$time<=horizon])
+  1-sf$upper[sf$time==year1]
+}
+fun.event <- function(lp, h0)
+{
+  h <- h0*exp(lp)
+  p <- 1-exp(-h)
+  return(p)
+}
+missings.table <- function(data){
+  missing.values.table <- data.frame(m=rep(0, ncol(data)))
+  rownames(missing.values.table) <- colnames(data)
+  for (variable in colnames(data)){
+    missing.values.table[variable,] <- sum(is.na(data[,variable]))
+  }
+  print(missing.values.table[order(missing.values.table$m, decreasing=TRUE), , drop=FALSE])
+}

### 2022-11-29 Code/check_probabilities.R

@@ -1,25 +1,0 @@
-h0.MSM <- 0.1438848
-MSM.cal.fact <- 1.0973015
-
-rscript <- function(status, age, ulceration, location, breslow, tumor_burden) {
-  h0.Rec <- 0.2534864
-  coef.Rec <- c(1.0264235, 0.3514440, 0.4914014, 0.2621760, 0.3803103,
-                0.7062802, 0.8212041, 0.1866868, -0.3838754)
-  names(coef.Rec) <- c("SNstatus=Positive", "Age.SN", "Ulceration=Yes",
-                       "Loc_CAT=leg", "Loc_CAT=trunk", "Loc_CAT=headneck",
-                       "Breslow", "Rdamcrit", "SNstatus=Positive * Breslow")
-  c.Breslow <- 0.7154091
-  center.Rec <- 2.031039
-  lp <- coef.Rec["SNstatus=Positive"]*status+
-    coef.Rec["Age.SN"]*log(age)+
-    coef.Rec["Ulceration=Yes"]*ulceration+
-    coef.Rec["Loc_CAT=leg"]*I(location == 1)+
-    coef.Rec["Loc_CAT=trunk"]*I(location == 2)+
-    coef.Rec["Loc_CAT=headneck"]*I(location == 3)+
-    coef.Rec["Breslow"]*(log(breslow)-c.Breslow)+
-    coef.Rec["Rdamcrit"]*log(tumor_burden)+
-    coef.Rec["SNstatus=Positive * Breslow"]*(log(breslow)-c.Breslow)*status-center.Rec
-  p.Rec <- 1-exp(-h0.Rec*exp(lp))
-  return(as.numeric(p.Rec))
-}
-rscript(0, 55, 0, 0, 7, 1)

### 2022-11-29 Code/old_analysis.R

@@ -1,1533 +1,0 @@
-# De belangrijkste primaire factoren die we mee zouden willen nemen zijn (waarvan de laatste 2 iets minder dan de eerste 6):
-# -	Leeftijd (continu)
-# -	Gender (male/female)
-# -	Locatie melanoom (o.a. arm, hoofd/hals)
-# -	Breslow dikte (continu)
-# -	Histologie (o.a. NM, ALM)
-# -	Ulceratie (yes/no). (Er zijn slechts 128 unknown = 4.1%, ik was in de war met andere factor, dus eigenlijk weinig unknown!).
-# -	Clark level. Dit was altijd een zeer sterke prognostische factor in allerlei studies, mogelijk wordt deze factor in de toekomst iets minder van belang maar we denken dat hij misschien toch wel meegenomen zou moeten worden.
-# -	Hoeveelheid verwijderde sentinel nodes (=Tot_nr_SNs)
-#
-# Daarnaast mogelijk interessant:
-# -	Hoeveelheid SN fields (totaal aantal locaties waar de SN is afgenomen, er zijn er een aantal met 2 of zelfs 3, zoals beide oksels en/of ??n lies)? En/of locatie SN field. Maar wij denken dat deze beide factoren te maken hebben met de locatie van het melanoom (bijv. een melanoom op de rug geeft de kans op meerdere SN fields, terwijl een melanoom op het been altijd maar ??n SN field heeft namelijk in de lies aan dezelfde kant). Locatie melanoom staat reeds hierboven genoemd om mee te nemen.
[…]

### 2022-11-29 Code/table.missing.values.R

@@ -1,8 +1,0 @@
-missings.table <- function(data){
-  missing.values.table <- data.frame(m=rep(0, ncol(data)))
-  rownames(missing.values.table) <- colnames(data)
-  for (variable in colnames(data)){
-    missing.values.table[variable,] <- sum(is.na(data[,variable]))
-  }
-  print(missing.values.table[order(missing.values.table$m, decreasing=TRUE), , drop=FALSE])  
-}

### 2022-11-29 Code/val_surv_mi_v3.R

@@ -1,199 +1,0 @@
-# # Input for val.surv.mi
-# p<-as.matrix(prediction)   # matrix with predicted probabilities for imputation i in columns (complete case analyis: one column)
-# y<-Surv(time,status)  # Time to event outcome as Surv object (time,status)
-# g<-4 # number of risk groups; default=5
-# main<-"Plot label" # Default = ""
-# save_plots<-FALSE # save plots to file? default=FALSE
-# time<-30 #time point at which to evaluate the predicted probabilities; default=NULL (not entered), the maximum time point will be taken
-#
-# val.surv.mi(p=p,y=y,g=g,main=main,save_plots = save_plots,time=time)
-
-Rubin.combine<-function(est,se){
-  m<-length(est)
-  est.mi<-mean(est)
-  var.w<-mean(se^2)
-  var.b<-0
-  if (m>1) var.b<-sum((est - est.mi)^2)/(m - 1)
-  se.mi<-sqrt(var.w+(1+1/m)*var.b)
-  return(list(est=est.mi,se=se.mi))
-}
-
-val.surv.mi<-function(p,y,g=5,main="",file_path="",save_plots=FALSE,time=NULL,
-                      lim=c(0,1),dist=TRUE,CI=FALSE, df=3){
-  if (!is.null(time)){
-    y[y[,1]>time,2]<-0
-    y[y[,1]>time,1]<-time
-  }
-
-  lp<-log(-log(1-p))
-
-  n<-length(y)
-  m.imp.val<-ncol(lp)
-
-  cindex<-rep(0,m.imp.val)
-  cindex.se<-rep(0,m.imp.val)
-  slope<-rep(0,m.imp.val)
-  slope.se<-rep(0,m.imp.val)
-  int<-rep(0,m.imp.val)
-  int.se<-rep(0,m.imp.val)
-  cindex<-rep(0,m.imp.val)
-  cindex.se<-rep(0,m.imp.val)
-
-  p.groups<-array(rep(0,g*m.imp.val),dim=c(m.imp.val,g),dimnames=list(1:m.imp.val,1:g))
-  y.groups<-array(rep(0,2*g*m.imp.val),dim=c(m.imp.val,g,2),dimnames=list(1:m.imp.val,1:g,c("obs","se")))
-  lp.range<-min(lp)+0:100*(max(lp)-min(lp))/100
[…]

### 2022-12-19 Code/0_PrepareTrainingData.Rmd
Validaiton added

@@
 author: "Carolien C.H.M. Maas"
-date: "14-11-2022"
+date: "December 2022"
 output: html_document
@@
 ```{r, eval=TRUE}
+# 11 patients for which it doesn"t hold that Tot_nr_SNs = nr_SNs_Field_1 + nr_SNs_Field_2 + nr_SNs_Field_3
 sel.train.data$multiple.fields <- ifelse(!is.na(sel.train.data$nr_SNs_Field_2)&sel.train.data$nr_SNs_Field_2>0,1,0)
@@
 ```{r, eval=TRUE}
-cat('Number of patients dead with recurrence without Date_1st_Rec:', sum(is.na(sel.train.data$Date_1st_Rec)&sel.train.data$Recurrence=="recurred"&sel.train.data$dead=="Yes"), '\n')
-cat('Number of patients alive with recurrence without Date_1st_Rec:', sum(is.na(sel.train.data$Date_1st_Rec)&sel.train.data$Recurrence=="recurred"&sel.train.data$dead=="No"))
+cat("Number of patients dead with recurrence without Date_1st_Rec:", sum(is.na(sel.train.data$Date_1st_Rec)&sel.train.data$Recurrence=="recurred"&sel.train.data$dead=="Yes"), "\n")
+cat("Number of patients alive with recurrence without Date_1st_Rec:", sum(is.na(sel.train.data$Date_1st_Rec)&sel.train.data$Recurrence=="recurred"&sel.train.data$dead=="No"))
 ```
@@
 dd.training<-rms::datadist(training.data.to.be.imputed)
-options(datadist='dd.training')
+options(datadist="dd.training")
 options(digits=8)
@@
 missings.table(training.data.to.be.imputed)
-save(list=c("dd.training", "training.data.to.be.imputed"),file='Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata', compress=TRUE)
+save(list=c("dd.training", "training.data.to.be.imputed"),file="Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata", compress=TRUE)
 ```

### 2022-12-19 Code/0_PrepareValidationData.Rmd
Validaiton added

@@
                                                to.data.frame=TRUE, origin="01-01-1970")
+cat("Number of patients in data set:", nrow(original.validation.data), "\n")
 
-# remove space after Negative # Bespreken met Robert
+# remove space after Negative
 levels(original.validation.data$SNstatus)[levels(original.validation.data$SNstatus)=="Negative "] <- "Negative"
 
-# all patients may be included
-sel.val.data <- original.validation.data
+# exclude 4 patients without Status (don"t know if they alive or dead)
+cat("Number of patients without Status:", sum(original.validation.data$Status=="Unknown              "), "\n")
+sel.val.data <- original.validation.data[original.validation.data$Status!="Dead, MCC            ",]
+# exculde 1 patient with cause of death MCC
+sel.val.data <- sel.val.data[sel.val.data$Status!="Unknown              ",]
+cat("Number of included patients:", nrow(sel.val.data))
 ```
@@
 ```{r, eval=TRUE}
-sel.val.data$Histology[sel.val.data$Histology=="LLM"] <- "other" # Bespreken met Robert, is LLM and LMM the same?
+sel.val.data$Histology[sel.val.data$Histology=="LLM"] <- "other" # LLM and LMM is the same
 ```
@@
 ```{r, eval=TRUE}
-sel.val.data$multiple.fields <- as.numeric(!is.na(sel.val.data$Tot_nr_SNs)&sel.val.data$Tot_nr_SNs>1) # Bespreken met ask Robert
+sel.val.data$multiple.fields <- as.numeric(!is.na(sel.val.data$Tot_nr_SNs)&sel.val.data$Tot_nr_SNs>1)
 ```
@@
 #### Recurrence 
+Set recurrence indicator
+```{r, eval=TRUE}
+# Recurrence if patient has Date_1st_Rec, Alive with melanoma, or dead with melanoma
+sel.val.data$Recurrence <- as.numeric(!is.na(sel.val.data$Date_1st_Rec)) # n=1352
+# sel.val.data$Recurrence <- as.numeric(sel.val.data$Status=="Alive with Melanoma  " | sel.val.data$Status=="Dead, melanoma       " | !is.na(sel.val.data$Date_1st_Rec)) # n=1352
+sel.val.data$Recurrence <- factor(sel.val.data$Recurrence, labels=c("no recurrence", "recurred"))
+```
+
 Define time since diagnosis in years (follow-up time)
@@
 sel.val.data$FU.time <- (sel.val.data$Last_FU-sel.val.data$SNdate)/(24*60*60)/365.25
+# median: 4.1629021
 ```
@@
 
-Impute recurrence time for 45 patients with missing Rec.time but Recurrence=="recurred"
+Set recurrence time equal to follow-up time for those without recurrence
 ```{r, eval=TRUE}
-cat('Number of patients dead with recurrence without Date_1st_Rec:', sum(is.na(sel.val.data$Date_1st_Rec)&sel.val.data$Status=="Dead, melanoma       "), '\n')
-cat('Number of patients alive with recurrence without Date_1st_Rec:', sum(is.na(sel.val.data$Date_1st_Rec)&sel.val.data$Status=="Alive with Melanoma  "))
-```
-
-Set recurrence indicator
-```{r, eval=TRUE}
-Rec.date <- as.numeric(!is.na(sel.val.data$Date_1st_Rec)) # n=1294
-Rec.date <- factor(Rec.date, labels=c("no recurrence", "recurred"))
-table(sel.val.data$Status, Rec.date) # Bespreken met Robert
-
-cat("Number of alive patients with Date_1st_Rec, but status NSR:", sum(!is.na(sel.val.data$Date_1st_Rec)&sel.val.data$Status=="Alive NSR            "), "\n")
-cat("Number of alive patients with Date_1st_Rec, but status unknown:", sum(!is.na(sel.val.data$Date_1st_Rec)&sel.val.data$Status=="Alive, status unknown"), "\n")
-cat("Number of dead patients with Date_1st_Rec, but status unknown:", sum(!is.na(sel.val.data$Date_1st_Rec)&sel.val.data$Status=="Dead, cause unknown  "), "\n")
-cat("Number of dead patients with Date_1st_Rec, but status not melanoma:", sum(!is.na(sel.val.data$Date_1st_Rec)&sel.val.data$Status=="Dead, not melanoma   "), "\n")
-
-# Recurrence if patient has Date_1st_Rec, Alive with melanoma, or dead with melanoma
-sel.val.data$Recurrence <- as.numeric(sel.val.data$Status=="Alive with Melanoma  " | sel.val.data$Status=="Dead, melanoma       " | !is.na(sel.val.data$Date_1st_Rec)) # n=1352
-sum(sel.val.data$Recurrence)
+sel.val.data$Rec.time[sel.val.data$Recurrence=="no recurrence"|is.na(sel.val.data$SNdate)]<-sel.val.data$FU.time[sel.val.data$Recurrence=="no recurrence"|is.na(sel.val.data$SNdate)]
+table(sel.val.data$Status, sel.val.data$Recurrence)
 ```
@@
 sel.val.data$dead <- as.numeric(stringr::str_detect(sel.val.data$Status, "Dead"))
-
-# 504 patients died due to melanoma, 102 patients died from other causes
+sel.val.data$dead <- factor(sel.val.data$dead, labels=c("No", "Yes"))
 table(sel.val.data$Status, sel.val.data$dead)
-
 sel.val.data$MSM <- as.numeric(sel.val.data$Status=="Dead, melanoma       ")
@@
 
-sel.val.data$Recurrence <- factor(sel.val.data$Recurrence, labels=c("no recurrence", "recurred"))
-sel.val.data$dead <- factor(sel.val.data$dead, labels=c("No", "Yes"))
 sel.val.data$MSM <- factor(sel.val.data$MSM, labels=c("Other", "Death by melanoma"))
@@
                              "Mitosis",
-                             "Recurrence", "dead", "MSM")
+                             "Recurrence", "dead", "MSM", "Rec.time", "FU.time")
 validation.data.to.be.imputed <- sel.val.data[, variables.to.be.imputed]
@@
 dd.validation<-rms::datadist(validation.data.to.be.imputed)
-options(datadist='dd.validation')
+options(datadist="dd.validation")
 options(digits=8)
@@
 missings.table(validation.data.to.be.imputed)
-save(list=c("dd.validation", "validation.data.to.be.imputed"),file='Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.Rdata', compress=TRUE)
+save(list=c("dd.validation", "validation.data.to.be.imputed"),file="Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.Rdata", compress=TRUE)
 ```

### 2022-12-19 Code/10_AJCC.Rmd
Validaiton added

--- 2022-11-29 Code/4_AJCC.Rmd
@@
 # load libraries
-# remotes::install_github("CHMMaas/PredictionTools")
 library(PredictionTools)  # to plot calibration
+library(compareC)         # compare AJCC to model
 ```
@@
 ```{r, eval=TRUE}
-calculate.AJCC.7.8 <- function(imputed.data=NULL, type=NULL, S.Rec.5=NULL, S.MSM.5=NULL){
+calculate.AJCC.7.8 <- function(imputed.data=NULL, type=NULL, S.Rec.5=NULL){
   ####
@@
   }
+
   # combine C-index for each imputed data set
@@
   # save C-index and confidence interval
-  utils::write.table(rbind(c("All patients:", sprintf('%.2f', cindex.mi.AJCC.7$est),
-                paste0("[", sprintf('%.2f', cindex.mi.AJCC.7$est+qnorm(.025)*cindex.mi.AJCC.7$se),
-                        "; ", sprintf('%.2f', cindex.mi.AJCC.7$est+qnorm(.975)*cindex.mi.AJCC.7$se), "]")),
-              c("Positive SN patients:", sprintf('%.2f', cindex.mi.AJCC.7.pos$est),
-                    paste0("[", sprintf('%.2f', cindex.mi.AJCC.7.pos$est+qnorm(.025)*cindex.mi.AJCC.7.pos$se),
-                           "; ", sprintf('%.2f', cindex.mi.AJCC.7.pos$est+qnorm(.975)*cindex.mi.AJCC.7.pos$se), "]")),
-              c("Negative SN patients:", sprintf('%.2f', cindex.mi.AJCC.7.neg$est),
-                    paste0("[", sprintf('%.2f', cindex.mi.AJCC.7.neg$est+qnorm(.025)*cindex.mi.AJCC.7.neg$se),
-                           "; ", sprintf('%.2f', cindex.mi.AJCC.7.neg$est+qnorm(.975)*cindex.mi.AJCC.7.neg$se), "]"))),
-              file=paste0('Z:/Project Melanoom/PaperMelanoma/Results/AJCC.7.Cindex.', type, '.txt'), 
-              row.names=FALSE, col.names=FALSE, sep=",")
+  AJCC.7.table <- rbind(c("All patients:", sprintf("%.2f", cindex.mi.AJCC.7$est),
+                paste0("[", sprintf("%.2f", cindex.mi.AJCC.7$est+qnorm(.025)*cindex.mi.AJCC.7$se),
+                        "; ", sprintf("%.2f", cindex.mi.AJCC.7$est+qnorm(.975)*cindex.mi.AJCC.7$se), "]")),
+              c("Negative SN patients:", sprintf("%.2f", cindex.mi.AJCC.7.neg$est),
+                    paste0("[", sprintf("%.2f", cindex.mi.AJCC.7.neg$est+qnorm(.025)*cindex.mi.AJCC.7.neg$se),
+                           "; ", sprintf("%.2f", cindex.mi.AJCC.7.neg$est+qnorm(.975)*cindex.mi.AJCC.7.neg$se), "]")),
+              c("Positive SN patients:", sprintf("%.2f", cindex.mi.AJCC.7.pos$est),
+                    paste0("[", sprintf("%.2f", cindex.mi.AJCC.7.pos$est+qnorm(.025)*cindex.mi.AJCC.7.pos$se),
+                           "; ", sprintf("%.2f", cindex.mi.AJCC.7.pos$est+qnorm(.975)*cindex.mi.AJCC.7.pos$se), "]")))
 
@@
   # save C-index and confidence interval
-  utils::write.table(rbind(c("All patients:", sprintf('%.2f', cindex.mi.AJCC.8$est),
-                          paste0("[", sprintf('%.2f', cindex.mi.AJCC.8$est+qnorm(.025)*cindex.mi.AJCC.8$se),
-                                 "; ", sprintf('%.2f', cindex.mi.AJCC.8$est+qnorm(.975)*cindex.mi.AJCC.8$se), "]")),
-                    c("Positive SN patients:", sprintf('%.2f', cindex.mi.AJCC.8.pos$est),
-                          paste0("[", sprintf('%.2f', cindex.mi.AJCC.8.pos$est+qnorm(.025)*cindex.mi.AJCC.8.pos$se),
-                                 "; ", sprintf('%.2f', cindex.mi.AJCC.8.pos$est+qnorm(.975)*cindex.mi.AJCC.8.pos$se), "]")),
-                    c("Negative SN patients", sprintf('%.2f', cindex.mi.AJCC.8.neg$est),
-                          paste0("[", sprintf('%.2f', cindex.mi.AJCC.8.neg$est+qnorm(.025)*cindex.mi.AJCC.8.neg$se),
-                                 "; ", sprintf('%.2f', cindex.mi.AJCC.8.neg$est+qnorm(.975)*cindex.mi.AJCC.8.neg$se), "]"))),
-              file=paste0('Z:/Project Melanoom/PaperMelanoma/Results/AJCC.8.Cindex.', type, '.txt'), 
-              row.names=FALSE, col.names=FALSE, sep=",")
+  AJCC.8.table <- rbind(c("All patients:", sprintf("%.2f", cindex.mi.AJCC.8$est),
+                          paste0("[", sprintf("%.2f", cindex.mi.AJCC.8$est+qnorm(.025)*cindex.mi.AJCC.8$se),
+                                 "; ", sprintf("%.2f", cindex.mi.AJCC.8$est+qnorm(.975)*cindex.mi.AJCC.8$se), "]")),
+                    c("Negative SN patients", sprintf("%.2f", cindex.mi.AJCC.8.neg$est),
+                          paste0("[", sprintf("%.2f", cindex.mi.AJCC.8.neg$est+qnorm(.025)*cindex.mi.AJCC.8.neg$se),
+                                 "; ", sprintf("%.2f", cindex.mi.AJCC.8.neg$est+qnorm(.975)*cindex.mi.AJCC.8.neg$se), "]")),
+                    c("Positive SN patients:", sprintf("%.2f", cindex.mi.AJCC.8.pos$est),
+                          paste0("[", sprintf("%.2f", cindex.mi.AJCC.8.pos$est+qnorm(.025)*cindex.mi.AJCC.8.pos$se),
+                                 "; ", sprintf("%.2f", cindex.mi.AJCC.8.pos$est+qnorm(.975)*cindex.mi.AJCC.8.pos$se), "]")))
 
@@
   if (type=="training"){
-    table.AJCC.7.8 <- cbind(c('1.1', '1.2', '2.1', '2.2', '2.3', '3.1', '3.2', '3.3', '3.4'),
+    AJCC.7.8.count <- cbind(c("1.1", "1.2", "2.1", "2.2", "2.3", "3.1", "3.2", "3.3", "3.4"),
                           c(round(rowMeans(table.AJCC.7), 0), 0),
@@
   } else if (type=="validation"){
-    table.AJCC.7.8 <- cbind(c('1.1', '1.2', '2.1', '2.2', '2.3', '3.1', '3.2', '3.3', '3.4'),
+    AJCC.7.8.count <- cbind(c("1.1", "1.2", "2.1", "2.2", "2.3", "3.1", "3.2", "3.3", "3.4"),
                           c(round(rowMeans(table.AJCC.7), 0), 0),
@@
   }
-  utils::write.table(table.AJCC.7.8,
-              file=paste0('Z:/Project Melanoom/PaperMelanoma/Results/table.AJCC.7.and.8.', type, '.txt'),
+  return(list(AJCC.7=AJCC.7, AJCC.7.table=AJCC.7.table,
+              AJCC.8=AJCC.8, AJCC.8.table=AJCC.8.table, AJCC.7.8.count=AJCC.7.8.count))
+}
+load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData")
+train.AJCC <- calculate.AJCC.7.8(imputed.data=imputed.training.data, S.Rec.5=S.Rec.5.training, type="training")
+load("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData")
+validation.AJCC <- calculate.AJCC.7.8(imputed.data=imputed.validation.data, S.Rec.5=S.Rec.5.validation, type="validation")
+```
+
+# Compare C-index of AJCC and developed model
+```{r, eval=TRUE}
+for (edition in c(7, 8)){
+  for (patient.type in c("All", "Negative", "Positive")){
+    model.C <- c()
+    se.C <- c()
+    diff <- c()
+    se <- c()
+    for (i in 1:m){
+      # compute linear predictor
+      lp.i <- predict(f.mi.BS.Rec.5,
+                    newdata=mice::complete(imputed.validation.data, i),
+                    type="lp")
+
+      if (patient.type=="All"){
+        lp.i <- lp.i
+        S <- S.Rec.5.validation
+        AJCC <- eval(parse(text=paste0("validation.AJCC$AJCC.", edition, "[, i]")))
+      } else if (patient.type=="Positive"){
+        lp.i <- lp.i[validation.data.to.be.imputed$SNstatus=="Positive"]
+        S <- S.Rec.5.validation[validation.data.to.be.imputed$SNstatus=="Positive",]
+        AJCC <- eval(parse(text=paste0("validation.AJCC$AJCC.", edition, "[validation.data.to.be.imputed$SNstatus==\"Positive\", i]")))
+      } else if (patient.type=="Negative"){
+        lp.i <- lp.i[validation.data.to.be.imputed$SNstatus=="Negative"]
+        S <- S.Rec.5.validation[validation.data.to.be.imputed$SNstatus=="Negative",]
+        AJCC <- eval(parse(text=paste0("validation.AJCC$AJCC.", edition, "[validation.data.to.be.imputed$SNstatus==\"Negative\", i]")))
+      }
+
+      # compute predictions from model
+      p.i <- 1-exp(-h0.Rec*exp(lp.i))
+
+      # compare AJCC to model
+      compare.i <- compareC::compareC(S[,1], S[,2], -p.i, -AJCC)
+
+      model.C <- c(model.C, compare.i$est.c["Cxy"])
+      se.C <- c(se.C, sqrt(compare.i$est.varCxy))
+
+      # save differences and SE of difference
+      diff <- c(diff, compare.i$est.diff_c)
+      se <- c(se, sqrt(compare.i$est.vardiff_c))
+    }
+    assign(paste0("diff.", patient.type, ".", edition), Rubin.combine(diff, se))
+    assign(paste0("model.", patient.type), Rubin.combine(model.C, se.C))
+  }
+}
+
+# save in Table
+compare.AJCC.model <- rbind(c("All patients:", 
+                sprintf("%.2f", diff.All.7$est),
+                paste0("[", sprintf("%.2f", diff.All.7$est+qnorm(.025)*diff.All.7$se),
+                        "; ", sprintf("%.2f", diff.All.7$est+qnorm(.975)*diff.All.7$se), "]"),
+                sprintf("%.2f", diff.All.8$est),
+                paste0("[", sprintf("%.2f", diff.All.8$est+qnorm(.025)*diff.All.8$se),
+                        "; ", sprintf("%.2f", diff.All.8$est+qnorm(.975)*diff.All.8$se), "]")),
+              c("Negative SN patients:", 
+                sprintf("%.2f", diff.Negative.7$est),
+                paste0("[", sprintf("%.2f", diff.Negative.7$est+qnorm(.025)*diff.Negative.7$se),
+                       "; ", sprintf("%.2f", diff.Negative.7$est+qnorm(.975)*diff.Negative.7$se), "]"),
+                sprintf("%.2f", diff.Negative.8$est),
+                paste0("[", sprintf("%.2f", diff.Negative.8$est+qnorm(.025)*diff.Negative.8$se),
+                       "; ", sprintf("%.2f", diff.Negative.8$est+qnorm(.975)*diff.Negative.8$se), "]")),
+              c("Positive SN patients:", 
+                sprintf("%.2f", diff.Positive.7$est),
+                paste0("[", sprintf("%.2f", diff.Positive.7$est+qnorm(.025)*diff.Positive.7$se),
+                       "; ", sprintf("%.2f", diff.Positive.7$est+qnorm(.975)*diff.Positive.7$se), "]"),
+                sprintf("%.2f", diff.Positive.8$est),
+                paste0("[", sprintf("%.2f", diff.Positive.8$est+qnorm(.025)*diff.Positive.8$se),
+                       "; ", sprintf("%.2f", diff.Positive.8$est+qnorm(.975)*diff.Positive.8$se), "]")))
+
+CIndex.val <- rbind(c("All patients:",
+                sprintf("%.2f", model.All$est),
+                paste0("[", sprintf("%.2f", model.All$est+qnorm(.025)*model.All$se), 
+                      "; ", sprintf("%.2f", model.All$est+qnorm(.975)*model.All$se), "]")),
+                c("Negative SN patients:",
+                sprintf("%.2f", model.Negative$est), 
+                paste0("[", sprintf("%.2f", model.Negative$est+qnorm(.025)*model.Negative$se), 
+                      "; ", sprintf("%.2f", model.Negative$est+qnorm(.975)*model.Negative$se), "]")),
+                c("Positve SN patients:",
+                sprintf("%.2f", model.Positive$est), 
+                paste0("[", sprintf("%.2f", model.Positive$est+qnorm(.025)*model.Positive$se), 
+                      "; ", sprintf("%.2f", model.Positive$est+qnorm(.975)*model.Positive$se), "]")))
+```
+
+# Write to Excel
+```{r, eval=TRUE}
+utils::write.table(cbind(train.AJCC$AJCC.7.table, rep("", 3), train.AJCC$AJCC.8.table[,-1]),
+              file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/AJCC.train.txt"),
               row.names=FALSE, col.names=FALSE, sep=",")
-}
-load('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData')
-calculate.AJCC.7.8(imputed.data=imputed.training.data, S.Rec.5=S.Rec.5.training, S.MSM.5=S.MSM.5.training, type='training')
-load('Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData')
-calculate.AJCC.7.8(imputed.data=imputed.validation.data, S.Rec.5=S.Rec.5.validation, S.MSM.5=S.MSM.5.validation, type='validation')
+utils::write.table(cbind(CIndex.val, rep("", 3),
+                   validation.AJCC$AJCC.7.table[, -1], rep("", 3), 
+                   compare.AJCC.model[, 2:3], rep("", 3),
+                   validation.AJCC$AJCC.8.table[,-1], rep("", 3),
+                   compare.AJCC.model[, 4:5]),
+              file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/AJCC.validation.txt"),
+              row.names=FALSE, col.names=FALSE, sep=",")
+utils::write.table(cbind(train.AJCC$AJCC.7.8.count, rep("", 3), validation.AJCC$AJCC.7.8.count[,-1]),
+              file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/AJCC.7.8.count.txt"),
+              row.names=FALSE, col.names=FALSE, sep=",")
 ```

### 2022-12-19 Code/11_TestModel.Rmd
Validaiton added

@@
 ### Check calibration
-Intercept of recurrence should be the same as when setting covariates to 'zero' values minus the center
+Intercept of recurrence should be the same as when setting covariates to "zero" values minus the center
 ```{r, eval=TRUE}
-load('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData')
+load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData")
+options(datadist="dd.training")
 
+# int.Rec only for last imputed data set
 int.Rec
+lp.BS <- f.mi.BS.Rec.5$linear
+f.mi.BS.MSM.5 <- rms::cph(S.MSM.5.training~lp.BS,y=TRUE,x=TRUE)
 20*f.mi.BS.Rec.5$coef["Age.SN"]+
@@
 ```{r, eval=TRUE}
-sum(h.Rec$density*fun.event(lp=lp.points.Rec.5, h0=h0.Rec))
-# equal to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
+# h0 for only last imputed data set
+f.Rec.basehaz.5 <- survival::basehaz(f.mi.BS.Rec.5,TRUE)
+h0.Rec.5 <- f.Rec.basehaz.5$hazard[f.Rec.basehaz.5$time==max(f.Rec.basehaz.5$time[f.Rec.basehaz.5$time<=horizon])]
+
+# MSM.cal.fact for only last imputed data set
+lp.5<-predict(f.mi.BS.Rec.5,
+              newdata=mice::complete(imputed.training.data,5), 
+              type="lp")
+f.mi.5 <- rms::cph(S.MSM.5.training~lp.5,y=TRUE,x=TRUE) 
+MSM.cal.fact.5 <- stats::coef(f.mi.5)
+
+sum(h.Rec$density*fun.event(lp=lp.points.Rec.5, h0=h0.Rec.5))
+# similar to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
 SF<-survival::survfit(S.Rec.5.training~1,data=training.data.to.be.imputed)
@@
 # average probability of recurrence
-mean(fun.event(lp=lp.score.Rec, h0=h0.Rec))
+mean(fun.event(lp=lp.Rec, h0=h0.Rec.5))
 ```
@@
 ```{r, eval=TRUE}
-sum(h.Rec$density*fun.event(lp=lp.points.MSM.5, h0=h0.MSM))
+lp.BS <- f.mi.BS.Rec.5$linear
+f.mi.BS.MSM.5 <- rms::cph(S.MSM.5.training~lp.BS,y=TRUE,x=TRUE)
+f.MSM.basehaz.5 <- survival::basehaz(f.mi.BS.MSM.5,TRUE)
+h0.MSM.5 <- f.MSM.basehaz.5$hazard[f.MSM.basehaz.5$time==max(f.MSM.basehaz.5$time[f.MSM.basehaz.5$time<=horizon])]
+
+sum(h.Rec$density*fun.event(lp=lp.points.MSM.5, h0=h0.MSM.5))
 # equal to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
@@
 # average probability of MSM
-mean(fun.event(lp=lp.MSM, h0=h0.MSM))
+mean(fun.event(lp=lp.MSM, h0=h0.MSM.5))
 ```

### 2022-12-19 Code/12_Validation.Rmd
Validaiton added

@@
 library(mice)
-library(survival)         # needed to use
-# remotes::install_github("CHMMaas/PredictionTools")
-library(PredictionTools)  # to plot calibration
-# import helper functions
+library(survival)
+library(PredictionTools)
 source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")
@@
 
-# Validation
-# Compute predictions for each multiple imputation and make calibration plots with metrics
+# Training data
 ```{r, eval=TRUE}
-# laod data
-load('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData')
-options(datadist='dd.training')
-options(digits=8)
-load('Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData')
 # load model
-load('Z:/Project Melanoom/PaperMelanoma/Results/h0.Rec.MSM.Rdata')
-p.Rec.val <- matrix(NA, nrow(validation.data.to.be.imputed), m)
-p.MSM.val <- matrix(NA, nrow(validation.data.to.be.imputed), m)
+load("Z:/Project Melanoom/PaperMelanoma/Results/h0.Rec.MSM.Rdata")
+
+# load data
+load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData")
+
+# predictions on training data
+p.Rec.5.training <- matrix(NA, nrow(training.data.to.be.imputed), m)
+p.MSM.5.training <- matrix(NA, nrow(training.data.to.be.imputed), m)
 for (i in 1:m){
-  # Linear predictor
+  lp.train.i <- predict(f.mi.BS.Rec.5,
+                      newdata=mice::complete(imputed.training.data, i),
+                      type="lp")
+  p.Rec.5.training[, i] <- fun.event(h0=h0.Rec, lp=lp.train.i)
+  p.MSM.5.training[, i] <- fun.event(h0=h0.MSM, lp=MSM.cal.fact*lp.train.i)
+}
+```
+
+# Validation data
+```{r, eval=TRUE}
+load("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData")
+p.Rec.5.validation <- matrix(NA, nrow(validation.data.to.be.imputed), m)
+p.MSM.5.validation <- matrix(NA, nrow(validation.data.to.be.imputed), m)
+for (i in 1:m){
   lp.val.i <- predict(f.mi.BS.Rec.5,
@@
                       type="lp")
+  p.Rec.5.validation[, i] <- 1-exp(-h0.Rec*exp(lp.val.i))
+  p.MSM.5.validation[, i] <- 1-exp(-h0.MSM*exp(MSM.cal.fact*lp.val.i))
+}
+```
 
-  # Recurrence
-  p.Rec.val[, i] <- 1-exp(-h0.Rec*exp(lp.val.i))
+```{r, eval=TRUE}
+# calibration curve for recurrence
+# TODO: validation based on dd.training or dd.validation?
+for (data.type in c("training", "validation")){
+  for (outcome in c("Rec.5", "MSM.5")){
+    options(datadist=paste0("dd.", data.type))
 
-  # MSM
-  p.MSM.val[, i] <- 1-exp(-h0.MSM*exp(MSM.cal.fact*lp.val.i))
-}
+    S <- eval(parse(text=paste0("S.", outcome, ".", data.type)))
+    p <- eval(parse(text=paste0("p.", outcome, ".", data.type)))
+    data <- eval(parse(text=paste0(data.type, ".data.to.be.imputed")))
 
-# calibration curve for recurrence, TODO: based on dd.training?
-png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/val.Rec.png"))
-PredictionTools::val.surv.mi(p.Rec.val,S.Rec.5.validation,g=5,main="Calibration plot of recurrence for all patients",
-                             time=NULL,lim=c(0,1),dist=TRUE)
+    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", data.type, ".", outcome, ".All.png"))
+    par(mar=rep(0, 4))
+    PredictionTools::val.surv.mi(p, S,
+                                 g=5,main="All patients", time=NULL,lim=c(0,1),dist=TRUE)
     grDevices::dev.off()
 
-# calibration curve for MSM
-png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/val.MSM.png"))
-PredictionTools::val.surv.mi(p.MSM.val,S.MSM.5.validation,g=5,main="Calibration plot of MSM for all patients",
-                             time=NULL,lim=c(0,1),dist=TRUE)
+    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", data.type, ".", outcome, ".Positive.png"))
+    par(mar=rep(0, 4))
+    PredictionTools::val.surv.mi(p[data$SNstatus=="Positive",],
+                                 S[data$SNstatus=="Positive",],
+                                 g=5,main="Patients with positive SN", time=NULL,lim=c(0,1),dist=TRUE)
     grDevices::dev.off()
 
+    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", data.type, ".", outcome, ".Negative.png"))
+    par(mar=rep(0, 4))
+    PredictionTools::val.surv.mi(p[data$SNstatus=="Negative",],
+                                 S[data$SNstatus=="Negative",],
+                                 g=5,main="Patients with negative SN", time=NULL,lim=c(0,1),dist=TRUE)
+    grDevices::dev.off()
+
+    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", data.type, ".", outcome, ".Neg.Pos.png"), 
+        width=32, height=16, units="cm", res=600)
+    par(mar=rep(0, 4))
+    layout(matrix(1:2, ncol=2, byrow=TRUE))
+    for (select.value in c("Positive", "Negative")){
+      plot(NA, xlim=0:1, ylim=0:1, xaxt="n", yaxt="n", bty="n")
+      img <- png::readPNG(paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", data.type, ".", outcome, ".", select.value, ".png"))
+      rasterImage(img, 0, 0, 1, 1)
+    }
+    grDevices::dev.off()
+
+    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/", data.type, ".", outcome, ".png"), width=32, height=45, units="cm", res=600)
+    par(mar=rep(0, 4))
+    layout(matrix(1:2, ncol=1, nrow=2, byrow=TRUE), heights=c(2, 1), widths=c(2, 2))
+    for (select.value in c("All", "Neg.Pos")){
+      plot(NA, xlim=0:1, ylim=0:1, xaxt="n", yaxt="n", bty="n")
+      img <- png::readPNG(paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", data.type, ".", outcome, ".", select.value, ".png"))
+      rasterImage(img, 0, 0, 1, 1)
+    }
+    grDevices::dev.off()
+  }
+}
 ```

### 2022-12-19 Code/1_DescriptivesTable.Rmd
Validaiton added

@@
 if(!is.null(dev.list())) dev.off()
-cat('\014')
+cat("\014")
 
@@
 
-load('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata')
-load('Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.Rdata')
+load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata")
+load("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.Rdata")
 descr.table <- arsenal::tableby(Center ~ SNstatus
@@
                                 +logRec.time
-                                +logFU.time, data=rbind(training.data.to.be.imputed, validation.data.to.be.imputed), 
+                                +logFU.time, data=rbind(training.data.to.be.imputed, validation.data.to.be.imputed[, -c(18, 19)]), # training.data.to.be.imputed,
                                 total.pos="before",

### 2022-12-19 Code/2_ImputeData.Rmd
Validaiton added

@@
   # load data
-  load(paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/', type, '.data.Rdata'))
+  load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/", type, "/", type, ".data.Rdata"))
 
@@
     # impute data
-    data.to.be.imputed <- eval(parse(text=paste0(type, '.data.to.be.imputed')))
+    data.to.be.imputed <- eval(parse(text=paste0(type, ".data.to.be.imputed")))
     imputed.data <- mice::mice(data.to.be.imputed, m=m, maxit=5, print=FALSE, seed=500)
-    assign(paste0('imputed.', type, '.data'), imputed.data)
-    save(list=c(paste0('imputed.', type, '.data')),
-         file=paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/imputed.', type, '.data.Rdata'), compress=TRUE)
+    # print(imputed.data$method)
+    # print(imputed.data$loggedEvents)
+    assign(paste0("imputed.", type, ".data"), imputed.data)
+    save(list=c(paste0("imputed.", type, ".data")),
+         file=paste0("Z:/Project Melanoom/PaperMelanoma/Data/", type, "/imputed.", type, ".data.Rdata"), compress=TRUE)
   }
-  load(paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/imputed.', type, '.data.Rdata'))
-  save(list=c(paste0('dd.', type), paste0(type, '.data.to.be.imputed'), 'm', paste0('imputed.', type, '.data')),
-       file=paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/', type, '.data.Rdata'), compress=TRUE)
+  load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/", type, "/imputed.", type, ".data.Rdata"))
+  save(list=c(paste0("dd.", type), paste0(type, ".data.to.be.imputed"), "m", paste0("imputed.", type, ".data")),
+       file=paste0("Z:/Project Melanoom/PaperMelanoma/Data/", type, "/", type, ".data.Rdata"), compress=TRUE)
 }
@@
 m <- 5
-new.imputation <- FALSE
-impute.missing.values(type='training', m=m, new.imputation=new.imputation)
-impute.missing.values(type='validation', m=m, new.imputation=new.imputation)
+new.imputation <- TRUE
+# Impute: Ulceration, Loc_CAT, Histology, Breslow, Tot_SNs_neg, Tot_SNs_pos, Rdamcrit, Dewar, Mitosis, logRec.time
+impute.missing.values(type="training", m=m, new.imputation=new.imputation)
+# Impute: Ulceration, Histology, Rdamcrit, Dewar, logRec.time
+impute.missing.values(type="validation", m=m, new.imputation=new.imputation)
 

### 2022-12-19 Code/3_Descriptives.Rmd
Validaiton added

@@
   # Load imputed data set 
-  load(paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/', type, '.data.RData'))
+  load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/", type, "/", type, ".data.RData"))
   imputed.data <- eval(parse(text=paste0("imputed.", type, ".data")))
@@
 
-  assign(paste0('S.Rec.', type), survival::Surv(exp(last.imputed.data.set$logRec.time), last.imputed.data.set$Recurrence=="recurred"))
-  assign(paste0('S.MSM.', type), survival::Surv(exp(last.imputed.data.set$logFU.time), last.imputed.data.set$MSM=="Death by melanoma"))
+  assign(paste0("S.Rec.", type), survival::Surv(exp(last.imputed.data.set$logRec.time), last.imputed.data.set$Recurrence=="recurred"))
+  assign(paste0("S.MSM.", type), survival::Surv(exp(last.imputed.data.set$logFU.time), last.imputed.data.set$MSM=="Death by melanoma"))
 
@@
               paste0("S.MSM.", type)),
-       file=paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/', type, '.data.Rdata'), compress=TRUE)
+       file=paste0("Z:/Project Melanoom/PaperMelanoma/Data/", type, "/", type, ".data.Rdata"), compress=TRUE)
 }
-prepare.data(type='training')
-prepare.data(type='validation')
+prepare.data(type="training")
+prepare.data(type="validation")
 ```
@@
 ```{r, eval=TRUE}
-load(paste0('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData'))
+load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData"))
 survminer::ggsurvplot(survival::survfit(S.Rec.training ~ Center, data=last.imputed.training.data.set), conf.int=TRUE, risk.table=TRUE)
 survminer::ggsurvplot(survival::survfit(S.MSM.training ~ Center, data=last.imputed.training.data.set), conf.int=TRUE, risk.table=TRUE)
-load(paste0('Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData'))
-survminer::ggsurvplot(survival::survfit(S.Rec.validation ~ Center, data=last.imputed.validation.data.set), conf.int=TRUE, risk.table=TRUE)
-survminer::ggsurvplot(survival::survfit(S.MSM.validation ~ Center, data=last.imputed.validation.data.set), conf.int=TRUE, risk.table=TRUE)
+load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData"))
+survminer::ggsurvplot(survival::survfit(S.Rec.validation ~ 1, data=last.imputed.validation.data.set), conf.int=TRUE, risk.table=TRUE)
+survminer::ggsurvplot(survival::survfit(S.MSM.validation ~ 1, data=last.imputed.validation.data.set), conf.int=TRUE, risk.table=TRUE)
 ```
 
-# Median follow-up time
+# Median time to recurrence and follow-up time
 ```{r, eval=TRUE}
-median.FU.time <- function(data=NA){
-  median.FU.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM=="Death by melanoma")~1, data=data[data$Center!="Australia",], reverse=TRUE))
-  tab.FU.time <- c("Total Development", paste(sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "quantile"]),
-                   paste0("[", sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "lower"]), "; ",
-                          sprintf("%.2f", median.FU.time.total$quantiles.survival[3, "upper"]), "]")))
-  median.FU.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM=="Death by melanoma")~Center, data=data, reverse=TRUE))
+median.time <- function(data=NA, type=NA){
+  # time in total
+  if (type=="Rec"){
+    median.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logRec.time), Recurrence=="recurred")~1, data=data[data$Center!="Australia",], reverse=TRUE))
+    median.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logRec.time), Recurrence=="recurred")~Center, data=data, reverse=TRUE))
+  } else if (type=="FU"){
+    median.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM=="Death by melanoma")~1, data=data[data$Center!="Australia",], reverse=TRUE))
+    median.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM=="Death by melanoma")~Center, data=data, reverse=TRUE))
+  }
+
+  # save in tab.time
+  tab.time <- c("Total Development", paste(sprintf("%.2f", median.time.total$quantiles.survival[3, "quantile"]),
+                   paste0("[", sprintf("%.2f", median.time.total$quantiles.survival[3, "lower"]), "; ",
+                          sprintf("%.2f", median.time.total$quantiles.survival[3, "upper"]), "]")))
+
+  # time per center
   for (center in c("Amsterdam", "Berlin", "Rotterdam", "Warsaw", "Australia")){
-    med.FU.table <- eval(parse(text=paste0("median.FU.time$`Center=", center, "`")))
-    med.FU <- med.FU.table[3,"quantile"]
-    lower.FU <- med.FU.table[3,"lower"]
-    upper.FU <- med.FU.table[3,"upper"]
-    tab.FU.time <- rbind(tab.FU.time, c(center, paste(sprintf("%.2f", med.FU), paste0("[", sprintf("%.2f", lower.FU), "; ", sprintf("%.2f", upper.FU), "]"))))
+    med.table <- eval(parse(text=paste0("median.time$`Center=", center, "`")))
+    med <- med.table[3,"quantile"]
+    lower <- med.table[3,"lower"]
+    upper <- med.table[3,"upper"]
+    tab.time <- rbind(tab.time, c(center, paste(sprintf("%.2f", med), paste0("[", sprintf("%.2f", lower), "; ", sprintf("%.2f", upper), "]"))))
   }
-  write.table(as.data.frame(tab.FU.time),
-              file=paste0('Z:/Project Melanoom/PaperMelanoma/Results/med.FU.table.txt'),
+
+  return(tab.time)
+}
+med.Rec <- median.time(data=rbind(training.data.to.be.imputed[, c("logRec.time", "Recurrence", "Center")], 
+                       validation.data.to.be.imputed[, c("logRec.time", "Recurrence", "Center")]), type="Rec")
+med.FU <- median.time(data=rbind(training.data.to.be.imputed[, c("logFU.time", "MSM", "Center")], 
+                       validation.data.to.be.imputed[, c("logFU.time", "MSM", "Center")]), type="FU")
+# save to .txt
+write.table(cbind(c("Recurrence time", "Follow-up time"), 
+                  rbind(c(med.Rec[, -1]), c(med.FU[,-1]))),
+              file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/med.table.txt"),
               row.names=FALSE, sep=",")
-  return(tab.FU.time)
-}
-median.FU.time(data=rbind(training.data.to.be.imputed[, c("logFU.time", "MSM", "Center")], 
-                          validation.data.to.be.imputed[, c("logFU.time", "MSM", "Center")]))
 ```
 
-# Recurrence at 5 years
+# Kaplan-Meier estimates of recurrence and MSM at 5 years
 ```{r, eval=TRUE}
@@
   # all patients, Recurrence
-  S.Rec <- eval(parse(text=paste0('S.Rec.', type)))
-  SF.Rec.all<-survfit(S.Rec~1, data=data.to.be.imputed)
+  S.Rec <- eval(parse(text=paste0("S.Rec.", type)))
+  SF.Rec.all <- survival::survfit(S.Rec~1, data=data.to.be.imputed)
   Rec.all <- 100*(1-SF.Rec.all$surv[SF.Rec.all$time==max(SF.Rec.all$time[SF.Rec.all$time<=horizon])])
+  n.Rec.all <- sum(SF.Rec.all$n.event[SF.Rec.all$time<=horizon])
 
   # only negative patients, Recurrence
-  SF.Rec.neg<-survfit(S.Rec[data.to.be.imputed$SNstatus=="Negative"]~1, data=data.to.be.imputed)
+  SF.Rec.neg <-survival::survfit(S.Rec[data.to.be.imputed$SNstatus=="Negative"]~1, data=data.to.be.imputed)
   Rec.neg <- 100*(1-SF.Rec.neg$surv[SF.Rec.neg$time==max(SF.Rec.neg$time[SF.Rec.neg$time<=horizon])])
+  n.Rec.neg <- sum(SF.Rec.neg$n.event[SF.Rec.neg$time<=horizon])
 
   # only positive patients, Recurrence
-  SF.Rec.pos<-survfit(S.Rec[data.to.be.imputed$SNstatus=="Positive"]~1, data=data.to.be.imputed)
+  SF.Rec.pos <- survival::survfit(S.Rec[data.to.be.imputed$SNstatus=="Positive"]~1, data=data.to.be.imputed)
   Rec.pos <- 100*(1-SF.Rec.pos$surv[SF.Rec.pos$time==max(SF.Rec.pos$time[SF.Rec.pos$time<=horizon])])
+  n.Rec.pos <- sum(SF.Rec.pos$n.event[SF.Rec.pos$time<=horizon])
 
   # all patients, MSM
-  S.MSM <- eval(parse(text=paste0('S.MSM.', type)))
-  SF.MSM.all<-survfit(S.MSM~1, data=data.to.be.imputed)
+  S.MSM <- eval(parse(text=paste0("S.MSM.", type)))
+  SF.MSM.all <- survival::survfit(S.MSM~1, data=data.to.be.imputed)
   MSM.all <- 100*(1-SF.MSM.all$surv[SF.MSM.all$time==max(SF.MSM.all$time[SF.MSM.all$time<=horizon])])
+  n.MSM.all <- sum(SF.MSM.all$n.event[SF.MSM.all$time<=horizon])
 
   # negative patients, MSM
-  SF.MSM.neg<-survfit(S.MSM[data.to.be.imputed$SNstatus=="Negative"]~1, data=data.to.be.imputed)
+  SF.MSM.neg <- survival::survfit(S.MSM[data.to.be.imputed$SNstatus=="Negative"]~1, data=data.to.be.imputed)
   MSM.neg <- 100*(1-SF.MSM.neg$surv[SF.MSM.neg$time==max(SF.MSM.neg$time[SF.MSM.neg$time<=horizon])])
+  n.MSM.neg <- sum(SF.MSM.neg$n.event[SF.MSM.neg$time<=horizon])
 
   # positive patients, MSM
-  SF.MSM.pos<-survfit(S.MSM[data.to.be.imputed$SNstatus=="Positive"]~1, data=data.to.be.imputed)
+  SF.MSM.pos <- survival::survfit(S.MSM[data.to.be.imputed$SNstatus=="Positive"]~1, data=data.to.be.imputed)
   MSM.pos <- 100*(1-SF.MSM.pos$surv[SF.MSM.pos$time==max(SF.MSM.pos$time[SF.MSM.pos$time<=horizon])])
+  n.MSM.pos <- sum(SF.MSM.pos$n.event[SF.MSM.pos$time<=horizon])
 
@@
                                   sum(data.to.be.imputed$SNstatus=="Positive")),
-                              perc=c(100,
-                                     sprintf("%.1f", sum(data.to.be.imputed$SNstatus=="Negative")/nrow(data.to.be.imputed)*100),
-                                     sprintf("%.1f", sum(data.to.be.imputed$SNstatus=="Positive")/nrow(data.to.be.imputed)*100)),
+                              e1=rep("", 3),
+                              n.Rec=c(n.Rec.all, n.Rec.pos, n.Rec.neg),
+                              n.MSM=c(n.MSM.all, n.MSM.pos, n.MSM.neg),
+                              e2=rep("", 3),
                               Rec=sprintf("%.1f", c(Rec.all, Rec.neg, Rec.pos)),
@@
   row.names(Rec.MSM.table) <- c("All", "Negative", "Positive")
-  utils::write.table(as.data.frame(Rec.MSM.table),file=paste0('Z:/Project Melanoom/PaperMelanoma/Results/Rec.MSM.table.', type, '.txt'),
-                     row.names=TRUE, sep=",")
+
   return(Rec.MSM.table)
 }
+horizon <- 5
+train.table <- Rec.MSM.table(horizon=horizon, type="training", data.to.be.imputed=training.data.to.be.imputed)
+val.table <- Rec.MSM.table(horizon=horizon, type="validation", data.to.be.imputed=validation.data.to.be.imputed)
 
-horizon <- 5
-Rec.MSM.table(horizon=horizon, type='training', data.to.be.imputed=training.data.to.be.imputed)
-Rec.MSM.table(horizon=horizon, type='validation', data.to.be.imputed=validation.data.to.be.imputed)
+utils::write.table(cbind(train.table, rep("", 3), val.table),
+                   file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Rec.MSM.table.txt"),
+                   row.names=TRUE, sep=",")
 ```
@@
 ```{r, eval=TRUE}
-for (type in c('training',  'validation')){
-  load(paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/', type, '.data.Rdata'))
-  for (outcome in c('Rec', 'MSM')){
-    S <- eval(parse(text=paste0('S.', outcome, '.', type)))
+for (type in c("training",  "validation")){
+  load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/", type, "/", type, ".data.Rdata"))
+  for (outcome in c("Rec", "MSM")){
+    S <- eval(parse(text=paste0("S.", outcome, ".", type)))
+    cat("Number of patients with follow-up for", outcome, "after 5 years:", sum(S[, 1]>5), "in", type, " data \n")
     S.5 <- S
@@
 
-    assign(paste0('S.', outcome, '.5.', type), S.5)
+    assign(paste0("S.", outcome, ".5.", type), S.5)
   }
@@
               paste0("S.MSM.5.", type)),
-     file=paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/', type, '.data.Rdata'), compress=TRUE)
+     file=paste0("Z:/Project Melanoom/PaperMelanoma/Data/", type, "/", type, ".data.Rdata"), compress=TRUE)
 }

### 2022-12-19 Code/4_TestNonlinearity.Rmd
Validaiton added

--- 2022-11-29 Code/5_TestNonlinearity.Rmd
@@
 ```{r, eval=TRUE}
-load('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData')
-options(datadist='dd.training')
-options(digits=8)
+load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData")
+options(datadist="dd.training")
 

### 2022-12-19 Code/5_FitModels.Rmd
Validaiton added

--- 2022-11-29 Code/6_FitModels.Rmd
@@
 ---
-title: "Fit models"
+title: "Fit models and plots"
 author: "Carolien C.H.M. Maas"
-date: "14-11-2022"
+date: "December 2022"
 output: html_document
@@
 library(utils)
+library(PredictionTools)
 ```
@@
 ```{r, eval=TRUE}
-c.Breslow <- mean(log(last.imputed.training.data.set$Breslow))
+# center Breslow by using all imputated data sets
+c.Breslow.all <- c()
+for (i in 1:m){
+  c.Breslow.all <- c(c.Breslow.all, mean(log(mice::complete(imputed.training.data, i)$Breslow)))
+}
+c.Breslow <- mean(c.Breslow.all)
+
 form.BS.Rec.5 <- S.Rec.5.training ~
@@
 # backward selection
-# round(stats::anova(f.mi.BS.Rec.5)[order(stats::anova(f.mi.BS.Rec.5)[, 'P']),], 3)
+# round(stats::anova(f.mi.BS.Rec.5)[order(stats::anova(f.mi.BS.Rec.5)[, "P"]),], 3)
 summary(f.mi.BS.Rec.5, Breslow=exp(c.Breslow))
@@
 ```{r, eval=TRUE}
-lp.BS <- f.mi.BS.Rec.5$linear
-f.mi.BS.MSM.5 <- rms::cph(S.MSM.5.training~lp.BS,y=TRUE,x=TRUE)
-MSM.cal.fact <- stats::coef(f.mi.BS.MSM.5)
-MSM.cal.fact
-stats::confint(f.mi.BS.MSM.5)
-utils::write.table(paste0(sprintf("%.2f", stats::coef(f.mi.BS.MSM.5)), " [",
-            sprintf("%.2f", as.numeric(stats::confint(f.mi.BS.MSM.5))[1]), "; ",
-            sprintf("%.2f", as.numeric(stats::confint(f.mi.BS.MSM.5))[2]), "]"),
-            file='Z:/Project Melanoom/PaperMelanoma/Results/calibration.MSS.txt', 
+# Combine calibration coefficients for all imputated data sets 
+MSM.cal.fact.all <- c()
+MSM.cal.fact.SE <- c()
+for (i in 1:m){
+  lp.i<-predict(f.mi.BS.Rec.5,
+                newdata=mice::complete(imputed.training.data,i), 
+                type="lp")
+  f.mi.i <- rms::cph(S.MSM.5.training~lp.i,y=TRUE,x=TRUE) 
+  MSM.cal.fact.all <- c(MSM.cal.fact.all, stats::coef(f.mi.i))
+  MSM.cal.fact.SE <- c(MSM.cal.fact.SE, sqrt(diag(vcov(f.mi.i))))
+}
+MSM.cal.fact.combined<-Rubin.combine(MSM.cal.fact.all,MSM.cal.fact.SE)
+MSM.cal.fact <- MSM.cal.fact.combined$est
+
+utils::write.table(paste0(sprintf("%.2f", MSM.cal.fact), " [",
+            sprintf("%.2f", MSM.cal.fact.combined$est+qnorm(.025)*MSM.cal.fact.combined$se), "; ",
+            sprintf("%.2f", MSM.cal.fact.combined$est+qnorm(.975)*MSM.cal.fact.combined$se), "]"),
+            file="Z:/Project Melanoom/PaperMelanoma/Results/calibration.MSS.txt", 
             row.names=FALSE, col.names=FALSE, sep=",")
 
-save(list=c("dd.training", "training.data.to.be.imputed", "m", "imputed.training.data", "last.imputed.training.data.set", 
-            "horizon", "S.Rec.training", "S.MSM.training", "S.Rec.5.training", "S.MSM.5.training",
-            "f.mi.full.Rec.5", "f.mi.BS.Rec.5", "f.mi.full.MSM.5.refit", "f.mi.BS.MSM.5.refit", "f.mi.BS.MSM.5", "MSM.cal.fact", "c.Breslow", 
+save(list=c("dd.training", "training.data.to.be.imputed", "m", 
+            "imputed.training.data", "last.imputed.training.data.set", 
+            "horizon", "S.Rec.5.training", "S.MSM.5.training",
+            "f.mi.full.Rec.5", "f.mi.BS.Rec.5", "f.mi.full.MSM.5.refit", 
+            "f.mi.BS.MSM.5.refit", "MSM.cal.fact", "c.Breslow", 
             "form.full.Rec.5", "form.BS.Rec.5", "form.full.MSM.5.refit", "form.BS.MSM.5.refit"),
-     file='Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata', compress=TRUE)
-load('Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData')
-save(list=c("dd.validation", "validation.data.to.be.imputed", "m", "imputed.validation.data", "last.imputed.validation.data.set", "horizon", 
+     file="Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata", compress=TRUE)
+load("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData")
+save(list=c("dd.validation", "validation.data.to.be.imputed", "m", 
+            "imputed.validation.data", "last.imputed.validation.data.set", "horizon", 
             "S.Rec.5.validation", "S.MSM.5.validation", "f.mi.BS.Rec.5"),
-     file=paste0('Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.Rdata'), compress=TRUE)
+     file=paste0("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.Rdata"), compress=TRUE)
 ```

### 2022-12-19 Code/6_CreateTablesandPlots.Rmd
Validaiton added

--- 2022-11-29 Code/7_CreateTablesandPlots.Rmd
@@
 library(png)
-# remotes::install_github("CHMMaas/PredictionTools")
-library(PredictionTools)  # to plot calibration
-# import helper functions
+library(PredictionTools)
 source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")
@@
 ### Create variables with results
-##### Put hazard ratio's and anova Chi-square statistics of positive and negative SN patients in Table
+##### Put hazard ratio"s and anova Chi-square statistics of positive and negative SN patients in Table of final recurrence model and refitted MSM model
 ```{r, eval=TRUE}
-load('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData')
-for (model in c('Rec.5', 'MSM.5.refit')){
+load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData")
+options(datadist="dd.training")
+
+for (model in c("Rec.5", "MSM.5.refit")){
   # load model
-  f.mi.full <- eval(parse(text=paste0('f.mi.full.', model)))
+  f.mi.full <- eval(parse(text=paste0("f.mi.full.", model)))
 
@@
   colnames(final.table) <- c("HR.neg", "CI.neg", "HR.pos", "CI.pos", "Main", "Int")
-  assign(paste0('final.full.table.', model), as.data.frame(final.table))
+  assign(paste0("final.full.table.", model), as.data.frame(final.table))
 }
@@
 
-##### Final model
-##### Put hazard ratio's and anova Chi-square statistics of positive and negative SN patients in Table
+##### Put hazard ratio"s and anova Chi-square statistics of positive and negative SN patients in Table of final recurrence model and refitted MSM model
 ```{r, eval=TRUE}
-for (model in c('Rec.5', 'MSM.5.refit')){
-  f.mi.BS <- eval(parse(text=paste0('f.mi.BS.', model)))
+for (model in c("Rec.5", "MSM.5.refit")){
+  f.mi.BS <- eval(parse(text=paste0("f.mi.BS.", model)))
 
@@
   colnames(BS.table) <- c("HR.neg", "CI.neg", "HR.pos", "CI.pos", "Main", "Int")
-  assign(paste0('final.BS.table.', model), as.data.frame(BS.table))
+  assign(paste0("final.BS.table.", model), as.data.frame(BS.table))
 }
@@
 
-# Calculate optimism using bootstrap
+# Calculate optimism using bootstrap using full model and doing backward selection for each bootstrap
 ```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
@@
 for (i in 1:m){
-  v<-validate(f.mi.full.Rec.5$fits[[i]],method='boot',bw=TRUE,rule='p',sls=0.05,B=100,pr=FALSE,type='individual')
+  v<-validate(f.mi.full.Rec.5$fits[[i]],method="boot",bw=TRUE,rule="p",sls=0.05,B=100,pr=FALSE,type="individual")
   slope<-c(slope,v["Slope","test"]) # slope of test
@@
 ```
+
+# Show optimism
 ```{r, eval=TRUE}
@@
 ```
-# Plot calibration of all centers combined and all centers seperate for recurrence and melanoma specific mortality refit 
-#### Add C-index to Recurrence and MSM refit Table
+
+# Add C-index to model for Recurrence and MSM refit Table
 ```{r, eval=TRUE}
-for (model in list('Rec.5', 'MSM.5.refit')){
-  if (model == 'Rec.5'){
+for (model in list("Rec.5", "MSM.5.refit")){
+  if (model == "Rec.5"){
     # plot predict
@@
     # load model
-    f.mi <- eval(parse(text=paste0('f.mi.', type, '.', model)))
+    f.mi <- eval(parse(text=paste0("f.mi.", type, ".", model)))
 
     # load outcome
-    if (model=='Rec.5'){
+    if (model=="Rec.5"){
       S <- S.Rec.5.training
     }
-    else if (model=="MSM.refit"){
+    else if (model=="MSM.5.refit"){
       S <- S.MSM.5.training
@@
 
-    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/all.centers.", model, '.', type, ".png"))
+    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/all.centers.", model, ".", type, ".png"))
     out <- PredictionTools::val.surv.mi(p,S,g=5,main="All centers",
@@
 
-    final.full.table <- eval(parse(text=paste0('final.', type, '.table.', model)))
-    final.full.table["C-index", "HR.neg"] <- sprintf('%.2f', out$cindex-optimism)
-    final.full.table["C-index", "CI.neg"] <- paste0("[", sprintf('%.2f', out$cindex.lower-optimism),
-                                                   "; ", sprintf('%.2f', out$cindex.upper-optimism), "]")
+    final.full.table <- eval(parse(text=paste0("final.", type, ".table.", model)))
+    final.full.table["C-index", "HR.neg"] <- sprintf("%.2f", out$cindex-optimism)
+    final.full.table["C-index", "CI.neg"] <- paste0("[", sprintf("%.2f", out$cindex.lower-optimism),
+                                                   "; ", sprintf("%.2f", out$cindex.upper-optimism), "]")
 
-    assign(paste0('final.', type, '.table.', model), final.full.table)
+    assign(paste0("final.", type, ".table.", model), final.full.table)
+
+    if (type=="BS"&model=="Rec.5"){
+      CIndex.BS.Rec.5.All <- c(sprintf("%.2f", out$cindex-optimism), 
+                              paste0("[", sprintf("%.2f", out$cindex.lower-optimism), 
+                                     "; ", sprintf("%.2f", out$cindex.upper-optimism), "]"))
+    }
   }
@@
 
-Add the C-index for MSM for model of recurrence
+# Add the C-index for calibrated MSM model to Table for recurrence
 ```{r, eval=TRUE}
 for (type in c("full", "BS")){
-  f.mi <- eval(parse(text=paste0('f.mi.', type, '.Rec.5')))
+  f.mi <- eval(parse(text=paste0("f.mi.", type, ".Rec.5")))
   cindex.MSM.refit <- rep(0, m)
@@
 
-  final.full.table <- eval(parse(text=paste0('final.', type, '.table.Rec.5')))
-  final.full.table["C-index MSM", "HR.neg"] <- sprintf('%.2f', cindex.mi.MSM.refit$est-optimism)
-  final.full.table["C-index MSM", "CI.neg"] <- paste0("[", sprintf('%.2f', cindex.mi.MSM.refit$est+qnorm(.025)*cindex.mi.MSM.refit$se-optimism),
-                                                 "; ", sprintf('%.2f', cindex.mi.MSM.refit$est+qnorm(.975)*cindex.mi.MSM.refit$se-optimism), "]")
+  final.full.table <- eval(parse(text=paste0("final.", type, ".table.Rec.5")))
+  final.full.table["C-index MSM", "HR.neg"] <- sprintf("%.2f", cindex.mi.MSM.refit$est-optimism)
+  final.full.table["C-index MSM", "CI.neg"] <- paste0("[", sprintf("%.2f", cindex.mi.MSM.refit$est+qnorm(.025)*cindex.mi.MSM.refit$se-optimism),
+                                                 "; ", sprintf("%.2f", cindex.mi.MSM.refit$est+qnorm(.975)*cindex.mi.MSM.refit$se-optimism), "]")
 
-  assign(paste0('final.', type, '.table.Rec.5'), final.full.table)
+  assign(paste0("final.", type, ".table.Rec.5"), final.full.table)
 }
@@
 write.table(final.full.table.Rec.5,
-            file='Z:/Project Melanoom/PaperMelanoma/Results/final.full.table.Rec.txt', 
+            file="Z:/Project Melanoom/PaperMelanoma/Results/final.full.table.Rec.txt", 
             sep=",")
 write.table(final.full.table.MSM.5.refit,
-            file='Z:/Project Melanoom/PaperMelanoma/Results/final.full.table.MSM.refit.txt', 
+            file="Z:/Project Melanoom/PaperMelanoma/Results/final.full.table.MSM.refit.txt", 
             sep=",")
 write.table(final.BS.table.Rec.5,
-            file='Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.Rec.txt', 
+            file="Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.Rec.txt", 
             sep=",")
 write.table(final.BS.table.MSM.5.refit,
-            file='Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.MSM.refit.txt', 
+            file="Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.MSM.refit.txt", 
             sep=",")
@@
 
-# C-index stratified by SN status
+# C-index of final model for recurrence stratified by SN status
 ```{r, eval=TRUE}
-for (outcome in c("Rec.5", "MSM.5.refit")){
 for (arm in c("Positive", "Negative")){
-    f.mi.stratified <- eval(parse(text=paste0('f.mi.BS.', outcome)))
-    if (outcome=="Rec.5"){
-      S.5 <- S.Rec.5.training
-    } else if (outcome=="MSM.5.refit"){
-      S.5 <- S.MSM.5.training
-    }
-    S.5.subset <- S.5[last.imputed.training.data.set$SNstatus==arm,]
+  f.mi.stratified <- eval(parse(text=paste0("f.mi.BS.Rec.5")))
+  S.5.subset <- S.Rec.5.training[last.imputed.training.data.set$SNstatus==arm,]
 
@@
 
-    print(c(outcome, arm, sprintf('%.2f', cindex.mi.stratified$est),
-    paste0("[", sprintf('%.2f', cindex.mi.stratified$est+qnorm(.025)*cindex.mi.stratified$se), 
-           "; ", sprintf('%.2f', cindex.mi.stratified$est+qnorm(.975)*cindex.mi.stratified$se), "]")))
+  assign(paste0("CIndex.BS.Rec.5.", arm), 
+         c(sprintf("%.2f", cindex.mi.stratified$est), 
+           paste0("[", sprintf("%.2f", cindex.mi.stratified$est+qnorm(.025)*cindex.mi.stratified$se), 
+                  "; ", sprintf("%.2f", cindex.mi.stratified$est+qnorm(.975)*cindex.mi.stratified$se), "]")))
+}
+
+CIndex.Rec.Table <- rbind(CIndex.BS.Rec.5.All, CIndex.BS.Rec.5.Positive, CIndex.BS.Rec.5.Negative)
+CIndex.Rec.Table
+utils::write.table(CIndex.Rec.Table,
+              file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/CIndex.train.txt"),
+              row.names=FALSE, col.names=FALSE, sep=",")
+```
+# Plot cross-validated calibration of all centers combined and all centers seperate for recurrence and calibrated melanoma specific mortality model
+```{r, eval=TRUE}
+Centers <- sort(as.character(unique(training.data.to.be.imputed$Center)))
+Centers.anonimized <- c("A", "B", "C", "D")
+cuts <- 5
+for (model in c("Rec.5", "MSM.5")){
+  if (model == "Rec.5"){
+    S.model <- S.Rec.5.training
+    fact <- 1
+  }
+  if (model == "MSM.5"){
+    S.model <- S.MSM.5.training
+    fact <- MSM.cal.fact
+  }
+
+  for (type in c("full", "BS")){
+    for (j in 1:4){
+      S.j<-S.model[training.data.to.be.imputed$Center==Centers[j],]
+      S.notj<-S.model[training.data.to.be.imputed$Center!=Centers[j],]
+      S.Rec.5.notj<-S.Rec.5.training[training.data.to.be.imputed$Center!=Centers[j],]
+
+      form.model<-eval(parse(text=paste0("form.", type, ".Rec.5")))
+      form.notj<-update(form.model, S.Rec.5.notj  ~ . )
+      f.mi.Rec.notj<-Hmisc::fit.mult.impute(form.notj,cph,xtrans=imputed.training.data,data=training.data.to.be.imputed,
+                                 n.impute=m,pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE,
+                                 sub=training.data.to.be.imputed$Center!=Centers[j])
+
+      if (model=="Rec.5"){
+        f.basehaz.notj<-survival::basehaz(f.mi.Rec.notj,TRUE)
+        h0.notj<-f.basehaz.notj$hazard[f.basehaz.notj$time==max(f.basehaz.notj$time[f.basehaz.notj$time<=horizon])]
+      } else if (model=="MSM.5"){
+        lp.notj<-f.mi.Rec.notj$linear
+        f.MSM.notj<-rms::cph(S.notj~lp.notj,y=TRUE,x=TRUE)
+
+        f.basehaz.notj<-basehaz(f.MSM.notj,TRUE)
+        h0.notj<-f.basehaz.notj$hazard[f.basehaz.notj$time==max(f.basehaz.notj$time[f.basehaz.notj$time<=horizon])]
+      }
+
+      for (select.value in c("Positive", "Negative", "All")){
+        # identify those patients that are selected
+        if (select.value == "All"){
+          select.patients <- TRUE
+        } else if (select.value != "All"){
+          select.patients <- training.data.to.be.imputed[training.data.to.be.imputed$Center==Centers[j],]$SNstatus==select.value
+        }
+
+        # calibration with metrics
+        S <- S.j[select.patients,]
+        p <- matrix(NA, nrow(S), m)
+        for (i in 1:m){
+          lp.i<-predict(f.mi.Rec.notj,newdata=mice::complete(imputed.training.data,i)[training.data.to.be.imputed$Center==Centers[j],], type="lp")
+
+          if (model=="MSM.5"){
+            lp.i<-predict(f.MSM.notj,newdata=lp.i)
+          }
+
+          if (select.value != "All"){
+            lp.i <- lp.i[select.patients]
+          }
+
+          # baseline hazard
+          if (model=="Rec.5"){
+            f.basehaz<-basehaz(f.mi.Rec.notj,TRUE)
+          }
+          else if (model=="MSM.5"){
+            f.basehaz<-basehaz(f.MSM.notj,TRUE)
+          }
+          h0.i <- f.basehaz$hazard[f.basehaz$time==max(f.basehaz$time[f.basehaz$time<=horizon])]
+
+          # probabilities
+          p.i <- fun.event(lp=fact*lp.i, h0=h0.i)
+          p[, i] <- p.i
+        }
+
+        # omit the observation with predicted probability = 1
+        omit <- unique(c(which(p[,1]==1), which(p[,2]==1), which(p[,3]==1), which(p[,4]==1), which(p[,5]==1)))
+        if (length(omit)>0){
+          cat("Omit:", omit, "\n")
+          p <- p[-omit,]
+          S <- S[-omit,]
+        }
+
+        png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", model, ".", select.value, ".", type, ".", Centers[j], ".png"))
+        out <- PredictionTools::val.surv.mi(p, S, g=5, main=paste("Center", Centers.anonimized[j]),
+                           time=NULL,lim=c(0,1),dist=TRUE)
+        grDevices::dev.off()
+      }
+    }
   }
@@
 
+# combine calibration plots into one
+for (select.value in c("All", "Positive", "Negative")){
+  for (y in c("Rec.5", "MSM.5")){
+    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/", y, ".", select.value, ".png"), width=16, height=16, units="cm", res=300)
+    par(mar=rep(0, 4))
+    layout(matrix(1:4, ncol=2, byrow=TRUE))
+    Centers.val <- rep(1:4, 3)
+    for (i in 1:4){
+      plot(NA, xlim=0:1, ylim=0:1, xaxt="n", yaxt="n", bty="n")
+      img <- readPNG(paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", y, ".", select.value, ".BS.", Centers[Centers.val[i]], ".png"))
+      rasterImage(img, 0, 0, 1, 1)
+    }
+    grDevices::dev.off()
+  }
+}
 ```

### 2022-12-19 Code/7_CreateCrossValidatedCalibrationPlots.Rmd
Validaiton added

--- 2022-12-19 Code/6_CreateTablesandPlots.Rmd
@@
 ---
-title: "Create Tables"
+title: "Create calibration plots"
 author: "Carolien C.H.M. Maas"
@@
 
-## Create Tables
+## Create calibration plots
 
@@
 # load libraries
-library(foreign)
+library(Hmisc)
+library(survival)
+library(rms)
 library(mice)
-library(survminer)
-library(prodlim)
-library(rms)
-library(Hmisc)
-library(png)
-library(PredictionTools)
+library(grDevices)
 source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")
+# remotes::install_github("CHMMaas/PredictionTools")
+library(PredictionTools)  # to plot calibration
 ```
 
-### Create variables with results
-##### Put hazard ratio"s and anova Chi-square statistics of positive and negative SN patients in Table of final recurrence model and refitted MSM model
+# Cross-validation recurrence 5-years
 ```{r, eval=TRUE}
@@
 
-for (model in c("Rec.5", "MSM.5.refit")){
-  # load model
-  f.mi.full <- eval(parse(text=paste0("f.mi.full.", model)))
-
-  # Coefficients with CI
-  # Negative results
-  neg.results <- summary(f.mi.full, SNstatus="Negative", Loc_CAT="arm")
-  neg.table <- rbind(rep(NA, 3), # SN status
-                     exp(neg.results[c("Sex - male:female",
-                                       "Age.SN",
-                                       "Ulceration - Yes:No",
-                                       "Loc_CAT - leg:arm",
-                                       "Loc_CAT - trunk:arm",
-                                       "Loc_CAT - head & neck:arm",
-                                       "Histology - NM:SSM",
-                                       "Histology - ALM:SSM",
-                                       "Histology - other:SSM",
-                                       "Breslow",
-                                       "multiple.fields - Multiple SN fields:Other",
-                                       "Tot_SNs_neg"), c("Effect", "Lower 0.95", "Upper 0.95")]),
-                     matrix(NA, nrow=6, ncol=3)) # Tot_SNs_pos, Rdamcrit, Combined, Parenchymal, Multifocal, Extensive
-  # add CI
-  neg.table <- cbind(sprintf("%.2f", neg.table[, 1]),
-                     paste0("[", sprintf("%.2f", neg.table[, 2]), "; ", sprintf("%.2f", neg.table[, 3]), "]"))
-
-  # Positive results
-  pos.results <- summary(f.mi.full, SNstatus="Positive", Loc_CAT="arm", Tot_SNs_pos=c(0, 1))
-  pos.table <- rbind(exp(neg.results["SNstatus - Positive:Negative", c("Effect", "Lower 0.95", "Upper 0.95")]),
-                     exp(pos.results[c("Sex - male:female",
-                                       "Age.SN",
-                                       "Ulceration - Yes:No",
-                                       "Loc_CAT - leg:arm",
-                                       "Loc_CAT - trunk:arm",
-                                       "Loc_CAT - head & neck:arm",
-                                       "Histology - NM:SSM",
-                                       "Histology - ALM:SSM",
-                                       "Histology - other:SSM",
-                                       "Breslow",
-                                       "multiple.fields - Multiple SN fields:Other",
-                                       "Tot_SNs_neg",
-                                       "Tot_SNs_pos",
-                                       "Rdamcrit",
-                                       "Dewar - combined:subcap",
-                                       "Dewar - parenchymal:subcap",
-                                       "Dewar - multifocal:subcap",
-                                       "Dewar - extensive:subcap"), c("Effect", "Lower 0.95", "Upper 0.95")]))
-  # add CI
-  pos.table <- cbind(sprintf("%.2f", pos.table[, 1]),
-                     paste0("[", sprintf("%.2f", pos.table[, 2]), "; ", sprintf("%.2f", pos.table[, 3]), "]"))
-
-  # combine postive with negative
-  full.table <- cbind(neg.table, pos.table)
-
-  # add reference rows
-  final.table <- rbind(full.table[1:4,],
-                       rep(NA, 4),     # location
-                       c("(ref)", NA, "(ref)", NA), # ref arm
-                       full.table[5:7,],   # other locations
-                       rep(NA, 4),     # histology
-                       c("(ref)", NA, "(ref)", NA), # ref SMM
-                       full.table[8:15,],
-                       rep(NA, 4),     # Dewar
-                       c(rep(NA, 2), "(ref)", NA), # ref subcap
-                       full.table[16:nrow(full.table),]) # other Dewar
-
-  # add Chi-square statistics
-  anova.out <- anova(f.mi.full)[, "Chi-Square"]
-  anova.main <- c(anova.out[c("SNstatus  (Factor+Higher Order Factors)",
-                              "Sex  (Factor+Higher Order Factors)",
-                              "Age.SN  (Factor+Higher Order Factors)",
-                              "Ulceration  (Factor+Higher Order Factors)",
-                              "Loc_CAT  (Factor+Higher Order Factors)")],
-                  rep(NA, 4),
-                  anova.out["Histology  (Factor+Higher Order Factors)"],
-                  rep(NA, 4),
-                  anova.out[c("Breslow  (Factor+Higher Order Factors)",
-                              "multiple.fields  (Factor+Higher Order Factors)",
-                              "Tot_SNs_neg",
-                              "Tot_SNs_pos",
-                              "Rdamcrit",
-                              "Dewar")],
-                  rep(NA, 5))
-  anova.int <- c(anova.out[c("TOTAL INTERACTION",
-                             "SNstatus * Sex  (Factor+Higher Order Factors)",
-                             "SNstatus * Age.SN  (Factor+Higher Order Factors)",
-                             "SNstatus * Ulceration  (Factor+Higher Order Factors)",
-                             "SNstatus * Loc_CAT  (Factor+Higher Order Factors)")],
-                 rep(NA, 4),
-                 anova.out["SNstatus * Histology  (Factor+Higher Order Factors)"],
-                 rep(NA, 4),
-                 anova.out[c("SNstatus * Breslow  (Factor+Higher Order Factors)",
-                             "SNstatus * multiple.fields  (Factor+Higher Order Factors)")],
-                 rep(NA, 9))
-  final.table <- cbind(final.table, sprintf("%.0f", anova.main), sprintf("%.0f", anova.int))
-
-  colnames(final.table) <- c("HR.neg", "CI.neg", "HR.pos", "CI.pos", "Main", "Int")
-  assign(paste0("final.full.table.", model), as.data.frame(final.table))
-}
-```
-
-##### Put hazard ratio"s and anova Chi-square statistics of positive and negative SN patients in Table of final recurrence model and refitted MSM model
-```{r, eval=TRUE}
-for (model in c("Rec.5", "MSM.5.refit")){
-  f.mi.BS <- eval(parse(text=paste0("f.mi.BS.", model)))
-
-  # Coefficients with CI
-  # Negative results
-  neg.results.BS <- summary(f.mi.BS, SNstatus="Negative", Breslow=exp(c.Breslow), Loc_CAT="arm")
-  neg.table.BS <- rbind(rep(NA, 3), # SN status
-                        exp(neg.results.BS[c("Age.SN",
-                                             "Ulceration - Yes:No",
-                                             "Loc_CAT - leg:arm",
-                                             "Loc_CAT - trunk:arm",
-                                             "Loc_CAT - head & neck:arm",
-                                             "Breslow"), c("Effect", "Lower 0.95", "Upper 0.95")]),
-                        rep(NA, 3)) # Rdamcrit
-  # add CI
-  neg.table.BS <- cbind(sprintf("%.2f", neg.table.BS[, 1]),
-                        paste0("[", sprintf("%.2f", neg.table.BS[, 2]), "; ", sprintf("%.2f", neg.table.BS[, 3]), "]"))
-
-  # Positive results
-  pos.results.BS <- summary(f.mi.BS, SNstatus="Positive", Breslow=exp(c.Breslow), Loc_CAT="arm")
-  pos.table.BS <- rbind(exp(neg.results.BS["SNstatus - Positive:Negative", c("Effect", "Lower 0.95", "Upper 0.95")]),
-                        exp(pos.results.BS[c("Age.SN",
-                                             "Ulceration - Yes:No",
-                                             "Loc_CAT - leg:arm",
-                                             "Loc_CAT - trunk:arm",
-                                             "Loc_CAT - head & neck:arm",
-                                             "Breslow",
-                                             "Rdamcrit"), c("Effect", "Lower 0.95", "Upper 0.95")]))
-  # add CI
-  pos.table.BS <- cbind(sprintf("%.2f", pos.table.BS[, 1]),
-                        paste0("[", sprintf("%.2f", pos.table.BS[, 2]), "; ", sprintf("%.2f", pos.table.BS[, 3]), "]"))
-
-  # combine postive with negative
-  BS.table <- cbind(neg.table.BS, pos.table.BS)
-
-  # add reference rows
-  BS.table <- rbind(BS.table[1:3,],
-                    rep(NA, 4),
-                    c("(ref)", NA, "(ref)", NA),
-                    BS.table[4:nrow(BS.table),])
-
-  # add Chi-square statistics
-  anova.out <- anova(f.mi.BS)[, "Chi-Square"]
-  anova.main <- c(anova.out[c("SNstatus  (Factor+Higher Order Factors)",
-                              "Age.SN",
-                              "Ulceration",
-                              "Loc_CAT")],
-                  rep(NA, 4),
-                  anova.out[c("Breslow  (Factor+Higher Order Factors)",
-                              "Rdamcrit")])
-  anova.int <- c(rep(NA, 8),
-                 anova.out["SNstatus * Breslow  (Factor+Higher Order Factors)"],
-                 NA)
-  BS.table <- cbind(BS.table, sprintf("%.0f", anova.main), sprintf("%.0f", anova.int))
-
-  colnames(BS.table) <- c("HR.neg", "CI.neg", "HR.pos", "CI.pos", "Main", "Int")
-  assign(paste0("final.BS.table.", model), as.data.frame(BS.table))
-}
-```
-
-# Calculate optimism using bootstrap using full model and doing backward selection for each bootstrap
-```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
-set.seed(100)
-slope<-NULL
-cindex.orig<-NULL
-opt<-NULL
-cindex.B<-NULL
-for (i in 1:m){
-  v<-validate(f.mi.full.Rec.5$fits[[i]],method="boot",bw=TRUE,rule="p",sls=0.05,B=100,pr=FALSE,type="individual")
-  slope<-c(slope,v["Slope","test"]) # slope of test
-  cindex.orig<-c(cindex.orig,(v["Dxy","index.orig"]+1)/2) # original C-index
-  opt<-c(opt,v["Dxy","optimism"]/2) # optimism C-index
-  cindex.B<-c(cindex.B,(v["Dxy","index.corrected"]+1)/2) # index corrected C-index
-}
-```
-
-# Show optimism
-```{r, eval=TRUE}
-mean(cindex.orig)
-optimism <- mean(opt)
-optimism
-mean(cindex.B)
-mean(cindex.orig)-optimism
-```
-
-# Add C-index to model for Recurrence and MSM refit Table
-```{r, eval=TRUE}
-for (model in list("Rec.5", "MSM.5.refit")){
-  if (model == "Rec.5"){
-    # plot predict
-    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/plot.predict.Rec.png"))
-    plot.predict <- plot(Predict(f.mi.full.Rec.5))
-    show(plot.predict)
-    grDevices::dev.off()
-
-    # plot predict Breslow
-    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/plot.predict.Breslow.Rec.png"))
-    plot.Breslow <- plot(Predict(f.mi.full.Rec.5, Breslow=0:20, SNstatus), ~ Breslow, groups="SNstatus",
-                         adj.subtitle=FALSE, lwd=2, labels=FALSE,
-                         par.settings=list(superpose.line=list(lty=c(1, 2), col=c("red", "blue"))),
-                         key=list(cex=.8,corner = c(0.1, .9),lines=list(col=c("red","blue"), lty=c(1,2), lwd=2),text=list(c("Negative","Positive"))))
-    show(plot.Breslow)
-    grDevices::dev.off()
-  }
-
-  # discrimination full and backward selected model
-  for (type in c("full", "BS")){
-    # load model
-    f.mi <- eval(parse(text=paste0("f.mi.", type, ".", model)))
-
-    # load outcome
-    if (model=="Rec.5"){
-      S <- S.Rec.5.training
-    }
-    else if (model=="MSM.5.refit"){
-      S <- S.MSM.5.training
-    }
-
-    # obtain predictions for each imputed data set
-    p <- matrix(NA, nrow(training.data.to.be.imputed), m)
-    for (i in 1:m){
-      # predicted lp for imputation i
-      lp.i <- predict(f.mi,newdata=mice::complete(imputed.training.data,i), type="lp")
-      # baseline hazard
-      f.basehaz <- survival::basehaz(f.mi,TRUE)
-      h0.i <- f.basehaz$hazard[f.basehaz$time==max(f.basehaz$time[f.basehaz$time<=horizon])]
-      # probabilities
-      p.i <- fun.event(lp=lp.i, h0=h0.i)
-      p[, i] <- p.i
-    }
-
-    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/all.centers.", model, ".", type, ".png"))
-    out <- PredictionTools::val.surv.mi(p,S,g=5,main="All centers",
-                       time=NULL,lim=c(0,1),dist=TRUE)
-    grDevices::dev.off()
-
-    final.full.table <- eval(parse(text=paste0("final.", type, ".table.", model)))
-    final.full.table["C-index", "HR.neg"] <- sprintf("%.2f", out$cindex-optimism)
-    final.full.table["C-index", "CI.neg"] <- paste0("[", sprintf("%.2f", out$cindex.lower-optimism),
-                                                   "; ", sprintf("%.2f", out$cindex.upper-optimism), "]")
-
-    assign(paste0("final.", type, ".table.", model), final.full.table)
-
-    if (type=="BS"&model=="Rec.5"){
-      CIndex.BS.Rec.5.All <- c(sprintf("%.2f", out$cindex-optimism), 
-                              paste0("[", sprintf("%.2f", out$cindex.lower-optimism), 
-                                     "; ", sprintf("%.2f", out$cindex.upper-optimism), "]"))
-    }
-  }
-}
-```
-
-# Add the C-index for calibrated MSM model to Table for recurrence
-```{r, eval=TRUE}
-for (type in c("full", "BS")){
-  f.mi <- eval(parse(text=paste0("f.mi.", type, ".Rec.5")))
-  cindex.MSM.refit <- rep(0, m)
-  cindex.se.MSM.refit <- rep(0, m)
-  for (i in 1:m){
-    rc <- Hmisc::rcorr.cens(-predict(f.mi,newdata=mice::complete(imputed.training.data,i), type="lp"), S.MSM.5.training)
-    cindex.MSM.refit[i]<-rc["C Index"]
-    cindex.se.MSM.refit[i]<-rc["S.D."]/2
-  }
-  cindex.mi.MSM.refit<-Rubin.combine(cindex.MSM.refit,cindex.se.MSM.refit)
-
-  final.full.table <- eval(parse(text=paste0("final.", type, ".table.Rec.5")))
-  final.full.table["C-index MSM", "HR.neg"] <- sprintf("%.2f", cindex.mi.MSM.refit$est-optimism)
-  final.full.table["C-index MSM", "CI.neg"] <- paste0("[", sprintf("%.2f", cindex.mi.MSM.refit$est+qnorm(.025)*cindex.mi.MSM.refit$se-optimism),
-                                                 "; ", sprintf("%.2f", cindex.mi.MSM.refit$est+qnorm(.975)*cindex.mi.MSM.refit$se-optimism), "]")
-
-  assign(paste0("final.", type, ".table.Rec.5"), final.full.table)
-}
-```
-
-# Save dataframes of result Tables
-```{r, eval=TRUE}
-# replace NA by blank
-for (df in c("final.full.table.Rec.5", "final.full.table.MSM.5.refit", "final.BS.table.Rec.5", "final.BS.table.MSM.5.refit")){
-  final.df <- eval(parse(text=df))
-  final.df[final.df=="[NA; NA]"] <- NA
-  final.df[final.df=="NA"] <- NA
-  final.df[is.na(final.df)] <- ""
-  assign(df, final.df)
-}
-
-# set rownames
-final.full.table.Rec.5 <- as.data.frame(final.full.table.Rec.5)
-row.names(final.full.table.Rec.5) <- c("Positive SN status", "Male",
-                                    "Age iqr", "Ulceration", "Location",
-                                    "Arm", "Leg", "Trunk", "Head and neck",
-                                    "Histology", "SSM", "NM", "ALM", "Other",
-                                    "Breslow iqr", "Multiple fields", 
-                                    "Total number of negative nodes",
-                                    "Total number of positive nodes",
-                                    "SN tumour burden iqr", 
-                                    "Location metastasis in lymph node",
-                                    "Subcap", "Combined", "Parenchymal",
-                                    "Multifocal", "Extensive",
-                                    rownames(final.full.table.Rec.5)[(nrow(final.full.table.Rec.5)-1):nrow(final.full.table.Rec.5)])
-final.full.table.MSM.5.refit <- as.data.frame(final.full.table.MSM.5.refit)
-row.names(final.full.table.MSM.5.refit) <- c("Positive SN status", "Male",
-                                          "Age", "Ulceration", "Location",
-                                          "Arm", "Leg", "Trunk", "Head and neck",
-                                          "Histology", "SSM", "NM", "ALM", "Other",
-                                          "Breslow", "Multiple fields", 
-                                          "Total number of negative nodes",
-                                          "Total number of positive nodes",
-                                          "SN tumour burden", 
-                                          "Location metastasis in lymph node",
-                                          "Subcap", "Combined", "Parenchymal",
-                                          "Multifocal", "Extensive",
-                                          row.names(final.full.table.MSM.5.refit)[nrow(final.full.table.MSM.5.refit)])
-final.BS.table.Rec.5 <- as.data.frame(final.BS.table.Rec.5)
-row.names(final.BS.table.Rec.5) <- c("Positive SN status", "Age", "Ulceration",
-                                   "Location", "Arm", "Leg", "Trunk", 
-                                   "Head and neck",
-                                   "Breslow", "SN tumour burden", 
-                                   row.names(final.BS.table.Rec.5)[(nrow(final.BS.table.Rec.5)-1):nrow(final.BS.table.Rec.5)])
-final.BS.table.MSM.5.refit <- as.data.frame(final.BS.table.MSM.5.refit)
-row.names(final.BS.table.MSM.5.refit) <- c("Positive SN status", "Age", "Ulceration",
-                                   "Location", "Arm", "Leg", "Trunk", 
-                                   "Head and neck",
-                                   "Breslow", "SN tumour burden", 
-                                   row.names(final.BS.table.MSM.5.refit)[nrow(final.BS.table.MSM.5.refit)])
-
-# save tables
-write.table(final.full.table.Rec.5,
-            file="Z:/Project Melanoom/PaperMelanoma/Results/final.full.table.Rec.txt", 
-            sep=",")
-write.table(final.full.table.MSM.5.refit,
-            file="Z:/Project Melanoom/PaperMelanoma/Results/final.full.table.MSM.refit.txt", 
-            sep=",")
-write.table(final.BS.table.Rec.5,
-            file="Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.Rec.txt", 
-            sep=",")
-write.table(final.BS.table.MSM.5.refit,
-            file="Z:/Project Melanoom/PaperMelanoma/Results/final.BS.table.MSM.refit.txt", 
-            sep=",")
-
-# show tables
-final.full.table.Rec.5
-final.full.table.MSM.5.refit
-final.BS.table.Rec.5
-final.BS.table.MSM.5.refit
-```
-
-# C-index of final model for recurrence stratified by SN status
-```{r, eval=TRUE}
-for (arm in c("Positive", "Negative")){
-  f.mi.stratified <- eval(parse(text=paste0("f.mi.BS.Rec.5")))
-  S.5.subset <- S.Rec.5.training[last.imputed.training.data.set$SNstatus==arm,]
-
-  cindex.stratified <- rep(0, m)
-  cindex.se.stratified <- rep(0, m)
-  for (i in 1:m){
-    subset.imputed.training.data <- as.data.frame(mice::complete(imputed.training.data,i)[last.imputed.training.data.set$SNstatus==arm,])
-    rc <- Hmisc::rcorr.cens(-predict(f.mi.stratified,newdata=subset.imputed.training.data, type="lp"), S.5.subset)
-    cindex.stratified[i]<-rc["C Index"]
-    cindex.se.stratified[i]<-rc["S.D."]/2
-  }
-  cindex.mi.stratified<-Rubin.combine(cindex.stratified,cindex.se.stratified)
-
-  assign(paste0("CIndex.BS.Rec.5.", arm), 
-         c(sprintf("%.2f", cindex.mi.stratified$est), 
-           paste0("[", sprintf("%.2f", cindex.mi.stratified$est+qnorm(.025)*cindex.mi.stratified$se), 
-                  "; ", sprintf("%.2f", cindex.mi.stratified$est+qnorm(.975)*cindex.mi.stratified$se), "]")))
-}
-
-CIndex.Rec.Table <- rbind(CIndex.BS.Rec.5.All, CIndex.BS.Rec.5.Positive, CIndex.BS.Rec.5.Negative)
-CIndex.Rec.Table
-utils::write.table(CIndex.Rec.Table,
-              file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/CIndex.train.txt"),
-              row.names=FALSE, col.names=FALSE, sep=",")
-```
-# Plot cross-validated calibration of all centers combined and all centers seperate for recurrence and calibrated melanoma specific mortality model
-```{r, eval=TRUE}
 Centers <- sort(as.character(unique(training.data.to.be.imputed$Center)))
@@
     S.model <- S.Rec.5.training
-    fact <- 1
   }
@@
     S.model <- S.MSM.5.training
-    fact <- MSM.cal.fact
   }
@@
           # probabilities
-          p.i <- fun.event(lp=fact*lp.i, h0=h0.i)
+          p.i <- fun.event(lp=lp.i, h0=h0.i)
           p[, i] <- p.i
@@
       plot(NA, xlim=0:1, ylim=0:1, xaxt="n", yaxt="n", bty="n")
-      img <- readPNG(paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", y, ".", select.value, ".BS.", Centers[Centers.val[i]], ".png"))
+      img <- png::readPNG(paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", y, ".", select.value, ".BS.", Centers[Centers.val[i]], ".png"))
       rasterImage(img, 0, 0, 1, 1)

### 2022-12-19 Code/8_CreateNomogram.Rmd
Validaiton added

---
title: "Create Nomogram"
author: "Carolien C.H.M. Maas"
date: "December 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Create Nomogram

### Load packages, functions, and data 
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# erase memory
rm(list = ls(all.names = TRUE))

# for reproducibility
set.seed(100)

# load libraries
library(rms)
library(survival)
source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")
```

# Nomogram
```{r, eval=TRUE}
load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData")
options(datadist="dd.training")

# Limits of Age, Breslow, Rdamcrit
Age.class <- c(20, 50, 80)
Breslow.class <- c(0.1, 0.3, 0.6, 1, 2, 3, 7)
Rdamcrit.class <- c(0.01, 0.1, 0.3, 1, 3, 10)

# initial nomogram
nom.0 <- rms::nomogram(f.mi.BS.Rec.5,maxscale=10,
                  Breslow=Breslow.class, Rdamcrit=Rdamcrit.class, Age.SN=Age.class)

# baseline hazard of recurrence averaged over all imputed data sets
h0.Rec.all <- c()
for (i in 1:m){
  f.Rec.basehaz.i <- basehaz(cph(form.BS.Rec.5, data=mice::complete(imputed.training.data, i), se.fit=TRUE, x=TRUE, y=TRUE))
  h0.Rec.i <- f.Rec.basehaz.i$hazard[f.Rec.basehaz.i$time==max(f.Rec.basehaz.i$time[f.Rec.basehaz.i$time<=horizon])]
  h0.Rec.all <- c(h0.Rec.all, h0.Rec.i)
}
h0.Rec <- mean(h0.Rec.all)

# lp range Recurrence
rc.Rec <- (1/attributes(nom.0)$info$sc)
int.Rec <- attributes(nom.0)$info$Intercept
lp.range <- int.Rec+rc.Rec*(0:10)*2

# nomogram
nom <- rms::nomogram(f.mi.BS.Rec.5, maxscale=10,
                Breslow=Breslow.class, Rdamcrit=Rdamcrit.class, Age.SN=Age.class,
                lp=TRUE, fun.at=fun.event(lp=lp.range, h0=h0.Rec))
attributes(nom)$names[attributes(nom)$names=="Age.SN"]<-"Age"
attributes(nom)$names[attributes(nom)$names=="Loc_CAT"]<-"Location"
attributes(nom)$names[attributes(nom)$names=="Rdamcrit"]<-"Tumor burden"
attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Positive)"]<-"Breslow + positive SN"
attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Negative)"]<-"Breslow + negative SN"
# save width: 1000, height: 800
par(mar = rep(0, 4))
plot(nom, total.sep.page=TRUE, col.grid=gray(c(0.8, 0.95)))
nom.0
nom

save(list=c("dd.training", "training.data.to.be.imputed", "m", "imputed.training.data", 
            "last.imputed.training.data.set", "horizon", "S.Rec.5.training", "S.MSM.5.training",
            "f.mi.full.Rec.5", "f.mi.BS.Rec.5", "f.mi.full.MSM.5.refit", "f.mi.BS.MSM.5.refit", "MSM.cal.fact", "c.Breslow", 
            "form.full.Rec.5", "form.BS.Rec.5", "form.full.MSM.5.refit", "form.BS.MSM.5.refit", 
            "rc.Rec", "int.Rec", "h0.Rec"),
     file="Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata", compress=TRUE)
```

### 2022-12-19 Code/9_CreateRiskScoreHistogram.Rmd
Validaiton added

---
title: "Create histogram with risk scores"
author: "Carolien C.H.M. Maas"
date: "December 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Create calibration plots

### Load packages, functions, and data 
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# erase memory
rm(list = ls(all.names = TRUE))

# for reproducibility
set.seed(100)

# load libraries
library(grDevices)
library(stats)
library(survival)
source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")
```

# Risk score distribution
```{r, eval=TRUE}
load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData")
options(datadist="dd.training")

# baseline hazard of MSM
h0.MSM.all <- c()
lp.MSM.all <- c()
for (i in 1:m){
  lp.i<-predict(f.mi.BS.Rec.5,
                newdata=mice::complete(imputed.training.data,i), 
                type="lp")
  lp.MSM.all <- cbind(lp.MSM.all, lp.i)
  f.MSM.basehaz.i <- basehaz(rms::cph(S.MSM.5.training~lp.i,y=TRUE,x=TRUE))
  h0.MSM.i <- f.MSM.basehaz.i$hazard[f.MSM.basehaz.i$time==max(f.MSM.basehaz.i$time[f.MSM.basehaz.i$time<=horizon])]
  h0.MSM.all <- c(h0.MSM.all, h0.MSM.i)
}
h0.MSM <- mean(h0.MSM.all)
lp.MSM <- rowMeans(lp.MSM.all)

# TODO: Is this correct?
lp.MSM.all[3,]

# lp range MSM
rc.MSM <- MSM.cal.fact*rc.Rec
int.MSM <- MSM.cal.fact*int.Rec
max.range <- 16
points.range <- 0:max.range
lp.points.Rec.5 <- int.Rec+rc.Rec*points.range
lp.points.MSM.5 <- int.MSM+rc.MSM*points.range

# Recurrence lp scores
Breslow.trunc <- last.imputed.training.data.set$Breslow
Breslow.trunc[last.imputed.training.data.set$Breslow<0.1] <- .1
Breslow.trunc[last.imputed.training.data.set$Breslow>7] <- 7
lp.Rec <- predict(f.mi.BS.Rec.5,newdata=data.frame(SNstatus=last.imputed.training.data.set$SNstatus,
                                                         Age.SN=last.imputed.training.data.set$Age.SN,
                                                         Ulceration=last.imputed.training.data.set$Ulceration,
                                                         Loc_CAT=last.imputed.training.data.set$Loc_CAT,
                                                         Breslow=Breslow.trunc,
                                                         Rdamcrit=last.imputed.training.data.set$Rdamcrit))
mean(lp.Rec)
score.Rec <- round((lp.Rec-int.Rec)/rc.Rec,0)
max.breaks <- max.range+1
h.Rec <- hist(score.Rec,plot=FALSE,breaks=0:max.breaks,right=FALSE, beside=TRUE)

# score table
data.frame(table(score.Rec)/length(score.Rec))
```

# Plot of risk score distribution
```{r, eval=TRUE, message=FALSE}
h.plot <- h.Rec
h.plot$density <- 100*h.Rec$density
p.Rec <- fun.event(lp=lp.points.Rec.5, h0=h0.Rec)
p.MSM <- fun.event(lp=lp.points.MSM.5, h0=h0.MSM)
x.lim <- c(0, max.range+1)
y.lim <- c(0, 100)
y.lim.hist <- c(0, 25)
png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Risk.distribution.png"))
par(mar = c(5,5,2,5))
plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col="white")
abline(h=(0:30)*1,col="light grey",lty=2)
axis(side = 4)
mtext(side = 4, line = 3, "Risk score distribution (%)")
par(new = TRUE)
plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col="white")
par(new = TRUE)
plot(points.range,p.Rec*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=15, xlab="Risk score",ylab="Predicted 5-year risk (%)")
points(points.range,p.MSM*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=16)
legend(x=0,y=100,legend=c("Recurrence","Melanoma-specific mortality"),pch=c(15,16),cex=.9,bg="white")
grDevices::dev.off()
```

```{r, eval=TRUE, echo=FALSE}
# table of risk distribution
cbind(points.range, round(p.Rec*100, 1), round(p.MSM*100, 1))[c(1, 9, 10, 12, 13),]
```
# Save for Rshiny app
```{r, eval=TRUE}
coef.Rec <- stats::coef(f.mi.BS.Rec.5)
center.Rec <- f.mi.BS.Rec.5$center

save(h0.Rec, h0.MSM, coef.Rec, c.Breslow, center.Rec, MSM.cal.fact,
     file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/h0.Rec.MSM.Rdata"), compress=TRUE)
save(list=c("dd.training", "training.data.to.be.imputed", "m", "imputed.training.data", 
            "last.imputed.training.data.set", "horizon", "S.Rec.5.training", "S.MSM.5.training",
            "f.mi.full.Rec.5", "f.mi.BS.Rec.5", "f.mi.full.MSM.5.refit", "f.mi.BS.MSM.5.refit", "MSM.cal.fact", "c.Breslow", 
            "form.full.Rec.5", "form.BS.Rec.5", "form.full.MSM.5.refit", "form.BS.MSM.5.refit", 
            "rc.Rec", "int.Rec", "h0.Rec", "lp.points.Rec.5",
            "rc.MSM", "rc.Rec", "h0.MSM", "lp.points.MSM.5",
            "lp.Rec", "lp.MSM", "max.breaks", "h.Rec", 
            "h.plot", "y.lim.hist", "x.lim", "coef.Rec", "center.Rec"),
     file="Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata", compress=TRUE)
```

# Risk distribution and scores with ggplot
```{r, eval=TRUE}
library(ggplot2)
last.imputed.training.data.set$score.Rec <- score.Rec # TODO: over 5 imputations?

plot <- ggplot(data=data.frame(), aes(x=0:max.range, y=0:100))+
  geom_bar(data=last.imputed.training.data.set, 
           aes(x=score.Rec, y=(..count..)/sum(..count..)*400, fill=SNstatus))+
  geom_line(aes(x=points.range, y=p.Rec*100))+
  geom_point(aes(x=points.range, y=p.Rec*100, alpha="Recurrence"), shape=15)+
  geom_line(aes(x=points.range, y=p.MSM*100))+
  geom_point(aes(x=points.range, y=p.MSM*100, alpha="Melanoma-specific mortality"), shape=19)+
  scale_y_continuous(limits=c(0, 100), minor_breaks=seq(0, 100, 5),
                     sec.axis=sec_axis(~./4, name="Risk score distribution (%)"))+
  ylab("Predicted 5-year risk (%)")+
  xlab("Risk score")+
  theme_bw()+
  scale_fill_discrete(name=NULL,
                      labels=c("Negative SN status", "Positive SN status"))+
  scale_alpha_manual(name=NULL, 
                     values=c(15, 19),
                     breaks=c("Recurrence", "Melanoma-specific mortality"),
                     guide=guide_legend(override.aes=list(linetype=c(0, 1), 
                                                          shape=c(15, 19), 
                                                          color="black")))+
  theme(legend.position=c(0.25, 0.80),
        legend.title=element_blank())
png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Risk.distribution.png"), 
    width=15, height=15, units="cm", res=300)
show(plot)
grDevices::dev.off()
```

### 2023-01-05 Code/3_Descriptives.Rmd
Validation added
Median time, Q1, Q3 corrected

@@
 ---
-title: "Multiple imputation"
+title: "Descriptive statistics"
 author: "Carolien C.H.M. Maas"
-date: "December 2022"
+date: "January 2023"
 output: html_document
@@
   tab.time <- c("Total Development", paste(sprintf("%.2f", median.time.total$quantiles.survival[3, "quantile"]),
-                   paste0("[", sprintf("%.2f", median.time.total$quantiles.survival[3, "lower"]), "; ",
-                          sprintf("%.2f", median.time.total$quantiles.survival[3, "upper"]), "]")))
+                   paste0("[", sprintf("%.2f", median.time.total$quantiles.survival[4, "quantile"]), "; ",
+                          sprintf("%.2f", median.time.total$quantiles.survival[2, "quantile"]), "]")))
 
@@
     med <- med.table[3,"quantile"]
-    lower <- med.table[3,"lower"]
-    upper <- med.table[3,"upper"]
+    lower <- med.table[4,"quantile"]
+    upper <- med.table[2,"quantile"]
     tab.time <- rbind(tab.time, c(center, paste(sprintf("%.2f", med), paste0("[", sprintf("%.2f", lower), "; ", sprintf("%.2f", upper), "]"))))

### 2023-02-02 Code/0_PrepareValidationData.Rmd
Validation added
Comment mb.c binary

@@
 author: "Carolien C.H.M. Maas"
-date: "December 2022"
+date: "January 2023"
 output: html_document
@@
 sel.val.data$Rec.time[sel.val.data$Recurrence=="no recurrence"|is.na(sel.val.data$SNdate)]<-sel.val.data$FU.time[sel.val.data$Recurrence=="no recurrence"|is.na(sel.val.data$SNdate)]
+
+sel.val.data$Rec.time[sel.val.data$Recurrence=="recurrence"&is.na(sel.val.data$Date_1st_Rec)]<-NA
 table(sel.val.data$Status, sel.val.data$Recurrence)

### 2023-02-09 Code/0_PrepareValidationData.Rmd
Added analysis with pooled data

@@
 sel.val.data <- droplevels(sel.val.data)
-sel.val.data$Center <- "Australia"
+sel.val.data$Center <- 5
+sel.val.data$Center <- factor(sel.val.data$Center, labels = c("Australia"))
 ```
@@
                              "Mitosis",
-                             "Recurrence", "dead", "MSM", "Rec.time", "FU.time")
+                             "Recurrence", "dead", "MSM")
 validation.data.to.be.imputed <- sel.val.data[, variables.to.be.imputed]

### 2023-02-09 Code/13_Check_Pooled_Data.Rmd
Added analysis with pooled data

---
title: "Fit models and plots"
author: "Carolien C.H.M. Maas"
date: "January 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Fit models

### Load packages, functions, and data 
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# erase memory
rm(list = ls(all.names = TRUE))

# for reproducibility
set.seed(100)

# load libraries
library(mice)
library(dplyr)
library(rms)
```

# Check pooled data
```{r, eval=TRUE}
load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData")

# model based on training data
c.Breslow.all <- c()
for (i in 1:m){
  c.Breslow.all <- c(c.Breslow.all, mean(log(mice::complete(imputed.training.data, i)$Breslow)))
}
c.Breslow <- mean(c.Breslow.all)

# prepare data
load("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData")
# combine two raw (without imputation) data sets
pooled.data.to.be.imputed <- rbind(training.data.to.be.imputed, validation.data.to.be.imputed)
# add continent
continent.df <- data.frame(Continent=c(rep("Europe", nrow(complete(imputed.training.data, 5))), 
                                     rep("Australia", nrow(complete(imputed.validation.data, 5)))))

# set datadist
dd.pooled<-rms::datadist(cbind(pooled.data.to.be.imputed, continent.df))
options(datadist="dd.pooled")

# combine two imputed data sets into one mids object
imputed.pooled.data <- mice::rbind(imputed.training.data, imputed.validation.data)
# add continent variable
imputed.pooled.data <- mice::cbind(imputed.pooled.data, continent.df)

# define the outcome based on last imputed data set
last.imputed.pooled.data.set <- rbind(mice::complete(imputed.training.data, m), mice::complete(imputed.validation.data, m))
S.Rec.training.validation <- survival::Surv(exp(last.imputed.pooled.data.set$logRec.time), last.imputed.pooled.data.set$Recurrence=="recurred")
S.Rec.5.training.validation <- S.Rec.training.validation
S.Rec.5.training.validation[S.Rec.training.validation[,1]>5, 1] <- 5 # set time to 5
S.Rec.5.training.validation[S.Rec.training.validation[,1]>5, 2] <- 0 # set censoring
```

```{r, eval=TRUE}
# model is based on training data
# plus a variable for continent
# model is fitted on pooled data
form.BS.Rec.5.pooled.continent <- S.Rec.5.training.validation ~
  SNstatus +
  Age.SN +
  Ulceration +
  Loc_CAT +
  I(log(Breslow)-c.Breslow) +
  log(Rdamcrit) +
  SNstatus:I(log(Breslow)-c.Breslow) +
  Continent*(SNstatus + Age.SN + Ulceration + Loc_CAT + I(log(Breslow)-c.Breslow) + log(Rdamcrit))

# fit model
f.mi.full.pooled.continent <- Hmisc::fit.mult.impute(form.BS.Rec.5.pooled.continent, cph, xtrans=imputed.pooled.data,
                                 data=pooled.data.to.be.imputed, n.impute=m, pr=FALSE,
                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
anova(f.mi.full.pooled.continent)
```
```{r, eval=TRUE}
# model is based on training data
# plus a variable for continent
# model is fitted on pooled data
form.BS.Rec.5.pooled <- S.Rec.5.training.validation ~
  SNstatus +
  Age.SN +
  Ulceration +
  Loc_CAT +
  I(log(Breslow)-c.Breslow) +
  log(Rdamcrit) +
  SNstatus:I(log(Breslow)-c.Breslow) +
  Continent

# fit model
f.mi.full.pooled <- Hmisc::fit.mult.impute(form.BS.Rec.5.pooled, cph, xtrans=imputed.pooled.data,
                                 data=pooled.data.to.be.imputed, n.impute=m, pr=FALSE,
                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
summary(f.mi.full.pooled)
anova(f.mi.full.pooled)
```

```{r, eval=TRUE}
# Coefficients with CI
# Negative results
neg.results.BS <- summary(f.mi.full.pooled, SNstatus="Negative", Breslow=exp(c.Breslow), Loc_CAT="arm", Continent="Europe")
neg.table.BS <- rbind(rep(NA, 3), # SN status
                      exp(neg.results.BS[c("Age.SN",
                                           "Ulceration - Yes:No",
                                           "Loc_CAT - leg:arm",
                                           "Loc_CAT - trunk:arm",
                                           "Loc_CAT - head & neck:arm",
                                           "Breslow"), c("Effect", "Lower 0.95", "Upper 0.95")]),
                      rep(NA, 3), # Rdamcrit
                      exp(neg.results.BS["Continent - Australia:Europe", c("Effect", "Lower 0.95", "Upper 0.95")])) 
# add CI
neg.table.BS <- cbind(sprintf("%.2f", neg.table.BS[, 1]),
                      paste0("[", sprintf("%.2f", neg.table.BS[, 2]), "; ", sprintf("%.2f", neg.table.BS[, 3]), "]"))

# Positive results
pos.results.BS <- summary(f.mi.full.pooled, SNstatus="Positive", Breslow=exp(c.Breslow), Loc_CAT="arm", Continent="Europe")
pos.table.BS <- rbind(exp(neg.results.BS["SNstatus - Positive:Negative", c("Effect", "Lower 0.95", "Upper 0.95")]),
                      exp(pos.results.BS[c("Age.SN",
                                           "Ulceration - Yes:No",
                                           "Loc_CAT - leg:arm",
                                           "Loc_CAT - trunk:arm",
                                           "Loc_CAT - head & neck:arm",
                                           "Breslow",
                                           "Rdamcrit",
                                           "Continent - Australia:Europe"), c("Effect", "Lower 0.95", "Upper 0.95")]))
# add CI
pos.table.BS <- cbind(sprintf("%.2f", pos.table.BS[, 1]),
                      paste0("[", sprintf("%.2f", pos.table.BS[, 2]), "; ", sprintf("%.2f", pos.table.BS[, 3]), "]"))

# combine postive with negative
BS.table <- cbind(neg.table.BS, pos.table.BS)

BS.table <- rbind(BS.table[1:3,],
                    rep(NA, 4),
                    c("(ref)", NA, "(ref)", NA),
                    BS.table[4:nrow(BS.table),])

# add blanks
BS.table[BS.table=="[NA; NA]"] <- NA
BS.table[BS.table=="NA"] <- NA
BS.table[is.na(BS.table)] <- ""

# write to Excel
write.table(BS.table,
            file="Z:/Project Melanoom/PaperMelanoma/Results/pooled.final.full.table.Rec.txt",
            row.names=FALSE,
            sep=",")
```

### 2023-02-09 Code/1_DescriptivesTable.Rmd
Validation added
Added analysis with pooled data

@@
 
+# load packages
 library(arsenal)
@@
                                 +logRec.time
-                                +logFU.time, data=rbind(training.data.to.be.imputed, validation.data.to.be.imputed[, -c(18, 19)]), # training.data.to.be.imputed,
+                                +logFU.time, data=rbind(training.data.to.be.imputed, validation.data.to.be.imputed), # training.data.to.be.imputed,
                                 total.pos="before",

### 2023-02-09 Code/2_ImputeData.Rmd
Validation added
Added analysis with pooled data

@@
   # load data
+  if (type=="pooled"){
+    load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata"))
+    load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.Rdata"))
+    pooled.data.to.be.imputed <- rbind(training.data.to.be.imputed, validation.data.to.be.imputed)
+    dd.pooled<-rms::datadist(pooled.data.to.be.imputed)
+  } else{
     load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/", type, "/", type, ".data.Rdata"))
+  }
 
@@
 impute.missing.values(type="validation", m=m, new.imputation=new.imputation)
+# Impute pooled data
+# impute.missing.values(type="pooled", m=m, new.imputation=new.imputation)
 

### 2023-02-09 Code/13_Check_Pooled_Data.Rmd
Pooled data analysis

@@
 
+# set datadist
+options(datadist="dd.training")
+
 # model based on training data
@@
                                      rep("Australia", nrow(complete(imputed.validation.data, 5)))))
-
-# set datadist
-dd.pooled<-rms::datadist(cbind(pooled.data.to.be.imputed, continent.df))
-options(datadist="dd.pooled")
 
@@
                                  fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
-summary(f.mi.full.pooled)
+summary(f.mi.full.pooled, Continent="Europe")
 anova(f.mi.full.pooled)

### 2023-02-16 Code/12_Validation.Rmd
Validation added
Final results

@@
                       type="lp")
+  print(head(lp.train.i))
   p.Rec.5.training[, i] <- fun.event(h0=h0.Rec, lp=lp.train.i)
@@
 }
+print(coef(f.mi.BS.Rec.5))
 ```

### 2023-02-16 Code/2_ImputeData.Rmd
Final results

--- 2022-12-19 Code/2_ImputeData.Rmd
@@
     imputed.data <- mice::mice(data.to.be.imputed, m=m, maxit=5, print=FALSE, seed=500)
-    # print(imputed.data$method)
-    # print(imputed.data$loggedEvents)
+    print(imputed.data$method)
+    print(imputed.data$loggedEvents)
     assign(paste0("imputed.", type, ".data"), imputed.data)

### 2023-02-16 Code/5_FitModels.Rmd
Validation added
Final results

@@
 }
-MSM.cal.fact.combined<-Rubin.combine(MSM.cal.fact.all,MSM.cal.fact.SE)
+MSM.cal.fact.combined<-Rubin.combine(MSM.cal.fact.all, MSM.cal.fact.SE)
 MSM.cal.fact <- MSM.cal.fact.combined$est

### 2023-02-16 Code/6_CreateTablesandPlots.Rmd
Validation added
Final results

@@
     f.mi <- eval(parse(text=paste0("f.mi.", type, ".", model)))
+    form.mi <- eval(parse(text=paste0("form.", type, ".", model)))
+
+    # obtain predictions for each imputed data set
+    p <- matrix(NA, nrow(training.data.to.be.imputed), m)
+    for (i in 1:m){
+      # baseline hazard
+      # f.Rec.basehaz.i <- basehaz(f.mi, TRUE)
+      f.Rec.basehaz.i <- basehaz(cph(form.mi, data=mice::complete(imputed.training.data, i), se.fit=TRUE, x=TRUE, y=TRUE))
+      h0.Rec.i <- f.Rec.basehaz.i$hazard[f.Rec.basehaz.i$time==max(f.Rec.basehaz.i$time[f.Rec.basehaz.i$time<=horizon])]
+
+      # probabilities
+      lp.i <- predict(f.mi,newdata=mice::complete(imputed.training.data, i), type="lp")
+      p.i <- fun.event(lp=lp.i, h0=h0.Rec.i)
+      p[, i] <- p.i
+    }
 
@@
       S <- S.MSM.5.training
-    }
-
-    # obtain predictions for each imputed data set
-    p <- matrix(NA, nrow(training.data.to.be.imputed), m)
-    for (i in 1:m){
-      # predicted lp for imputation i
-      lp.i <- predict(f.mi,newdata=mice::complete(imputed.training.data,i), type="lp")
-      # baseline hazard
-      f.basehaz <- survival::basehaz(f.mi,TRUE)
-      h0.i <- f.basehaz$hazard[f.basehaz$time==max(f.basehaz$time[f.basehaz$time<=horizon])]
-      # probabilities
-      p.i <- fun.event(lp=lp.i, h0=h0.i)
-      p[, i] <- p.i
     }
@@
 ```
-# Plot cross-validated calibration of all centers combined and all centers seperate for recurrence and calibrated melanoma specific mortality model
-```{r, eval=TRUE}
-Centers <- sort(as.character(unique(training.data.to.be.imputed$Center)))
-Centers.anonimized <- c("A", "B", "C", "D")
-cuts <- 5
-for (model in c("Rec.5", "MSM.5")){
-  if (model == "Rec.5"){
-    S.model <- S.Rec.5.training
-    fact <- 1
-  }
-  if (model == "MSM.5"){
-    S.model <- S.MSM.5.training
-    fact <- MSM.cal.fact
-  }
-
-  for (type in c("full", "BS")){
-    for (j in 1:4){
-      S.j<-S.model[training.data.to.be.imputed$Center==Centers[j],]
-      S.notj<-S.model[training.data.to.be.imputed$Center!=Centers[j],]
-      S.Rec.5.notj<-S.Rec.5.training[training.data.to.be.imputed$Center!=Centers[j],]
-
-      form.model<-eval(parse(text=paste0("form.", type, ".Rec.5")))
-      form.notj<-update(form.model, S.Rec.5.notj  ~ . )
-      f.mi.Rec.notj<-Hmisc::fit.mult.impute(form.notj,cph,xtrans=imputed.training.data,data=training.data.to.be.imputed,
-                                 n.impute=m,pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE,
-                                 sub=training.data.to.be.imputed$Center!=Centers[j])
-
-      if (model=="Rec.5"){
-        f.basehaz.notj<-survival::basehaz(f.mi.Rec.notj,TRUE)
-        h0.notj<-f.basehaz.notj$hazard[f.basehaz.notj$time==max(f.basehaz.notj$time[f.basehaz.notj$time<=horizon])]
-      } else if (model=="MSM.5"){
-        lp.notj<-f.mi.Rec.notj$linear
-        f.MSM.notj<-rms::cph(S.notj~lp.notj,y=TRUE,x=TRUE)
-
-        f.basehaz.notj<-basehaz(f.MSM.notj,TRUE)
-        h0.notj<-f.basehaz.notj$hazard[f.basehaz.notj$time==max(f.basehaz.notj$time[f.basehaz.notj$time<=horizon])]
-      }
-
-      for (select.value in c("Positive", "Negative", "All")){
-        # identify those patients that are selected
-        if (select.value == "All"){
-          select.patients <- TRUE
-        } else if (select.value != "All"){
-          select.patients <- training.data.to.be.imputed[training.data.to.be.imputed$Center==Centers[j],]$SNstatus==select.value
-        }
-
-        # calibration with metrics
-        S <- S.j[select.patients,]
-        p <- matrix(NA, nrow(S), m)
-        for (i in 1:m){
-          lp.i<-predict(f.mi.Rec.notj,newdata=mice::complete(imputed.training.data,i)[training.data.to.be.imputed$Center==Centers[j],], type="lp")
-
-          if (model=="MSM.5"){
-            lp.i<-predict(f.MSM.notj,newdata=lp.i)
-          }
-
-          if (select.value != "All"){
-            lp.i <- lp.i[select.patients]
-          }
-
-          # baseline hazard
-          if (model=="Rec.5"){
-            f.basehaz<-basehaz(f.mi.Rec.notj,TRUE)
-          }
-          else if (model=="MSM.5"){
-            f.basehaz<-basehaz(f.MSM.notj,TRUE)
-          }
-          h0.i <- f.basehaz$hazard[f.basehaz$time==max(f.basehaz$time[f.basehaz$time<=horizon])]
-
-          # probabilities
-          p.i <- fun.event(lp=fact*lp.i, h0=h0.i)
-          p[, i] <- p.i
-        }
-
-        # omit the observation with predicted probability = 1
-        omit <- unique(c(which(p[,1]==1), which(p[,2]==1), which(p[,3]==1), which(p[,4]==1), which(p[,5]==1)))
-        if (length(omit)>0){
-          cat("Omit:", omit, "\n")
-          p <- p[-omit,]
-          S <- S[-omit,]
-        }
-
-        png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", model, ".", select.value, ".", type, ".", Centers[j], ".png"))
-        out <- PredictionTools::val.surv.mi(p, S, g=5, main=paste("Center", Centers.anonimized[j]),
-                           time=NULL,lim=c(0,1),dist=TRUE)
-        grDevices::dev.off()
-      }
-    }
-  }
-}
-
-# combine calibration plots into one
-for (select.value in c("All", "Positive", "Negative")){
-  for (y in c("Rec.5", "MSM.5")){
-    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/", y, ".", select.value, ".png"), width=16, height=16, units="cm", res=300)
-    par(mar=rep(0, 4))
-    layout(matrix(1:4, ncol=2, byrow=TRUE))
-    Centers.val <- rep(1:4, 3)
-    for (i in 1:4){
-      plot(NA, xlim=0:1, ylim=0:1, xaxt="n", yaxt="n", bty="n")
-      img <- readPNG(paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", y, ".", select.value, ".BS.", Centers[Centers.val[i]], ".png"))
-      rasterImage(img, 0, 0, 1, 1)
-    }
-    grDevices::dev.off()
-  }
-}
-```

### 2023-02-16 Code/7_CreateCrossValidatedCalibrationPlots.Rmd
Validation added
Final results

@@
 
-      if (model=="Rec.5"){
-        f.basehaz.notj<-survival::basehaz(f.mi.Rec.notj,TRUE)
-        h0.notj<-f.basehaz.notj$hazard[f.basehaz.notj$time==max(f.basehaz.notj$time[f.basehaz.notj$time<=horizon])]
-      } else if (model=="MSM.5"){
-        lp.notj<-f.mi.Rec.notj$linear
-        f.MSM.notj<-rms::cph(S.notj~lp.notj,y=TRUE,x=TRUE)
-
-        f.basehaz.notj<-basehaz(f.MSM.notj,TRUE)
-        h0.notj<-f.basehaz.notj$hazard[f.basehaz.notj$time==max(f.basehaz.notj$time[f.basehaz.notj$time<=horizon])]
-      }
-
       for (select.value in c("Positive", "Negative", "All")){
@@
         for (i in 1:m){
-          lp.i<-predict(f.mi.Rec.notj,newdata=mice::complete(imputed.training.data,i)[training.data.to.be.imputed$Center==Centers[j],], type="lp")
+          data.i <- mice::complete(imputed.training.data, i)
+          lp.i <- predict(f.mi.Rec.notj,newdata=data.i[training.data.to.be.imputed$Center==Centers[j],], type="lp")
 
           if (model=="MSM.5"){
-            lp.i<-predict(f.MSM.notj,newdata=lp.i)
+            lp.notj<-f.mi.Rec.notj$linear
+            f.MSM.notj<-rms::cph(S.notj~lp.notj, data=data.i[which(data.i$Center!=Centers[j]),],y=TRUE,x=TRUE)
+            lp.i<- predict(f.MSM.notj,newdata=lp.i)
           }
@@
           if (model=="Rec.5"){
-            f.basehaz<-basehaz(f.mi.Rec.notj,TRUE)
+            f.basehaz.i <- basehaz(cph(form.notj, data=data.i[which(data.i$Center!=Centers[j]),], se.fit=TRUE, x=TRUE, y=TRUE))
           }
           else if (model=="MSM.5"){
-            f.basehaz<-basehaz(f.MSM.notj,TRUE)
+            f.basehaz.i <- basehaz(f.MSM.notj)
           }
-          h0.i <- f.basehaz$hazard[f.basehaz$time==max(f.basehaz$time[f.basehaz$time<=horizon])]
+          h0.i <- f.basehaz.i$hazard[f.basehaz.i$time==max(f.basehaz.i$time[f.basehaz.i$time<=horizon])]
 

### 2023-09-13 Code/0_PrepareTrainingData.Rmd
Validation added
Update

@@
 table(sel.train.data$Status, sel.train.data$dead)
-sel.train.data$MSM<-1*(sel.train.data$Status=="DOD")
+sel.train.data$MSM <- 1*(sel.train.data$Status=="DOD")
 ```
@@
 
+#### Describe years
+```{r, eval=TRUE}
+dates <- as.POSIXct(sel.train.data$SNdate, origin="1582-10-14")
+Years <- as.numeric(format(dates, format="%Y"))
+utils::write.table(cbind(table(Years), table(Years, sel.train.data$Recurrence)[,2], 
+                         table(Years, sel.train.data$MSM)[, 2]),
+                   file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Time.Rec.MSM.txt"),
+                   row.names=TRUE, sep=",")
+hist(Years, main="")
+```
+
 # Multiple imputation

### 2023-09-13 Code/13_Check_Pooled_Data.Rmd
Update

@@
                                  data=pooled.data.to.be.imputed, n.impute=m, pr=FALSE,
-                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
+                                 fit.reps=TRUE, fitargs=list(y=TRUE, x=TRUE, se.fit=TRUE))
 anova(f.mi.full.pooled.continent)
@@
                                  data=pooled.data.to.be.imputed, n.impute=m, pr=FALSE,
-                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
+                                 fit.reps=TRUE, fitargs=list(y=TRUE, x=TRUE, se.fit=TRUE))
 summary(f.mi.full.pooled, Continent="Europe")

### 2023-09-13 Code/3_Descriptives.Rmd
Update

@@
 load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData"))
-survminer::ggsurvplot(survival::survfit(S.Rec.training ~ Center, data=last.imputed.training.data.set), conf.int=TRUE, risk.table=TRUE)
-survminer::ggsurvplot(survival::survfit(S.MSM.training ~ Center, data=last.imputed.training.data.set), conf.int=TRUE, risk.table=TRUE)
+png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.Rec.training.png"), width=16, height=16, units="cm", res=300)
+survminer::ggsurvplot(survival::survfit(S.Rec.training ~ Center, data=last.imputed.training.data.set), 
+                      legend.title="",
+                      conf.int=TRUE, risk.table=TRUE)+
+  ggplot2::guides(colour=guide_legend(nrow=2))
+grDevices::dev.off()
+png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.MSM.training.png"), width=16, height=16, units="cm", res=300)
+survminer::ggsurvplot(survival::survfit(S.MSM.training ~ Center, data=last.imputed.training.data.set), 
+                      legend.title="",
+                      conf.int=TRUE, risk.table=TRUE)+
+  ggplot2::guides(colour=guide_legend(nrow=2))
+grDevices::dev.off()
+
 load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData"))
-survminer::ggsurvplot(survival::survfit(S.Rec.validation ~ 1, data=last.imputed.validation.data.set), conf.int=TRUE, risk.table=TRUE)
-survminer::ggsurvplot(survival::survfit(S.MSM.validation ~ 1, data=last.imputed.validation.data.set), conf.int=TRUE, risk.table=TRUE)
+png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.Rec.validation.png"), width=16, height=16, units="cm", res=300)
+survminer::ggsurvplot(survival::survfit(S.Rec.validation ~ 1, data=last.imputed.validation.data.set), 
+                      legend.title="",
+                      legend.labs="Center=Australia",
+                      conf.int=TRUE, risk.table=TRUE)
+grDevices::dev.off()
+png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.MSM.validation.png"), width=16, height=16, units="cm", res=300)
+survminer::ggsurvplot(survival::survfit(S.MSM.validation ~ 1, data=last.imputed.validation.data.set), 
+                      legend.title="",
+                      legend.labs="Center=Australia",
+                      conf.int=TRUE, risk.table=TRUE)
+grDevices::dev.off()
 ```

### 2023-09-13 Code/4_TestNonlinearity.Rmd
Validation added
Update

@@
   SNstatus:multiple.fields
-linear.full.Rec <- Hmisc::fit.mult.impute(form.linear.full.Rec, cph, xtrans=imputed.training.data,
-                                          data=training.data.to.be.imputed,n.impute=m, pr=FALSE,
-                                          fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
+linear.full.Rec <- Hmisc::fit.mult.impute(form.linear.full.Rec, cph, 
+                                          xtrans=imputed.training.data,
+                                          data=training.data.to.be.imputed,
+                                          n.impute=m, pr=FALSE,
+                                          fit.reps=TRUE, 
+                                          fitargs=list(y=TRUE, x=TRUE, 
+                                                       se.fit=TRUE))
 nonlinear.age.full.Rec <- Hmisc::fit.mult.impute(form.nonlinear.age.full.Rec, cph, xtrans=imputed.training.data,
                                                  data=training.data.to.be.imputed,n.impute=m, pr=FALSE,
-                                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
+                                                 fit.reps=TRUE, 
+                                                 fitargs=list(y=TRUE, x=TRUE, 
+                                                       se.fit=TRUE))
 # distribution of Age
@@
                                                      data=training.data.to.be.imputed,n.impute=m, pr=FALSE,
-                                                     fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
+                                                     fit.reps=TRUE, 
+                                                     fitargs=list(y=TRUE, x=TRUE, 
+                                                       se.fit=TRUE))
 # distribution of Breslow
@@
                                                       data=training.data.to.be.imputed,n.impute=m, pr=FALSE,
-                                                      fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
+                                                      fit.reps=TRUE, 
+                                                      fitargs=list(y=TRUE, x=TRUE, 
+                                                       se.fit=TRUE))
 # distribution of Rdamcrit

### 2023-09-13 Code/5_FitModels.Rmd
Update

@@
   SNstatus:multiple.fields
-f.mi.full.Rec.5 <- Hmisc::fit.mult.impute(form.full.Rec.5, cph, xtrans=imputed.training.data,
-                                 data=training.data.to.be.imputed,n.impute=m, pr=FALSE,
-                                 fit.reps=TRUE, y=TRUE, x=TRUE, se.fit=TRUE)
+f.mi.full.Rec.5 <- Hmisc::fit.mult.impute(form.full.Rec.5, cph, 
+                                          xtrans=imputed.training.data,
+                                          data=training.data.to.be.imputed,
+                                          n.impute=m, pr=FALSE, fit.reps=TRUE, 
+                                          fitargs=list(y=TRUE, x=TRUE, 
+                                                       se.fit=TRUE))
 stats::anova(f.mi.full.Rec.5)
@@
 for (i in 1:m){
-  c.Breslow.all <- c(c.Breslow.all, mean(log(mice::complete(imputed.training.data, i)$Breslow)))
+  c.Breslow.all <- c(c.Breslow.all, 
+                     mean(log(mice::complete(imputed.training.data, i)$Breslow)))
 }
@@
   SNstatus:I(log(Breslow)-c.Breslow)
-f.mi.BS.Rec.5 <- Hmisc::fit.mult.impute(form.BS.Rec.5,cph,xtrans=imputed.training.data,data=training.data.to.be.imputed,n.impute=m,
-                           pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
+f.mi.BS.Rec.5 <- Hmisc::fit.mult.impute(form.BS.Rec.5,cph,
+                                        xtrans=imputed.training.data,
+                                        data=training.data.to.be.imputed,
+                                        n.impute=m, pr=FALSE,fit.reps=TRUE,
+                                        fitargs=list(y=TRUE,x=TRUE,se.fit=TRUE))
 # backward selection
@@
   SNstatus:multiple.fields
-f.mi.full.MSM.5.refit <- Hmisc::fit.mult.impute(form.full.MSM.5.refit,cph,xtrans=imputed.training.data,data=training.data.to.be.imputed,n.impute=m,
-                                 pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
+f.mi.full.MSM.5.refit <- Hmisc::fit.mult.impute(form.full.MSM.5.refit,cph,
+                                                xtrans=imputed.training.data,
+                                                data=training.data.to.be.imputed,
+                                                n.impute=m, pr=FALSE,
+                                                fit.reps=TRUE,
+                                                fitargs=list(y=TRUE,x=TRUE,
+                                                             se.fit=TRUE))
 summary(f.mi.full.MSM.5.refit, Tot_SNs_pos=c(0, 1))
@@
   SNstatus:I(log(Breslow)-c.Breslow)
-f.mi.BS.MSM.5.refit <- Hmisc::fit.mult.impute(form.BS.MSM.5.refit,cph,xtrans=imputed.training.data,data=training.data.to.be.imputed,n.impute=m,
-                                  pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE)
+f.mi.BS.MSM.5.refit <- Hmisc::fit.mult.impute(form.BS.MSM.5.refit,cph,
+                                              xtrans=imputed.training.data,
+                                              data=training.data.to.be.imputed,
+                                              n.impute=m,
+                                              pr=FALSE,fit.reps=TRUE,
+                                              fitargs=list(y=TRUE,x=TRUE,
+                                                           se.fit=TRUE))
 summary(f.mi.BS.MSM.5.refit, Breslow=exp(c.Breslow))
@@
 ```
+
+# Redevelop model
+```{r, eval=TRUE}
+form.full.MSM.5.redev <- S.MSM.5.training ~
+  SNstatus +
+  Sex +
+  Age.SN +
+  Ulceration +
+  # Loc_CAT +
+  # Histology +
+  log(Breslow) +
+  # multiple.fields +
+  # Tot_SNs_neg +
+  # Tot_SNs_pos +
+  log(Rdamcrit) +
+  # Dewar +
+  # SNstatus:Sex +
+  SNstatus:Age.SN +
+  # SNstatus:Ulceration +
+  # SNstatus:Loc_CAT +
+  # SNstatus:Histology +
+  SNstatus:log(Breslow) 
+  # SNstatus:multiple.fields
+f.mi.full.MSM.5.redev <- Hmisc::fit.mult.impute(form.full.MSM.5.redev,cph,
+                                                xtrans=imputed.training.data,
+                                                data=training.data.to.be.imputed,
+                                                n.impute=m,pr=FALSE,fit.reps=TRUE,
+                                                fitargs=list(y=TRUE,x=TRUE,se.fit=TRUE))
+# summary(f.mi.full.MSM.5.redev)
+stats::anova(f.mi.full.MSM.5.redev)
+
+# backward selection
+# round(stats::anova(f.mi.full.MSM.5.redev)[order(stats::anova(f.mi.full.MSM.5.redev)[, "P"]),], 3)
+
+# c-index
+p.MSM.5.redev<- matrix(NA, nrow(training.data.to.be.imputed), m)
+f.MSM.basehaz <- survival::basehaz(cph(form.full.MSM.5.redev, 
+                             data=mice::complete(imputed.training.data, 1), 
+                             se.fit=TRUE, x=TRUE, y=TRUE))
+h0.MSM.redev <- f.MSM.basehaz$hazard[f.MSM.basehaz$time==max(f.MSM.basehaz$time[f.MSM.basehaz$time<=horizon])]
+for (i in 1:m){
+  lp.val.i <- predict(f.mi.full.MSM.5.redev,
+                      newdata=mice::complete(imputed.training.data, i),
+                      type="lp")
+  p.MSM.5.redev[, i] <- 1-exp(-h0.MSM.redev*exp(lp.val.i))
+}
+PredictionTools::val.surv.mi(p.MSM.5.redev, S.MSM.5.training, g=5,
+                             main="All patients",time=NULL,lim=c(0,1),dist=TRUE)
+```

### 2023-09-13 Code/6_CreateTablesandPlots.Rmd
Update

@@
       # f.Rec.basehaz.i <- basehaz(f.mi, TRUE)
-      f.Rec.basehaz.i <- basehaz(cph(form.mi, data=mice::complete(imputed.training.data, i), se.fit=TRUE, x=TRUE, y=TRUE))
+      f.Rec.basehaz.i <- survival::basehaz(cph(form.mi, data=mice::complete(imputed.training.data, i), se.fit=TRUE, x=TRUE, y=TRUE))
       h0.Rec.i <- f.Rec.basehaz.i$hazard[f.Rec.basehaz.i$time==max(f.Rec.basehaz.i$time[f.Rec.basehaz.i$time<=horizon])]

### 2023-09-13 Code/7_CreateCrossValidatedCalibrationPlots.Rmd
Update

@@
       form.notj<-update(form.model, S.Rec.5.notj  ~ . )
-      f.mi.Rec.notj<-Hmisc::fit.mult.impute(form.notj,cph,xtrans=imputed.training.data,data=training.data.to.be.imputed,
-                                            n.impute=m,pr=FALSE,fit.reps=TRUE,y=TRUE,x=TRUE,se.fit=TRUE,
+      f.mi.Rec.notj<-Hmisc::fit.mult.impute(form.notj,cph,
+                                            xtrans=imputed.training.data,
+                                            data=training.data.to.be.imputed,
+                                            n.impute=m,pr=FALSE,fit.reps=TRUE,
+                                            fitargs=list(y=TRUE, x=TRUE, se.fit=TRUE),
                                             sub=training.data.to.be.imputed$Center!=Centers[j])

### 2023-10-06 Code/0_PrepareTrainingData.Rmd
Final edits

@@
 # load data file
-original.train.data <- foreign::read.spss(file="Z:/Project Melanoom/PaperMelanoma/Data/training/Dataset Predictiemodel.sav",
+original.training.data <- foreign::read.spss(file="Z:/Project Melanoom/PaperMelanoma/Data/training/Dataset Predictiemodel.sav",
                                     to.data.frame=TRUE, origin="01-01-1970")
@@
 ```{r, eval=TRUE}
-sel.train.data <- original.train.data[original.train.data$EXCLUSIE!="Exclusion", selected.vars]
-cat("Number of patients excluded:", nrow(original.train.data)-nrow(sel.train.data), "\n")
+sel.train.data <- original.training.data[original.training.data$EXCLUSIE!="Exclusion", selected.vars]
+cat("Number of patients excluded:", nrow(original.training.data)-nrow(sel.train.data), "\n")
 ```
@@
 
-#### Describe years
-```{r, eval=TRUE}
-dates <- as.POSIXct(sel.train.data$SNdate, origin="1582-10-14")
-Years <- as.numeric(format(dates, format="%Y"))
-utils::write.table(cbind(table(Years), table(Years, sel.train.data$Recurrence)[,2], 
-                         table(Years, sel.train.data$MSM)[, 2]),
-                   file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Time.Rec.MSM.txt"),
-                   row.names=TRUE, sep=",")
-hist(Years, main="")
-```
-
 # Multiple imputation
@@
                              "Tot_SNs_neg", "Tot_SNs_pos", "Rdamcrit", "Dewar",
-                             "Mitosis",
-                             "Recurrence", "dead", "MSM")
+                             "Mitosis", "Status", "Recurrence", "dead", "MSM")
 training.data.to.be.imputed <- sel.train.data[, variables.to.be.imputed]
@@
 missings.table(training.data.to.be.imputed)
-save(list=c("dd.training", "training.data.to.be.imputed"),file="Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata", compress=TRUE)
+# save(list=c("dd.training", "training.data.to.be.imputed"),file="Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata", compress=TRUE)
 ```

### 2023-10-06 Code/0_PrepareValidationData.Rmd
Final edits

@@
 ```{r, eval=TRUE}
-sel.val.data$Mitosis <- as.numeric(!is.na(sel.val.data$Mitosis))
+sel.val.data$Mitosis <- as.numeric(sel.val.data$Mitosis)
 ```
@@
 sel.val.data$Ulceration <- factor(sel.val.data$Ulceration, labels=c("No", "Yes"))
-sel.val.data$Mitosis <- factor(sel.val.data$Mitosis, labels=c("Yes", "No"))
 
@@
                              "Tot_SNs_neg", "Tot_SNs_pos", "Rdamcrit", "Dewar",
-                             "Mitosis",
-                             "Recurrence", "dead", "MSM")
+                             "Mitosis", "Recurrence", "dead", "MSM")
 validation.data.to.be.imputed <- sel.val.data[, variables.to.be.imputed]

### 2023-10-06 Code/10_AJCC.Rmd
Final edits

@@
       ulceration.patient <- as.character(imputation.i[patient, "Ulceration"])
+      if (type=="training"){
         mitosis.patient <- as.character(imputation.i[patient, "Mitosis"])
+      } else{
+        mitosis.patient <- ifelse(imputation.i[patient, "Mitosis"]<1, "Yes", "No")
+      }
       nodes.patient <- imputation.i[patient, "Tot_SNs_pos"]

### 2023-10-06 Code/1_DescriptivesTable.Rmd
Update
Final edits

@@
 load("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.Rdata")
-descr.table <- arsenal::tableby(Center ~ SNstatus
+validation.data.to.be.imputed$Mitosis <- validation.data.to.be.imputed$Mitosis>0
+validation.data.to.be.imputed$Mitosis <- factor(validation.data.to.be.imputed$Mitosis, labels=c("No", "Yes"))
+pooled.data.to.be.imputed <- rbind(training.data.to.be.imputed, validation.data.to.be.imputed)
+pooled.data.to.be.imputed$Continent <- c(rep("Europe", nrow(training.data.to.be.imputed)), 
+                                         rep("Australie", nrow(validation.data.to.be.imputed)))
+
+descr.table <- arsenal::tableby(Continent ~ SNstatus
                                 +Sex
@@
                                 +multiple.fields
-                                +Tot_SNs_neg
-                                +Tot_SNs_pos
-                                +Rdamcrit
-                                +arsenal::includeNA(Dewar)
                                 +arsenal::includeNA(Mitosis)
@@
                                 +logRec.time
-                                +logFU.time, data=rbind(training.data.to.be.imputed, validation.data.to.be.imputed), # training.data.to.be.imputed,
+                                +logFU.time, 
+                                data=pooled.data.to.be.imputed,
+                                total.pos="before",
+                                digits = 1,
+                                test = F,
+                                total = F,
+                                numeric.test = "kwt", cat.test = "chisq",
+                                numeric.stats = c("meansd", "medianq1q3", "range", "Nmiss2"),
+                                cat.stats = c("countpct"),
+                                stats.labels = list(
+                                  meansd = "Mean (SD)",
+                                  medianq1q3 = "Median (Q1, Q3)",
+                                  range = "Min - Max",
+                                  Nmiss2 = "Missing"
+                                ))
+
+descr.table.pos <- arsenal::tableby(Continent ~ Tot_SNs_pos
+                                +Rdamcrit
+                                +arsenal::includeNA(Dewar),
+                                data=pooled.data.to.be.imputed[pooled.data.to.be.imputed$SNstatus=="Positive",],
+                                total.pos="before",
+                                digits = 1,
+                                test = F,
+                                total = T,
+                                numeric.test = "kwt", cat.test = "chisq",
+                                numeric.stats = c("meansd", "medianq1q3", "range", "Nmiss2"),
+                                cat.stats = c("countpct"),
+                                stats.labels = list(
+                                  meansd = "Mean (SD)",
+                                  medianq1q3 = "Median (Q1, Q3)",
+                                  range = "Min - Max",
+                                  Nmiss2 = "Missing"
+                                ))
+
+descr.table.neg <- arsenal::tableby(Continent ~ Tot_SNs_neg,
+                                data=pooled.data.to.be.imputed[pooled.data.to.be.imputed$SNstatus=="Negative",],
                                 total.pos="before",
@@
 knitr::kable(summary(descr.table, labelTranslations=my_labels))
+
+# only negative
+knitr::kable(summary(descr.table.neg, labelTranslations=my_labels))
+
+# only positive
+knitr::kable(summary(descr.table.pos, labelTranslations=my_labels))
 ```

### 2023-10-06 Code/3_Descriptives.Rmd
Final edits

--- 2023-01-05 Code/3_Descriptives.Rmd
@@
 ```
-
-### Kaplan-Meier plots
-```{r, eval=TRUE}
-load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData"))
-survminer::ggsurvplot(survival::survfit(S.Rec.training ~ Center, data=last.imputed.training.data.set), conf.int=TRUE, risk.table=TRUE)
-survminer::ggsurvplot(survival::survfit(S.MSM.training ~ Center, data=last.imputed.training.data.set), conf.int=TRUE, risk.table=TRUE)
-load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData"))
-survminer::ggsurvplot(survival::survfit(S.Rec.validation ~ 1, data=last.imputed.validation.data.set), conf.int=TRUE, risk.table=TRUE)
-survminer::ggsurvplot(survival::survfit(S.MSM.validation ~ 1, data=last.imputed.validation.data.set), conf.int=TRUE, risk.table=TRUE)
-```
-
 # Median time to recurrence and follow-up time
@@
   }
-
   # save in tab.time
-  tab.time <- c("Total Development", paste(sprintf("%.2f", median.time.total$quantiles.survival[3, "quantile"]),
-                   paste0("[", sprintf("%.2f", median.time.total$quantiles.survival[4, "quantile"]), "; ",
-                          sprintf("%.2f", median.time.total$quantiles.survival[2, "quantile"]), "]")))
+  tab.time <- c("Total Development", paste(sprintf("%.2f", median.time.total$quantile[3]),
+                   paste0("[", sprintf("%.2f", median.time.total$quantile[4]), "; ",
+                          sprintf("%.2f", median.time.total$quantile[2]), "]")))
 
@@
   for (center in c("Amsterdam", "Berlin", "Rotterdam", "Warsaw", "Australia")){
-    med.table <- eval(parse(text=paste0("median.time$`Center=", center, "`")))
-    med <- med.table[3,"quantile"]
-    lower <- med.table[4,"quantile"]
-    upper <- med.table[2,"quantile"]
+    sel.center <- eval(parse(text=paste0("median.time$Center==\"", center, "\"")))
+    q.center <- median.time$quantile[sel.center]
+    med <- q.center[3]
+    lower <- q.center[4]
+    upper <- q.center[2]
     tab.time <- rbind(tab.time, c(center, paste(sprintf("%.2f", med), paste0("[", sprintf("%.2f", lower), "; ", sprintf("%.2f", upper), "]"))))
@@
 }
+load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData"))
+load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData"))
 med.Rec <- median.time(data=rbind(training.data.to.be.imputed[, c("logRec.time", "Recurrence", "Center")], 
@@
   SF.Rec.all <- survival::survfit(S.Rec~1, data=data.to.be.imputed)
-  Rec.all <- 100*(1-SF.Rec.all$surv[SF.Rec.all$time==max(SF.Rec.all$time[SF.Rec.all$time<=horizon])])
+  RFS.all <- 100*SF.Rec.all$surv[SF.Rec.all$time==max(SF.Rec.all$time[SF.Rec.all$time<=horizon])]
   n.Rec.all <- sum(SF.Rec.all$n.event[SF.Rec.all$time<=horizon])
@@
   SF.Rec.neg <-survival::survfit(S.Rec[data.to.be.imputed$SNstatus=="Negative"]~1, data=data.to.be.imputed)
-  Rec.neg <- 100*(1-SF.Rec.neg$surv[SF.Rec.neg$time==max(SF.Rec.neg$time[SF.Rec.neg$time<=horizon])])
+  RFS.neg <- 100*SF.Rec.neg$surv[SF.Rec.neg$time==max(SF.Rec.neg$time[SF.Rec.neg$time<=horizon])]
   n.Rec.neg <- sum(SF.Rec.neg$n.event[SF.Rec.neg$time<=horizon])
@@
   SF.Rec.pos <- survival::survfit(S.Rec[data.to.be.imputed$SNstatus=="Positive"]~1, data=data.to.be.imputed)
-  Rec.pos <- 100*(1-SF.Rec.pos$surv[SF.Rec.pos$time==max(SF.Rec.pos$time[SF.Rec.pos$time<=horizon])])
+  RFS.pos <- 100*SF.Rec.pos$surv[SF.Rec.pos$time==max(SF.Rec.pos$time[SF.Rec.pos$time<=horizon])]
   n.Rec.pos <- sum(SF.Rec.pos$n.event[SF.Rec.pos$time<=horizon])
@@
   SF.MSM.all <- survival::survfit(S.MSM~1, data=data.to.be.imputed)
-  MSM.all <- 100*(1-SF.MSM.all$surv[SF.MSM.all$time==max(SF.MSM.all$time[SF.MSM.all$time<=horizon])])
+  MSS.all <- 100*SF.MSM.all$surv[SF.MSM.all$time==max(SF.MSM.all$time[SF.MSM.all$time<=horizon])]
   n.MSM.all <- sum(SF.MSM.all$n.event[SF.MSM.all$time<=horizon])
@@
   SF.MSM.neg <- survival::survfit(S.MSM[data.to.be.imputed$SNstatus=="Negative"]~1, data=data.to.be.imputed)
-  MSM.neg <- 100*(1-SF.MSM.neg$surv[SF.MSM.neg$time==max(SF.MSM.neg$time[SF.MSM.neg$time<=horizon])])
+  MSS.neg <- 100*SF.MSM.neg$surv[SF.MSM.neg$time==max(SF.MSM.neg$time[SF.MSM.neg$time<=horizon])]
   n.MSM.neg <- sum(SF.MSM.neg$n.event[SF.MSM.neg$time<=horizon])
@@
   SF.MSM.pos <- survival::survfit(S.MSM[data.to.be.imputed$SNstatus=="Positive"]~1, data=data.to.be.imputed)
-  MSM.pos <- 100*(1-SF.MSM.pos$surv[SF.MSM.pos$time==max(SF.MSM.pos$time[SF.MSM.pos$time<=horizon])])
+  MSS.pos <- 100*SF.MSM.pos$surv[SF.MSM.pos$time==max(SF.MSM.pos$time[SF.MSM.pos$time<=horizon])]
   n.MSM.pos <- sum(SF.MSM.pos$n.event[SF.MSM.pos$time<=horizon])
@@
                               e2=rep("", 3),
-                              Rec=sprintf("%.1f", c(Rec.all, Rec.neg, Rec.pos)),
-                              MSM=sprintf("%.1f", c(MSM.all, MSM.neg, MSM.pos)))
+                              RFS=sprintf("%.1f", c(RFS.all, RFS.neg, RFS.pos)),
+                              MSS=sprintf("%.1f", c(MSS.all, MSS.neg, MSS.pos)))
   row.names(Rec.MSM.table) <- c("All", "Negative", "Positive")
@@
 ```
+### Kaplan-Meier plots
+```{r, eval=TRUE}
+png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.Rec.training.png"), width=16, height=16, units="cm", res=300)
+survminer::ggsurvplot(survival::survfit(S.Rec.5.training ~ Center, data=last.imputed.training.data.set), 
+                      legend.title="",
+                      legend.labs=c("Rotterdam", "Berlin", "Warsaw", "Amsterdam"),
+                      title="Development",
+                      ylab="Recurrence-free survival probability",
+                      xlab="Time in years",
+                      conf.int=TRUE, risk.table=TRUE)+
+  ggplot2::guides(colour=guide_legend(nrow=2))
+grDevices::dev.off()
+png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.MSM.training.png"), width=16, height=16, units="cm", res=300)
+survminer::ggsurvplot(survival::survfit(S.MSM.5.training ~ Center, data=last.imputed.training.data.set), 
+                      legend.title="",
+                      legend.labs=c("Rotterdam", "Berlin", "Warsaw", "Amsterdam"),
+                      title="Development",
+                      ylab="Melanoma-specific survival probability",
+                      xlab="Time in years",
+                      conf.int=TRUE, risk.table=TRUE)+
+  ggplot2::guides(colour=guide_legend(nrow=2))
+grDevices::dev.off()
+
+load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData"))
+png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.Rec.validation.png"), width=16, height=16, units="cm", res=300)
+survminer::ggsurvplot(survival::survfit(S.Rec.5.validation ~ 1, data=last.imputed.validation.data.set), 
+                      legend.title="",
+                      legend.labs="Australia",
+                      title="Validation",
+                      ylab="Recurrence-free survival probability",
+                      xlab="Time in years",
+                      conf.int=TRUE, risk.table=TRUE)
+grDevices::dev.off()
+png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.MSM.validation.png"), width=16, height=16, units="cm", res=300)
+survminer::ggsurvplot(survival::survfit(S.MSM.5.validation ~ 1, data=last.imputed.validation.data.set), 
+                      legend.title="",
+                      legend.labs="Australia",
+                      title="Validation",
+                      ylab="Melanoma-specific survival probability",
+                      xlab="Time in years",
+                      conf.int=TRUE, risk.table=TRUE)
+grDevices::dev.off()
+```

### 2023-10-06 Code/6_CreateTablesandPlots.Rmd
Final edits

@@
   neg.results.BS <- summary(f.mi.BS, SNstatus="Negative", Breslow=exp(c.Breslow), Loc_CAT="arm")
-  neg.table.BS <- rbind(rep(NA, 3), # SN status
+  pos.results.BS <- summary(f.mi.BS, SNstatus="Positive", Breslow=exp(c.Breslow), Loc_CAT="arm")
+  table.BS <- rbind(exp(neg.results.BS["SNstatus - Positive:Negative", c("Effect", "Lower 0.95", "Upper 0.95")]),
                     exp(neg.results.BS[c("Age.SN",
@@
                                              "Breslow"), c("Effect", "Lower 0.95", "Upper 0.95")]),
-                        rep(NA, 3)) # Rdamcrit
-  # add CI
-  neg.table.BS <- cbind(sprintf("%.2f", neg.table.BS[, 1]),
-                        paste0("[", sprintf("%.2f", neg.table.BS[, 2]), "; ", sprintf("%.2f", neg.table.BS[, 3]), "]"))
-
-  # Positive results
-  pos.results.BS <- summary(f.mi.BS, SNstatus="Positive", Breslow=exp(c.Breslow), Loc_CAT="arm")
-  pos.table.BS <- rbind(exp(neg.results.BS["SNstatus - Positive:Negative", c("Effect", "Lower 0.95", "Upper 0.95")]),
-                        exp(pos.results.BS[c("Age.SN",
-                                             "Ulceration - Yes:No",
-                                             "Loc_CAT - leg:arm",
-                                             "Loc_CAT - trunk:arm",
-                                             "Loc_CAT - head & neck:arm",
-                                             "Breslow",
+                    exp(pos.results.BS[c("Breslow",
                                          "Rdamcrit"), c("Effect", "Lower 0.95", "Upper 0.95")]))
   # add CI
-  pos.table.BS <- cbind(sprintf("%.2f", pos.table.BS[, 1]),
-                        paste0("[", sprintf("%.2f", pos.table.BS[, 2]), "; ", sprintf("%.2f", pos.table.BS[, 3]), "]"))
-
-  # combine postive with negative
-  BS.table <- cbind(neg.table.BS, pos.table.BS)
+  BS.table <- cbind(sprintf("%.2f", table.BS[, 1]),
+                        paste0("[", sprintf("%.2f", table.BS[, 2]), "; ", 
+                               sprintf("%.2f", table.BS[, 3]), "]"))
 
   # add reference rows
-  BS.table <- rbind(BS.table[1:3,],
-                    rep(NA, 4),
-                    c("(ref)", NA, "(ref)", NA),
-                    BS.table[4:nrow(BS.table),])
+  BS.table <- rbind(rep(NA, 2),             # SN status title
+                    c("(ref)", NA),         # Negative SN status
+                    BS.table[1:3,],         # Positive SN status, Age, Ulceration
+                    rep(NA, 2),             # Location title
+                    c("(ref)", NA),         # Upper limb ref
+                    BS.table[4:6,],         # Lower limb, tunk, head&neck
+                    rep(NA, 2),             # Breslow title
+                    BS.table[7:9,])         # Breslow negative and positive, tumour burden
 
@@
   anova.out <- anova(f.mi.BS)[, "Chi-Square"]
-  anova.main <- c(anova.out[c("SNstatus  (Factor+Higher Order Factors)",
-                              "Age.SN",
-                              "Ulceration",
-                              "Loc_CAT")],
-                  rep(NA, 4),
-                  anova.out[c("Breslow  (Factor+Higher Order Factors)",
-                              "Rdamcrit")])
-  anova.int <- c(rep(NA, 8),
-                 anova.out["SNstatus * Breslow  (Factor+Higher Order Factors)"],
-                 NA)
-  BS.table <- cbind(BS.table, sprintf("%.0f", anova.main), sprintf("%.0f", anova.int))
+  anova.main <- c(anova.out[c("SNstatus  (Factor+Higher Order Factors)")], # SN status
+                  NA,               # negative SN
+                  NA,               # positive SN
+                  anova.out[c("Age.SN",     # Age
+                              "Ulceration", # Ulceration
+                              "Loc_CAT")],  # Location
+                  NA,               # upper limb
+                  NA,               # lower limb
+                  NA,               # trunk
+                  NA,               # head & neck
+                  NA,               # Breslow title
+                  anova.out[c("Breslow  (Factor+Higher Order Factors)")], # Breslow negative
+                  anova.out["SNstatus * Breslow  (Factor+Higher Order Factors)"], # Breslow positive
+                  anova.out[c("Rdamcrit")])
+  BS.table <- cbind(BS.table, sprintf("%.0f", anova.main))
 
-  colnames(BS.table) <- c("HR.neg", "CI.neg", "HR.pos", "CI.pos", "Main", "Int")
+  colnames(BS.table) <- c("HR", "CI", "Imp.")
   assign(paste0("final.BS.table.", model), as.data.frame(BS.table))
@@
                          adj.subtitle=FALSE, lwd=2, labels=FALSE,
-                         par.settings=list(superpose.line=list(lty=c(1, 2), col=c("red", "blue"))),
-                         key=list(cex=.8,corner = c(0.1, .9),lines=list(col=c("red","blue"), lty=c(1,2), lwd=2),text=list(c("Negative","Positive"))))
+                         par.settings=list(superpose.line=list(lty=c(1, 2), 
+                                                               col=c("red", "blue"))),
+                         key=list(cex=.8,corner = c(0.1, .9),
+                                  lines=list(col=c("red","blue"), 
+                                             lty=c(1,2), lwd=2),
+                                  text=list(c("Negative","Positive"))))
     show(plot.Breslow)
@@
       # f.Rec.basehaz.i <- basehaz(f.mi, TRUE)
-      f.Rec.basehaz.i <- survival::basehaz(cph(form.mi, data=mice::complete(imputed.training.data, i), se.fit=TRUE, x=TRUE, y=TRUE))
+      f.Rec.basehaz.i <- survival::basehaz(cph(form.mi, 
+                                               data=mice::complete(imputed.training.data, i), 
+                                               se.fit=TRUE, x=TRUE, y=TRUE))
       h0.Rec.i <- f.Rec.basehaz.i$hazard[f.Rec.basehaz.i$time==max(f.Rec.basehaz.i$time[f.Rec.basehaz.i$time<=horizon])]
@@
     final.full.table <- eval(parse(text=paste0("final.", type, ".table.", model)))
-    final.full.table["C-index", "HR.neg"] <- sprintf("%.2f", out$cindex-optimism)
-    final.full.table["C-index", "CI.neg"] <- paste0("[", sprintf("%.2f", out$cindex.lower-optimism),
+    final.full.table["C-index", "HR"] <- sprintf("%.2f", out$cindex-optimism)
+    final.full.table["C-index", "CI"] <- paste0("[", sprintf("%.2f", out$cindex.lower-optimism),
                                                    "; ", sprintf("%.2f", out$cindex.upper-optimism), "]")
@@
   f.mi <- eval(parse(text=paste0("f.mi.", type, ".Rec.5")))
-  cindex.MSM.refit <- rep(0, m)
-  cindex.se.MSM.refit <- rep(0, m)
+  cindex.MSM.cal <- rep(0, m)
+  cindex.se.MSM.cal <- rep(0, m)
   for (i in 1:m){
-    rc <- Hmisc::rcorr.cens(-predict(f.mi,newdata=mice::complete(imputed.training.data,i), type="lp"), S.MSM.5.training)
-    cindex.MSM.refit[i]<-rc["C Index"]
-    cindex.se.MSM.refit[i]<-rc["S.D."]/2
+    rc <- Hmisc::rcorr.cens(-predict(f.mi,
+                                     newdata=mice::complete(imputed.training.data,i), type="lp"), 
+                            S.MSM.5.training)
+    cindex.MSM.cal[i] <- rc["C Index"]
+    cindex.se.MSM.cal[i] <- rc["S.D."]/2
   }
-  cindex.mi.MSM.refit<-Rubin.combine(cindex.MSM.refit,cindex.se.MSM.refit)
+  cindex.mi.MSM.cal <- Rubin.combine(cindex.MSM.cal, cindex.se.MSM.cal)
 
   final.full.table <- eval(parse(text=paste0("final.", type, ".table.Rec.5")))
-  final.full.table["C-index MSM", "HR.neg"] <- sprintf("%.2f", cindex.mi.MSM.refit$est-optimism)
-  final.full.table["C-index MSM", "CI.neg"] <- paste0("[", sprintf("%.2f", cindex.mi.MSM.refit$est+qnorm(.025)*cindex.mi.MSM.refit$se-optimism),
-                                                 "; ", sprintf("%.2f", cindex.mi.MSM.refit$est+qnorm(.975)*cindex.mi.MSM.refit$se-optimism), "]")
+  final.full.table["C-index MSM", "HR"] <- sprintf("%.2f", cindex.mi.MSM.cal$est-optimism)
+  final.full.table["C-index MSM", "CI"] <- paste0("[", sprintf("%.2f", cindex.mi.MSM.cal$est+qnorm(.025)*cindex.mi.MSM.cal$se-optimism),
+                                                 "; ", sprintf("%.2f", cindex.mi.MSM.cal$est+qnorm(.975)*cindex.mi.MSM.cal$se-optimism), "]")
 
@@
 final.BS.table.Rec.5 <- as.data.frame(final.BS.table.Rec.5)
-row.names(final.BS.table.Rec.5) <- c("Positive SN status", "Age", "Ulceration",
-                                   "Location", "Arm", "Leg", "Trunk", 
+row.names(final.BS.table.Rec.5) <- c("SN status",
+                                     "Negative SN status",
+                                     "Positive SN status", 
+                                     "Age", 
+                                     "Ulceration",
+                                     "Location", 
+                                     "Upper limb", 
+                                     "Lower limb", 
+                                     "Trunk", 
                                      "Head and neck",
-                                   "Breslow", "SN tumour burden", 
+                                     "Breslow",
+                                     "Breslow (negative)",
+                                     "Breslow (positive)",
+                                     "SN tumour burden", 
                                    row.names(final.BS.table.Rec.5)[(nrow(final.BS.table.Rec.5)-1):nrow(final.BS.table.Rec.5)])
 final.BS.table.MSM.5.refit <- as.data.frame(final.BS.table.MSM.5.refit)
-row.names(final.BS.table.MSM.5.refit) <- c("Positive SN status", "Age", "Ulceration",
-                                   "Location", "Arm", "Leg", "Trunk", 
+row.names(final.BS.table.MSM.5.refit) <- c("SN status",
+                                     "Negative SN status",
+                                     "Positive SN status", 
+                                     "Age", 
+                                     "Ulceration",
+                                     "Location", 
+                                     "Upper limb", 
+                                     "Lower limb", 
+                                     "Trunk", 
                                      "Head and neck",
-                                   "Breslow", "SN tumour burden", 
+                                     "Breslow",
+                                     "Breslow (negative)",
+                                     "Breslow (positive)",
+                                     "SN tumour burden",  
                                    row.names(final.BS.table.MSM.5.refit)[nrow(final.BS.table.MSM.5.refit)])
@@
   for (i in 1:m){
-    subset.imputed.training.data <- as.data.frame(mice::complete(imputed.training.data,i)[last.imputed.training.data.set$SNstatus==arm,])
-    rc <- Hmisc::rcorr.cens(-predict(f.mi.stratified,newdata=subset.imputed.training.data, type="lp"), S.5.subset)
-    cindex.stratified[i]<-rc["C Index"]
-    cindex.se.stratified[i]<-rc["S.D."]/2
+    subset.imputed.training.data <- as.data.frame(mice::complete(imputed.training.data, i)[last.imputed.training.data.set$SNstatus==arm,])
+    rc <- Hmisc::rcorr.cens(-predict(f.mi.stratified,
+                                     newdata=subset.imputed.training.data, 
+                                     type="lp"), 
+                            S.5.subset)
+    cindex.stratified[i] <- rc["C Index"]
+    cindex.se.stratified[i] <- rc["S.D."]/2
   }
-  cindex.mi.stratified<-Rubin.combine(cindex.stratified,cindex.se.stratified)
+  cindex.mi.stratified <- Rubin.combine(cindex.stratified, cindex.se.stratified)
 

### 2023-10-06 Code/8_CreateNomogram.Rmd
Validation added
Final edits

@@
 attributes(nom)$names[attributes(nom)$names=="Loc_CAT"]<-"Location"
-attributes(nom)$names[attributes(nom)$names=="Rdamcrit"]<-"Tumor burden"
-attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Positive)"]<-"Breslow + positive SN"
-attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Negative)"]<-"Breslow + negative SN"
+attributes(nom)$names[attributes(nom)$names=="Rdamcrit"]<-"Tumour burden"
+attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Positive)"]<-"Breslow (positive SN)"
+attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Negative)"]<-"Breslow (negative SN)"
 # save width: 1000, height: 800

### 2023-10-06 Code/AdditionalAnalysis_MitosisValidation.Rmd
Final edits

---
title: "Fit models and plots"
author: "Carolien C.H.M. Maas"
date: "January 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load packages, functions, and data 
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# erase memory
rm(list = ls(all.names = TRUE))

# for reproducibility
set.seed(100)

# load libraries
library(mice)
library(dplyr)
library(rms)
```

# Full model
```{r, eval=TRUE}
load("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData")

# set datadist
options(datadist="dd.validation")

# define the outcome based on last imputed data set
last.imputed.data.set <- mice::complete(imputed.validation.data, m)
S.Rec.validation <- survival::Surv(exp(last.imputed.data.set$logRec.time), last.imputed.data.set$Recurrence=="recurred")
S.Rec.5.validation <- S.Rec.validation
S.Rec.5.validation[S.Rec.validation[,1]>5, 1] <- 5 # set time to 5
S.Rec.5.validation[S.Rec.validation[,1]>5, 2] <- 0 # set censoring

# fit model
form.full.Rec.5.Mitosis.log <- S.Rec.5.validation ~
  SNstatus +
  Sex +
  Age.SN +
  Ulceration +
  Loc_CAT +
  Histology +
  log(Breslow) +
  multiple.fields +
  log(Mitosis+1) +
  Tot_SNs_neg +
  Tot_SNs_pos +
  log(Rdamcrit) +
  Dewar +
  SNstatus:Sex +
  SNstatus:Age.SN +
  SNstatus:Ulceration +
  SNstatus:Loc_CAT +
  SNstatus:Histology +
  SNstatus:log(Breslow) +
  SNstatus:multiple.fields +
  SNstatus:log(Mitosis+1)
f.mi.full.Rec.5.Mitosis.log <- Hmisc::fit.mult.impute(form.full.Rec.5.Mitosis.log, cph, 
                                          xtrans=imputed.validation.data,
                                          data=validation.data.to.be.imputed,
                                          n.impute=m, pr=FALSE, fit.reps=TRUE, 
                                          fitargs=list(y=TRUE, x=TRUE, 
                                                       se.fit=TRUE))
f.mi.full.Rec.5.Mitosis.log
stats::anova(f.mi.full.Rec.5.Mitosis.log)
plot(Predict(f.mi.full.Rec.5.Mitosis.log))

C.Mitosis.log.full <- c()
for (i in 1:m){
  C.Mitosis.log.full <- c(C.Mitosis.log.full, Hmisc::rcorr.cens(-predict(f.mi.full.Rec.5.Mitosis.log,
                                                       newdata=mice::complete(imputed.validation.data,i), type="lp"), 
                                                       S.Rec.5.validation)["C Index"])
}
mean(C.Mitosis.log.full)
```

# Final model after backward selection
```{r, eval=TRUE}
form.BS.Rec.5.Mitosis.log <- S.Rec.5.validation ~
  SNstatus +
  Sex +
  Age.SN +
  Ulceration +
  Loc_CAT +
  # Histology +
  log(Breslow) +
  # multiple.fields +
  log(Mitosis+1) +
  # Tot_SNs_neg +
  # Tot_SNs_pos +
  log(Rdamcrit) 
  # Dewar +
  # SNstatus:Sex +
  # SNstatus:Age.SN +
  # SNstatus:Ulceration +
  # SNstatus:Loc_CAT +
  # SNstatus:Histology +
  # SNstatus:log(Breslow) +
  # SNstatus:multiple.fields +
  # SNstatus:log(Mitosis+1)
f.mi.BS.Rec.5.Mitosis.log <- Hmisc::fit.mult.impute(form.BS.Rec.5.Mitosis.log, cph, 
                                          xtrans=imputed.validation.data,
                                          data=validation.data.to.be.imputed,
                                          n.impute=m, pr=FALSE, fit.reps=TRUE, 
                                          fitargs=list(y=TRUE, x=TRUE, 
                                                       se.fit=TRUE))
f.mi.BS.Rec.5.Mitosis.log
stats::anova(f.mi.BS.Rec.5.Mitosis.log)
plot(Predict(f.mi.BS.Rec.5.Mitosis.log))

C.Mitosis.log <- c()
for (i in 1:m){
  C.Mitosis.log <- c(C.Mitosis.log, Hmisc::rcorr.cens(-predict(f.mi.BS.Rec.5.Mitosis.log,
                                                       newdata=mice::complete(imputed.validation.data,i), type="lp"), 
                                                       S.Rec.5.validation)["C Index"])
}
mean(C.Mitosis.log)
```

```{r,eval=TRUE}
results.BS <- summary(f.mi.BS.Rec.5.Mitosis.log, SNstatus="Negative", Loc_CAT="arm")
table.BS <- rbind(exp(results.BS[c("SNstatus - Positive:Negative",
                                 "Mitosis", 
                                 "Age.SN",
                                 "Sex - female:male",
                                 "Ulceration - Yes:No",
                                 "Loc_CAT - leg:arm",
                                 "Loc_CAT - trunk:arm",
                                 "Loc_CAT - head & neck:arm",
                                 "Breslow",
                                 "Rdamcrit"), c("Effect", "Lower 0.95", "Upper 0.95")]))
BS.table <- cbind(sprintf("%.3f", table.BS[, 1]),
                      paste0("[", sprintf("%.3f", table.BS[, 2]), "; ", 
                             sprintf("%.3f", table.BS[, 3]), "]"))

# add reference rows
BS.table <- rbind(rep(NA, 2),             # SN status title
                  c("(ref)", NA),         # Negative SN status
                  BS.table[1:5,],         # Positive SN status, Mitosis, Age, Sex, Ulceration
                  rep(NA, 2),             # Location title
                  c("(ref)", NA),         # Upper limb ref
                  BS.table[6:10,])        # Lower limb, tunk, head&neck, Breslow, tumour burden

# add Chi-square statistics
anova.out <- anova(f.mi.BS.Rec.5.Mitosis.log)[, "Chi-Square"]
anova.main <- c(anova.out[c("SNstatus")], # SN status
                NA,               # negative SN
                NA,               # positive SN
                anova.out[c("Mitosis",
                            "Age.SN",     # Age
                            "Sex",
                            "Ulceration", # Ulceration
                            "Loc_CAT")],  # Location
                NA,               # upper limb
                NA,               # lower limb
                NA,               # trunk
                NA,               # head & neck
                anova.out[c("Breslow")], # Breslow
                anova.out[c("Rdamcrit")])
BS.table <- cbind(BS.table, sprintf("%.0f", anova.main))

colnames(BS.table) <- c("HR", "CI", "Imp.")
rownames(BS.table) <- c("SN status",
                         "Negative SN status",
                         "Positive SN status", 
                         "Mitosis (2 vs 7)",
                         "Age (46 vs. 68)", 
                         "Females",
                         "Ulceration",
                         "Location", 
                         "Upper limb", 
                         "Lower limb", 
                         "Trunk", 
                         "Head and neck",
                         "Breslow (1.3 vs 3.2)",
                         "SN tumour burden (0.01 vs. 23.0)")
BS.table[BS.table=="NA"] <- NA
BS.table[is.na(BS.table)] <- ""
BS.table
openxlsx::write.xlsx(as.data.frame(BS.table),
            rowNames=TRUE,
            file="Z:/Project Melanoom/PaperMelanoma/Results/Review/Final.Model.Australian.xlsx")
```
# Final model without Mitosis
```{r, eval=TRUE}
form.BS.Rec.5.without.Mitosis <- S.Rec.5.validation ~
  SNstatus +
  Sex + 
  Age.SN +
  Ulceration +
  Loc_CAT +
  log(Breslow) +
  log(Rdamcrit)
f.mi.BS.Rec.5.without.Mitosis <- Hmisc::fit.mult.impute(form.BS.Rec.5.without.Mitosis, cph, 
                                          xtrans=imputed.validation.data,
                                          data=validation.data.to.be.imputed,
                                          n.impute=m, pr=FALSE, fit.reps=TRUE, 
                                          fitargs=list(y=TRUE, x=TRUE, 
                                                       se.fit=TRUE))

C.without.Mitosis <- c()
for (i in 1:m){
  C.without.Mitosis <- c(C.without.Mitosis, Hmisc::rcorr.cens(-predict(f.mi.BS.Rec.5.without.Mitosis,
                                                       newdata=mice::complete(imputed.validation.data,i), type="lp"), 
                                                       S.Rec.5.validation)["C Index"])
}
mean(C.without.Mitosis)
``` 
# Test model with or without Mitosis
```{r, eval=TRUE}
AIC(f.mi.BS.Rec.5.Mitosis.log)
AIC(f.mi.BS.Rec.5.without.Mitosis)
LR.test.log <- lrtest(f.mi.BS.Rec.5.Mitosis.log, f.mi.BS.Rec.5.without.Mitosis)
LR.test.log
```

# Calculate optimism using bootstrap using full model and doing backward selection for each bootstrap
```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(100)
slope<-NULL
cindex.orig<-NULL
opt<-NULL
cindex.B<-NULL
for (i in 1:m){
  v<-validate(f.mi.full.Rec.5.Mitosis.log$fits[[i]],method="boot",bw=TRUE,rule="p",sls=0.05,B=100,pr=FALSE,type="individual")
  slope<-c(slope,v["Slope","test"]) # slope of test
  cindex.orig<-c(cindex.orig,(v["Dxy","index.orig"]+1)/2) # original C-index
  opt<-c(opt,v["Dxy","optimism"]/2) # optimism C-index
  cindex.B<-c(cindex.B,(v["Dxy","index.corrected"]+1)/2) # index corrected C-index
}
mean(cindex.orig)
optimism <- mean(opt)
optimism
mean(cindex.B)
mean(cindex.orig)-optimism
```
```{r, eval=TRUE}
load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData")
m <- 5
horizon <- 5
p.Rec.5.Europe <- matrix(NA, nrow(training.data.to.be.imputed), m)
p.Rec.5.Australia <- matrix(NA, nrow(validation.data.to.be.imputed), m)
for (i in 1:m){
  f.Rec.basehaz.i <- survival::basehaz(cph(form.BS.Rec.5.Mitosis.log, 
                                               data=mice::complete(imputed.validation.data, i), 
                                               se.fit=TRUE, x=TRUE, y=TRUE))
  h0.Rec.i <- f.Rec.basehaz.i$hazard[f.Rec.basehaz.i$time==max(f.Rec.basehaz.i$time[f.Rec.basehaz.i$time<=horizon])]

  lp.AUS.i <- predict(f.mi.BS.Rec.5.Mitosis.log,
                      newdata=mice::complete(imputed.validation.data, i),
                      type="lp")
  p.Rec.5.Australia[, i] <- 1-exp(-h0.Rec.i*exp(lp.AUS.i))

  imputed.training.data.i <- mice::complete(imputed.training.data, i)
  imputed.training.data.i$Mitosis <- as.numeric(imputed.training.data.i$Mitosis=="Yes")

  lp.EU.i <- predict(f.mi.BS.Rec.5.Mitosis.log,
                      newdata=imputed.training.data.i,
                      type="lp")
  p.Rec.5.Europe[, i] <- 1-exp(-h0.Rec.i*exp(lp.EU.i))
}
PredictionTools::val.surv.mi(p.Rec.5.Australia, S.Rec.5.validation,
                             g=5,main="Performance of Australian model for Australian patients", time=NULL,lim=c(0,1),dist=TRUE)
PredictionTools::val.surv.mi(p.Rec.5.Europe, S.Rec.5.training,
                             g=5,main="Performance of Australian model for European patients", time=NULL,lim=c(0,1),dist=TRUE)
```

### 2023-10-06 Code/AdditionalSensitivityAnalysis_CheckYears.Rmd
Final edits

---
title: "Sensitivity Analysis"
author: "Carolien C.H.M. Maas"
date: "January 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Fit models

### Load packages, functions, and data
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# erase memory
rm(list = ls(all.names = TRUE))

# for reproducibility
set.seed(100)

library(ggplot2)
library(rms)
```

# Load data 
```{r, eval=TRUE}
load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata")
original.training.data <- foreign::read.spss(file="Z:/Project Melanoom/PaperMelanoma/Data/training/Dataset Predictiemodel.sav",
                                    to.data.frame=TRUE, 
                                    origin="01-01-1970")
load("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.Rdata")
original.validation.data <- foreign::read.spss(file="Z:/Project Melanoom/PaperMelanoma/Data/validation/Validation_cohort_MIA.sav",
                                               to.data.frame=TRUE, 
                                               origin="01-01-1970")
```
# Inclusion development data
```{r, eval=TRUE}
sel.train.data <- original.training.data[original.training.data$EXCLUSIE!="Exclusion", ]
sel.train.data <- sel.train.data[!is.na(sel.train.data$Last_FU),]
sel.train.data$Rec.time <- (sel.train.data$Date_1st_Rec-sel.train.data$SNdate)/(24*60*60)/365.25
sel.train.data <- sel.train.data[!(sel.train.data$Rec.time<0&!is.na(sel.train.data$Rec.time)),]
```
# Describe years
```{r, eval=TRUE}
dates.train <- as.POSIXct(sel.train.data$SNdate, origin="1582-10-15")
dates.val <- as.POSIXct(original.validation.data$SNdate, origin="1582-10-15")
Years.train <- as.numeric(format(dates.train, format="%Y"))
Years.train[Years.train==2014] <- 2013
Years.val <- as.numeric(format(dates.val, format="%Y"))
hist(Years.train, main="Development")
hist(Years.val, main="Validation")
```
# Make plot
```{r, eval=TRUE}
dats <- rbind(data.frame(Years=Years.train, Data="Development"),
              data.frame(Years=Years.val, Data="Validation"))
ggplot(dats, aes(Years, fill=Data))+
  geom_histogram(alpha=1, position="dodge", bins=50)+
  ylab("Frequency")+xlab("Calendar year of sentinel node procedure")+
  theme_set(theme_bw())
```
# Add Years to model linearly
```{r, eval=TRUE}
Years.train.stand <- Years.train-1993

form.full.Rec.5.Years <- S.Rec.5.training ~
  SNstatus +
  Sex +
  Age.SN +
  Ulceration +
  Loc_CAT +
  Histology +
  log(Breslow) +
  multiple.fields +
  Tot_SNs_neg +
  Tot_SNs_pos +
  log(Rdamcrit) +
  Dewar +
  SNstatus:Sex +
  SNstatus:Age.SN +
  SNstatus:Ulceration +
  SNstatus:Loc_CAT +
  SNstatus:Histology +
  SNstatus:log(Breslow) +
  SNstatus:multiple.fields +
  Years.train.stand
f.mi.full.Rec.5.Years <- Hmisc::fit.mult.impute(form.full.Rec.5.Years, cph, 
                                          xtrans=imputed.training.data,
                                          data=training.data.to.be.imputed,
                                          n.impute=m, pr=FALSE, fit.reps=TRUE, 
                                          fitargs=list(y=TRUE, x=TRUE, 
                                                       se.fit=TRUE))

form.BS.Rec.5.Years <- S.Rec.5.training ~
  SNstatus +
  Age.SN +
  Ulceration +
  Loc_CAT +
  I(log(Breslow)-c.Breslow) +
  log(Rdamcrit) +
  SNstatus:I(log(Breslow)-c.Breslow)+
  Years.train.stand
f.mi.BS.Rec.5.Years <- Hmisc::fit.mult.impute(form.BS.Rec.5.Years,cph,
                                        xtrans=imputed.training.data,
                                        data=training.data.to.be.imputed,
                                        n.impute=m, pr=FALSE,fit.reps=TRUE,
                                        fitargs=list(y=TRUE,x=TRUE,se.fit=TRUE))
```

# Melanoma-specific survival
```{r, eval=TRUE}
form.BS.MSM.5.refit.Years <- S.MSM.5.training ~
  SNstatus +
  Age.SN +
  Ulceration +
  Loc_CAT +
  I(log(Breslow)-c.Breslow) +
  log(Rdamcrit) +
  SNstatus:I(log(Breslow)-c.Breslow)+
  Years.train.stand
f.mi.BS.MSM.5.refit.Years <- Hmisc::fit.mult.impute(form.BS.MSM.5.refit.Years,cph,
                                              xtrans=imputed.training.data,
                                              data=training.data.to.be.imputed,
                                              n.impute=m,
                                              pr=FALSE,fit.reps=TRUE,
                                              fitargs=list(y=TRUE,x=TRUE,
                                                           se.fit=TRUE))
```
# Add Years to non-linear
```{r, eval=TRUE}
form.full.Rec.5.Years.nonlinear <- S.Rec.5.training ~
  SNstatus +
  Sex +
  Age.SN +
  Ulceration +
  Loc_CAT +
  Histology +
  log(Breslow) +
  multiple.fields +
  Tot_SNs_neg +
  Tot_SNs_pos +
  log(Rdamcrit) +
  Dewar +
  SNstatus:Sex +
  SNstatus:Age.SN +
  SNstatus:Ulceration +
  SNstatus:Loc_CAT +
  SNstatus:Histology +
  SNstatus:log(Breslow) +
  SNstatus:multiple.fields +
  rcs(Years.train.stand)
f.mi.full.Rec.5.Years.nonlinear <- Hmisc::fit.mult.impute(form.full.Rec.5.Years.nonlinear, cph, 
                                          xtrans=imputed.training.data,
                                          data=training.data.to.be.imputed,
                                          n.impute=m, pr=FALSE, fit.reps=TRUE, 
                                          fitargs=list(y=TRUE, x=TRUE, 
                                                       se.fit=TRUE))

form.BS.Rec.5.Years.nonlinear <- S.Rec.5.training ~
  SNstatus +
  Age.SN +
  Ulceration +
  Loc_CAT +
  I(log(Breslow)-c.Breslow) +
  log(Rdamcrit) +
  SNstatus:I(log(Breslow)-c.Breslow)+
  rcs(Years.train.stand)
f.mi.BS.Rec.5.Years.nonlinear <- Hmisc::fit.mult.impute(form.BS.Rec.5.Years.nonlinear,cph,
                                        xtrans=imputed.training.data,
                                        data=training.data.to.be.imputed,
                                        n.impute=m, pr=FALSE,fit.reps=TRUE,
                                        fitargs=list(y=TRUE,x=TRUE,se.fit=TRUE))

form.BS.MSM.5.refit.Years.nonlinear <- S.MSM.5.training ~
  SNstatus +
  Age.SN +
  Ulceration +
  Loc_CAT +
  I(log(Breslow)-c.Breslow) +
  log(Rdamcrit) +
  SNstatus:I(log(Breslow)-c.Breslow)+
  rcs(Years.train.stand)
f.mi.BS.MSM.5.refit.Years.nonlinear <- Hmisc::fit.mult.impute(form.BS.MSM.5.refit.Years.nonlinear,cph,
                                              xtrans=imputed.training.data,
                                              data=training.data.to.be.imputed,
                                              n.impute=m,
                                              pr=FALSE,fit.reps=TRUE,
                                              fitargs=list(y=TRUE,x=TRUE,
                                                           se.fit=TRUE))
```
# Make Table
```{r,eval=TRUE}
results.Years <- c()
for (outcome in c("Rec.5", "MSM.5.refit")){
  if (outcome=="Rec.5"){
    models <- c("full", "BS")
  } else{
      models <- c("BS")
  }

  for (model in models){
    form <- eval(parse(text=paste0("f.mi.", model, ".", outcome)))
    results.Years <- rbind(results.Years, c(paste(model, outcome, "- no years"), 
                                            sprintf("%.1f", AIC(form)), "", ""))

    for (linearity in c("Years", "Years.nonlinear")){
        form.years <- eval(parse(text=paste0("f.mi.", model, ".", outcome, ".", linearity)))
        LR.Years <- lrtest(form, form.years)
        results.Years <- rbind(results.Years, c(paste(model, outcome, "-", linearity),
                                                sprintf("%.1f", AIC(form.years)), 
                                                sprintf("%.1f", as.numeric(LR.Years$stats["L.R. Chisq"])), 
                                                sprintf("%.3f", as.numeric(LR.Years$stats["P"]))))
    }
  }
}
openxlsx::write.xlsx(as.data.frame(results.Years),
            rowNames=FALSE,
            file="Z:/Project Melanoom/PaperMelanoma/Results/Review/SA.Years.xlsx")
```

### 2023-11-20 Code/3_Descriptives.Rmd
Final edits

@@
 
-  assign(paste0("S.Rec.", type), survival::Surv(exp(last.imputed.data.set$logRec.time), last.imputed.data.set$Recurrence=="recurred"))
-  assign(paste0("S.MSM.", type), survival::Surv(exp(last.imputed.data.set$logFU.time), last.imputed.data.set$MSM=="Death by melanoma"))
+  Rec.event.composite <- ifelse(last.imputed.data.set$Recurrence=="recurred", 1, 0) # last.imputed.data.set$dead=="Yes"
+  assign(paste0("S.Rec.", type), survival::Surv(exp(last.imputed.data.set$logRec.time), 
+                                                Rec.event.composite))
+  assign(paste0("S.MSM.", type), survival::Surv(exp(last.imputed.data.set$logFU.time), 
+                                                last.imputed.data.set$MSM=="Death by melanoma"))
 
@@
   if (type=="Rec"){
-    median.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logRec.time), Recurrence=="recurred")~1, data=data[data$Center!="Australia",], reverse=TRUE))
-    median.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logRec.time), Recurrence=="recurred")~Center, data=data, reverse=TRUE))
+    median.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logRec.time), 
+                                                                        ifelse(Recurrence=="recurred" , 1, 0))~1, data=data[data$Center!="Australia",], reverse=TRUE)) # | dead=="Yes"
+    median.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logRec.time), ifelse(Recurrence=="recurred" , 1, 0))~Center, data=data, reverse=TRUE)) # | dead=="Yes"
   } else if (type=="FU"){
@@
 load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData"))
-med.Rec <- median.time(data=rbind(training.data.to.be.imputed[, c("logRec.time", "Recurrence", "Center")], 
-                       validation.data.to.be.imputed[, c("logRec.time", "Recurrence", "Center")]), type="Rec")
+med.Rec <- median.time(data=rbind(training.data.to.be.imputed[, c("logRec.time", "Recurrence", "dead", "Center")], 
+                       validation.data.to.be.imputed[, c("logRec.time", "Recurrence", "dead", "Center")]), type="Rec")
 med.FU <- median.time(data=rbind(training.data.to.be.imputed[, c("logFU.time", "MSM", "Center")], 
@@
 ```{r, eval=TRUE}
+est.CI.n <- function(KM=NULL, horizon=5){
+  t <- max(KM$time[KM$time<=horizon])
+  est <- 100*KM$surv[KM$time==t]
+  est.LCI <- 100*KM$lower[KM$time==t]
+  est.UCI <- 100*KM$upper[KM$time==t]
+  n <- sum(KM$n.event[KM$time<=horizon])
+  return(list(t=t, est=est, est.LCI=est.LCI, est.UCI=est.UCI, n=n))
+}
+
 Rec.MSM.table <- function(horizon=5, type=NULL, data.to.be.imputed=NULL){
+  # select last imputed data set
+  imputed.data <- eval(parse(text=paste0("imputed.", type, ".data")))
+  last.imputed.data.set <- mice::complete(imputed.data, m)
+
   # all patients, Recurrence
   S.Rec <- eval(parse(text=paste0("S.Rec.", type)))
-  SF.Rec.all <- survival::survfit(S.Rec~1, data=data.to.be.imputed)
-  RFS.all <- 100*SF.Rec.all$surv[SF.Rec.all$time==max(SF.Rec.all$time[SF.Rec.all$time<=horizon])]
-  n.Rec.all <- sum(SF.Rec.all$n.event[SF.Rec.all$time<=horizon])
+  KM.Rec.all <- survival::survfit(S.Rec~1, data=data.to.be.imputed)
+  out.Rec.all <- est.CI.n(KM=KM.Rec.all, horizon=horizon)
 
   # only negative patients, Recurrence
-  SF.Rec.neg <-survival::survfit(S.Rec[data.to.be.imputed$SNstatus=="Negative"]~1, data=data.to.be.imputed)
-  RFS.neg <- 100*SF.Rec.neg$surv[SF.Rec.neg$time==max(SF.Rec.neg$time[SF.Rec.neg$time<=horizon])]
-  n.Rec.neg <- sum(SF.Rec.neg$n.event[SF.Rec.neg$time<=horizon])
+  KM.Rec.neg <-survival::survfit(S.Rec[data.to.be.imputed$SNstatus=="Negative"]~1, data=data.to.be.imputed)
+  out.Rec.neg <- est.CI.n(KM=KM.Rec.neg, horizon=horizon)
 
   # only positive patients, Recurrence
-  SF.Rec.pos <- survival::survfit(S.Rec[data.to.be.imputed$SNstatus=="Positive"]~1, data=data.to.be.imputed)
-  RFS.pos <- 100*SF.Rec.pos$surv[SF.Rec.pos$time==max(SF.Rec.pos$time[SF.Rec.pos$time<=horizon])]
-  n.Rec.pos <- sum(SF.Rec.pos$n.event[SF.Rec.pos$time<=horizon])
+  KM.Rec.pos <- survival::survfit(S.Rec[data.to.be.imputed$SNstatus=="Positive"]~1, data=data.to.be.imputed)
+  out.Rec.pos <- est.CI.n(KM=KM.Rec.pos, horizon=horizon)
 
@@
   S.MSM <- eval(parse(text=paste0("S.MSM.", type)))
-  SF.MSM.all <- survival::survfit(S.MSM~1, data=data.to.be.imputed)
-  MSS.all <- 100*SF.MSM.all$surv[SF.MSM.all$time==max(SF.MSM.all$time[SF.MSM.all$time<=horizon])]
-  n.MSM.all <- sum(SF.MSM.all$n.event[SF.MSM.all$time<=horizon])
+  KM.MSM.all <- survival::survfit(S.MSM~1, data=data.to.be.imputed)
+  out.MSM.all <- est.CI.n(KM=KM.MSM.all, horizon=horizon)
 
   # negative patients, MSM
-  SF.MSM.neg <- survival::survfit(S.MSM[data.to.be.imputed$SNstatus=="Negative"]~1, data=data.to.be.imputed)
-  MSS.neg <- 100*SF.MSM.neg$surv[SF.MSM.neg$time==max(SF.MSM.neg$time[SF.MSM.neg$time<=horizon])]
-  n.MSM.neg <- sum(SF.MSM.neg$n.event[SF.MSM.neg$time<=horizon])
+  KM.MSM.neg <- survival::survfit(S.MSM[data.to.be.imputed$SNstatus=="Negative"]~1, data=data.to.be.imputed)
+  out.MSM.neg <- est.CI.n(KM=KM.MSM.neg, horizon=horizon)
 
   # positive patients, MSM
-  SF.MSM.pos <- survival::survfit(S.MSM[data.to.be.imputed$SNstatus=="Positive"]~1, data=data.to.be.imputed)
-  MSS.pos <- 100*SF.MSM.pos$surv[SF.MSM.pos$time==max(SF.MSM.pos$time[SF.MSM.pos$time<=horizon])]
-  n.MSM.pos <- sum(SF.MSM.pos$n.event[SF.MSM.pos$time<=horizon])
+  KM.MSM.pos <- survival::survfit(S.MSM[data.to.be.imputed$SNstatus=="Positive"]~1, data=data.to.be.imputed)
+  out.MSM.pos <- est.CI.n(KM=KM.MSM.pos, horizon=horizon)
 
@@
                               e1=rep("", 3),
-                              n.Rec=c(n.Rec.all, n.Rec.pos, n.Rec.neg),
-                              n.MSM=c(n.MSM.all, n.MSM.pos, n.MSM.neg),
+                              n.Rec=c(out.Rec.all$n, 
+                                      out.Rec.neg$n, 
+                                      out.Rec.pos$n),
+                              n.MSM=c(out.MSM.all$n, 
+                                      out.MSM.neg$n, 
+                                      out.MSM.pos$n),
                               e2=rep("", 3),
-                              RFS=sprintf("%.1f", c(RFS.all, RFS.neg, RFS.pos)),
-                              MSS=sprintf("%.1f", c(MSS.all, MSS.neg, MSS.pos)))
+                              Rec=sprintf("%.1f", c(out.Rec.all$est,
+                                                    out.Rec.neg$est,
+                                                    out.Rec.pos$est)),
+                              Rec.LCI=sprintf("%.1f", c(out.Rec.all$est.LCI,
+                                                        out.Rec.neg$est.LCI,
+                                                        out.Rec.pos$est.LCI)),
+                              Rec.UCI=sprintf("%.1f", c(out.Rec.all$est.UCI,
+                                                        out.Rec.neg$est.UCI,
+                                                        out.Rec.pos$est.UCI)),
+                              e3=rep("", 3),
+                              MSM=sprintf("%.1f", c(out.MSM.all$est,
+                                                    out.MSM.neg$est,
+                                                    out.MSM.pos$est)),
+                              MSM.LCI=sprintf("%.1f", c(out.MSM.all$est.LCI,
+                                                        out.MSM.neg$est.LCI,
+                                                        out.MSM.pos$est.LCI)),
+                              MSM.UCI=sprintf("%.1f", c(out.MSM.all$est.UCI, 
+                                                        out.MSM.neg$est.UCI, 
+                                                        out.MSM.pos$est.UCI)))
   row.names(Rec.MSM.table) <- c("All", "Negative", "Positive")

### 2023-11-20 Code/8_CreateNomogram.Rmd
Final edits

@@
 attributes(nom)$names[attributes(nom)$names=="Loc_CAT"]<-"Location"
-attributes(nom)$names[attributes(nom)$names=="Rdamcrit"]<-"Tumour burden"
+attributes(nom)$names[attributes(nom)$names=="Rdamcrit"]<-"Maximum diameter of largest SN metastasis"
 attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Positive)"]<-"Breslow (positive SN)"

### 2023-11-20 Code/9_CreateRiskScoreHistogram.Rmd
Validation added
Final edits

@@
 points(points.range,p.MSM*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=16)
-legend(x=0,y=100,legend=c("Recurrence","Melanoma-specific mortality"),pch=c(15,16),cex=.9,bg="white")
+legend(x=0,y=100,legend=c("Recurrence or death","Melanoma-specific mortality"),pch=c(15,16),cex=.9,bg="white")
 grDevices::dev.off()
@@
   geom_line(aes(x=points.range, y=p.Rec*100))+
-  geom_point(aes(x=points.range, y=p.Rec*100, alpha="Recurrence"), shape=15)+
+  geom_point(aes(x=points.range, y=p.Rec*100, alpha="Recurrence or death"), shape=15)+
   geom_line(aes(x=points.range, y=p.MSM*100))+
@@
                      values=c(15, 19),
-                     breaks=c("Recurrence", "Melanoma-specific mortality"),
+                     breaks=c("Recurrence or death", "Melanoma-specific mortality"),
                      guide=guide_legend(override.aes=list(linetype=c(0, 1), 

### 2023-11-23 Code/0_PrepareValidationData.Rmd
Final commit

@@
 author: "Carolien C.H.M. Maas"
-date: "January 2023"
+date: "November 2023"
 output: html_document
@@
 ```{r, eval=TRUE}
-sel.val.data$Rec.time <- (sel.val.data$Date_1st_Rec-sel.val.data$SNdate)/(24*60*60)/365.25
+sel.val.data$Recurrence.time <- (sel.val.data$Date_1st_Rec-sel.val.data$SNdate)/(24*60*60)/365.25
 ```
@@
 ```{r, eval=TRUE}
-sum(!(!(sel.val.data$Rec.time<0&!is.na(sel.val.data$Rec.time))))
+sum(!(!(sel.val.data$Recurrence.time<0&!is.na(sel.val.data$Recurrence.time))))
 ```
-
-Set recurrence time equal to follow-up time for those without recurrence
-```{r, eval=TRUE}
-sel.val.data$Rec.time[sel.val.data$Recurrence=="no recurrence"|is.na(sel.val.data$SNdate)]<-sel.val.data$FU.time[sel.val.data$Recurrence=="no recurrence"|is.na(sel.val.data$SNdate)]
-
-sel.val.data$Rec.time[sel.val.data$Recurrence=="recurrence"&is.na(sel.val.data$Date_1st_Rec)]<-NA
-table(sel.val.data$Status, sel.val.data$Recurrence)
-```
-
-#### Melanoma specific mortality (MSM)
+# Define death by any cause
 ```{r, eval=TRUE}
@@
 sel.val.data$dead <- factor(sel.val.data$dead, labels=c("No", "Yes"))
+``` 
+<!-- Set recurrence time equal to follow-up time for those without recurrence -->
+<!-- ```{r, eval=TRUE} -->
+<!-- sel.val.data$Recurrence.time[sel.val.data$Recurrence=="no recurrence"|is.na(sel.val.data$SNdate)]<-sel.val.data$FU.time[sel.val.data$Recurrence=="no recurrence"|is.na(sel.val.data$SNdate)] -->
+
+<!-- sel.val.data$Recurrence.time[sel.val.data$Recurrence=="recurrence"&is.na(sel.val.data$Date_1st_Rec)]<-NA -->
+<!-- table(sel.val.data$Status, sel.val.data$Recurrence) -->
+<!-- ``` -->
+Create composite Recurrence
+```{r, eval=TRUE}
+# event = 1 if recurrence or death, whichever occurs first
+sel.val.data$Rec.composite <- ifelse(sel.val.data$Recurrence=="recurred" | sel.val.data$dead=="Yes", 1, 0)
+
+# if patient has recurrence, set time to Recurrence.time
+# if patient has no recurrence, but died, set time to FU.time
+# if patient has no recurrence, and did not die, set to FU.time
+sel.val.data$Rec.composite.time <- ifelse(sel.val.data$Recurrence=="recurred", 
+       sel.val.data$Recurrence.time,
+       sel.val.data$FU.time)
+```
+#### Melanoma specific mortality (MSM)
+```{r, eval=TRUE}
 table(sel.val.data$Status, sel.val.data$dead)
@@
 
+Relation between outcome and covariates
+```{r, eval=TRUE}
+sel.val.data$logRec.composite.time <- log(sel.val.data$Rec.composite.time+1/365.25)
+sel.val.data$logFU.time <- log(sel.val.data$FU.time+1/365.25)
+```
+
 # Multiple imputation
@@
                              "Tot_SNs_neg", "Tot_SNs_pos", "Rdamcrit", "Dewar",
-                             "Mitosis", "Recurrence", "dead", "MSM")
+                             "Mitosis", "Status",
+                             "Rec.composite", "logRec.composite.time", 
+                             "MSM", "logFU.time")
 validation.data.to.be.imputed <- sel.val.data[, variables.to.be.imputed]
+missings.table(validation.data.to.be.imputed)
 ```
@@
 options(digits=8)
-```
-
-Relation between outcome and covariates
-```{r, eval=TRUE}
-validation.data.to.be.imputed$logRec.time <- log(sel.val.data$Rec.time+1/365.25)
-validation.data.to.be.imputed$logFU.time <- log(sel.val.data$FU.time+1/365.25)
-missings.table(validation.data.to.be.imputed)
 save(list=c("dd.validation", "validation.data.to.be.imputed"),file="Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.Rdata", compress=TRUE)

### 2023-11-23 Code/12_Validation.Rmd
Final commit

--- 2022-12-19 Code/12_Validation.Rmd
@@
 ---
-title: "Create histogram with risk scores"
+title: "Validate model"
 author: "Carolien C.H.M. Maas"
-date: "December 2022"
+date: "November 2023"
 output: html_document
@@
 ```{r, eval=TRUE}
-# load model
-load("Z:/Project Melanoom/PaperMelanoma/Results/h0.Rec.MSM.Rdata")
-
 # load data
@@
     PredictionTools::val.surv.mi(p, S,
-                                 g=5,main="All patients", time=NULL,lim=c(0,1),dist=TRUE)
+                                 g=5, main="All patients", time=NULL, 
+                                 lim=c(0,1), dist=TRUE,
+                                 optimism.C=ifelse(data.type=="training", optimism, 0))
     grDevices::dev.off()
@@
                                  S[data$SNstatus=="Positive",],
-                                 g=5,main="Patients with positive SN", time=NULL,lim=c(0,1),dist=TRUE)
+                                 g=5, main="Patients with positive SN", 
+                                 time=NULL, lim=c(0,1), dist=TRUE,
+                                 optimism.C=ifelse(data.type=="training", optimism, 0))
     grDevices::dev.off()
@@
                                  S[data$SNstatus=="Negative",],
-                                 g=5,main="Patients with negative SN", time=NULL,lim=c(0,1),dist=TRUE)
+                                 g=5, main="Patients with negative SN", 
+                                 time=NULL, lim=c(0,1), dist=TRUE,
+                                 optimism.C=ifelse(data.type=="training", optimism, 0))
     grDevices::dev.off()

### 2023-11-23 Code/5_FitModels.Rmd
Final commit

@@
 }
-MSM.cal.fact.combined<-Rubin.combine(MSM.cal.fact.all, MSM.cal.fact.SE)
+MSM.cal.fact.combined <- Rubin.combine(MSM.cal.fact.all, MSM.cal.fact.SE)
 MSM.cal.fact <- MSM.cal.fact.combined$est
@@
 PredictionTools::val.surv.mi(p.MSM.5.redev, S.MSM.5.training, g=5,
-                             main="All patients",time=NULL,lim=c(0,1),dist=TRUE)
+                             main="All patients", time=NULL, lim=c(0, 1),
+                             dist=TRUE)
 ```

### 2023-11-23 Code/6_CreateTablesandPlots.Rmd
Final commit

@@
 library(png)
+# remotes::install_github("CHMMaas/PredictionTools")
 library(PredictionTools)
@@
   neg.table <- cbind(sprintf("%.2f", neg.table[, 1]),
-                     paste0("[", sprintf("%.2f", neg.table[, 2]), "; ", sprintf("%.2f", neg.table[, 3]), "]"))
+                     paste0("[", sprintf("%.2f", neg.table[, 2]), "; ", 
+                            sprintf("%.2f", neg.table[, 3]), "]"))
 
@@
 for (i in 1:m){
-  v<-validate(f.mi.full.Rec.5$fits[[i]],method="boot",bw=TRUE,rule="p",sls=0.05,B=100,pr=FALSE,type="individual")
-  slope<-c(slope,v["Slope","test"]) # slope of test
-  cindex.orig<-c(cindex.orig,(v["Dxy","index.orig"]+1)/2) # original C-index
-  opt<-c(opt,v["Dxy","optimism"]/2) # optimism C-index
-  cindex.B<-c(cindex.B,(v["Dxy","index.corrected"]+1)/2) # index corrected C-index
+  v <- validate(f.mi.full.Rec.5$fits[[i]], method="boot", bw=TRUE, rule="p",
+                sls=0.05, B=100, pr=FALSE, type="individual")
+  slope <- c(slope, v["Slope","test"]) # slope of test
+  cindex.orig <- c(cindex.orig, (v["Dxy","index.orig"]+1)/2) # original C-index
+  opt <- c(opt, v["Dxy","optimism"]/2) # optimism C-index
+  cindex.B <- c(cindex.B, (v["Dxy","index.corrected"]+1)/2) # index corrected C-index
 }
@@
 # replace NA by blank
-for (df in c("final.full.table.Rec.5", "final.full.table.MSM.5.refit", "final.BS.table.Rec.5", "final.BS.table.MSM.5.refit")){
+for (df in c("final.full.table.Rec.5", "final.full.table.MSM.5.refit", 
+             "final.BS.table.Rec.5", "final.BS.table.MSM.5.refit")){
   final.df <- eval(parse(text=df))
@@
               row.names=FALSE, col.names=FALSE, sep=",")
+
+save(list=c("dd.training", "training.data.to.be.imputed", "m", 
+            "imputed.training.data", "last.imputed.training.data.set", 
+            "horizon", "S.Rec.5.training", "S.MSM.5.training",
+            "f.mi.full.Rec.5", "f.mi.BS.Rec.5", "f.mi.full.MSM.5.refit", 
+            "f.mi.BS.MSM.5.refit", "MSM.cal.fact", "c.Breslow", "optimism",
+            "form.full.Rec.5", "form.BS.Rec.5", 
+            "form.full.MSM.5.refit", "form.BS.MSM.5.refit"),
+     file="Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata", compress=TRUE)
 ```

### 2023-11-23 Code/7_CreateCrossValidatedCalibrationPlots.Rmd
Final commit

@@
 library(grDevices)
-source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")
 # remotes::install_github("CHMMaas/PredictionTools")
 library(PredictionTools)  # to plot calibration
+source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")
 ```

### 2023-11-23 Code/9_CreateRiskScoreHistogram.Rmd
Final commit

@@
 int.MSM <- MSM.cal.fact*int.Rec
-max.range <- 16
+max.range <- 17
 points.range <- 0:max.range
@@
 mean(lp.Rec)
-score.Rec <- round((lp.Rec-int.Rec)/rc.Rec,0)
+score.Rec <- round((lp.Rec-int.Rec)/rc.Rec, 0)
 max.breaks <- max.range+1
-h.Rec <- hist(score.Rec,plot=FALSE,breaks=0:max.breaks,right=FALSE, beside=TRUE)
+h.Rec <- hist(score.Rec, 
+              plot=FALSE,
+              breaks=0:max.breaks,
+              right=FALSE, 
+              beside=TRUE)
 
@@
 
-save(h0.Rec, h0.MSM, coef.Rec, c.Breslow, center.Rec, MSM.cal.fact,
-     file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/h0.Rec.MSM.Rdata"), compress=TRUE)
-save(list=c("dd.training", "training.data.to.be.imputed", "m", "imputed.training.data", 
-            "last.imputed.training.data.set", "horizon", "S.Rec.5.training", "S.MSM.5.training",
-            "f.mi.full.Rec.5", "f.mi.BS.Rec.5", "f.mi.full.MSM.5.refit", "f.mi.BS.MSM.5.refit", "MSM.cal.fact", "c.Breslow", 
-            "form.full.Rec.5", "form.BS.Rec.5", "form.full.MSM.5.refit", "form.BS.MSM.5.refit", 
+save(rc.Rec, int.Rec, f.mi.BS.Rec.5, MSM.cal.fact, max.range,
+     file="Z:/Project Melanoom/PaperMelanoma/Results/model.Rdata", 
+     compress=TRUE)
+
+save(list=c("dd.training", "training.data.to.be.imputed", "m", 
+            "imputed.training.data", "last.imputed.training.data.set", 
+            "horizon", "S.Rec.5.training", "S.MSM.5.training",
+            "f.mi.full.Rec.5", "f.mi.BS.Rec.5", 
+            "f.mi.full.MSM.5.refit", "f.mi.BS.MSM.5.refit", 
+            "MSM.cal.fact", "c.Breslow", "optimism",
+            "form.full.Rec.5", "form.BS.Rec.5", 
+            "form.full.MSM.5.refit", "form.BS.MSM.5.refit", 
             "rc.Rec", "int.Rec", "h0.Rec", "lp.points.Rec.5",

### 2023-11-23 Code/AdditionalAnalysis_Review.R

@@ -1,0 +1,93 @@
+#libraries
+library(Hmisc)
+library(survAUC)
+library(pROC)
+library(mice)
+
+# load data
+load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData")
+load("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData")
+file.path <- "Z:/Project Melanoom/PaperMelanoma/Results/"
+
+# calculate Harrell's C and Uno's C
+Harrell.Rec.train <- c()
+Uno.Rec.train <- c()
+Harrell.Rec.val <- c()
+Uno.Rec.val <- c()
+for (i in 1:m){
+  lp.train <- predict(f.mi.BS.Rec.5, newdata=mice::complete(imputed.training.data, i), type="lp")
+  Harrell.Rec.train <- c(Harrell.Rec.train, Hmisc::rcorr.cens(-lp.train, S=S.Rec.5.training)["C Index"])
+  Uno.Rec.train <- c(Uno.Rec.train, survAUC::UnoC(Surv.rsp=S.Rec.5.training, Surv.rsp.new=S.Rec.5.training, lpnew=lp.train, time=5))
+
+  lp.val <- predict(f.mi.BS.Rec.5, newdata=mice::complete(imputed.validation.data, i), type="lp")
+  Harrell.Rec.val <- c(Harrell.Rec.val, Hmisc::rcorr.cens(-lp.val, S=S.Rec.5.validation)["C Index"])
+  Uno.Rec.val <- c(Uno.Rec.val, survAUC::UnoC(Surv.rsp=S.Rec.5.validation, Surv.rsp.new=S.Rec.5.validation, lpnew=lp.val, time=5))
+}
+write.table(round(data.frame(Harrell=c(mean(Harrell.Rec.train),
+                                 mean(Uno.Rec.train)),
+                       Uno=c(mean(Harrell.Rec.val),
+                             mean(Uno.Rec.val))), 2),
+            file="Z:/Project Melanoom/PaperMelanoma/Results/Harrell.Uno.txt",
+            sep=",")
+
+# ROC curve
+labels <- data.frame(outcomes=c("Rec", "MSM"),
+                     long.outcomes=c("Recurrence-free survival", "Melanoma-specific survival"),
+                     types=c("training", "validation"),
+                     long.types=c("(Europe)", "(Australia)"))
+for (type in c("training", "validation")){
+  if (type=="training"){
[…]

### 2023-11-23 Code/00_PrepareTrainingData.Rmd
Rename files

--- 2023-10-06 Code/0_PrepareTrainingData.Rmd
@@
 #### SN tumour burden
-11 patients with positive SN have SN tumour burden of exactly 1
+11 patients with positive SN have SN tumour burden (maximum diameter) of exactly 1
 ```{r, eval=TRUE}
@@
 ```{r, eval=TRUE}
-sel.train.data$Rec.time <- (sel.train.data$Date_1st_Rec-sel.train.data$SNdate)/(24*60*60)/365.25
+sel.train.data$Recurrence.time <- (sel.train.data$Date_1st_Rec-sel.train.data$SNdate)/(24*60*60)/365.25
 ```
@@
 old.n <- nrow(sel.train.data)
-sel.train.data <- sel.train.data[!(sel.train.data$Rec.time<0&!is.na(sel.train.data$Rec.time)),]
+sel.train.data <- sel.train.data[!(sel.train.data$Recurrence.time<0&!is.na(sel.train.data$Recurrence.time)),]
 cat("Number of patients excluded:", old.n-nrow(sel.train.data), "\n")
 ```
-
-Set recurrence time equal to follow-up time for those without recurrence
-```{r, eval=TRUE}
-sel.train.data$Rec.time[sel.train.data$Recurrence=="no recurrence"|is.na(sel.train.data$SNdate)]<-sel.train.data$FU.time[sel.train.data$Recurrence=="no recurrence"|is.na(sel.train.data$SNdate)]
-table(sel.train.data$Status, sel.train.data$Recurrence)
-```
-
-Impute recurrence time for 45 patients with missing Rec.time but Recurrence=="recurred"
+Impute recurrence time for 45 patients with missing Recurrence.time but Recurrence=="recurred"
 ```{r, eval=TRUE}
@@
 ```
+Create composite Recurrence
+```{r, eval=TRUE}
+# event = 1 if recurrence or death, whichever occurs first
+sel.train.data$Rec.composite <- ifelse(sel.train.data$Recurrence=="recurred" | sel.train.data$dead=="Yes", 1, 0)
 
+# if patient has recurrence, set time to Recurrence.time
+# if patient has no recurrence, but died, set time to FU.time
+# if patient has no recurrence, and did not die, set to FU.time
+sel.train.data$Rec.composite.time <- ifelse(sel.train.data$Recurrence=="recurred", 
+       sel.train.data$Recurrence.time,
+       sel.train.data$FU.time)
+```
 #### Melanoma specific mortality (MSM)
@@
 sel.train.data$MSM <- 1*(sel.train.data$Status=="DOD")
+```
+#### Relation between outcome and covariates
+```{r, eval=TRUE}
+sel.train.data$logRec.composite.time <- log(sel.train.data$Rec.composite.time+1/365.25)
+sel.train.data$logFU.time <- log(sel.train.data$FU.time+1/365.25)
 ```
@@
 # set levels for multiple.fields
-sel.train.data$multiple.fields <- factor(sel.train.data$multiple.fields, labels=c("Other", "Multiple SN fields"))
-sel.train.data$MSM <- factor(sel.train.data$MSM, labels=c("Other", "Death by melanoma"))
+sel.train.data$multiple.fields <- factor(sel.train.data$multiple.fields, 
+                                         labels=c("Other", "Multiple SN fields"))
+sel.train.data$MSM <- factor(sel.train.data$MSM, 
+                             labels=c("Other", "Death by melanoma"))
 ```
 
-# Multiple imputation
+# Prepare Multiple imputation
 ```{r, eval=TRUE}
@@
                              "Tot_SNs_neg", "Tot_SNs_pos", "Rdamcrit", "Dewar",
-                             "Mitosis", "Status", "Recurrence", "dead", "MSM")
+                             "Mitosis", "Status", 
+                             "Rec.composite", "logRec.composite.time",
+                             "MSM", "logFU.time")
 training.data.to.be.imputed <- sel.train.data[, variables.to.be.imputed]
+missings.table(training.data.to.be.imputed)
 ```
@@
 options(digits=8)
+save(list=c("dd.training", "training.data.to.be.imputed"),
+     file="Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata", 
+     compress=TRUE)
 ```
-
-Relation between outcome and covariates
-```{r, eval=TRUE}
-training.data.to.be.imputed$logRec.time <- log(sel.train.data$Rec.time+1/365.25)
-training.data.to.be.imputed$logFU.time <- log(sel.train.data$FU.time+1/365.25)
-missings.table(training.data.to.be.imputed)
-# save(list=c("dd.training", "training.data.to.be.imputed"),file="Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata", compress=TRUE)
-```

### 2023-11-23 Code/01_DescriptivesTable.Rmd
Rename files

--- 2023-10-06 Code/1_DescriptivesTable.Rmd
@@
                                 +multiple.fields
-                                +arsenal::includeNA(Mitosis)
-                                +Recurrence
-                                +dead
-                                +MSM
-                                +logRec.time
-                                +logFU.time, 
+                                +arsenal::includeNA(Mitosis),
                                 data=pooled.data.to.be.imputed,

### 2023-11-23 Code/03_Descriptives.Rmd
Rename files

--- 2023-11-20 Code/3_Descriptives.Rmd
@@
 
-  Rec.event.composite <- ifelse(last.imputed.data.set$Recurrence=="recurred", 1, 0) # last.imputed.data.set$dead=="Yes"
-  assign(paste0("S.Rec.", type), survival::Surv(exp(last.imputed.data.set$logRec.time), 
-                                                Rec.event.composite))
+  assign(paste0("S.Rec.", type), survival::Surv(exp(last.imputed.data.set$logRec.composite.time), 
+                                                last.imputed.data.set$Rec.composite))
   assign(paste0("S.MSM.", type), survival::Surv(exp(last.imputed.data.set$logFU.time), 
@@
   if (type=="Rec"){
-    median.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logRec.time), 
-                                                                        ifelse(Recurrence=="recurred" , 1, 0))~1, data=data[data$Center!="Australia",], reverse=TRUE)) # | dead=="Yes"
-    median.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logRec.time), ifelse(Recurrence=="recurred" , 1, 0))~Center, data=data, reverse=TRUE)) # | dead=="Yes"
+    median.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logRec.composite.time), 
+                                                                        Rec.composite)~1, data=data[data$Center!="Australia",], reverse=TRUE))
+    median.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logRec.composite.time), 
+                                                                  Rec.composite)~Center, data=data, reverse=TRUE))
   } else if (type=="FU"){
-    median.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM=="Death by melanoma")~1, data=data[data$Center!="Australia",], reverse=TRUE))
-    median.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM=="Death by melanoma")~Center, data=data, reverse=TRUE))
+    median.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), 
+                                                                        MSM=="Death by melanoma")~1, 
+                                                          data=data[data$Center!="Australia",], 
+                                                          reverse=TRUE))
+    median.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), 
+                                                                  MSM=="Death by melanoma")~Center, 
+                                                    data=data, 
+                                                    reverse=TRUE))
   }
   # save in tab.time
-  tab.time <- c("Total Development", paste(sprintf("%.2f", median.time.total$quantile[3]),
-                   paste0("[", sprintf("%.2f", median.time.total$quantile[4]), "; ",
-                          sprintf("%.2f", median.time.total$quantile[2]), "]")))
+  tab.time <- c("Total Development", paste(sprintf("%.1f", median.time.total$quantile[3]),
+                   paste0("[", sprintf("%.1f", median.time.total$quantile[4]), "; ",
+                          sprintf("%.1f", median.time.total$quantile[2]), "]")))
 
@@
     upper <- q.center[2]
-    tab.time <- rbind(tab.time, c(center, paste(sprintf("%.2f", med), paste0("[", sprintf("%.2f", lower), "; ", sprintf("%.2f", upper), "]"))))
+    tab.time <- rbind(tab.time, c(center, paste(sprintf("%.1f", med), paste0("[", sprintf("%.1f", lower), "; ", sprintf("%.1f", upper), "]"))))
   }
@@
 load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData"))
-med.Rec <- median.time(data=rbind(training.data.to.be.imputed[, c("logRec.time", "Recurrence", "dead", "Center")], 
-                       validation.data.to.be.imputed[, c("logRec.time", "Recurrence", "dead", "Center")]), type="Rec")
+med.Rec <- median.time(data=rbind(training.data.to.be.imputed[, c("logRec.composite.time", "Rec.composite", "Center")], 
+                       validation.data.to.be.imputed[, c("logRec.composite.time", "Rec.composite", "Center")]), type="Rec")
 med.FU <- median.time(data=rbind(training.data.to.be.imputed[, c("logFU.time", "MSM", "Center")], 
@@
 horizon <- 5
-train.table <- Rec.MSM.table(horizon=horizon, type="training", data.to.be.imputed=training.data.to.be.imputed)
-val.table <- Rec.MSM.table(horizon=horizon, type="validation", data.to.be.imputed=validation.data.to.be.imputed)
+train.table <- Rec.MSM.table(horizon=horizon, type="training", 
+                             data.to.be.imputed=training.data.to.be.imputed)
+val.table <- Rec.MSM.table(horizon=horizon, type="validation", 
+                           data.to.be.imputed=validation.data.to.be.imputed)
 
@@
 ```{r, eval=TRUE}
-png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.Rec.training.png"), width=16, height=16, units="cm", res=300)
-survminer::ggsurvplot(survival::survfit(S.Rec.5.training ~ Center, data=last.imputed.training.data.set), 
+pooled.data <- rbind(last.imputed.training.data.set, last.imputed.validation.data.set)
+S.Rec.pooled <- survival::Surv(exp(pooled.data$logRec.composite.time),
+                               pooled.data$Rec.composite)
+S.Rec.5.pooled <- S.Rec.pooled
+S.Rec.5.pooled[S.Rec.pooled[,1]>5, 1] <- 5 # set time to 5
+S.Rec.5.pooled[S.Rec.pooled[,1]>5, 2] <- 0 # set censoring
+
+png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.Rec.composite.png"), width=16, height=16, units="cm", res=300)
+survminer::ggsurvplot(survival::survfit(S.Rec.5.pooled ~ Center, 
+                                        data=pooled.data), 
                       legend.title="",
-                      legend.labs=c("Rotterdam", "Berlin", "Warsaw", "Amsterdam"),
-                      title="Development",
+                      legend.labs=c("Rotterdam", "Berlin", "Warsaw", "Amsterdam", "Australia"),
+                      title="Recurrence-free survival",
                       ylab="Recurrence-free survival probability",
@@
 grDevices::dev.off()
-png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.MSM.training.png"), width=16, height=16, units="cm", res=300)
-survminer::ggsurvplot(survival::survfit(S.MSM.5.training ~ Center, data=last.imputed.training.data.set), 
+
+S.MSM.pooled <- survival::Surv(exp(pooled.data$logFU.time),
+                               pooled.data$MSM=="Death by melanoma")
+S.MSM.5.pooled <- S.MSM.pooled
+S.MSM.5.pooled[S.MSM.pooled[,1]>5, 1] <- 5 # set time to 5
+S.MSM.5.pooled[S.MSM.pooled[,1]>5, 2] <- 0 # set censoring
+
+png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.MSM.png"), width=16, height=16, units="cm", res=300)
+survminer::ggsurvplot(survival::survfit(S.MSM.5.pooled ~ Center, 
+                                        data=pooled.data), 
                       legend.title="",
-                      legend.labs=c("Rotterdam", "Berlin", "Warsaw", "Amsterdam"),
-                      title="Development",
+                      legend.labs=c("Rotterdam", "Berlin", "Warsaw", "Amsterdam", "Australia"),
+                      title="Melanoma-specific survival",
                       ylab="Melanoma-specific survival probability",
@@
 grDevices::dev.off()
-
-load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData"))
-png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.Rec.validation.png"), width=16, height=16, units="cm", res=300)
-survminer::ggsurvplot(survival::survfit(S.Rec.5.validation ~ 1, data=last.imputed.validation.data.set), 
-                      legend.title="",
-                      legend.labs="Australia",
-                      title="Validation",
-                      ylab="Recurrence-free survival probability",
-                      xlab="Time in years",
-                      conf.int=TRUE, risk.table=TRUE)
-grDevices::dev.off()
-png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.MSM.validation.png"), width=16, height=16, units="cm", res=300)
-survminer::ggsurvplot(survival::survfit(S.MSM.5.validation ~ 1, data=last.imputed.validation.data.set), 
-                      legend.title="",
-                      legend.labs="Australia",
-                      title="Validation",
-                      ylab="Melanoma-specific survival probability",
-                      xlab="Time in years",
-                      conf.int=TRUE, risk.table=TRUE)
-grDevices::dev.off()
 ```

### 2023-11-23 Code/08_CreateNomogram.Rmd
Rename files

--- 2023-11-20 Code/8_CreateNomogram.Rmd
@@
 
-save(list=c("dd.training", "training.data.to.be.imputed", "m", "imputed.training.data", 
-            "last.imputed.training.data.set", "horizon", "S.Rec.5.training", "S.MSM.5.training",
-            "f.mi.full.Rec.5", "f.mi.BS.Rec.5", "f.mi.full.MSM.5.refit", "f.mi.BS.MSM.5.refit", "MSM.cal.fact", "c.Breslow", 
-            "form.full.Rec.5", "form.BS.Rec.5", "form.full.MSM.5.refit", "form.BS.MSM.5.refit", 
+save(list=c("dd.training", "training.data.to.be.imputed", "m", 
+            "imputed.training.data",  "last.imputed.training.data.set", 
+            "horizon", "S.Rec.5.training", "S.MSM.5.training",
+            "f.mi.full.Rec.5", "f.mi.BS.Rec.5", "f.mi.full.MSM.5.refit", 
+            "f.mi.BS.MSM.5.refit", "MSM.cal.fact", "c.Breslow", "optimism",
+            "form.full.Rec.5", "form.BS.Rec.5", 
+            "form.full.MSM.5.refit", "form.BS.MSM.5.refit", 
             "rc.Rec", "int.Rec", "h0.Rec"),

### 2023-12-11 Code/06_CreateTablesandPlots.Rmd
Added Uno's C-index

--- 2023-11-23 Code/6_CreateTablesandPlots.Rmd
@@
     png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/all.centers.", model, ".", type, ".png"))
-    out <- PredictionTools::val.surv.mi(p,S,g=5,main="All centers",
-                                        time=NULL,lim=c(0,1),dist=TRUE)
+    out <- PredictionTools::val.surv.mi(p=p, y=S, g=5, main="All centers", CI.metrics=TRUE,
+                                        time=NULL, lim=c(0,1), dist=TRUE)
     grDevices::dev.off()
@@
     final.full.table <- eval(parse(text=paste0("final.", type, ".table.", model)))
-    final.full.table["C-index", "HR"] <- sprintf("%.2f", out$cindex-optimism)
-    final.full.table["C-index", "CI"] <- paste0("[", sprintf("%.2f", out$cindex.lower-optimism),
+    final.full.table["Harrell's C-index", "HR"] <- sprintf("%.2f", out$cindex-optimism)
+    final.full.table["Harrell's C-index", "CI"] <- paste0("[", sprintf("%.2f", out$cindex.lower-optimism),
                                                    "; ", sprintf("%.2f", out$cindex.upper-optimism), "]")
+    final.full.table["Uno's C-index", "HR"] <- sprintf("%.2f", out$uno.C-optimism)
 
@@
   f.mi <- eval(parse(text=paste0("f.mi.", type, ".Rec.5")))
+  unoC.MSM <- rep(0, m)
   cindex.MSM.cal <- rep(0, m)
@@
   for (i in 1:m){
-    rc <- Hmisc::rcorr.cens(-predict(f.mi,
-                                     newdata=mice::complete(imputed.training.data,i), type="lp"), 
+    lp.i <- predict(f.mi, newdata=mice::complete(imputed.training.data,i), type="lp")
+    rc <- Hmisc::rcorr.cens(-lp.i, 
                             S.MSM.5.training)
@@
     cindex.se.MSM.cal[i] <- rc["S.D."]/2
+
+    unoC.MSM[i] <- survC1::Est.Cval(mydata=cbind(S.MSM.5.training[, 1], 
+                                              S.MSM.5.training[, 2], 
+                                              lp.i), 
+                                 tau=5, nofit=TRUE)$Dhat
   }
@@
   final.full.table <- eval(parse(text=paste0("final.", type, ".table.Rec.5")))
-  final.full.table["C-index MSM", "HR"] <- sprintf("%.2f", cindex.mi.MSM.cal$est-optimism)
-  final.full.table["C-index MSM", "CI"] <- paste0("[", sprintf("%.2f", cindex.mi.MSM.cal$est+qnorm(.025)*cindex.mi.MSM.cal$se-optimism),
+  final.full.table["Harrell's C-index MSM", "HR"] <- sprintf("%.2f", 
+                                                cindex.mi.MSM.cal$est-optimism)
+  final.full.table["Harrell's C-index MSM", "CI"] <- paste0("[", 
+    sprintf("%.2f", cindex.mi.MSM.cal$est+qnorm(.025)*cindex.mi.MSM.cal$se-optimism),
     "; ", sprintf("%.2f", cindex.mi.MSM.cal$est+qnorm(.975)*cindex.mi.MSM.cal$se-optimism), "]")
+  final.full.table["Uno's C-index MSM", "HR"] <- sprintf("%.2f", mean(unoC.MSM)-optimism)
 
@@
                                     "Multifocal", "Extensive",
-                                    rownames(final.full.table.Rec.5)[(nrow(final.full.table.Rec.5)-1):nrow(final.full.table.Rec.5)])
+                                    rownames(final.full.table.Rec.5)[(nrow(final.full.table.Rec.5)-3):nrow(final.full.table.Rec.5)])
 final.full.table.MSM.5.refit <- as.data.frame(final.full.table.MSM.5.refit)
@@
                                           "Multifocal", "Extensive",
-                                          row.names(final.full.table.MSM.5.refit)[nrow(final.full.table.MSM.5.refit)])
+                                          row.names(final.full.table.MSM.5.refit)[(nrow(final.full.table.MSM.5.refit)-1):nrow(final.full.table.MSM.5.refit)])
 final.BS.table.Rec.5 <- as.data.frame(final.BS.table.Rec.5)
@@
                                      "SN tumour burden", 
-                                   row.names(final.BS.table.Rec.5)[(nrow(final.BS.table.Rec.5)-1):nrow(final.BS.table.Rec.5)])
+                                   row.names(final.BS.table.Rec.5)[(nrow(final.BS.table.Rec.5)-3):nrow(final.BS.table.Rec.5)])
 final.BS.table.MSM.5.refit <- as.data.frame(final.BS.table.MSM.5.refit)
@@
                                      "SN tumour burden",  
-                                   row.names(final.BS.table.MSM.5.refit)[nrow(final.BS.table.MSM.5.refit)])
+                                   row.names(final.BS.table.MSM.5.refit)[(nrow(final.BS.table.MSM.5.refit)-1):nrow(final.BS.table.MSM.5.refit)])
 

### 2023-12-11 Code/07_CreateCrossValidatedCalibrationPlots.Rmd
Added Uno's C-index

--- 2023-11-23 Code/7_CreateCrossValidatedCalibrationPlots.Rmd
@@
         out <- PredictionTools::val.surv.mi(p, S, g=5, main=paste("Center", Centers.anonimized[j]),
-                                            time=NULL,lim=c(0,1),dist=TRUE)
+                                            time=NULL,lim=c(0,1),dist=TRUE,
+                                            CI.metrics=TRUE)
         grDevices::dev.off()

### 2023-12-11 Code/10_AJCC.Rmd
Final commit
Added Uno's C-index

--- 2022-12-19 Code/10_AJCC.Rmd
@@
 author: "Carolien C.H.M. Maas"
-date: "December 2022"
+date: "November 2022"
 output: html_document
@@
   AJCC.7 <- c()
+  unoC.AJCC.7 <- rep(0, m)
   cindex.AJCC.7 <- rep(0, m)
   cindex.AJCC.7.se <- rep(0, m)
+  unoC.AJCC.7.pos <- rep(0, m)
   cindex.AJCC.7.pos <- rep(0, m)
   cindex.AJCC.7.se.pos <- rep(0, m)
+  unoC.AJCC.7.neg <- rep(0, m)
   cindex.AJCC.7.neg <- rep(0, m)
@@
 
-    # save TNM.7 and AJCC.7 for each imputated data set
+    # save TNM.7 and AJCC.7 for each imputed data set
     TNM.7 <- cbind(TNM.7, TNM.7.i)
@@
 
-    # compute C-index
+    # compute Harrell's C-index
     rc <- Hmisc::rcorr.cens(-AJCC.7.i, S.Rec.5) # add a minus to revert class order: from worst to best
@@
 
+    # Uno's c-index  
+    unoC.AJCC.7[i] <- survC1::Est.Cval(mydata=cbind(S.Rec.5[, 1], S.Rec.5[, 2], 
+                                              AJCC.7.i), 
+                                 tau=5, nofit=TRUE)$Dhat
+
     # compute C-index for only positive sentinel node patients
-    rc.pos <- Hmisc::rcorr.cens(-AJCC.7.i[imputation.i$SNstatus=="Positive"], S.Rec.5[imputation.i$SNstatus=="Positive",])
+    positives <- imputation.i$SNstatus=="Positive"
+    rc.pos <- Hmisc::rcorr.cens(-AJCC.7.i[positives], S.Rec.5[positives,])
     cindex.AJCC.7.pos[i]<-rc.pos["C Index"]
@@
 
+    # Uno's c-index  
+    unoC.AJCC.7.pos[i] <- survC1::Est.Cval(mydata=cbind(S.Rec.5[positives, 1], 
+                                              S.Rec.5[positives, 2], 
+                                              AJCC.7.i[positives]), 
+                                 tau=5, nofit=TRUE)$Dhat
+
     # compute C-index for only negative sentinel node patients
-    rc.neg <- Hmisc::rcorr.cens(-AJCC.7.i[imputation.i$SNstatus=="Negative"], S.Rec.5[imputation.i$SNstatus=="Negative",])
+    negatives <- imputation.i$SNstatus=="Negative"
+    rc.neg <- Hmisc::rcorr.cens(-AJCC.7.i[negatives], S.Rec.5[negatives,])
     cindex.AJCC.7.neg[i]<-rc.neg["C Index"]
     cindex.AJCC.7.se.neg[i]<-rc.neg["S.D."]/2
+
+    # Uno's c-index  
+    unoC.AJCC.7.neg[i] <- survC1::Est.Cval(mydata=cbind(S.Rec.5[negatives, 1], 
+                                              S.Rec.5[negatives, 2], 
+                                              AJCC.7.i[negatives]), 
+                                 tau=5, nofit=TRUE)$Dhat
   }
@@
                 paste0("[", sprintf("%.2f", cindex.mi.AJCC.7$est+qnorm(.025)*cindex.mi.AJCC.7$se),
-                        "; ", sprintf("%.2f", cindex.mi.AJCC.7$est+qnorm(.975)*cindex.mi.AJCC.7$se), "]")),
+                        "; ", sprintf("%.2f", cindex.mi.AJCC.7$est+qnorm(.975)*cindex.mi.AJCC.7$se), "]"),
+                sprintf("%.2f", mean(unoC.AJCC.7))),
+              c("Positive SN patients:", sprintf("%.2f", cindex.mi.AJCC.7.pos$est),
+                    paste0("[", sprintf("%.2f", cindex.mi.AJCC.7.pos$est+qnorm(.025)*cindex.mi.AJCC.7.pos$se),
+                           "; ", sprintf("%.2f", cindex.mi.AJCC.7.pos$est+qnorm(.975)*cindex.mi.AJCC.7.pos$se), "]"),
+                sprintf("%.2f", mean(unoC.AJCC.7.pos))),
               c("Negative SN patients:", sprintf("%.2f", cindex.mi.AJCC.7.neg$est),
                     paste0("[", sprintf("%.2f", cindex.mi.AJCC.7.neg$est+qnorm(.025)*cindex.mi.AJCC.7.neg$se),
-                           "; ", sprintf("%.2f", cindex.mi.AJCC.7.neg$est+qnorm(.975)*cindex.mi.AJCC.7.neg$se), "]")),
-              c("Positive SN patients:", sprintf("%.2f", cindex.mi.AJCC.7.pos$est),
-                    paste0("[", sprintf("%.2f", cindex.mi.AJCC.7.pos$est+qnorm(.025)*cindex.mi.AJCC.7.pos$se),
-                           "; ", sprintf("%.2f", cindex.mi.AJCC.7.pos$est+qnorm(.975)*cindex.mi.AJCC.7.pos$se), "]")))
+                           "; ", sprintf("%.2f", cindex.mi.AJCC.7.neg$est+qnorm(.975)*cindex.mi.AJCC.7.neg$se), "]"),
+                sprintf("%.2f", mean(unoC.AJCC.7.neg))))
 
@@
   AJCC.8 <- c()
+  unoC.AJCC.8 <- rep(0, m)
   cindex.AJCC.8 <- rep(0, m)
   cindex.AJCC.8.se <- rep(0, m)
+  unoC.AJCC.8.pos <- rep(0, m)
   cindex.AJCC.8.pos <- rep(0, m)
   cindex.AJCC.8.se.pos <- rep(0, m)
+  unoC.AJCC.8.neg <- rep(0, m)
   cindex.AJCC.8.neg <- rep(0, m)
@@
 
-    # save TNM.8 and AJCC.8 for each imputated data set
+    # save TNM.8 and AJCC.8 for each imputed data set
     TNM.8 <- cbind(TNM.8, TNM.8.i)
@@
 
-    # compute C-index
+    # compute Harrell's C-index
     rc <- Hmisc::rcorr.cens(-AJCC.8.i, S.Rec.5) # add a minus to revert class order: from worst to best
@@
 
+    # Uno's c-index  
+    unoC.AJCC.8[i] <- survC1::Est.Cval(mydata=cbind(S.Rec.5[, 1], S.Rec.5[, 2], 
+                                              AJCC.8.i), 
+                                 tau=5, nofit=TRUE)$Dhat
+
     # compute C-index for only positive sentinel node patients
-    rc.pos <- Hmisc::rcorr.cens(-AJCC.8.i[imputation.i$SNstatus=="Positive"], S.Rec.5[imputation.i$SNstatus=="Positive",])
+    positives <- imputation.i$SNstatus=="Positive"
+    rc.pos <- Hmisc::rcorr.cens(-AJCC.8.i[positives], S.Rec.5[positives,])
     cindex.AJCC.8.pos[i]<-rc.pos["C Index"]
     cindex.AJCC.8.se.pos[i]<-rc.pos["S.D."]/2
+    unoC.AJCC.8.pos[i] <- survC1::Est.Cval(mydata=cbind(S.Rec.5[positives, 1], 
+                                                    S.Rec.5[positives, 2], 
+                                                    AJCC.8.i[positives]), 
+                                 tau=5, nofit=TRUE)$Dhat
 
     # compute C-index for only negative sentinel node patients
-    rc.neg <- Hmisc::rcorr.cens(-AJCC.8.i[imputation.i$SNstatus=="Negative"], S.Rec.5[imputation.i$SNstatus=="Negative",])
+    negatives <- imputation.i$SNstatus=="Negative"
+    rc.neg <- Hmisc::rcorr.cens(-AJCC.8.i[negatives], S.Rec.5[negatives,])
     cindex.AJCC.8.neg[i]<-rc.neg["C Index"]
     cindex.AJCC.8.se.neg[i]<-rc.neg["S.D."]/2
+    unoC.AJCC.8.neg[i] <- survC1::Est.Cval(mydata=cbind(S.Rec.5[negatives, 1], 
+                                                    S.Rec.5[negatives, 2], 
+                                                    AJCC.8.i[negatives]), 
+                                 tau=5, nofit=TRUE)$Dhat
+
   }
@@
                           paste0("[", sprintf("%.2f", cindex.mi.AJCC.8$est+qnorm(.025)*cindex.mi.AJCC.8$se),
-                                 "; ", sprintf("%.2f", cindex.mi.AJCC.8$est+qnorm(.975)*cindex.mi.AJCC.8$se), "]")),
+                                 "; ", sprintf("%.2f", cindex.mi.AJCC.8$est+qnorm(.975)*cindex.mi.AJCC.8$se), "]"),
+                          sprintf("%.2f", mean(unoC.AJCC.8))),
+                    c("Positive SN patients:", sprintf("%.2f", cindex.mi.AJCC.8.pos$est),
+                          paste0("[", sprintf("%.2f", cindex.mi.AJCC.8.pos$est+qnorm(.025)*cindex.mi.AJCC.8.pos$se),
+                                 "; ", sprintf("%.2f", cindex.mi.AJCC.8.pos$est+qnorm(.975)*cindex.mi.AJCC.8.pos$se), "]"),
+                      sprintf("%.2f", mean(unoC.AJCC.8.pos))),
                     c("Negative SN patients", sprintf("%.2f", cindex.mi.AJCC.8.neg$est),
                           paste0("[", sprintf("%.2f", cindex.mi.AJCC.8.neg$est+qnorm(.025)*cindex.mi.AJCC.8.neg$se),
-                                 "; ", sprintf("%.2f", cindex.mi.AJCC.8.neg$est+qnorm(.975)*cindex.mi.AJCC.8.neg$se), "]")),
-                    c("Positive SN patients:", sprintf("%.2f", cindex.mi.AJCC.8.pos$est),
-                          paste0("[", sprintf("%.2f", cindex.mi.AJCC.8.pos$est+qnorm(.025)*cindex.mi.AJCC.8.pos$se),
-                                 "; ", sprintf("%.2f", cindex.mi.AJCC.8.pos$est+qnorm(.975)*cindex.mi.AJCC.8.pos$se), "]")))
+                                 "; ", sprintf("%.2f", cindex.mi.AJCC.8.neg$est+qnorm(.975)*cindex.mi.AJCC.8.neg$se), "]"),
+                      sprintf("%.2f", mean(unoC.AJCC.8.neg))))
 
@@
               row.names=FALSE, col.names=FALSE, sep=",")
-utils::write.table(cbind(CIndex.val, rep("", 3),
-                   validation.AJCC$AJCC.7.table[, -1], rep("", 3), 
-                   compare.AJCC.model[, 2:3], rep("", 3),
-                   validation.AJCC$AJCC.8.table[,-1], rep("", 3),
+utils::write.table(cbind(CIndex.val,
+                   validation.AJCC$AJCC.7.table[, -1],
+                   compare.AJCC.model[, 2:3],
+                   validation.AJCC$AJCC.8.table[,-1],
                    compare.AJCC.model[, 4:5]),

### 2023-12-11 Code/12_Validation.Rmd
Added Uno's C-index

@@
                                  g=5, main="All patients", time=NULL, 
-                                 lim=c(0,1), dist=TRUE,
+                                 lim=c(0,1), dist=TRUE, CI.metrics=TRUE,
                                  optimism.C=ifelse(data.type=="training", optimism, 0))
@@
                                  g=5, main="Patients with positive SN", 
-                                 time=NULL, lim=c(0,1), dist=TRUE,
+                                 time=NULL, lim=c(0,1), dist=TRUE, CI.metrics=TRUE,
                                  optimism.C=ifelse(data.type=="training", optimism, 0))
@@
                                  g=5, main="Patients with negative SN", 
-                                 time=NULL, lim=c(0,1), dist=TRUE,
+                                 time=NULL, lim=c(0,1), dist=TRUE, CI.metrics=TRUE,
                                  optimism.C=ifelse(data.type=="training", optimism, 0))

### 2023-12-11 Code/AdditionalAnalysis_Review.R

@@ -1,8 +1,15 @@
+# clear all
+rm(list = ls(all.names = TRUE))
+if(!is.null(dev.list())) dev.off()
+cat("\014")
+
 #libraries
 library(Hmisc)
-library(survAUC)
-library(pROC)
+library(survivalROC)
 library(mice)
+library(survAUC) # Uno's C-index not by Uno
+library(survC1)  # Uno's C-index by Uno
+library(dplyr)
 
 # load data
 load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData")
@@ -10,25 +17,87 @@ load("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData")
 file.path <- "Z:/Project Melanoom/PaperMelanoma/Results/"
 
 # calculate Harrell's C and Uno's C
+lp.train <- c()
+lp.val <- c()
 Harrell.Rec.train <- c()
 Uno.Rec.train <- c()
+Uno.Rec.train.correct <- c()
 Harrell.Rec.val <- c()
 Uno.Rec.val <- c()
+Uno.Rec.val.correct <- c()
 for (i in 1:m){
-  lp.train <- predict(f.mi.BS.Rec.5, newdata=mice::complete(imputed.training.data, i), type="lp")
-  Harrell.Rec.train <- c(Harrell.Rec.train, Hmisc::rcorr.cens(-lp.train, S=S.Rec.5.training)["C Index"])
-  Uno.Rec.train <- c(Uno.Rec.train, survAUC::UnoC(Surv.rsp=S.Rec.5.training, Surv.rsp.new=S.Rec.5.training, lpnew=lp.train, time=5))
+  lp.train.i <- predict(f.mi.BS.Rec.5,
+                      newdata=mice::complete(imputed.training.data, i),
+                      type="lp")
+  lp.train <- cbind(lp.train, lp.train.i)
+  Harrell.Rec.train <- c(Harrell.Rec.train,
+                         Hmisc::rcorr.cens(-lp.train.i, S=S.Rec.5.training)["C Index"])
+  Uno.Rec.train <- c(Uno.Rec.train, survAUC::UnoC(Surv.rsp=S.Rec.5.training,
+                                                  Surv.rsp.new=S.Rec.5.training,
+                                                  lpnew=lp.train.i, time=5))
 
[…]

### 2024-01-19 Code/05_FitModels.Rmd
Final edits

--- 2023-11-23 Code/5_FitModels.Rmd
@@
                                         fitargs=list(y=TRUE,x=TRUE,se.fit=TRUE))
+
+# check Cox Proportional Hazards Assumption
+PH.assumption <- survival::cox.zph(f.mi.BS.Rec.5)
+
 # backward selection
@@
             "imputed.training.data", "last.imputed.training.data.set", 
-            "horizon", "S.Rec.5.training", "S.MSM.5.training",
+            "horizon", 
+            "S.Rec.training", "S.MSM.training",
+            "S.Rec.5.training", "S.MSM.5.training",
             "f.mi.full.Rec.5", "f.mi.BS.Rec.5", "f.mi.full.MSM.5.refit", 
@@
             "imputed.validation.data", "last.imputed.validation.data.set", "horizon", 
+            "S.Rec.validation", "S.MSM.validation", 
             "S.Rec.5.validation", "S.MSM.5.validation", "f.mi.BS.Rec.5"),
@@
 }
-PredictionTools::val.surv.mi(p.MSM.5.redev, S.MSM.5.training, g=5,
-                             main="All patients", time=NULL, lim=c(0, 1),
+PredictionTools::val.surv.mi(p.MSM.5.redev, S.MSM.training, g=5,
+                             main="All patients", time=5, lim=c(0, 1),
                              dist=TRUE)

### 2024-01-19 Code/06_CreateTablesandPlots.Rmd
Final edits

--- 2023-11-23 Code/6_CreateTablesandPlots.Rmd
@@
     # plot predict
-    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/plot.predict.Rec.png"))
+    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/plot.predict.Rec.png"),
+        width=1800, height=1800, units="px", res=300)
     plot.predict <- plot(Predict(f.mi.full.Rec.5))
@@
     # plot predict Breslow
-    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/plot.predict.Breslow.Rec.png"))
+    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/plot.predict.Breslow.Rec.png"),
+        width=1800, height=1800, units="px", res=300)
     plot.Breslow <- plot(Predict(f.mi.full.Rec.5, Breslow=0:20, SNstatus), ~ Breslow, groups="SNstatus",
@@
     if (model=="Rec.5"){
-      S <- S.Rec.5.training
+      S <- S.Rec.training
     }
     else if (model=="MSM.5.refit"){
-      S <- S.MSM.5.training
+      S <- S.MSM.training
     }
 
-    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/all.centers.", model, ".", type, ".png"))
-    out <- PredictionTools::val.surv.mi(p,S,g=5,main="All centers",
-                                        time=NULL,lim=c(0,1),dist=TRUE)
+    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/all.centers.", model, ".", type, ".png"),
+        width=1800, height=1800, units="px", res=300)
+    out <- PredictionTools::val.surv.mi(p=p, y=S, g=5, time=5,
+                                        main="All centers", CI.metrics=TRUE,
+                                        lim=c(0,1), dist=TRUE, n.iter=40)
     grDevices::dev.off()
@@
     final.full.table <- eval(parse(text=paste0("final.", type, ".table.", model)))
-    final.full.table["C-index", "HR"] <- sprintf("%.2f", out$cindex-optimism)
-    final.full.table["C-index", "CI"] <- paste0("[", sprintf("%.2f", out$cindex.lower-optimism),
+    final.full.table["Harrell's C-index", "HR"] <- sprintf("%.2f", out$cindex)
+    final.full.table["Harrell's C-index", "CI"] <- paste0("[", sprintf("%.2f", out$cindex.lower),
+                                                   "; ", sprintf("%.2f", out$cindex.upper), "]")
+    final.full.table["Harrell's optimism-corrected C-index", "HR"] <- sprintf("%.2f", out$cindex-optimism)
+    final.full.table["Harrell's optimism-corrected C-index", "CI"] <- paste0("[", sprintf("%.2f", out$cindex.lower-optimism),
                                                    "; ", sprintf("%.2f", out$cindex.upper-optimism), "]")
+    final.full.table["Uno's C-index", "HR"] <- sprintf("%.2f", out$unoC)
+    final.full.table["Uno's C-index", "CI"] <- paste0("[", sprintf("%.2f", out$unoC.lower),
+                                                   "; ", sprintf("%.2f", out$unoC.upper), "]")
+    final.full.table["AUC", "HR"] <- sprintf("%.2f", out$AUC)
+    final.full.table["AUC", "CI"] <- paste0("[", sprintf("%.2f", out$AUC.lower),
+                                                   "; ", sprintf("%.2f", out$AUC.upper), "]")
 
@@
   f.mi <- eval(parse(text=paste0("f.mi.", type, ".Rec.5")))
-  cindex.MSM.cal <- rep(0, m)
-  cindex.se.MSM.cal <- rep(0, m)
+  cindex.MSM <- rep(0, m)
+  cindex.se.MSM <- rep(0, m)
+  unoC.MSM <- rep(0, m)
+  AUC.MSM <- rep(0, m)
   for (i in 1:m){
-    rc <- Hmisc::rcorr.cens(-predict(f.mi,
-                                     newdata=mice::complete(imputed.training.data,i), type="lp"), 
+    lp.i <- predict(f.mi, newdata=mice::complete(imputed.training.data,i), type="lp")
+
+    # Harrell's C-index
+    rc <- Hmisc::rcorr.cens(-lp.i, 
                             S.MSM.5.training)
-    cindex.MSM.cal[i] <- rc["C Index"]
-    cindex.se.MSM.cal[i] <- rc["S.D."]/2
+    cindex.MSM[i] <- rc["C Index"]
+    cindex.se.MSM[i] <- rc["S.D."]/2
+
+    # Uno's C-index
+    Uno.data <- cbind(S.MSM.5.training[, 1], S.MSM.5.training[, 2], lp.i)
+    unoC.MSM[i] <- survC1::Est.Cval(mydata=Uno.data, 
+                                    tau=horizon, nofit=TRUE)$Dhat
+    if (i == 1){
+      unoC.boot <- c()
+      nr.iter <- 40
+      for (itr in 1:nr.iter){
+        boot <- Uno.data[sample(1:nrow(Uno.data), nrow(Uno.data), replace=TRUE),]
+        unoC <- survC1::Est.Cval(mydata=boot,
+                                  tau=horizon,
+                                  nofit=TRUE)$Dhat
+        unoC.boot <- c(unoC.boot, unoC)
       }
-  cindex.mi.MSM.cal <- Rubin.combine(cindex.MSM.cal, cindex.se.MSM.cal)
+      unoC.MSM.se <- PredictionTools::Rubin.combine(mean(unoC.boot), sd(unoC.boot))$se
+    }
+
+    # time-dependent ROC
+    AUC.MSM.i <- timeROC::timeROC(T=S.MSM.training[, 1],
+                                   delta=S.MSM.training[, 2],
+                                   marker=lp.i,
+                                   times=horizon,
+                                   cause=1, 
+                                   iid=FALSE)
+
+    # time-dependent ROC SE
+    if (i == 1){
+      AUC.MSM.i <- timeROC::timeROC(T=S.MSM.training[, 1],
+                                   delta=S.MSM.training[, 2],
+                                   marker=lp.i,
+                                   times=horizon,
+                                   cause=1, 
+                                   iid=TRUE)
+      AUC.MSM.CI <- stats::confint(AUC.MSM.i)
+      AUC.MSM.se <- as.numeric((as.numeric(AUC.MSM.CI$CI_AUC[2])/100-as.numeric(AUC.MSM.i$AUC[2]))/AUC.MSM.CI$C.alpha)
+    }
+    AUC.MSM[i] <- as.numeric(AUC.MSM.i$AUC[2])
+  }
+  cindex.mi.MSM <- Rubin.combine(cindex.MSM, cindex.se.MSM)
 
   final.full.table <- eval(parse(text=paste0("final.", type, ".table.Rec.5")))
-  final.full.table["C-index MSM", "HR"] <- sprintf("%.2f", cindex.mi.MSM.cal$est-optimism)
-  final.full.table["C-index MSM", "CI"] <- paste0("[", sprintf("%.2f", cindex.mi.MSM.cal$est+qnorm(.025)*cindex.mi.MSM.cal$se-optimism),
-                                                 "; ", sprintf("%.2f", cindex.mi.MSM.cal$est+qnorm(.975)*cindex.mi.MSM.cal$se-optimism), "]")
+  final.full.table["Harrell's C-index MSM", "HR"] <- sprintf("%.2f", cindex.mi.MSM$est)
+  final.full.table["Harrell's C-index MSM", "CI"] <- paste0("[", 
+    sprintf("%.2f", cindex.mi.MSM$est+qnorm(.025)*cindex.mi.MSM$se),
+    "; ", sprintf("%.2f", cindex.mi.MSM$est+qnorm(.975)*cindex.mi.MSM$se), "]")
+  final.full.table["Harrell's optimism-corrected C-index MSM", "HR"] <- sprintf("%.2f", cindex.mi.MSM$est-optimism)
+  final.full.table["Harrell's optimism-corrected C-index MSM", "CI"] <- paste0("[", 
+    sprintf("%.2f", cindex.mi.MSM$est+qnorm(.025)*cindex.mi.MSM$se-optimism),
+    "; ", sprintf("%.2f", cindex.mi.MSM$est+qnorm(.975)*cindex.mi.MSM$se-optimism), "]")
+  final.full.table["Uno's C-index MSM", "HR"] <- sprintf("%.2f", mean(unoC.MSM))
+  final.full.table["Uno's C-index MSM", "CI"] <- paste0("[", 
+    sprintf("%.2f", mean(unoC.MSM)+qnorm(.025)*unoC.MSM.se),
+    "; ", sprintf("%.2f", mean(unoC.MSM)+qnorm(.975)*unoC.MSM.se), "]")
+  final.full.table["AUC MSM", "HR"] <- sprintf("%.2f", mean(AUC.MSM))
+  final.full.table["AUC MSM", "CI"] <- paste0("[", 
+    sprintf("%.2f", mean(AUC.MSM)+qnorm(.025)*AUC.MSM.se),
+    "; ", sprintf("%.2f", mean(AUC.MSM)+qnorm(.975)*AUC.MSM.se), "]")
 
@@
                                     "Multifocal", "Extensive",
-                                    rownames(final.full.table.Rec.5)[(nrow(final.full.table.Rec.5)-1):nrow(final.full.table.Rec.5)])
+                                    rownames(final.full.table.Rec.5)[(nrow(final.full.table.Rec.5)-7):nrow(final.full.table.Rec.5)])
 final.full.table.MSM.5.refit <- as.data.frame(final.full.table.MSM.5.refit)
@@
                                           "Multifocal", "Extensive",
-                                          row.names(final.full.table.MSM.5.refit)[nrow(final.full.table.MSM.5.refit)])
+                                          row.names(final.full.table.MSM.5.refit)[(nrow(final.full.table.MSM.5.refit)-3):nrow(final.full.table.MSM.5.refit)])
 final.BS.table.Rec.5 <- as.data.frame(final.BS.table.Rec.5)
@@
                                      "SN tumour burden", 
-                                   row.names(final.BS.table.Rec.5)[(nrow(final.BS.table.Rec.5)-1):nrow(final.BS.table.Rec.5)])
+                                   row.names(final.BS.table.Rec.5)[(nrow(final.BS.table.Rec.5)-7):nrow(final.BS.table.Rec.5)])
 final.BS.table.MSM.5.refit <- as.data.frame(final.BS.table.MSM.5.refit)
@@
                                      "SN tumour burden",  
-                                   row.names(final.BS.table.MSM.5.refit)[nrow(final.BS.table.MSM.5.refit)])
+                                   row.names(final.BS.table.MSM.5.refit)[(nrow(final.BS.table.MSM.5.refit)-3):nrow(final.BS.table.MSM.5.refit)])
 
@@
 
+# optimism-corrected
 CIndex.Rec.Table <- rbind(CIndex.BS.Rec.5.All, CIndex.BS.Rec.5.Positive, CIndex.BS.Rec.5.Negative)
 CIndex.Rec.Table
-utils::write.table(CIndex.Rec.Table,
-              file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/CIndex.train.txt"),
-              row.names=FALSE, col.names=FALSE, sep=",")
+# utils::write.table(CIndex.Rec.Table,
+#               file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/CIndex.train.txt"),
+#               row.names=FALSE, col.names=FALSE, sep=",")
 

### 2024-01-19 Code/07_CreateCrossValidatedCalibrationPlots.Rmd
Final edits

--- 2023-11-23 Code/7_CreateCrossValidatedCalibrationPlots.Rmd
@@
   if (model == "Rec.5"){
+    S.full <- S.Rec.training
     S.model <- S.Rec.5.training
@@
   if (model == "MSM.5"){
+    S.full <- S.MSM.training
     S.model <- S.MSM.5.training
@@
       S.j<-S.model[training.data.to.be.imputed$Center==Centers[j],]
+      S.full.j<-S.full[training.data.to.be.imputed$Center==Centers[j],]
       S.notj<-S.model[training.data.to.be.imputed$Center!=Centers[j],]
@@
         # calibration with metrics
-        S <- S.j[select.patients,]
+        # S.full <- S.full.j[select.patients,]
+        S <- S.full.j[select.patients,]
         p <- matrix(NA, nrow(S), m)
@@
         # omit the observation with predicted probability = 1
-        omit <- unique(c(which(p[,1]==1), which(p[,2]==1), which(p[,3]==1), which(p[,4]==1), which(p[,5]==1)))
-        if (length(omit)>0){
-          cat("Omit:", omit, "\n")
-          p <- p[-omit,]
-          S <- S[-omit,]
-        }
+        # omit <- unique(c(which(p[,1]==1), which(p[,2]==1), which(p[,3]==1), which(p[,4]==1), which(p[,5]==1)))
+        # if (length(omit)>0){
+        #   cat("Omit:", omit, "\n")
+        #   p <- p[-omit,]
+        #   S.full <- S.full[-omit,]
+        # }
 
-        png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", model, ".", select.value, ".", type, ".", Centers[j], ".png"))
-        out <- PredictionTools::val.surv.mi(p, S, g=5, main=paste("Center", Centers.anonimized[j]),
-                                            time=NULL,lim=c(0,1),dist=TRUE)
+        png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", model, ".", select.value, ".", type, ".", Centers[j], ".png"),
+            width=1800, height=1800, units="px", res=300)
+        out <- PredictionTools::val.surv.mi(p, S, g=5, 
+                                            main=paste("Center", Centers.anonimized[j]),
+                                            time=5, lim=c(0,1), dist=TRUE,
+                                            CI.metrics=TRUE, n.iter=40)
         grDevices::dev.off()

### 2024-01-19 Code/08_CreateNomogram.Rmd
Final edits

@@
             "imputed.training.data",  "last.imputed.training.data.set", 
-            "horizon", "S.Rec.5.training", "S.MSM.5.training",
+            "horizon", 
+            "S.Rec.training", "S.MSM.training",
+            "S.Rec.5.training", "S.MSM.5.training",
             "f.mi.full.Rec.5", "f.mi.BS.Rec.5", "f.mi.full.MSM.5.refit", 

### 2024-01-19 Code/09_CreateRiskScoreHistogram.Rmd
Final edits

--- 2023-11-23 Code/9_CreateRiskScoreHistogram.Rmd
@@
 y.lim.hist <- c(0, 25)
-png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Risk.distribution.png"))
+png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Risk.distribution.png"),
+    width=1800, height=1800, units="px", res=300)
 par(mar = c(5,5,2,5))
@@
             "imputed.training.data", "last.imputed.training.data.set", 
-            "horizon", "S.Rec.5.training", "S.MSM.5.training",
+            "horizon", 
+            "S.Rec.training", "S.MSM.training",
+            "S.Rec.5.training", "S.MSM.5.training",
             "f.mi.full.Rec.5", "f.mi.BS.Rec.5", 
@@
         legend.title=element_blank())
-png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Risk.distribution.png"), 
-    width=15, height=15, units="cm", res=300)
-show(plot)
-grDevices::dev.off()
+ggsave(file="Z:/Project Melanoom/PaperMelanoma/Results/Risk.distribution.pdf",
+       width=15, height=15, units="cm", dpi=300)
 ```

### 2024-01-19 Code/10_AJCC.Rmd
Final edits

--- 2022-12-19 Code/10_AJCC.Rmd
@@
 library(compareC)         # compare AJCC to model
+library(survival)
 ```
@@
 ```{r, eval=TRUE}
-calculate.AJCC.7.8 <- function(imputed.data=NULL, type=NULL, S.Rec.5=NULL){
+calculate.AJCC.7.8 <- function(imputed.data=NULL, type=NULL, 
+                               S.Rec.5=NULL, S.Rec=NULL){
   ####
@@
   AJCC.7 <- c()
-  cindex.AJCC.7 <- rep(0, m)
-  cindex.AJCC.7.se <- rep(0, m)
-  cindex.AJCC.7.pos <- rep(0, m)
-  cindex.AJCC.7.se.pos <- rep(0, m)
-  cindex.AJCC.7.neg <- rep(0, m)
-  cindex.AJCC.7.se.neg <- rep(0, m)
   for (i in 1:m){
-    imputation.i <- mice::complete(imputed.data,i)
+    # imputed data set i
+    imputation.i <- mice::complete(imputed.data, i)
+
+    # calcualte TNM and AJCC for each imputation
     TNM.7.i <- rep("", nrow(imputation.i))
@@
     }
-
-    # save TNM.7 and AJCC.7 for each imputated data set
+    # save TNM.7 and AJCC.7 for each imputed data set
     TNM.7 <- cbind(TNM.7, TNM.7.i)
     AJCC.7 <- cbind(AJCC.7, AJCC.7.i)
-
-    # compute C-index
-    rc <- Hmisc::rcorr.cens(-AJCC.7.i, S.Rec.5) # add a minus to revert class order: from worst to best
-    cindex.AJCC.7[i]<-rc["C Index"]
-    cindex.AJCC.7.se[i]<-rc["S.D."]/2
-
-    # compute C-index for only positive sentinel node patients
-    rc.pos <- Hmisc::rcorr.cens(-AJCC.7.i[imputation.i$SNstatus=="Positive"], S.Rec.5[imputation.i$SNstatus=="Positive",])
-    cindex.AJCC.7.pos[i]<-rc.pos["C Index"]
-    cindex.AJCC.7.se.pos[i]<-rc.pos["S.D."]/2
-
-    # compute C-index for only negative sentinel node patients
-    rc.neg <- Hmisc::rcorr.cens(-AJCC.7.i[imputation.i$SNstatus=="Negative"], S.Rec.5[imputation.i$SNstatus=="Negative",])
-    cindex.AJCC.7.neg[i]<-rc.neg["C Index"]
-    cindex.AJCC.7.se.neg[i]<-rc.neg["S.D."]/2
   }
-
-  # combine C-index for each imputed data set
-  cindex.mi.AJCC.7<-Rubin.combine(cindex.AJCC.7,cindex.AJCC.7.se)
-  cindex.mi.AJCC.7.pos<-Rubin.combine(cindex.AJCC.7.pos,cindex.AJCC.7.se.pos)
-  cindex.mi.AJCC.7.neg<-Rubin.combine(cindex.AJCC.7.neg,cindex.AJCC.7.se.neg)
-
-  # save C-index and confidence interval
-  AJCC.7.table <- rbind(c("All patients:", sprintf("%.2f", cindex.mi.AJCC.7$est),
-                paste0("[", sprintf("%.2f", cindex.mi.AJCC.7$est+qnorm(.025)*cindex.mi.AJCC.7$se),
-                        "; ", sprintf("%.2f", cindex.mi.AJCC.7$est+qnorm(.975)*cindex.mi.AJCC.7$se), "]")),
-              c("Negative SN patients:", sprintf("%.2f", cindex.mi.AJCC.7.neg$est),
-                    paste0("[", sprintf("%.2f", cindex.mi.AJCC.7.neg$est+qnorm(.025)*cindex.mi.AJCC.7.neg$se),
-                           "; ", sprintf("%.2f", cindex.mi.AJCC.7.neg$est+qnorm(.975)*cindex.mi.AJCC.7.neg$se), "]")),
-              c("Positive SN patients:", sprintf("%.2f", cindex.mi.AJCC.7.pos$est),
-                    paste0("[", sprintf("%.2f", cindex.mi.AJCC.7.pos$est+qnorm(.025)*cindex.mi.AJCC.7.pos$se),
-                           "; ", sprintf("%.2f", cindex.mi.AJCC.7.pos$est+qnorm(.975)*cindex.mi.AJCC.7.pos$se), "]")))
 
@@
   AJCC.8 <- c()
-  cindex.AJCC.8 <- rep(0, m)
-  cindex.AJCC.8.se <- rep(0, m)
-  cindex.AJCC.8.pos <- rep(0, m)
-  cindex.AJCC.8.se.pos <- rep(0, m)
-  cindex.AJCC.8.neg <- rep(0, m)
-  cindex.AJCC.8.se.neg <- rep(0, m)
   for (i in 1:m){
+    # imputed data set i
     imputation.i <- mice::complete(imputed.data,i)
+
+    # calculate TNM and AJCC for each imputed data set i
     TNM.8.i <- rep("", nrow(imputation.i))
@@
     }
-
-    # save TNM.8 and AJCC.8 for each imputated data set
+    # save TNM.8 and AJCC.8 for each imputed data set
     TNM.8 <- cbind(TNM.8, TNM.8.i)
     AJCC.8 <- cbind(AJCC.8, AJCC.8.i)
+  }
 
-    # compute C-index
-    rc <- Hmisc::rcorr.cens(-AJCC.8.i, S.Rec.5) # add a minus to revert class order: from worst to best
-    cindex.AJCC.8[i]<-rc["C Index"]
-    cindex.AJCC.8.se[i]<-rc["S.D."]/2
+  for (AJCC.nr in c(7, 8)){  
+    AJCC.table <- c()
+    for (select.value in c("All", "Positive", "Negative")){
+      cindex <- rep(0, m)
+      cindex.se <- rep(0, m)
+      unoC <- rep(0, m)
+      tdAUC <- rep(0, m)
+      for (i in 1:m){
+        # select the AJCC 7 or 8 for the ith imputed data set
+        AJCC.i <- eval(parse(text=paste0("AJCC.", AJCC.nr, "[, i]")))
 
-    # compute C-index for only positive sentinel node patients
-    rc.pos <- Hmisc::rcorr.cens(-AJCC.8.i[imputation.i$SNstatus=="Positive"], S.Rec.5[imputation.i$SNstatus=="Positive",])
-    cindex.AJCC.8.pos[i]<-rc.pos["C Index"]
-    cindex.AJCC.8.se.pos[i]<-rc.pos["S.D."]/2
+        # identify those patients that are selected
+        if (select.value == "All"){
+          select.patients <- TRUE
+        } else if (select.value != "All"){
+          select.patients <- imputation.i$SNstatus==select.value
+        }
 
-    # compute C-index for only negative sentinel node patients
-    rc.neg <- Hmisc::rcorr.cens(-AJCC.8.i[imputation.i$SNstatus=="Negative"], S.Rec.5[imputation.i$SNstatus=="Negative",])
-    cindex.AJCC.8.neg[i]<-rc.neg["C Index"]
-    cindex.AJCC.8.se.neg[i]<-rc.neg["S.D."]/2
+        # compute Harrell's C-index
+        # add a minus to revert class order: from worst to best
+        rc <- Hmisc::rcorr.cens(-AJCC.i[select.patients], S.Rec.5[select.patients,]) 
+        cindex[i] <- rc["C Index"]
+        cindex.se[i] <- rc["S.D."]/2
+
+        # Uno's C-index
+        Uno.data <- cbind(S.Rec.5[select.patients, 1], 
+                          S.Rec.5[select.patients, 2], 
+                          AJCC.i[select.patients])
+        unoC[i] <- survC1::Est.Cval(mydata=Uno.data, 
+                                    tau=horizon, 
+                                    nofit=TRUE)$Dhat
+        if (i == 1){
+          unoC.boot <- c()
+          n.iter <- 40
+          for (itr in 1:n.iter){
+            boot <- Uno.data[sample(1:nrow(Uno.data), nrow(Uno.data), replace=TRUE),]
+            unoC <- survC1::Est.Cval(mydata=boot,
+                                     tau=horizon,
+                                     nofit=TRUE)$Dhat
+            unoC.boot <- c(unoC.boot, unoC)
           }
-  # combine C-index for each imputed data set
-  cindex.mi.AJCC.8<-Rubin.combine(cindex.AJCC.8,cindex.AJCC.8.se)
-  cindex.mi.AJCC.8.pos<-Rubin.combine(cindex.AJCC.8.pos,cindex.AJCC.8.se.pos)
-  cindex.mi.AJCC.8.neg<-Rubin.combine(cindex.AJCC.8.neg,cindex.AJCC.8.se.neg)
+          unoC.se <- PredictionTools::Rubin.combine(mean(unoC.boot), sd(unoC.boot))$se
+        }
 
-  # save C-index and confidence interval
-  AJCC.8.table <- rbind(c("All patients:", sprintf("%.2f", cindex.mi.AJCC.8$est),
-                          paste0("[", sprintf("%.2f", cindex.mi.AJCC.8$est+qnorm(.025)*cindex.mi.AJCC.8$se),
-                                 "; ", sprintf("%.2f", cindex.mi.AJCC.8$est+qnorm(.975)*cindex.mi.AJCC.8$se), "]")),
-                    c("Negative SN patients", sprintf("%.2f", cindex.mi.AJCC.8.neg$est),
-                          paste0("[", sprintf("%.2f", cindex.mi.AJCC.8.neg$est+qnorm(.025)*cindex.mi.AJCC.8.neg$se),
-                                 "; ", sprintf("%.2f", cindex.mi.AJCC.8.neg$est+qnorm(.975)*cindex.mi.AJCC.8.neg$se), "]")),
-                    c("Positive SN patients:", sprintf("%.2f", cindex.mi.AJCC.8.pos$est),
-                          paste0("[", sprintf("%.2f", cindex.mi.AJCC.8.pos$est+qnorm(.025)*cindex.mi.AJCC.8.pos$se),
-                                 "; ", sprintf("%.2f", cindex.mi.AJCC.8.pos$est+qnorm(.975)*cindex.mi.AJCC.8.pos$se), "]")))
+        # time-dependent ROC
+        tdAUC.i <- timeROC::timeROC(T=S.Rec[select.patients, 1],
+                                       delta=S.Rec[select.patients, 2],
+                                       marker=AJCC.i[select.patients],
+                                       times=horizon,
+                                       cause=1, 
+                                       iid=FALSE)
+
+        # time-dependent ROC SE
+        if (i == 1){
+          tdAUC.i <- timeROC::timeROC(T=S.Rec[select.patients, 1],
+                                       delta=S.Rec[select.patients, 2],
+                                       marker=AJCC.i[select.patients],
+                                       times=horizon,
+                                       cause=1, 
+                                       iid=TRUE)
+          tdAUC.CI <- stats::confint(tdAUC.i)
+          tdAUC.se <- as.numeric((as.numeric(tdAUC.CI$CI_AUC[2])/100-as.numeric(tdAUC.i$AUC[2]))/tdAUC.CI$C.alpha)
+        }
+        tdAUC[i] <- as.numeric(tdAUC.i$AUC[2])
+      }
+      cindex.mi<-Rubin.combine(cindex, cindex.se)
+
+      # add to Table
+      AJCC.table <- rbind(AJCC.table, 
+                          c(paste(select.value, "patients"), sprintf("%.2f", cindex.mi$est),
+                            paste0("[", sprintf("%.2f", cindex.mi$est+qnorm(.025)*cindex.mi$se),
+                                    "; ", sprintf("%.2f", cindex.mi$est+qnorm(.975)*cindex.mi$se), "]"),
+                            sprintf("%.2f", mean(unoC)),
+                            paste0("[", sprintf("%.2f", mean(unoC)+qnorm(.025)*unoC.se),
+                                    "; ", sprintf("%.2f", mean(unoC)+qnorm(.975)*unoC.se), "]"),
+                            sprintf("%.2f", mean(tdAUC)),
+                            paste0("[", sprintf("%.2f", mean(tdAUC)+qnorm(.025)*tdAUC.se),
+                                    "; ", sprintf("%.2f", mean(tdAUC)+qnorm(.975)*tdAUC.se), "]")))
+    }
+    assign(paste0("AJCC.", AJCC.nr, ".table"), AJCC.table)
+  }
 
@@
   }
-
-  # print(sort(unique(AJCC.7[,1])))
-  # print(sort(unique(AJCC.8[,1])))
 
@@
   }
+
   return(list(AJCC.7=AJCC.7, AJCC.7.table=AJCC.7.table,
@@
 load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData")
-train.AJCC <- calculate.AJCC.7.8(imputed.data=imputed.training.data, S.Rec.5=S.Rec.5.training, type="training")
-load("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData")
-validation.AJCC <- calculate.AJCC.7.8(imputed.data=imputed.validation.data, S.Rec.5=S.Rec.5.validation, type="validation")
-```
-
-# Compare C-index of AJCC and developed model
-```{r, eval=TRUE}
-for (edition in c(7, 8)){
-  for (patient.type in c("All", "Negative", "Positive")){
-    model.C <- c()
-    se.C <- c()
-    diff <- c()
-    se <- c()
-    for (i in 1:m){
-      # compute linear predictor
-      lp.i <- predict(f.mi.BS.Rec.5,
-                    newdata=mice::complete(imputed.validation.data, i),
-                    type="lp")
-
-      if (patient.type=="All"){
-        lp.i <- lp.i
-        S <- S.Rec.5.validation
-        AJCC <- eval(parse(text=paste0("validation.AJCC$AJCC.", edition, "[, i]")))
-      } else if (patient.type=="Positive"){
-        lp.i <- lp.i[validation.data.to.be.imputed$SNstatus=="Positive"]
-        S <- S.Rec.5.validation[validation.data.to.be.imputed$SNstatus=="Positive",]
-        AJCC <- eval(parse(text=paste0("validation.AJCC$AJCC.", edition, "[validation.data.to.be.imputed$SNstatus==\"Positive\", i]")))
-      } else if (patient.type=="Negative"){
-        lp.i <- lp.i[validation.data.to.be.imputed$SNstatus=="Negative"]
-        S <- S.Rec.5.validation[validation.data.to.be.imputed$SNstatus=="Negative",]
-        AJCC <- eval(parse(text=paste0("validation.AJCC$AJCC.", edition, "[validation.data.to.be.imputed$SNstatus==\"Negative\", i]")))
-      }
-
-      # compute predictions from model
-      p.i <- 1-exp(-h0.Rec*exp(lp.i))
-
-      # compare AJCC to model
-      compare.i <- compareC::compareC(S[,1], S[,2], -p.i, -AJCC)
-
-      model.C <- c(model.C, compare.i$est.c["Cxy"])
-      se.C <- c(se.C, sqrt(compare.i$est.varCxy))
-
-      # save differences and SE of difference
-      diff <- c(diff, compare.i$est.diff_c)
-      se <- c(se, sqrt(compare.i$est.vardiff_c))
-    }
-    assign(paste0("diff.", patient.type, ".", edition), Rubin.combine(diff, se))
-    assign(paste0("model.", patient.type), Rubin.combine(model.C, se.C))
-  }
-}
-
-# save in Table
-compare.AJCC.model <- rbind(c("All patients:", 
-                sprintf("%.2f", diff.All.7$est),
-                paste0("[", sprintf("%.2f", diff.All.7$est+qnorm(.025)*diff.All.7$se),
-                        "; ", sprintf("%.2f", diff.All.7$est+qnorm(.975)*diff.All.7$se), "]"),
-                sprintf("%.2f", diff.All.8$est),
-                paste0("[", sprintf("%.2f", diff.All.8$est+qnorm(.025)*diff.All.8$se),
-                        "; ", sprintf("%.2f", diff.All.8$est+qnorm(.975)*diff.All.8$se), "]")),
-              c("Negative SN patients:", 
-                sprintf("%.2f", diff.Negative.7$est),
-                paste0("[", sprintf("%.2f", diff.Negative.7$est+qnorm(.025)*diff.Negative.7$se),
-                       "; ", sprintf("%.2f", diff.Negative.7$est+qnorm(.975)*diff.Negative.7$se), "]"),
-                sprintf("%.2f", diff.Negative.8$est),
-                paste0("[", sprintf("%.2f", diff.Negative.8$est+qnorm(.025)*diff.Negative.8$se),
-                       "; ", sprintf("%.2f", diff.Negative.8$est+qnorm(.975)*diff.Negative.8$se), "]")),
-              c("Positive SN patients:", 
-                sprintf("%.2f", diff.Positive.7$est),
-                paste0("[", sprintf("%.2f", diff.Positive.7$est+qnorm(.025)*diff.Positive.7$se),
-                       "; ", sprintf("%.2f", diff.Positive.7$est+qnorm(.975)*diff.Positive.7$se), "]"),
-                sprintf("%.2f", diff.Positive.8$est),
-                paste0("[", sprintf("%.2f", diff.Positive.8$est+qnorm(.025)*diff.Positive.8$se),
-                       "; ", sprintf("%.2f", diff.Positive.8$est+qnorm(.975)*diff.Positive.8$se), "]")))
-
-CIndex.val <- rbind(c("All patients:",
-                sprintf("%.2f", model.All$est),
-                paste0("[", sprintf("%.2f", model.All$est+qnorm(.025)*model.All$se), 
-                      "; ", sprintf("%.2f", model.All$est+qnorm(.975)*model.All$se), "]")),
-                c("Negative SN patients:",
-                sprintf("%.2f", model.Negative$est), 
-                paste0("[", sprintf("%.2f", model.Negative$est+qnorm(.025)*model.Negative$se), 
-                      "; ", sprintf("%.2f", model.Negative$est+qnorm(.975)*model.Negative$se), "]")),
-                c("Positve SN patients:",
-                sprintf("%.2f", model.Positive$est), 
-                paste0("[", sprintf("%.2f", model.Positive$est+qnorm(.025)*model.Positive$se), 
-                      "; ", sprintf("%.2f", model.Positive$est+qnorm(.975)*model.Positive$se), "]")))
+train.AJCC <- calculate.AJCC.7.8(imputed.data=imputed.training.data, 
+                                 S.Rec.5=S.Rec.5.training, 
+                                 S.Rec=S.Rec.training, type="training")
 ```
@@
 ```{r, eval=TRUE}
-utils::write.table(cbind(train.AJCC$AJCC.7.table, rep("", 3), train.AJCC$AJCC.8.table[,-1]),
+AJCC.Cindex.table <- rbind(train.AJCC$AJCC.7.table[1,], 
+                         train.AJCC$AJCC.8.table[1,],
+                         train.AJCC$AJCC.7.table[2,], 
+                         train.AJCC$AJCC.8.table[2,],
+                         train.AJCC$AJCC.7.table[3,], 
+                         train.AJCC$AJCC.8.table[3,])
+utils::write.table(AJCC.Cindex.table,
               file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/AJCC.train.txt"),
               row.names=FALSE, col.names=FALSE, sep=",")
-utils::write.table(cbind(CIndex.val, rep("", 3),
-                   validation.AJCC$AJCC.7.table[, -1], rep("", 3), 
-                   compare.AJCC.model[, 2:3], rep("", 3),
-                   validation.AJCC$AJCC.8.table[,-1], rep("", 3),
-                   compare.AJCC.model[, 4:5]),
-              file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/AJCC.validation.txt"),
-              row.names=FALSE, col.names=FALSE, sep=",")
-utils::write.table(cbind(train.AJCC$AJCC.7.8.count, rep("", 3), validation.AJCC$AJCC.7.8.count[,-1]),
-              file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/AJCC.7.8.count.txt"),
-              row.names=FALSE, col.names=FALSE, sep=",")
 ```

### 2024-01-19 Code/12_Validation.Rmd
Final edits

--- 2023-11-23 Code/12_Validation.Rmd
@@
 for (data.type in c("training", "validation")){
-  for (outcome in c("Rec.5", "MSM.5")){
+  for (outcome in c("Rec", "MSM")){
     options(datadist=paste0("dd.", data.type))
@@
     S <- eval(parse(text=paste0("S.", outcome, ".", data.type)))
-    p <- eval(parse(text=paste0("p.", outcome, ".", data.type)))
+    p <- eval(parse(text=paste0("p.", outcome, ".5.", data.type)))
     data <- eval(parse(text=paste0(data.type, ".data.to.be.imputed")))
 
-    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", data.type, ".", outcome, ".All.png"))
+    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", data.type, ".", outcome, ".All.png"),
+        width=1800, height=1800, units="px", res=300)
     par(mar=rep(0, 4))
     PredictionTools::val.surv.mi(p, S,
-                                 g=5, main="All patients", time=NULL, 
+                                 g=5, main="All patients", time=horizon, 
                                  lim=c(0,1), dist=TRUE, 
+                                 CI.metrics=TRUE, n.iter=40,
                                  optimism.C=ifelse(data.type=="training", optimism, 0))
@@
 
-    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", data.type, ".", outcome, ".Positive.png"))
+    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", data.type, ".", outcome, ".Positive.png"),
+        width=1800, height=1800, units="px", res=300)
     par(mar=rep(0, 4))
@@
                                  g=5, main="Patients with positive SN", 
-                                 time=NULL, lim=c(0,1), dist=TRUE,
+                                 time=horizon, lim=c(0,1), dist=TRUE, 
+                                 CI.metrics=TRUE, n.iter=40,
                                  optimism.C=ifelse(data.type=="training", optimism, 0))
@@
 
-    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", data.type, ".", outcome, ".Negative.png"))
+    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", data.type, ".", outcome, ".Negative.png"),
+        width=1800, height=1800, units="px", res=300)
     par(mar=rep(0, 4))
@@
                                  g=5, main="Patients with negative SN", 
-                                 time=NULL, lim=c(0,1), dist=TRUE,
+                                 time=horizon, lim=c(0,1), dist=TRUE, 
+                                 CI.metrics=TRUE, n.iter=40,
                                  optimism.C=ifelse(data.type=="training", optimism, 0))
@@
 
-    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/", data.type, ".", outcome, ".png"), width=32, height=45, units="cm", res=600)
+    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/", data.type, ".", outcome, ".png"), 
+        width=32, height=45, units="cm", res=600)
     par(mar=rep(0, 4))

### 2024-01-19 Code/AdditionalAnalysis_Review.R

@@ -1,187 +1,0 @@
-# clear all
-rm(list = ls(all.names = TRUE))
-if(!is.null(dev.list())) dev.off()
-cat("\014")
-
-#libraries
-library(Hmisc)
-library(survivalROC)
-library(mice)
-library(survAUC) # Uno's C-index not by Uno
-library(survC1)  # Uno's C-index by Uno
-library(dplyr)
-
-# load data
-load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData")
-load("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData")
-file.path <- "Z:/Project Melanoom/PaperMelanoma/Results/"
-
-# calculate Harrell's C and Uno's C
-lp.train <- c()
-lp.val <- c()
-Harrell.Rec.train <- c()
-Uno.Rec.train <- c()
-Uno.Rec.train.correct <- c()
-Harrell.Rec.val <- c()
-Uno.Rec.val <- c()
-Uno.Rec.val.correct <- c()
-for (i in 1:m){
-  lp.train.i <- predict(f.mi.BS.Rec.5,
-                      newdata=mice::complete(imputed.training.data, i),
-                      type="lp")
-  lp.train <- cbind(lp.train, lp.train.i)
-  Harrell.Rec.train <- c(Harrell.Rec.train,
-                         Hmisc::rcorr.cens(-lp.train.i, S=S.Rec.5.training)["C Index"])
-  Uno.Rec.train <- c(Uno.Rec.train, survAUC::UnoC(Surv.rsp=S.Rec.5.training,
-                                                  Surv.rsp.new=S.Rec.5.training,
-                                                  lpnew=lp.train.i, time=5))
-
-  # right function for Uno?
-  df.unoC.train <- cbind(S.Rec.5.training[,1],
-                         S.Rec.5.training[,2],
-                         lp.train.i)
-  unoC.train <- Est.Cval(mydata=df.unoC.train, tau=5, nofit=TRUE)
-  Uno.Rec.train.correct <- c(Uno.Rec.train.correct, unoC.train$Dhat)
-
-  lp.val.i <- predict(f.mi.BS.Rec.5,
-                    newdata=mice::complete(imputed.validation.data, i),
-                    type="lp")
-  lp.val <- cbind(lp.val, lp.val.i)
-  Harrell.Rec.val <- c(Harrell.Rec.val,
-                       Hmisc::rcorr.cens(-lp.val.i, S=S.Rec.5.validation)["C Index"])
[…]

### 2024-01-19 Code/AdditionalAnalysis_tdROC.r

@@ -1,0 +1,63 @@
+# clear all
+rm(list = ls(all.names = TRUE))
+if(!is.null(dev.list())) dev.off()
+cat("\014")
+
+#libraries
+library(Hmisc)
+library(survivalROC)
+library(mice)
+library(survAUC) # Uno's C-index not by Uno
+library(survC1)  # Uno's C-index by Uno
+library(dplyr)
+
+# load data
+load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData")
+load("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData")
+
+# time-dependent ROC for a single imputation
+labels <- data.frame(outcomes=c("Rec", "MSM"),
+                     long.outcomes=c("Recurrence-free survival", "Melanoma-specific survival"),
+                     types=c("training", "validation"),
+                     long.types=c("(Europe)", "(Australia)"))
+for (data.type in labels$types){
+  imputed.data <- eval(parse(text=paste0("imputed.", data.type, ".data")))
+  single.imputation <- complete(imputed.data, 1)
+  lp.train <- predict(f.mi.BS.Rec.5,
+                      newdata=single.imputation,
+                      type="lp")
+
+  for (outcome in labels$outcomes){
+    S <- eval(parse(text=paste0("S.", outcome, ".", data.type)))
+    time.ROC <- timeROC::timeROC(T=S[,1],
+                                 delta=S[,2],
+                                 marker=lp.train,
+                                 times=5,
+                                 cause=1,
+                                 iid=FALSE)
+    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/tdROC.", outcome, ".", data.type, ".png"),
+        width=1800, height=1800, units="px", res=300)
+    plot(y=time.ROC$TP[,2], x=time.ROC$FP[,1], type="l",
+         xlab="False positive fraction (1-specificity)",
+         ylab="True positive fraction (sensitivity)",
+         main=paste(labels[labels$outcomes==outcome, "long.outcomes"], labels[labels$types==data.type, "long.types"]))
+    abline(a=0, b=1, col="grey")
+    text(x=0.1, y=1, label=paste("AUC:", sprintf("%.2f", time.ROC$AUC[2])))
+    dev.off()
+  }
+}
+
[…]

### 2024-03-11 Code/03_Descriptives_save.Rmd
Final edits

--- 2023-11-23 Code/03_Descriptives.Rmd
@@
 grDevices::dev.off()
+png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.Rec.composite.zoomed.png"), width=16, height=16, units="cm", res=300)
+survminer::ggsurvplot(survival::survfit(S.Rec.5.pooled ~ Center, 
+                                        data=pooled.data), 
+                      legend.title="",
+                      legend.labs=c("Rotterdam", "Berlin", "Warsaw", "Amsterdam", "Australia"),
+                      title="Recurrence-free survival",
+                      ylab="Recurrence-free survival probability",
+                      xlab="Time in years",
+                      ylim=c(0.6, 1),
+                      conf.int=TRUE, risk.table=TRUE)+
+  ggplot2::guides(colour=guide_legend(nrow=2))
+grDevices::dev.off()
 
@@
 grDevices::dev.off()
+png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.MSM.zoomed.png"), width=16, height=16, units="cm", res=300)
+survminer::ggsurvplot(survival::survfit(S.MSM.5.pooled ~ Center, 
+                                        data=pooled.data), 
+                      legend.title="",
+                      legend.labs=c("Rotterdam", "Berlin", "Warsaw", "Amsterdam", "Australia"),
+                      title="Melanoma-specific survival",
+                      ylab="Melanoma-specific survival probability",
+                      xlab="Time in years",
+                      ylim=c(0.6, 1),
+                      conf.int=TRUE, risk.table=TRUE)+
+  ggplot2::guides(colour=guide_legend(nrow=2))
+grDevices::dev.off()
 ```

### 2024-03-11 Code/04_TestNonlinearity.Rmd
Final edits

---
title: "Test non-linearity"
author: "Carolien C.H.M. Maas"
date: "November 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Test non-linearity

### Load packages, functions, and data 
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# erase memory
rm(list = ls(all.names = TRUE))

# for reproducibility
set.seed(100)

# load libraries
library(foreign)
library(mice)
library(survminer)
library(prodlim)
library(rms)
library(Hmisc)
library(png)
# remotes::install_github("CHMMaas/PredictionTools")
library(PredictionTools)  # to plot calibration
# import helper functions
source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")
```

# Test non-lineairities
### Test if Age should be modelled non-linearly
```{r, eval=TRUE}
load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData")
options(datadist="dd.training")

form.linear.full.Rec <- S.Rec.5.training ~
  SNstatus +
  Sex +
  Age.SN +
  Ulceration +
  Loc_CAT +
  Histology +
  Breslow +
  multiple.fields +
  Tot_SNs_neg +
  Tot_SNs_pos +
  Rdamcrit +
  Dewar +
  SNstatus:Sex +
  SNstatus:Age.SN +
  SNstatus:Ulceration +
  SNstatus:Loc_CAT +
  SNstatus:Histology +
  SNstatus:Breslow +
  SNstatus:multiple.fields
form.log.age.full.Rec <- S.Rec.5.training ~
  SNstatus +
  Sex +
  log(Age.SN) +
  Ulceration +
  Loc_CAT +
  Histology +
  Breslow +
  multiple.fields +
  Tot_SNs_neg +
  Tot_SNs_pos +
  Rdamcrit +
  Dewar +
  SNstatus:Sex +
  SNstatus:log(Age.SN) +
  SNstatus:Ulceration +
  SNstatus:Loc_CAT +
  SNstatus:Histology +
  SNstatus:Breslow +
  SNstatus:multiple.fields
form.nonlinear.age.full.Rec <- S.Rec.5.training ~
  SNstatus +
  Sex +
  rcs(Age.SN) +
  Ulceration +
  Loc_CAT +
  Histology +
  Breslow +
  multiple.fields +
  Tot_SNs_neg +
  Tot_SNs_pos +
  Rdamcrit +
  Dewar +
  SNstatus:Sex +
  SNstatus:rcs(Age.SN) +
  SNstatus:Ulceration +
  SNstatus:Loc_CAT +
  SNstatus:Histology +
  SNstatus:Breslow +
  SNstatus:multiple.fields
linear.full.Rec <- Hmisc::fit.mult.impute(form.linear.full.Rec, cph, 
                                          xtrans=imputed.training.data,
                                          data=training.data.to.be.imputed,
                                          n.impute=m, pr=FALSE,
                                          fit.reps=TRUE, 
                                          fitargs=list(y=TRUE, x=TRUE, 
                                                       se.fit=TRUE))
log.age.full.Rec <- Hmisc::fit.mult.impute(form.log.age.full.Rec, cph, xtrans=imputed.training.data,
                                                 data=training.data.to.be.imputed,n.impute=m, pr=FALSE,
                                                 fit.reps=TRUE, 
                                                 fitargs=list(y=TRUE, x=TRUE, 
                                                       se.fit=TRUE))
nonlinear.age.full.Rec <- Hmisc::fit.mult.impute(form.nonlinear.age.full.Rec, cph, xtrans=imputed.training.data,
                                                 data=training.data.to.be.imputed,n.impute=m, pr=FALSE,
                                                 fit.reps=TRUE, 
                                                 fitargs=list(y=TRUE, x=TRUE, 
                                                       se.fit=TRUE))
# distribution of Age
hist(last.imputed.training.data.set$Age.SN)
# linear model
linear.full.Rec$loglik[2]
# using rcs(Age) gives slightly better log likelihood
nonlinear.age.full.Rec$loglik[2]
```

### Test if Breslow should be modelled non-linearly
```{r, eval=TRUE}
form.log.Breslow.full.Rec <- S.Rec.5.training ~
  SNstatus +
  Sex +
  Age.SN +
  Ulceration +
  Loc_CAT +
  Histology +
  log(Breslow) +
  multiple.fields +
  Tot_SNs_neg +
  Tot_SNs_pos +
  Rdamcrit +
  Dewar +
  SNstatus:Sex +
  SNstatus:Age.SN +
  SNstatus:Ulceration +
  SNstatus:Loc_CAT +
  SNstatus:Histology +
  SNstatus:log(Breslow) +
  SNstatus:multiple.fields
form.nonlinear.Breslow.full.Rec <- S.Rec.5.training ~
  SNstatus +
  Sex +
  Age.SN +
  Ulceration +
  Loc_CAT +
  Histology +
  rcs(Breslow) +
  multiple.fields +
  Tot_SNs_neg +
  Tot_SNs_pos +
  Rdamcrit +
  Dewar +
  SNstatus:Sex +
  SNstatus:Age.SN +
  SNstatus:Ulceration +
  SNstatus:Loc_CAT +
  SNstatus:Histology +
  SNstatus:rcs(Breslow) +
  SNstatus:multiple.fields
log.Breslow.full.Rec <- Hmisc::fit.mult.impute(form.log.Breslow.full.Rec, cph, xtrans=imputed.training.data,
                                                     data=training.data.to.be.imputed,n.impute=m, pr=FALSE,
                                                     fit.reps=TRUE, 
                                                     fitargs=list(y=TRUE, x=TRUE, 
                                                       se.fit=TRUE))
nonlinear.Breslow.full.Rec <- Hmisc::fit.mult.impute(form.nonlinear.Breslow.full.Rec, cph, xtrans=imputed.training.data,
                                                     data=training.data.to.be.imputed,n.impute=m, pr=FALSE,
                                                     fit.reps=TRUE, 
                                                     fitargs=list(y=TRUE, x=TRUE, 
                                                       se.fit=TRUE))
# distribution of Breslow
hist(last.imputed.training.data.set$Breslow)
# linear model
linear.full.Rec$loglik[2]
# using rcs(Breslow) gives better log likelihood
nonlinear.Breslow.full.Rec$loglik[2]
```

### Test if Rdamcrit should be modelled non-linearly
```{r, eval=TRUE}
form.log.Rdamcrit.full.Rec <- S.Rec.5.training ~
  SNstatus +
  Sex +
  Age.SN +
  Ulceration +
  Loc_CAT +
  Histology +
  Breslow +
  multiple.fields +
  Tot_SNs_neg +
  Tot_SNs_pos +
  log(Rdamcrit) +
  Dewar +
  SNstatus:Sex +
  SNstatus:Age.SN +
  SNstatus:Ulceration +
  SNstatus:Loc_CAT +
  SNstatus:Histology +
  SNstatus:Breslow +
  SNstatus:multiple.fields
form.nonlinear.Rdamcrit.full.Rec <- S.Rec.5.training ~
  SNstatus +
  Sex +
  Age.SN +
  Ulceration +
  Loc_CAT +
  Histology +
  Breslow +
  multiple.fields +
  Tot_SNs_neg +
  Tot_SNs_pos +
  rcs(Rdamcrit) +
  Dewar +
  SNstatus:Sex +
  SNstatus:Age.SN +
  SNstatus:Ulceration +
  SNstatus:Loc_CAT +
  SNstatus:Histology +
  SNstatus:Breslow +
  SNstatus:multiple.fields
log.Rdamcrit.full.Rec <- Hmisc::fit.mult.impute(form.log.Rdamcrit.full.Rec, cph, xtrans=imputed.training.data,
                                                      data=training.data.to.be.imputed,n.impute=m, pr=FALSE,
                                                      fit.reps=TRUE, 
                                                      fitargs=list(y=TRUE, x=TRUE, 
                                                       se.fit=TRUE))
nonlinear.Rdamcrit.full.Rec <- Hmisc::fit.mult.impute(form.nonlinear.Rdamcrit.full.Rec, cph, xtrans=imputed.training.data,
                                                      data=training.data.to.be.imputed,n.impute=m, pr=FALSE,
                                                      fit.reps=TRUE, 
                                                      fitargs=list(y=TRUE, x=TRUE, 
                                                       se.fit=TRUE))
# distribution of Rdamcrit
hist(last.imputed.training.data.set$Rdamcrit)
# linear model
linear.full.Rec$loglik[2]
# using rcs(Rdamcrit) gives better log likelihood
nonlinear.Rdamcrit.full.Rec$loglik[2]

print(cbind(c("Linear", "Age log", "Age rcs", "Breslow log", "Breslow rcs", 
              "Rdamcrit log", "Rdamcrit rcs"),
            round(c(BIC(linear.full.Rec), 
                    BIC(log.age.full.Rec), BIC(nonlinear.age.full.Rec), 
                    BIC(log.Breslow.full.Rec), BIC(nonlinear.Breslow.full.Rec), 
                    BIC(log.Rdamcrit.full.Rec), BIC(nonlinear.Rdamcrit.full.Rec)), 1)))
```

### 2024-03-11 Code/05_FitModels_save.Rmd
Final edits

--- 2024-01-19 Code/05_FitModels.Rmd
@@
 
-save(list=c("dd.training", "training.data.to.be.imputed", "m", 
-            "imputed.training.data", "last.imputed.training.data.set", 
-            "horizon", 
-            "S.Rec.training", "S.MSM.training",
-            "S.Rec.5.training", "S.MSM.5.training",
-            "f.mi.full.Rec.5", "f.mi.BS.Rec.5", "f.mi.full.MSM.5.refit", 
-            "f.mi.BS.MSM.5.refit", "MSM.cal.fact", "c.Breslow", 
-            "form.full.Rec.5", "form.BS.Rec.5", "form.full.MSM.5.refit", "form.BS.MSM.5.refit"),
-     file="Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata", compress=TRUE)
 load("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData")
@@
 ```
+
+# Calculate baseline hazards
+```{r, eval=TRUE}
+# baseline hazard of recurrence averaged over all imputed data sets
+h0.Rec.all <- c()
+for (i in 1:m){
+  f.Rec.basehaz.i <- survival::basehaz(cph(form.BS.Rec.5, data=mice::complete(imputed.training.data, i), se.fit=TRUE, x=TRUE, y=TRUE))
+  h0.Rec.i <- f.Rec.basehaz.i$hazard[f.Rec.basehaz.i$time==max(f.Rec.basehaz.i$time[f.Rec.basehaz.i$time<=horizon])]
+  h0.Rec.all <- c(h0.Rec.all, h0.Rec.i)
+}
+h0.Rec <- mean(h0.Rec.all)
+
+# baseline hazard of MSM
+h0.MSM.all <- c()
+for (i in 1:m){
+  lp.i<-predict(f.mi.BS.Rec.5,
+                newdata=mice::complete(imputed.training.data,i), 
+                type="lp")
+  f.MSM.basehaz.i <- survival::basehaz(rms::cph(S.MSM.5.training~lp.i,y=TRUE,x=TRUE))
+  h0.MSM.i <- f.MSM.basehaz.i$hazard[f.MSM.basehaz.i$time==max(f.MSM.basehaz.i$time[f.MSM.basehaz.i$time<=horizon])]
+  h0.MSM.all <- c(h0.MSM.all, h0.MSM.i)
+}
+h0.MSM <- mean(h0.MSM.all)
+```
+# Save for webapp
+```{r,eval=TRUE}
+save(f.mi.BS.Rec.5, MSM.cal.fact, h0.Rec, h0.MSM, c.Breslow,
+     file="Z:/Project Melanoom/PaperMelanoma/Results/model.Rdata", 
+     compress=TRUE)
+save(list=c("dd.training", "training.data.to.be.imputed", "m", 
+            "imputed.training.data", "last.imputed.training.data.set", 
+            "horizon", "S.Rec.training", "S.MSM.training",
+            "S.Rec.5.training", "S.MSM.5.training",
+            "f.mi.full.Rec.5", "f.mi.BS.Rec.5", "f.mi.full.MSM.5.refit", 
+            "f.mi.BS.MSM.5.refit", "MSM.cal.fact", "c.Breslow", "h0.Rec", "h0.MSM",
+            "form.full.Rec.5", "form.BS.Rec.5", "form.full.MSM.5.refit", "form.BS.MSM.5.refit"),
+     file="Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata", compress=TRUE)
+```

### 2024-03-11 Code/06_CreateTablesandPlots_save.Rmd
Final edits

--- 2024-01-19 Code/06_CreateTablesandPlots.Rmd
@@
 
-  colnames(BS.table) <- c("HR", "CI", "Imp.")
+  # add p-values
+  p.out <- anova(f.mi.BS)[, "P"]
+  p.main <- c(p.out[c("SNstatus  (Factor+Higher Order Factors)")], # SN status
+                  NA,               # negative SN
+                  NA,               # positive SN
+                  p.out[c("Age.SN",     # Age
+                              "Ulceration", # Ulceration
+                              "Loc_CAT")],  # Location
+                  NA,               # upper limb
+                  NA,               # lower limb
+                  NA,               # trunk
+                  NA,               # head & neck
+                  NA,               # Breslow title
+                  p.out[c("Breslow  (Factor+Higher Order Factors)")], # Breslow negative
+                  p.out["SNstatus * Breslow  (Factor+Higher Order Factors)"], # Breslow positive
+                  p.out[c("Rdamcrit")])
+  BS.table <- cbind(BS.table, sprintf("%.4f", p.main))
+
+  colnames(BS.table) <- c("HR", "CI", "Imp.", "p-value")
   assign(paste0("final.BS.table.", model), as.data.frame(BS.table))
@@
 for (model in list("Rec.5", "MSM.5.refit")){
+  print(model)
   if (model == "Rec.5"){
@@
   for (type in c("full", "BS")){
+    print(type)
     # load model
@@
     final.full.table <- eval(parse(text=paste0("final.", type, ".table.", model)))
-    final.full.table["Harrell's C-index", "HR"] <- sprintf("%.2f", out$cindex)
-    final.full.table["Harrell's C-index", "CI"] <- paste0("[", sprintf("%.2f", out$cindex.lower),
-                                                   "; ", sprintf("%.2f", out$cindex.upper), "]")
-    final.full.table["Harrell's optimism-corrected C-index", "HR"] <- sprintf("%.2f", out$cindex-optimism)
-    final.full.table["Harrell's optimism-corrected C-index", "CI"] <- paste0("[", sprintf("%.2f", out$cindex.lower-optimism),
-                                                   "; ", sprintf("%.2f", out$cindex.upper-optimism), "]")
-    final.full.table["Uno's C-index", "HR"] <- sprintf("%.2f", out$unoC)
-    final.full.table["Uno's C-index", "CI"] <- paste0("[", sprintf("%.2f", out$unoC.lower),
-                                                   "; ", sprintf("%.2f", out$unoC.upper), "]")
+    final.full.table["Harrell's C-index", "HR"] <- sprintf("%.2f", out$HarrellC)
+    final.full.table["Harrell's C-index", "CI"] <- paste0("[", sprintf("%.2f", out$HarrellC.lower),
+                                                   "; ", sprintf("%.2f", out$HarrellC.upper), "]")
+    final.full.table["Harrell's optimism-corrected C-index", "HR"] <- sprintf("%.2f", out$HarrellC-optimism)
+    final.full.table["Harrell's optimism-corrected C-index", "CI"] <- paste0("[", sprintf("%.2f", out$HarrellC.lower-optimism),
+                                                   "; ", sprintf("%.2f", out$HarrellC.upper-optimism), "]")
+    final.full.table["Uno's C-index", "HR"] <- sprintf("%.2f", out$UnoC)
+    final.full.table["Uno's C-index", "CI"] <- paste0("[", sprintf("%.2f", out$UnoC.lower),
+                                                   "; ", sprintf("%.2f", out$UnoC.upper), "]")
     final.full.table["AUC", "HR"] <- sprintf("%.2f", out$AUC)
@@
     if (type=="BS"&model=="Rec.5"){
-      CIndex.BS.Rec.5.All <- c(sprintf("%.2f", out$cindex-optimism), 
-                              paste0("[", sprintf("%.2f", out$cindex.lower-optimism), 
-                                     "; ", sprintf("%.2f", out$cindex.upper-optimism), "]"))
+      CIndex.BS.Rec.5.All <- c(sprintf("%.2f", out$HarrellC-optimism), 
+                              paste0("[", sprintf("%.2f", out$HarrellC.lower-optimism), 
+                                     "; ", sprintf("%.2f", out$HarrellC.upper-optimism), "]"))
     }
@@
 for (type in c("full", "BS")){
+  print(type)
   f.mi <- eval(parse(text=paste0("f.mi.", type, ".Rec.5")))
-  cindex.MSM <- rep(0, m)
-  cindex.se.MSM <- rep(0, m)
-  unoC.MSM <- rep(0, m)
+  HarrellC.MSM <- rep(0, m)
+  HarrellC.MSM.se <- rep(0, m)
+  UnoC.MSM <- rep(0, m)
+  UnoC.MSM.se <- rep(0, m)
   AUC.MSM <- rep(0, m)
+  AUC.MSM.se <- rep(0, m)
   for (i in 1:m){
+    # linear predictor ith prediction
     lp.i <- predict(f.mi, newdata=mice::complete(imputed.training.data,i), type="lp")
+    f.MSM.lp.i <- survival::coxph(S.MSM.5.training ~ lp.i)
 
     # Harrell's C-index
-    rc <- Hmisc::rcorr.cens(-lp.i, 
-                            S.MSM.5.training)
-    cindex.MSM[i] <- rc["C Index"]
-    cindex.se.MSM[i] <- rc["S.D."]/2
+    rc.H <- survival::concordance(f.MSM.lp.i, timewt = "n")
+    HarrellC.MSM[i] <- rc.H$concordance
+    HarrellC.MSM.se[i] <- sqrt(rc.H$var)
 
     # Uno's C-index
-    Uno.data <- cbind(S.MSM.5.training[, 1], S.MSM.5.training[, 2], lp.i)
-    unoC.MSM[i] <- survC1::Est.Cval(mydata=Uno.data, 
-                                    tau=horizon, nofit=TRUE)$Dhat
-    if (i == 1){
-      unoC.boot <- c()
-      nr.iter <- 40
-      for (itr in 1:nr.iter){
-        boot <- Uno.data[sample(1:nrow(Uno.data), nrow(Uno.data), replace=TRUE),]
-        unoC <- survC1::Est.Cval(mydata=boot,
-                                  tau=horizon,
-                                  nofit=TRUE)$Dhat
-        unoC.boot <- c(unoC.boot, unoC)
-      }
-      unoC.MSM.se <- PredictionTools::Rubin.combine(mean(unoC.boot), sd(unoC.boot))$se
-    }
+    rc.U <- survival::concordance(f.MSM.lp.i, timewt = "n/G2")
+    UnoC.MSM[i] <- rc.U$concordance
+    UnoC.MSM.se[i] <- sqrt(rc.U$var)
 
@@
                                    cause=1,
-                                   iid=FALSE)
-
-    # time-dependent ROC SE
-    if (i == 1){
-      AUC.MSM.i <- timeROC::timeROC(T=S.MSM.training[, 1],
-                                   delta=S.MSM.training[, 2],
-                                   marker=lp.i,
-                                   times=horizon,
-                                   cause=1, 
                                    iid=TRUE)
+    AUC.MSM[i] <- as.numeric(AUC.MSM.i$AUC[2])
     AUC.MSM.CI <- stats::confint(AUC.MSM.i)
-      AUC.MSM.se <- as.numeric((as.numeric(AUC.MSM.CI$CI_AUC[2])/100-as.numeric(AUC.MSM.i$AUC[2]))/AUC.MSM.CI$C.alpha)
+    AUC.MSM.se[i] <- as.numeric((as.numeric(AUC.MSM.CI$CI_AUC[2])/100-as.numeric(AUC.MSM.i$AUC[2]))/AUC.MSM.CI$C.alpha)
   }
-    AUC.MSM[i] <- as.numeric(AUC.MSM.i$AUC[2])
-  }
-  cindex.mi.MSM <- Rubin.combine(cindex.MSM, cindex.se.MSM)
+  HarrellC.mi.MSM <- Rubin.combine(HarrellC.MSM, HarrellC.MSM.se)
+  Uno.mi.MSM <- Rubin.combine(UnoC.MSM, UnoC.MSM.se)
+  AUC.mi.MSM <- Rubin.combine(AUC.MSM, AUC.MSM.se)
 
   final.full.table <- eval(parse(text=paste0("final.", type, ".table.Rec.5")))
-  final.full.table["Harrell's C-index MSM", "HR"] <- sprintf("%.2f", cindex.mi.MSM$est)
+  final.full.table["Harrell's C-index MSM", "HR"] <- sprintf("%.2f", HarrellC.mi.MSM$est)
   final.full.table["Harrell's C-index MSM", "CI"] <- paste0("[",
-    sprintf("%.2f", cindex.mi.MSM$est+qnorm(.025)*cindex.mi.MSM$se),
-    "; ", sprintf("%.2f", cindex.mi.MSM$est+qnorm(.975)*cindex.mi.MSM$se), "]")
-  final.full.table["Harrell's optimism-corrected C-index MSM", "HR"] <- sprintf("%.2f", cindex.mi.MSM$est-optimism)
+    sprintf("%.2f", HarrellC.mi.MSM$est+qnorm(.025)*HarrellC.mi.MSM$se),
+    "; ", sprintf("%.2f", HarrellC.mi.MSM$est+qnorm(.975)*HarrellC.mi.MSM$se), "]")
+  final.full.table["Harrell's optimism-corrected C-index MSM", "HR"] <- sprintf("%.2f", HarrellC.mi.MSM$est-optimism)
   final.full.table["Harrell's optimism-corrected C-index MSM", "CI"] <- paste0("[",
-    sprintf("%.2f", cindex.mi.MSM$est+qnorm(.025)*cindex.mi.MSM$se-optimism),
-    "; ", sprintf("%.2f", cindex.mi.MSM$est+qnorm(.975)*cindex.mi.MSM$se-optimism), "]")
-  final.full.table["Uno's C-index MSM", "HR"] <- sprintf("%.2f", mean(unoC.MSM))
+    sprintf("%.2f", HarrellC.mi.MSM$est+qnorm(.025)*HarrellC.mi.MSM$se-optimism),
+    "; ", sprintf("%.2f", HarrellC.mi.MSM$est+qnorm(.975)*HarrellC.mi.MSM$se-optimism), "]")
+  final.full.table["Uno's C-index MSM", "HR"] <- sprintf("%.2f", Uno.mi.MSM$est)
   final.full.table["Uno's C-index MSM", "CI"] <- paste0("[",
-    sprintf("%.2f", mean(unoC.MSM)+qnorm(.025)*unoC.MSM.se),
-    "; ", sprintf("%.2f", mean(unoC.MSM)+qnorm(.975)*unoC.MSM.se), "]")
-  final.full.table["AUC MSM", "HR"] <- sprintf("%.2f", mean(AUC.MSM))
+    sprintf("%.2f", Uno.mi.MSM$est+qnorm(.025)*Uno.mi.MSM$se),
+    "; ", sprintf("%.2f", Uno.mi.MSM$est+qnorm(.975)*Uno.mi.MSM$se), "]")
+  final.full.table["AUC MSM", "HR"] <- sprintf("%.2f", AUC.mi.MSM$est)
   final.full.table["AUC MSM", "CI"] <- paste0("[",
-    sprintf("%.2f", mean(AUC.MSM)+qnorm(.025)*AUC.MSM.se),
-    "; ", sprintf("%.2f", mean(AUC.MSM)+qnorm(.975)*AUC.MSM.se), "]")
+    sprintf("%.2f", AUC.mi.MSM$est+qnorm(.025)*AUC.mi.MSM$se),
+    "; ", sprintf("%.2f", AUC.mi.MSM$est+qnorm(.975)*AUC.mi.MSM$se), "]")
 
@@
 # C-index of final model for recurrence stratified by SN status
-```{r, eval=TRUE}
+```{r, eval=FALSE}
 for (arm in c("Positive", "Negative")){
@@
 #               row.names=FALSE, col.names=FALSE, sep=",")
+```
 
+```{r,eval=TRUE}
 save(list=c("dd.training", "training.data.to.be.imputed", "m", 
@@
             "f.mi.full.Rec.5", "f.mi.BS.Rec.5", "f.mi.full.MSM.5.refit", 
-            "f.mi.BS.MSM.5.refit", "MSM.cal.fact", "c.Breslow", "optimism",
+            "f.mi.BS.MSM.5.refit", "MSM.cal.fact", "h0.Rec", "h0.MSM",
+            "c.Breslow", "optimism",
             "form.full.Rec.5", "form.BS.Rec.5", 

### 2024-03-11 Code/08_CreateNomogram_save.Rmd
Final edits

--- 2023-11-23 Code/08_CreateNomogram.Rmd
@@
 
-# baseline hazard of recurrence averaged over all imputed data sets
-h0.Rec.all <- c()
-for (i in 1:m){
-  f.Rec.basehaz.i <- basehaz(cph(form.BS.Rec.5, data=mice::complete(imputed.training.data, i), se.fit=TRUE, x=TRUE, y=TRUE))
-  h0.Rec.i <- f.Rec.basehaz.i$hazard[f.Rec.basehaz.i$time==max(f.Rec.basehaz.i$time[f.Rec.basehaz.i$time<=horizon])]
-  h0.Rec.all <- c(h0.Rec.all, h0.Rec.i)
-}
-h0.Rec <- mean(h0.Rec.all)
-
 # lp range Recurrence
@@
             "imputed.training.data",  "last.imputed.training.data.set", 
-            "horizon", "S.Rec.5.training", "S.MSM.5.training",
+            "horizon", 
+            "S.Rec.training", "S.MSM.training",
+            "S.Rec.5.training", "S.MSM.5.training",
             "f.mi.full.Rec.5", "f.mi.BS.Rec.5", "f.mi.full.MSM.5.refit", 
-            "f.mi.BS.MSM.5.refit", "MSM.cal.fact", "c.Breslow", "optimism",
+            "f.mi.BS.MSM.5.refit", "MSM.cal.fact", "h0.Rec", "h0.MSM",
+            "c.Breslow", "optimism",
             "form.full.Rec.5", "form.BS.Rec.5", 

### 2024-03-11 Code/09_CreateRiskScoreHistogram_save.Rmd
Final edits

--- 2024-01-19 Code/09_CreateRiskScoreHistogram.Rmd
@@
 options(datadist="dd.training")
-
-# baseline hazard of MSM
-h0.MSM.all <- c()
-lp.MSM.all <- c()
-for (i in 1:m){
-  lp.i<-predict(f.mi.BS.Rec.5,
-                newdata=mice::complete(imputed.training.data,i), 
-                type="lp")
-  lp.MSM.all <- cbind(lp.MSM.all, lp.i)
-  f.MSM.basehaz.i <- basehaz(rms::cph(S.MSM.5.training~lp.i,y=TRUE,x=TRUE))
-  h0.MSM.i <- f.MSM.basehaz.i$hazard[f.MSM.basehaz.i$time==max(f.MSM.basehaz.i$time[f.MSM.basehaz.i$time<=horizon])]
-  h0.MSM.all <- c(h0.MSM.all, h0.MSM.i)
-}
-h0.MSM <- mean(h0.MSM.all)
-lp.MSM <- rowMeans(lp.MSM.all)
-
-# TODO: Is this correct?
-lp.MSM.all[3,]
 
@@
 
-save(rc.Rec, int.Rec, f.mi.BS.Rec.5, MSM.cal.fact, max.range,
-     file="Z:/Project Melanoom/PaperMelanoma/Results/model.Rdata", 
-     compress=TRUE)
-
 save(list=c("dd.training", "training.data.to.be.imputed", "m", 
@@
             "rc.MSM", "rc.Rec", "h0.MSM", "lp.points.MSM.5",
-            "lp.Rec", "lp.MSM", "max.breaks", "h.Rec",
+            "max.breaks", "h.Rec",
             "h.plot", "y.lim.hist", "x.lim", "coef.Rec", "center.Rec"),

### 2024-03-11 Code/11_TestModel.Rmd
Final commit
Final edits

@@
 ---
-title: "Create histogram with risk scores"
+title: "Check model"
 author: "Carolien C.H.M. Maas"
-date: "December 2022"
+date: "November 2023"
 output: html_document
@@
 
+### Check mean event rates of risk distribution
+#### Recurrence
+```{r, eval=TRUE}
+load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData")
+
+# h0 for only last imputed data set
+f.Rec.basehaz.5 <- survival::basehaz(f.mi.BS.Rec.5,TRUE)
+h0.Rec.5 <- f.Rec.basehaz.5$hazard[f.Rec.basehaz.5$time==max(f.Rec.basehaz.5$time[f.Rec.basehaz.5$time<=horizon])]
+
+# lp for last imputed data set
+lp.Rec<-predict(f.mi.BS.Rec.5,
+              newdata=mice::complete(imputed.training.data,5), 
+              type="lp")
+
+# calibration factor
+f.mi.5 <- rms::cph(S.MSM.5.training~lp.Rec,y=TRUE,x=TRUE) 
+MSM.cal.fact.5 <- stats::coef(f.mi.5)
+
+# sum(h.Rec$density*fun.event(lp=lp.points.Rec.5, h0=h0.Rec))
+# similar to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
+SF<-survival::survfit(S.Rec.5.training~1,data=training.data.to.be.imputed)
+100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
+# average probability of recurrence
+mean(fun.event(lp=lp.Rec, h0=h0.Rec))
+```
 # Test if model is correct
@@
 
-### Check mean event rates of risk distribution
-#### Recurrence
-```{r, eval=TRUE}
-# h0 for only last imputed data set
-f.Rec.basehaz.5 <- survival::basehaz(f.mi.BS.Rec.5,TRUE)
-h0.Rec.5 <- f.Rec.basehaz.5$hazard[f.Rec.basehaz.5$time==max(f.Rec.basehaz.5$time[f.Rec.basehaz.5$time<=horizon])]
-
-# MSM.cal.fact for only last imputed data set
-lp.5<-predict(f.mi.BS.Rec.5,
-              newdata=mice::complete(imputed.training.data,5), 
-              type="lp")
-f.mi.5 <- rms::cph(S.MSM.5.training~lp.5,y=TRUE,x=TRUE) 
-MSM.cal.fact.5 <- stats::coef(f.mi.5)
-
-sum(h.Rec$density*fun.event(lp=lp.points.Rec.5, h0=h0.Rec.5))
-# similar to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
-SF<-survival::survfit(S.Rec.5.training~1,data=training.data.to.be.imputed)
-100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
-# average probability of recurrence
-mean(fun.event(lp=lp.Rec, h0=h0.Rec.5))
-```
-
 #### MSM
@@
 
-sum(h.Rec$density*fun.event(lp=lp.points.MSM.5, h0=h0.MSM.5))
+# sum(h.Rec$density*fun.event(lp=lp.points.MSM.5, h0=h0.MSM.5))
 # equal to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
@@
 # average probability of MSM
-mean(fun.event(lp=lp.MSM, h0=h0.MSM.5))
+mean(fun.event(lp=MSM.cal.fact*lp.Rec, h0=h0.MSM))
 ```
@@
 ```
+### See if RFS always lower than MSS
+```{r, eval=TRUE}
+p.RFS <- fun.event(lp=predict(f.mi.BS.Rec.5), h0=h0.Rec)
+p.MSS <- fun.event(lp=MSM.cal.fact*predict(f.mi.BS.Rec.5), h0=h0.MSM)
+plot(p.RFS, p.MSS)
+abline(a=0, b=1, col="red")
+```

### 2024-03-27 Code/00_PrepareTrainingData_save.Rmd
Final edits
Added instability plot

--- 2023-11-23 Code/00_PrepareTrainingData.Rmd
@@
 # erase memory
-rm(list = ls(all.names = TRUE))
+# rm(list = ls(all.names = TRUE))
 
@@
                              "MSM", "logFU.time")
+                             # "SNdate", "Date_1st_Rec", "Last_FU")
 training.data.to.be.imputed <- sel.train.data[, variables.to.be.imputed]

### 2024-03-27 Code/11_TestModel.Rmd
Added instability plot

@@
 # load libraries
+library(dplyr)
 library(rms)
@@
 ```
+# Stability plot
+```{r, eval=TRUE}
+# give each individual a unique ID
+last.imputed.training.data.set$ID <- 1:nrow(last.imputed.training.data.set)
+
+# fit model
+final.model <- cph(form.BS.Rec.5, data=last.imputed.training.data.set, y=TRUE, x=TRUE, se.fit=TRUE)
+
+# get linear predictor
+lp.final <- predict(final.model, type="lp")
+
+# obtain baseline hazard
+f.basehaz.final <- survival::basehaz(final.model)
+h0.final <- f.basehaz.final$hazard[f.basehaz.final$time==max(f.basehaz.final$time[f.basehaz.final$time<=horizon])]
+
+# obtain probability prediction
+p.original <- fun.event(h0=h0.final, lp=lp.final)
+
+# check final model
+SF<-survival::survfit(S.Rec.5.training~1,data=last.imputed.training.data.set)
+cat(" KM      :", (1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])]), "\n",
+    "avg pred:", mean(p.original), "\n")
+
+# fit model on bootstrapped sample
+p.boot <- c()
+B <- 200
+set.seed(1)
+for (i in 1:B){
+  # bootstrap sample
+  boot.indices <- sample(last.imputed.training.data.set$ID, 
+                             nrow(last.imputed.training.data.set), 
+                             replace=TRUE)
+  bootstrap.sample <- last.imputed.training.data.set[boot.indices,]
+  S.boot <- S.Rec.5.training[boot.indices,]
+
+  # fit model on bootstrap
+  form.boot <- update(form.BS.Rec.5, S.boot ~ .)
+  boot.model <- cph(form.boot, data=bootstrap.sample, y=TRUE, x=TRUE, se.fit=TRUE)
+
+  # linear predictor
+  lp.b <- predict(boot.model, newdata=last.imputed.training.data.set, type="lp")
+
+  # baseline hazard
+  f.basehaz.boot <- survival::basehaz(boot.model)
+  h0.boot <- f.basehaz.boot$hazard[f.basehaz.boot$time==max(f.basehaz.boot$time[f.basehaz.boot$time<=horizon])]
+
+  # predict probability
+  p.boot <- c(p.boot, fun.event(h0=h0.boot, lp=lp.b))
+
+  # check bootstrapped model
+  SF.boot<-survival::survfit(S.Rec.5.training~1, data=bootstrap.sample)
+  # cat(" KM      :", (1-SF.boot$surv[SF.boot$time==max(SF.boot$time[SF.boot$time<=horizon])]), "\n",
+  #     "avg pred:", mean(p.boot), "\n")
+}
+p.df <- data.frame(p.original=rep(p.original, B), p.boot=p.boot)
+```
+```{r, eval=TRUE}
+library(ggplot2)
+ggplot(data=p.df, 
+       mapping=aes(x=p.original, y=p.boot))+
+  geom_point(alpha=0.3)+
+  geom_abline(intercept=0, slope=1)+
+  ylim(c(0, 1))+ylab("Predictions using bootstrap models")+
+  xlim(c(0, 1))+xlab("Prediction using original model")+
+  theme_bw()
+```
+```{r,eval=TRUE}
+library(dcurves)
+dca(S ~ Model, data=data.frame(S=S.Rec.training, Model=p.RFS), time=5) %>% plot(smooth=TRUE)
+dca(S ~ Model, data=data.frame(S=S.MSM.training, Model=p.MSS), time=5) %>% plot(smooth=TRUE)
+```

### 2024-06-20 Code/06_CreateTablesandPlots_save.Rmd
Final edits

@@
                                         main="All centers", CI.metrics=TRUE,
-                                        lim=c(0,1), dist=TRUE, n.iter=40)
+                                        lim=c(0,1), dist=TRUE)
     grDevices::dev.off()

### 2024-06-20 Code/12_Validation.Rmd
Final edits

@@
   for (outcome in c("Rec", "MSM")){
+    cat("Make calibration plot for", data.type, "data, for the outcome", outcome, "\n")
     options(datadist=paste0("dd.", data.type))
@@
                                  lim=c(0,1), dist=TRUE, 
-                                 CI.metrics=TRUE, n.iter=40,
+                                 CI.metrics=TRUE,
+                                 show.metrics=c(rep(TRUE, 5), FALSE),
                                  optimism.C=ifelse(data.type=="training", optimism, 0))
@@
                                  time=horizon, lim=c(0,1), dist=TRUE, 
-                                 CI.metrics=TRUE, n.iter=40,
+                                 CI.metrics=TRUE, 
+                                 show.metrics=c(rep(TRUE, 5), FALSE),
                                  optimism.C=ifelse(data.type=="training", optimism, 0))
